### 2019-11-29 manuscript.rmd
manuscript version

---
title: "Sampbias, a method to evaluate geographic sampling bias in species distribution data"
author: "Alexander Zizka^1,2,3^, Alexandre Antonelli^3,4,5^, Daniele Silvestro^3,4^"
bibliography: sampbias.bib
output: 
  bookdown::pdf_document2:
  # word_document:
    toc: false
    toc_depth: 1
    fig_caption: true
    number_sections: false
header-includes: |
    \usepackage{caption}
    \usepackage{fancyhdr}
    \pagestyle{fancy}
    \fancyhead[L]{sampbias accessibility bias}
    \fancyhead[R]{ } 
    \usepackage{float}
    \floatplacement{figure}{H}
    \usepackage{lineno}
    \linenumbers
    \usepackage[url=false]{biblatex}
linestretch: 2
fontsize: 12pt
always_allow_html: yes
---

1. German Center for Integrative Biodiversity Research, University of Leipzig, Leipzig, Germany
2. Naturalis Biodiversity Center, Leiden University, 2300RA Leiden, The Netherlands 
3. Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden
4. Department of Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden
5. Royal Botanical Gardens Kew, Richmond, Surrey, United Kingdom

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sampbias)
```

\newpage{}

# Abstract
Georeferenced species occurrences from public databases have become essential to biodiversity resaerch and conservation, but have limitations. Geographically biased sampling is a widely recognized issue that might severely affect analyses. Especially "roadside bias", i.e. differences in sampling intensity among localities caused by differences in accessibility for humans is ubiquitous and might differ in strength among taxonomic groups and datasets. Yet, no general methodology existed to quantify the effect of roadside bias on a dataset level. Here we present *sampbias*, a novel algorithm and software to estimates the biasing effect of human accessibility means (roads, rivers, airports, cities, or any user-defined structures) in species occurrence datasets, based on a null model of random sampling and exponential distance decay. Results of *sampbias* can be used to compare the biasing effect among factors and datasets. *Sampbias* is implemented as a user-friendly R package, and Shiny app. We exemplify the use of *sampbias* on an example dataset of  mammal occurrences from the Indonesian island of Borneo downloaded from www.gbif.org. *Sampbias* offers an efficient and largely automated means for biodiversity scientists and non-specialists alike to explore bias in species occurrence data. The output of *sampbias* may be used to identify priorities for further collection or digitalization efforts, provide bias surfaces for species distribution modelling, or assess the reliability of scientific results based on publicly available species distribution data.

# Keywords
Collection effort, Global biodiversity Information Facility (GBIF), Presence only data, Roadside bias, Sampling intensity

\newpage{}

# Introduction
Publicly available datasets of geo-referenced species occurrences, such as provided by the Global Biodiversity Information Facility (www.gbif.org) have become a fundamental resource  in biological sciences, especially in biogeography, conservation, macroecology, and systematics. However, because these dataset are "presence-only" data, rarely include information on collection effort, are not collected systematically and often compiled from a variety of collection purposes (e.g. scientific expeditions, census counts, genetic barcoding studies, and citizen science observations) they are subject to collection biases [@Meyer2016]. That is the number of data points available is biased by factors other than species' presence or abundance, including the under sampling of specific taxa ("taxonomic bias", e.g., birds *vs.* nematodes), specific geographic regions ("geographic bias", i.e. easily accessible *vs.* remote areas), and specific temporal periods ["temporal bias", i.e. wet season *vs.* dry season, @Isaac2015; @Boakes2010]. While these biases are broadly recognized, and approaches exist to account for them in some analyses (for instance for species-richness estimation [@Engemann2015] species distribution modelling [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014], occupancy modelling [@Kery2016], or abundance estimations [@Shimadzu2015]), few attempts have been made [e.g. @Ruete2015] to quantify their strength and to discern among different sources of bias and the strength of bias among datasets, at least on the large scale.

Geographic sampling bias, the fact that sampling effort is spatially biased, rather than equally distributed over a given study area is prevalent in all non-systematically collected datasets of species distributions. Many factors can affect sampling effort, such as socio-economic [i.e. national research spending, history of scientific research, @Meyer2016] and political factors [armed conflict, democratic rights, @Ryden2019] or physical accessibility [i.e. distance to a road or river, terrain conditions, slope, @Yang2014; @Botts2011]. Especially physical accessibility is omnipresent as a biasing factor [e.g. @Lin2015; @Engemann2015], across spatial scales, and the term "roadside bias has been coined for it. In practice, this means that most species observations (occurrence points) are made in or near cities, along roads and rivers, and near other human settlements (such as airports). Less observations come from the middle of a tropical rainforest or from a mountain top. Interestingly, since the observation of different taxonomic groups has different challenges, geographic sampling bias and the effect of accessibility may differ among taxonomic groups [@Vale2012].

The implications of not considering spatial collection bias in biodiversity research are likely to be substantial [@Meyer2016; @Rocchini2015; @Shimadzu2015;  @Yang2013; @Kramer-Schadt2013; @Barbosa2013]. While it is unrealistic to expect that spatial biases in biodiversity data will ever disappear, it is crucial that researchers realise the intrinsic biases associated with the biodiversity data they are dealing with. This is the first step towards estimating to which extent these biases may affect their analyses, results and conclusions drawn from such data. Therefore, it is advisable for any study dealing with species occurrence data assess the biases covered by *sampbias*.

Here, we present *sampbias*, a novel method to quantify accessibility bias in individual datasets of species occurrences, in a way that is comparable across datasets, implemented as an R-package. Specifically sampbias uses a null-model of random sampling to address two questions:

1) How strong is the geographic sampling bias in a given dataset?

2) How important is the bias caused by different means of accessibility, such as to airport, cities, rivers or roads.

3) How is sampling bias distributed in space, i.e. which area are a priority for targeted sampling?

# Description
## General concept
*Sampbias* evaluates the biasing effect of geographic features by comparing the statistical distance distribution of species' occurrence records to theses features observed in a user-provided dataset to a simulated distribution expected under equal sampling. Based on an user-defined number of gazetteers indicating the position of the geographic features of interest, for instance roads, the expected statistical distribution of distances in the study area is calculated. Random sampling assumes no impact of environment or ecology on the distribution of occurrence records; therefore, sampbias can only be applied to multi-species data sets.

*Sampbias* works on a user-defined scale, and any multi-species occurrence records can be tested against any set of geographic gazetteers (reliability increases with increasing dataset size). Default large-scale gazetteers for airports, cities, rivers and roads are provided with *sampbias*. Species occurrence data as downloaded from the data portal of GBIF can be directly used as input data for sampbias. The output of the package includes measures of bias effect, which are comparable between different gazetteers (e.g. comparing biasing effect of roads and rivers), different taxa (e.g. birds vs. flowering plants) and different data sets (e.g. specimens vs human observations).

CoordinateCleaner is implemented in R [@rcoreteam2019] based on standard tools for spatial statistics: ggplot2 [@Wickham2009], geosphere [@Hijmans2019], maptools [@Bivand2019], raster [@Hijmans2019], sp [@Pebesma2005; @Bivand2013], and viridis [@Garnier2018].

## Distance calculation
To estimate the statistical distribution of geographic distances to biasing features, *sampbias* first uses gazetteers of the geographic location of biasing structures (e.g. roads) to generate a gridded raster across the entire study area (the geographic extent of the dataset) for each gazetteer and then calculates the distance ("as the crow flies") of the midpoint of each raster cell to the closest cell occupied by a instance of the gazetteer. These distance rasters are then used to sample the observed statistical distribution of distances in the observed dataset and the null distribution in a reference dataset of equal size with randomly distributed records (the null model). The resolution of the raster defined the precision of the distance estimates, for instance a 1x1 degree raster will yield approximately a 100km precision at the equator.

## Quantifying sampling bias using maximum likelihood
Under the assumption that organisms exist across the entire area of interest, we can expect the number of sampled occurrences to be distributed uniformly in space (even though, of course, the density of individuals and the species composition may be heterogeneous).  With *sampbias* we assess if a set of occurrences significantly departs from a null uniform distribution and whether these discrepancies between expected and observed distributions can be explained by distance from factors that potentially bias their sampling probability (e.g. distance from cities or roads).

Given the placement of a particular bias sources in the area of interest and assuming a uniform distribution of samples, the probability of a sampled occurrence located at a distance $d$ from the closest bias source is a function of the amount of available area at that distance. That is, the larger the area located at distance $d$ from a bias source the more samples we expect. For simplicity we discretize the area of interest in a number of grid cells and indicate with $f(x)$ the function describing the number of available grid cells at any distance $x$, for $0 < x < max(x)$, where $max(x)$ is the maximum observed distance from a bias source. The function $f(x)$ is therefore calculated based on the distances of each grid cell from its closest bias source.

The distribution of samples, in the absence of bias, should therefore represent a random sample from $f(x)$ and reflect its shape, i.e. $d ~ f(x)$ (Fig. \ref{fig:distdi}). However, in the presence of a bias, we expect the probability of finding occurrences to decrease with increasing distance from a bias source. This in turn will alter the resulting distribution of samples that no longer match the expected distribution. Here, we model the effect of sampling bias by assuming that the probability of sampling an occurrence decreases exponentially with increasing distance (Fig \ref{fig:distdi}), following the function $b(x,l) = l exp(-lx)$, where $l$ is the rate parameter. Under these assumption the expected distribution of samples is given by $g(x,l) \propto f(x) b(x,l)$.

The rate parameter $l$ describes the strength of the bias effect. When  $l$ is large the expected probability of sampling occurrences decreases very quickly as you move away from a bias source (Fig \ref{fig:distdi}). In contrast, when the $l$ parameter is small, the resulting exponential distribution effectively becomes more and more similar to a uniform distribution (Fig \ref{fig:distdi}), indicating that increasing distance from e.g. a city does not affect the sampling probability. We treat $l$ as an unknown variable and estimate it using maximum likelihood from the data based on the probability density function described by $g(x,l)$. This essentially means finding the value of $l$ that best explains any discrepancies between the expected distribution ($f(x)$) and the observed occurrences.

Once we have an estimated value of $l$, we can infer the expected sample bias as a function of distance using $b(x,l)$. Since the function $b(x,l)$ describes the exponentially decreasing sampling probability in relation to distance, we can define a standardized bias function as: $B(x,l) = 1 - b(x,l)/b(0,l)$, where  the level of bias is set to 0 at distance 0 from the bias source. The standardized bias function tends asymptotically to 1 as the distance tends to infinity. However, since in any area $max(x) << \infty$, for small values of $l$ (i.e. little or no bias), $B(x,l)$ will look essentially like a uniform distribution with values very close to 0. Large values of $l$ (i.e. strong bias) will instead result in a curve that quickly approaches 1 with increasing distance. The values provided by the standardized bias function can be interpreted as the proportion of occurrences that are missing from the sample, compared to the observed samples at distance 0. Thus, if for a given estimated $l$ we have $B(50,l) = 0.20$, we can expect that at 50 Km from e.g. a road the number of occurrences per grid cell will be about 80% of the occurrences sampled at distance 0 from the road, with 20% missing due to sampling bias.

When running *sampbias*, we typically test different sources of biases, such as roads, cities, airports, etc. For each factor an independent expected distribution ($f(x)$) is computed and a parameter $l$ is estimated from $g(x,l)$. These estimates can be use to produce maps showing the intensity of potential biases across the area based on the standardized bias function. The bias values obtained from different sources can then be averaged in each grid cell to produce a map showing the combined effects of all sources.

# Running sampbias
A default *sampbias* analysis can be run with four lines of code in R. The main function `SamplingBias` creates an object of the class `"sampbias"`, for which the package provides a plotting and summary method. Based on a `data.frame` including species identity and geographic coordinates, sampbias provides a bias effect estimate for each gazetteer and an average bias. Additionally some options exist to provide custom gazetteers, custom distances for the bias estimation, a custom grain size of the analysis, as well as some operators for the calculation of the bias distances. A tutorial on how to use sampbias is available with the package and in the electronic supplement of this publication (Appendix S1).

```{r, eval = FALSE, tidy=TRUE}
library(sampbias)

#a data table with species identify, longitude, and latitude
example.in <- read.csv(system.file("extdata", 
                                   "mammals_borneo.csv",
                                   package="sampbias"), 
                       sep = "\t")

#running sampbias
example.out <- SamplingBias(x = example.in, res = 0.1)

# summarizing the results
summary(example.out)
plot(example.out)

```

For data exploration we implemented the basic functionalities in a shiny app graphical user interface (Fig. \ref{fig:GUI}). Analyses can be run based on a tab separated .txt file with occurrence information including the column headers "species", "decimallongitude" and "decimallatitude", as for instance files downloaded from www.gbif.org, using custom gazetteers. A tutorial on how to use the *sampbias* GUI is available online (https://ropensci.github.io/sampbias/) and in the electronic supplement of this publication (Appendix S2).

# Empirical example
To exemplify the use and output of sampbias, we downloaded the occurrence records of all mammals available from the Indonesian island of Borneo [@gbifdoi], and quantify the biasing effect of airports, cities and roads in the dataset. **Something on the results, also add a table** (Fig. \ref{fig:empirical})

# Assumptions and future prospective
Two fundamental theoretical assumptions of *sampbias* are a random sampling of occurrence records across the study area as null model and an exponential increase of the biasing effect with distance from the gazetteers. We considered both acceptable approximations for the purpose of the package, but future expansions of *sampbias* could relax these assumptions, for instance by allowing other distance decay functions, such as a gamma distribution, and by changing the sampling scheme of the background points. The first steps towards this end are already implemented in the current version of *sampbias* with the option to limit background points to a convex hull around the dataset or limiting background points to terrestrial surface.

A practical limitation of *sampbias* is the trade-off between the resolution of raster for the distance calculation and the geographic extent of the dataset. For instance, a 100m resolution for a global dataset would lead to the generation of gridded raster a for which distance calculation will become computationally prohibitive in most practical cases, Hence, *sampbias* is best suited for local or regional datasets at high resolution (c. 100 -- 10,000m) or continental datasets at low resolution (c. 10 -- 100km).

# Todo
re-run empirical analysis

test units

add badges to the repository

# Data accessibility
The software presented here is available under a GPL-3 license. The *sampbias* R package and the source code for the shiny app are available via https://github.com/azizka/sampbias, the shiny app can be accessed at https://azizka.shinyapps.io/sampbias/. The R package includes an example dataset as well as vignettes detailing the use of the R package, the use of the shiny app and possibly warnings produced by the package (Appendix S3).

# Acknowledgements
We thank the organizers of the 2016 Ebben Nielsen challenge for inspiring and recognizing this research. We thank all data collectors and contributors to GBIF for their effort.

# Author contributions
All authors conceived of this study, AZ and DS developed the statistical algorithm, AZ wrote the R-package and Shiny app, AZ and DS wrote the manuscript with contributions from AA.

# Figures

<!-- Figure 1 A sketch of the different distributions -->
```{r distdi, out.width = "7in", echo = FALSE, fig.cap = "Schematic representation of the sampbias algorithm. The upper panel shows a case without accessibility bias: the observed distribution of distances from the geographic feature of interest matches the expected distribution and hence the bias function is flat. The lower panel shows a case with bias: the expected and observed distribution differ and the bias is best represented by an exponential function.", eval = TRUE}
knitr::include_graphics(path = "ms_figures/Distance_distribution_example.png")
```

<!-- Figure 2 shiny interface -->
```{r GUI, out.width = "7in", echo = FALSE, fig.cap = "The interface of the sampbias shiny app. With a navigation panel on the left and visualization of the results on the right.", eval = TRUE}
knitr::include_graphics(path = "ms_figures/bild_gui_sampbias.png")
```

<!-- Figure 3 empirical results -->
```{r empirical, out.width = "7in", echo = FALSE, fig.cap = "The spatial projection of the accessibility bias in a dataset of mammal occurrences on the Indonesian island of Borneo from www.gbif.org. Sampbias finds a strong biasing effect by roads, cities and lower by airports, but surprisingly no effect of rivers.", eval = TRUE}
knitr::include_graphics(path = "ms_figures/visualization_rpackage_application_githubtutorial.jpg")
```

# Supplementary material

Appendix S1 - Tutorial running sampbias in R

Appendix S2 - Tutorial running the sampbias shiny app

Appendix S3 - Possible warnings and their solutions

# References

### 2019-11-29 R/DisRast.R

@@ -1,5 +1,60 @@
 # Distance raster for all gazeteers get empirical distace distribution from
 # gazetteers, rasterbased
+
+
+#' Distance Rasters from a List of Geographic Gazetteers
+#'
+#' Creates a list of distances rasters based on a list of geographic
+#' gazetteers, as SpatialPoints or SpatialLines objects, and a template raster,
+#' indicating the desired extent and resolution
+#'
+#'
+#' @param gaz an object of the class \code{list}, including one or more
+#' geographic gazetteers of the class \code{SpatialPointsDataFrame} or
+#' \code{SpatialLinesDataFrame}.
+#' @param ras an object of the class \code{raster}. Defining the extent and
+#' resolution of the distances rasters.
+#' @param buffer numerical.  The size of the geographic buffer around the
+#' extent of ras for the distance calculations in degrees, to account for
+#' geographic structures neighbouring the study area (such as a road right
+#' outside the study area) Default is to the resolution of \code{ras}.
+#' @param ncores numerical.  The number of cores used for parallel computing.
+#' Must be lower than the available number of cores. Not finally implemented in
+#' version 0.1.0.
+#' @return a \code{list} of \code{raster} objects of the same length as
+#' \code{gaz}. The values in each raster correspond to the planar geographic
+#' distance to the next feature in \code{gaz}, given the resolution of
+#' \code{ras}
+#' @note Check https://github.com/azizka/sampbias/wiki for a tutorial on
+#' sampbias.
+#' @seealso \code{\link{SamplingBias}}
+#' @keywords spatial
+#' @examples
+#'
+#' #create raster for resolution and extent
+#' ras <- raster(extent(-5,5,-4,4), res = 1)
+#'
+#' #create point gazeteer
+#' pts <- data.frame(long = runif(n = 5, min = -5, max = 5),
+#'                   lat = runif(n = 5, min = -4, max = 4),
+#'                   dat = rep("A", 5))
+#'
+#' pts <- SpatialPointsDataFrame(coords = pts[,1:2], data = data.frame(pts[,3]))
+#'
+#' lin <- data.frame(long = seq(-5, 5, by = 1),
+#'                   lat = rep(2, times = 11))
+#' lin <- SpatialLinesDataFrame(sl = SpatialLines(list(Lines(Line(lin), ID="B1"))),
[…]

### 2019-11-29 R/SamplingBias.R

@@ -1,3 +1,120 @@
+#' Evaluating Sampling Bias in Species Distribution Data
+#'
+#' The major function of the package, calculating the bias effect of sampling
+#' bias due to geographic structures, such as the vicinity to cities, airports,
+#' rivers and roads. Results are projected to space, and can be compared
+#' numerically.
+#'
+#' The default gazetteers delivered with the package are simplified from
+#' http://www.naturalearthdata.com/downloads/. They include major features, and
+#' for small scale analyses custom gazetteers should be used.
+#'
+#' For computational convenience, the gazetteers are cropped to the extent of
+#' the point occurrence data sets. To account for the fact, that, relevant
+#' structures might lay directly outside this extent, but still influencing the
+#' distribution of samples in the study area, the buffer option, gives the
+#' area, around the extent that should be included in the distance calculation.
+#'
+#' The \code{biasdist} options defines at which distance the bias effect should
+#' be recorded. These values are comparable between different biasing factors
+#' and give an estimate of bias severity. For instance, if \code{biasdist} =
+#' c(0, 10000), the \code{biastable} will indicate the bias effect at distances
+#' 0 and 10000 meters from all sources of bias. The bias effect is the fraction
+#' of samples expected to be missed at a given distance relative to distance 0.
+#' A high bias effect means a high fraction of missed records. The fractions
+#' are comparable between sources of bias and different datasets and the bias
+#' at any given distance can thus inform on the relative severity of a biasing
+#' source.
+#'
+#' Visit \url{https://github.com/azizka/sampbias/wiki} for more information on
+#' distance calculation and the algorithm behind sampbias.
+#'
+#' @param x an object of the class \code{data.frame}, with one species
+#' occurrence record per line, and at least three columns, named
+#' \sQuote{species}, \sQuote{decimallongitude}, and \sQuote{decimallatitude}.
+#' @param gaz a list of geographic gazetteers as \code{SpatialPointDataFrame}
+#' or \code{SpatialLinesDataFrame}.  If NULL, a set of default gazetteers,
+#' representing large scale occurrence of airports, cities, rivers and roads is
+#' used.  See Details.
+#' @param res numerical.  The raster resolution for the distance calculation to
[…]

### 2019-11-29 R/internal_data_preparation_raster.R

@@ -1,3 +1,10 @@
+
+#'@importFrom stats na.omit
+#'@importFrom graphics hist
+#'@importFrom sp CRS Polygons SpatialPoints
+#'@importFrom raster extract rasterize
+
+
 #Occurrence raster
 .OccRast <- function(x, ras){
   rast <- raster::rasterize(x, ras, fun = "count")
@@ -29,7 +36,6 @@
     parallel::clusterExport(cl, "ras")
     parallel::clusterEvalQ(cl, library(raster))
     parallel::clusterEvalQ(cl, library(sp))
-    parallel::clusterEvalQ(cl, library(rgeos))
 
     rast <- parallel::parLapply(cl,
                                   inp,

### 2019-11-29 R/internal_likelihood_calculation.R

@@ -1,3 +1,7 @@
+#'@importFrom stats dexp median
+#'@importFrom graphics par plot
+#'@importFrom grDevices dev.off pdf
+
 .RunSampBias <- function(dists, distmats, biasdist = c(0, 10), plotextra = F, plotextrafile = "samp_bias_extra_plots.pdf") {
 
   names.dists <- names(dists)

### 2019-11-29 R/is.sampbias.R

@@ -1,2 +1,42 @@
+#' Is Method for Class sampbias
+#'
+#' Check class of \code{sampbias} objects.
+#'
+#' With two arguments, tests whether object can be treated as from class2.
+#' With one argument, returns all the super-classes of this object's class.
+#'
+#' @param object an object of the class \code{sampbias}
+#' @param class2 the names of the class to which is relations are to be
+#' examined defined, or (more efficiently) the class definition objects for the
+#' classes.
+#' @seealso Check https://github.com/azizka/sampbias/wiki for a tutorial on
+#' sampbias.
+#' @keywords methods
+#' @examples
+#'
+#' #simulate data
+#' occ <- data.frame(species = rep(sample(x = LETTERS, size = 5), times = 10),
+#'                   decimallongitude = runif(n = 50, min = -5, max = 5),
+#'                   decimallatitude = runif(n = 50, min = -4, max = 4))
+#'
+#'
+#' #create point gazetteer
+#' pts <- data.frame(long = runif(n = 5, min = -5, max = 5),
+#'                   lat = runif(n = 5, min = -4, max = 4),
+#'                   dat = rep("A", 5))
+#' pts <- SpatialPointsDataFrame(coords = pts[,1:2], data = data.frame(pts[,3]))
+#'
+#' lin <- data.frame(long = seq(-5, 5, by = 1),
+#'                   lat = rep(2, times = 11))
+#' lin <- SpatialLinesDataFrame(sl = SpatialLines(list(Lines(Line(lin), ID="B1"))),
+#'                              data = data.frame("B", row.names = "B1"))
+#'
+#' gaz <- list(lines.strucutre = lin, point.structure = pts)
+#'
+#' out <- SamplingBias(x = occ, gaz = gaz, terrestrial = FALSE)
+#' is(out)
+#'
+#'@export
+#'
 is.sampbias <- function(object, class2){
   inherits(object, "sampbias")}

### 2019-11-29 R/plot.sampbias.R

@@ -1,3 +1,55 @@
+#' Mapping Occurrence and Bias Rasters in an Object of the Class Sampbias
+#'
+#' Plotting method for class \code{sampbias}, generating a rasterized plots
+#' showing the number of occurrences and the number of species per gridcell,
+#' one bias raster indicating the effect strength for each gazetteer provided to
+#' \code{SamplingBias}, and a average bias raster indicating the average bias
+#' for each gridcell.
+#'
+#'
+#' @param x an object of the class \code{sampbias}.
+#' @param gaz a list of SpatialObjects, to be printed on the maps.  Should be
+#' the same objects provided to \code{SamplingBias} when creating the Object.
+#' If \code{gaz} is not supplied, the sampbias package standard gazetteers are
+#' used.
+#' @param sealine logical. Should the sealine be added to the plots? Default is
+#' to TRUE.
+#' @param ...  Additional arguments to be passed to \code{\link{plot}}.
+#' @return A series of R plots based on ggplot2.
+#' @note Check https://github.com/azizka/sampbias/wiki for a tutorial on
+#' sampbias.
+#' @seealso \code{\link{SamplingBias}}, \code{\link{summary.sampbias}}
+#' @keywords methods
+#' @examples
+#'
+#' #simulate data
+#' occ <- data.frame(species = rep(sample(x = LETTERS, size = 5), times = 10),
+#'                   decimallongitude = runif(n = 50, min = -5, max = 5),
+#'                   decimallatitude = runif(n = 50, min = -4, max = 4))
+#'
+#'
+#' #create point gazetteer
+#' pts <- data.frame(long = runif(n = 5, min = -5, max = 5),
+#'                   lat = runif(n = 5, min = -4, max = 4),
+#'                   dat = rep("A", 5))
+#' pts <- SpatialPointsDataFrame(coords = pts[,1:2], data = data.frame(pts[,3]))
+#'
+#' lin <- data.frame(long = seq(-5, 5, by = 1),
+#'                   lat = rep(2, times = 11))
+#' lin <- SpatialLinesDataFrame(sl = SpatialLines(list(Lines(Line(lin), ID="B1"))),
+#'                              data = data.frame("B", row.names = "B1"))
+#'
+#' gaz <- list(lines.strucutre = lin, point.structure = pts)
+#'
+#' out <- SamplingBias(x = occ, gaz = gaz, terrestrial = FALSE)
+#' plot(out)
+#'
+#'@export
[…]

### 2019-11-29 R/sampbias.R

@@ -1,0 +1,114 @@
+
+
+#' Global Airport Coordinates
+#' 
+#' A gazetteer of major global airport locations.
+#' 
+#' 
+#' @name airports
+#' @docType data
+#' @format A data frame with 891 observations on the following 4 variables.
+#' \describe{ \item{list("code")}{a factor, IATA-3 code.} \item{list("name")}{a
+#' character vector, airport names.} \item{list("longitude")}{a numeric
+#' vector.} \item{list("latitude")}{a numeric vector.} }
+#' @source \url{http://www.naturalearthdata.com/}
+#' @keywords datasets
+#' @examples
+#' 
+#' data(airports)
+#' str(airports)
+#' 
+NULL
+
+
+
+
+
+#' Global Major Populated Places
+#' 
+#' A gazetteer of global country and province capitals, as well as major cities
+#' and smaller cities in less populated areas.
+#' 
+#' 
+#' @name cities
+#' @docType data
+#' @format A data frame with 7322 observations on the following 4 variables.
+#' \describe{ \item{list("nameascii")}{a factor; city name.}
+#' \item{list("adm0_a3")}{a factor; ISO-3 code of country name.}
+#' \item{list("longitude")}{a numeric vector.} \item{list("latitude")}{a
+#' numeric vector.} }
+#' @source
+#' \url{http://www.naturalearthdata.com/downloads/10m-cultural-vectors/}
+#' @keywords datasets
+#' @examples
+#' 
+#' data(cities)
+#' str(cities)
+#' 
+NULL
+
+
+
+
+
+#' Global Coastlines
+#' 
+#' A \code{SpatialPolygonsDataFrame} with global coastlines.
+#' 
+#' 
+#' @name landmass
+#' @docType data
+#' @source
+#' \url{http://www.naturalearthdata.com/downloads/10m-physical-vectors/}
+#' @keywords datasets
+#' @examples
+#' 
+#' data("landmass")
+#' plot(landmass)
+#' 
+NULL
+
+
+
+
+
+#' Global Network of Major Roads
+#' 
+#' A gazetteer of major global roads as \code{SpatialPolygonsDataFrame}.
+#' 
+#' 
+#' @name roads
+#' @docType data
+#' @source
+#' \url{http://www.naturalearthdata.com/downloads/10m-cultural-vectors/}
+#' @keywords datasets
+#' @examples
+#' 
[…]

### 2019-11-29 R/summary.sampbias.R

@@ -1,3 +1,40 @@
+#' Summary Method for Class sampbias
+#'
+#' Summary method for objects of the class \code{sampbias}.
+#'
+#'
+#' @param object An object of the class \code{sampbias}
+#' @param \dots Additional arguments passed to summary.
+#' @return Summary printed to screen.
+#' @seealso \code{\link{SamplingBias}} \code{\link{is.sampbias}}
+#' \code{\link{plot.sampbias}}
+#' @keywords methods
+#' @examples
+#'
+#' #simulate data
+#' occ <- data.frame(species = rep(sample(x = LETTERS, size = 5), times = 10),
+#'                   decimallongitude = runif(n = 50, min = -5, max = 5),
+#'                   decimallatitude = runif(n = 50, min = -4, max = 4))
+#'
+#'
+#' #create point gazetteer
+#' pts <- data.frame(long = runif(n = 5, min = -5, max = 5),
+#'                   lat = runif(n = 5, min = -4, max = 4),
+#'                   dat = rep("A", 5))
+#' pts <- SpatialPointsDataFrame(coords = pts[,1:2], data = data.frame(pts[,3]))
+#'
+#' lin <- data.frame(long = seq(-5, 5, by = 1),
+#'                   lat = rep(2, times = 11))
+#' lin <- SpatialLinesDataFrame(sl = SpatialLines(list(Lines(Line(lin), ID="B1"))),
+#'                              data = data.frame("B", row.names = "B1"))
+#'
+#' gaz <- list(lines.strucutre = lin, point.structure = pts)
+#'
+#' out <- SamplingBias(x = occ, gaz = gaz, terrestrial = FALSE)
+#' summary(out)
+#'
+#' @export
+#'
 summary.sampbias <- function(object, ...) {
   cat("Number of occurences: ", object$summa$total_occ, "\n")
   cat("Number of species: ", object$summa$total_sp, "\n")

### 2019-11-29 pre_submission_tests.R

@@ -1,0 +1,20 @@
+
+pkgdown::build_site()
+
+source("https://install-github.me/MangoTheCat/goodpractice")
+library(goodpractice)
+gp("C:/Users/az64mycy/Dropbox (iDiv)/research_projects/29_CoordinateCleaner/CoordinateCleaner")
+
+#R CMD check
+devtools::check_built(path = ".",
+                      manual = T, cran = T)
+
+devtools::spell_check()
+
+devtools::check_rhub()
+
+devtools::check_win_devel()
+
+codemetar::write_codemeta()
+
+devtools::release()

### 2019-11-29 vignettes/Using_the_samp_bias_rpackage.R

@@ -1,0 +1,34 @@
+## ----setup, include=FALSE------------------------------------------------
+knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
+
+## ---- eval = FALSE-------------------------------------------------------
+#  require(devtools)
+#  install_github(repo = "azizka/sampbias")
+#  library(sampbias)
+#  library(maptools)
+
+## ---- echo = FALSE-------------------------------------------------------
+library(sampbias)
+library(maptools)
+
+## ------------------------------------------------------------------------
+#loading a text file to R
+occ <-read.csv(system.file("extdata", "mammals_borneo.csv", package="sampbias"), sep = "\t")
+
+## ------------------------------------------------------------------------
+cit <- readShapeSpatial(system.file("extdata", "Borneo_major_cities.shp", package="sampbias"))
+roa <- readShapeSpatial(system.file("extdata", "Borneo_major_roads.shp", package="sampbias"))
+
+gazetteers <- list(cities = cit,
+                   roads = roa)
+
+## ------------------------------------------------------------------------
+bias.out <- SamplingBias(x = occ, gaz = gazetteers)
+
+## ------------------------------------------------------------------------
+summary(bias.out)
+#plot(bias.out)
+
+## ------------------------------------------------------------------------
+#plot(bias.out, gaz = gazetteers)
+

### 2019-11-29 manuscript.rmd
gitignore and bibliogrphy file

@@
 Many factors can affect sampling effort, such as socio-economic [i.e.
-national research spending, history of scientific research, @Meyer2016] and political factors [armed conflict, democratic rights, @Ryden2019] or physical accessibility [i.e.
+national research spending, history of scientific research, @Meyer2015, @Daru2018] and political factors [armed conflict, democratic rights, @Ryden2019] or physical accessibility [i.e.
 distance to a road or river, terrain conditions, slope, @Yang2014; @Botts2011].
@@
 # Author contributions
-All authors conceived of this study, AZ and DS developed the statistical algorithm, AZ wrote the R-package and Shiny app, AZ and DS wrote the manuscript with contributions from AA.
+All authors conceived of this study, AZ and DS developed the statistical algorithm, AZ and DS wrote the R-package and AZ the Shiny app, AZ and DS wrote the manuscript with contributions from AA.
 

### 2019-12-02 manuscript.rmd
manuscript changes
new manuscript version

@@
 1. German Center for Integrative Biodiversity Research, University of Leipzig, Leipzig, Germany
-2. Naturalis Biodiversity Center, Leiden University, 2300RA Leiden, The Netherlands 
+2. Naturalis Biodiversity Center, Leiden University, Leiden, The Netherlands 
 3. Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden
@@
 Yet, no general methodology existed to quantify the effect of roadside bias on a dataset level.
-Here we present *sampbias*, a novel algorithm and software to estimates the biasing effect of human accessibility means (roads, rivers, airports, cities, or any user-defined structures) in species occurrence datasets, based on a null model of random sampling and exponential distance decay.
-Results of *sampbias* can be used to compare the biasing effect among factors and datasets.
+Here we present *sampbias*, a novel algorithm and software to estimate the biasing effect of accessibility (by roads, rivers, airports, cities, or any user-defined structures) in species occurrence datasets.
+*Sampbias* is based on a null model of equal sampling and exponential distance decay and the resultsare compareable among biasing factors and datasets.
 *Sampbias* is implemented as a user-friendly R package, and Shiny app.
-We exemplify the use of *sampbias* on an example dataset of  mammal occurrences from the Indonesian island of Borneo downloaded from www.gbif.org.
+We exemplify the use of *sampbias* on a dataset of  mammal occurrences from the Indonesian island of Borneo, downloaded from www.gbif.org.
 *Sampbias* offers an efficient and largely automated means for biodiversity scientists and non-specialists alike to explore bias in species occurrence data.
@@
 scientific expeditions, census counts, genetic barcoding studies, and citizen science observations) they are subject to collection biases [@Meyer2016].
-That is the number of data points available is biased by factors other than species' presence or abundance, including the under sampling of specific taxa ("taxonomic bias", e.g., birds *vs.* nematodes), specific geographic regions ("geographic bias", i.e.
+That is the number of data points available is biased by factors other than species' presence or abundance, including the under-sampling of specific taxa ("taxonomic bias", e.g., birds *vs.* nematodes), specific geographic regions ("geographic bias", i.e.
 easily accessible *vs.* remote areas), and specific temporal periods ["temporal bias", i.e.
 wet season *vs.* dry season, @Isaac2015; @Boakes2010].
-While these biases are broadly recognized, and approaches exist to account for them in some analyses (for instance for species-richness estimation [@Engemann2015] species distribution modelling [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014], occupancy modelling [@Kery2016], or abundance estimations [@Shimadzu2015]), few attempts have been made [e.g.
-@Ruete2015] to quantify their strength and to discern among different sources of bias and the strength of bias among datasets, at least on the large scale.
+While these biases are broadly recognized, and approaches exist to account for them in some analyses (for instance for species-richness estimation [@Engemann2015] species distribution modelling [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014], occupancy modelling [@Kery2016], or abundance estimations [@Shimadzu2015]), few attempts have been made [but see @Ruete2015] to discern among different sources of bias or to compare the strength of bias among datasets.
 
 Geographic sampling bias, the fact that sampling effort is spatially biased, rather than equally distributed over a given study area is prevalent in all non-systematically collected datasets of species distributions.
-Many factors can affect sampling effort, such as socio-economic [i.e.
-national research spending, history of scientific research, @Meyer2015, @Daru2018] and political factors [armed conflict, democratic rights, @Ryden2019] or physical accessibility [i.e.
-distance to a road or river, terrain conditions, slope, @Yang2014; @Botts2011].
+Many factors can affect sampling effort, such as socio-economic factors [i.e.
+national research spending, history of scientific research; @Meyer2015, @Daru2018] and political factors [armed conflict, democratic rights; @Ryden2019] or physical accessibility [i.e.
+distance to a road or river, terrain conditions, slope; @Yang2014; @Botts2011].
 Especially physical accessibility is omnipresent as a biasing factor [e.g.
-@Lin2015; @Engemann2015], across spatial scales, and the term "roadside bias has been coined for it.
+@Lin2015; @Engemann2015], across spatial scales, and the term "roadside bias" has been coined for it.
 In practice, this means that most species observations (occurrence points) are made in or near cities, along roads and rivers, and near other human settlements (such as airports).
@@
 This is the first step towards estimating to which extent these biases may affect their analyses, results and conclusions drawn from such data.
-Therefore, it is advisable for any study dealing with species occurrence data assess the biases covered by *sampbias*.
+Therefore, it is advisable for any study dealing with species occurrence data to assess the strength of accessibility bias in the underlaying data.
 
-Here, we present *sampbias*, a novel method to quantify accessibility bias in individual datasets of species occurrences, in a way that is comparable across datasets, implemented as an R-package.
-Specifically sampbias uses a null-model of random sampling to address two questions:
+Here, we present *sampbias*, a novel method to quantify accessibility bias in individual datasets of species occurrences, in a way that is comparable across datasets.
+*Sampbias* is implemented as an R-package.
+Specifically, *sampbias* uses a null-model of random sampling to address two questions:
 
-1) How strong is the geographic sampling bias in a given dataset?
+1) How strong is the accessibility bias in a given dataset?
 
-2) How important is the bias caused by different means of accessibility, such as to airport, cities, rivers or roads.
+2) How important are different means of human accessibility, such as to airport, cities, rivers or roads, in causing this bias?
 
-3) How is sampling bias distributed in space, i.e. which area are a priority for targeted sampling?
+3) How is sampling bias distributed in space, i.e. which areas are a priority for targeted sampling?
 
@@
 ## General concept
-*Sampbias* evaluates the biasing effect of geographic features by comparing the statistical distance distribution of species' occurrence records to theses features observed in a user-provided dataset to a simulated distribution expected under equal sampling.
-Based on an user-defined number of gazetteers indicating the position of the geographic features of interest, for instance roads, the expected statistical distribution of distances in the study area is calculated.
-Random sampling assumes no impact of environment or ecology on the distribution of occurrence records; therefore, sampbias can only be applied to multi-species data sets.
+Under the assumption that organisms exist across the entire area of interest, we can expect the number of sampled occurrences to be distributed uniformly in space (even though, of course, the density of individuals and the species composition may be heterogeneous).
+With *sampbias* we assess if a set of occurrences significantly departs from a null uniform distribution and whether these discrepancies between expected and observed distributions can be explained by distance from factors that potentially bias their sampling probability (e.g.
+distance from cities or roads).
 
-*Sampbias* works on a user-defined scale, and any multi-species occurrence records can be tested against any set of geographic gazetteers (reliability increases with increasing dataset size).
+*Sampbias* works on a user-defined scale, and any dataset of multi-species occurrence records can be tested against any geographic gazetteer (reliability increases with increasing dataset size).
 Default large-scale gazetteers for airports, cities, rivers and roads are provided with *sampbias*.
@@
 comparing biasing effect of roads and rivers), different taxa (e.g.
-birds vs.
-flowering plants) and different data sets (e.g.
-specimens vs human observations).
+birds *vs.* flowering plants) and different data sets (e.g.
+specimens *vs.* human observations).
 
@@
 ## Distance calculation
-To estimate the statistical distribution of geographic distances to biasing features, *sampbias* first uses gazetteers of the geographic location of biasing structures (e.g.
-roads) to generate a gridded raster across the entire study area (the geographic extent of the dataset) for each gazetteer and then calculates the distance ("as the crow flies") of the midpoint of each raster cell to the closest cell occupied by a instance of the gazetteer.
-These distance rasters are then used to sample the observed statistical distribution of distances in the observed dataset and the null distribution in a reference dataset of equal size with randomly distributed records (the null model).
-The resolution of the raster defined the precision of the distance estimates, for instance a 1x1 degree raster will yield approximately a 100km precision at the equator.
+*Sampbias* uses gazetteers of the geographic location of bias sources (e.g.
+roads) to generate a grid across the study area (the geographic extent of the dataset) for each gazetteer and then calculates the distance ("as the crow flies") of the midpoint of each grid cell to the closest cell containing an instance of the gazetteer.
+We then use these distance grids to sample the distribution of distances in the observed dataset and the null distribution in a reference dataset of equal size with randomly distributed records (the null model).
+The resolution of the grid defined the precision of the distance estimates, for instance a 1x1 degree raster will yield approximately a 100km precision at the equator.
 
-## Quantifying sampling bias using maximum likelihood
-Under the assumption that organisms exist across the entire area of interest, we can expect the number of sampled occurrences to be distributed uniformly in space (even though, of course, the density of individuals and the species composition may be heterogeneous).
-With *sampbias* we assess if a set of occurrences significantly departs from a null uniform distribution and whether these discrepancies between expected and observed distributions can be explained by distance from factors that potentially bias their sampling probability (e.g.
-distance from cities or roads).
-
+## Quantifying accessibility bias using maximum likelihood
 Given the placement of a particular bias sources in the area of interest and assuming a uniform distribution of samples, the probability of a sampled occurrence located at a distance $d$ from the closest bias source is a function of the amount of available area at that distance.
@@
 This in turn will alter the resulting distribution of samples that no longer match the expected distribution.
-Here, we model the effect of sampling bias by assuming that the probability of sampling an occurrence decreases exponentially with increasing distance (Fig \ref{fig:distdi}), following the function $b(x,l) = l exp(-lx)$, where $l$ is the rate parameter.
+Here, we model the effect of sampling bias by assuming that the probability of sampling an occurrence decreases exponentially with increasing distance (Fig.
+\ref{fig:distdi}), following the function $b(x,l) = l exp(-lx)$, where $l$ is the rate parameter.
 Under these assumption the expected distribution of samples is given by $g(x,l) \propto f(x) b(x,l)$.
@@
 The rate parameter $l$ describes the strength of the bias effect.
-When  $l$ is large the expected probability of sampling occurrences decreases very quickly as you move away from a bias source (Fig \ref{fig:distdi}).
-In contrast, when the $l$ parameter is small, the resulting exponential distribution effectively becomes more and more similar to a uniform distribution (Fig \ref{fig:distdi}), indicating that increasing distance from e.g.
+When  $l$ is large the expected probability of sampling occurrences decreases very quickly as you move away from a bias source (Fig.
+\ref{fig:distdi}).
+In contrast, when the $l$ parameter is small, the resulting exponential distribution effectively becomes more and more similar to a uniform distribution (Fig.
+\ref{fig:distdi}), indicating that increasing distance from e.g.
 a city does not affect the sampling probability.
@@
 
-Once we have an estimated value of $l$, we can infer the expected sample bias as a function of distance using $b(x,l)$.
+Once we have an estimated value of $l$, we can infer the expected accessibility bias as a function of distance using $b(x,l)$.
 Since the function $b(x,l)$ describes the exponentially decreasing sampling probability in relation to distance, we can define a standardized bias function as: $B(x,l) = 1 - b(x,l)/b(0,l)$, where  the level of bias is set to 0 at distance 0 from the bias source.
@@
 # Running sampbias
-A default *sampbias* analysis can be run with four lines of code in R.
+A default *sampbias* analysis can be run with few lines of code in R.
 The main function `SamplingBias` creates an object of the class `"sampbias"`, for which the package provides a plotting and summary method.
@@
 
-For data exploration we implemented the basic functionalities in a shiny app graphical user interface (Fig.
+For data exploration we implemented the basic functionalities in a shiny app as graphical user interface (Fig.
 \ref{fig:GUI}).
@@
 # Assumptions and future prospective
-Two fundamental theoretical assumptions of *sampbias* are a random sampling of occurrence records across the study area as null model and an exponential increase of the biasing effect with distance from the gazetteers.
+Two assumptions of *sampbias* are a equal sampling of occurrence records across the study area as null model and an exponential increase of the biasing effect with distance from the gazetteers.
 We considered both acceptable approximations for the purpose of the package, but future expansions of *sampbias* could relax these assumptions, for instance by allowing other distance decay functions, such as a gamma distribution, and by changing the sampling scheme of the background points.
-The first steps towards this end are already implemented in the current version of *sampbias* with the option to limit background points to a convex hull around the dataset or limiting background points to terrestrial surface.
+The first steps towards these goals are already implemented in the current version of *sampbias* with the option to limit background points to a convex hull around the dataset or limiting background points to terrestrial surface.
 
-A practical limitation of *sampbias* is the trade-off between the resolution of raster for the distance calculation and the geographic extent of the dataset.
-For instance, a 100m resolution for a global dataset would lead to the generation of gridded raster a for which distance calculation will become computationally prohibitive in most practical cases, Hence, *sampbias* is best suited for local or regional datasets at high resolution (c.
+A practical limitation of *sampbias* is the trade-off between the resolution of the grid for the distance calculation and the geographic extent of the dataset.
+For instance, a 100m resolution for a global dataset would lead to the generation of grid  for which distance calculation will become computationally prohibitive in most practical cases, Hence, *sampbias* is best suited for local or regional datasets at high resolution (c.
 100 -- 10,000m) or continental datasets at low resolution (c.
@@
 test units
-
-add badges to the repository
 
@@
 <!-- Figure 1 A sketch of the different distributions -->
-```{r distdi, out.width = "7in", echo = FALSE, fig.cap = "Schematic representation of the sampbias algorithm.
-The upper panel shows a case without accessibility bias: the observed distribution of distances from the geographic feature of interest matches the expected distribution and hence the bias function is flat.
-The lower panel shows a case with bias: the expected and observed distribution differ and the bias is best represented by an exponential function.", eval = TRUE}
+```{r distdi, out.width = "7in", echo = FALSE, fig.cap = "Schematic representation of the \\textit{sampbias} algorithm.
+The upper panel (A-C) shows a case without accessibility bias: the observed distribution of distances from the geographic feature of interest matches the expected distribution and hence the bias function is flat.
+The lower panel (D-F) shows a case with bias: the expected and observed distribution differ and the bias is best represented by an exponential function.", eval = TRUE}
 knitr::include_graphics(path = "ms_figures/Distance_distribution_example.png")
@@
 <!-- Figure 3 empirical results -->
-```{r empirical, out.width = "7in", echo = FALSE, fig.cap = "The spatial projection of the accessibility bias in a dataset of mammal occurrences on the Indonesian island of Borneo from www.gbif.org.
-Sampbias finds a strong biasing effect by roads, cities and lower by airports, but surprisingly no effect of rivers.", eval = TRUE}
+```{r empirical, out.width = "7in", echo = FALSE, fig.cap = "The spatial projection of the accessibility bias in an empirical example dataset of mammal occurrences on the Indonesian island of Borneo from www.gbif.org.
+\\textit{Sampbias} finds a strong biasing effect by roads, cities and lower by airports, but, surprisingly, no effect of rivers.", eval = TRUE}
 knitr::include_graphics(path = "ms_figures/visualization_rpackage_application_githubtutorial.jpg")

### 2019-12-03 manuscript.rmd
Update manuscript.rmd

@@
 # Abstract
-Georeferenced species occurrences from public databases have become essential to biodiversity resaerch and conservation, but have limitations.
+Georeferenced species occurrences from public databases have become essential to biodiversity research and conservation, but have limitations.
 Geographically biased sampling is a widely recognized issue that might severely affect analyses.
@@
 differences in sampling intensity among localities caused by differences in accessibility for humans is ubiquitous and might differ in strength among taxonomic groups and datasets.
-Yet, no general methodology existed to quantify the effect of roadside bias on a dataset level.
+Yet, no general methodology exists to quantify the effect of roadside or other sources of bias on a dataset level.
 Here we present *sampbias*, a novel algorithm and software to estimate the biasing effect of accessibility (by roads, rivers, airports, cities, or any user-defined structures) in species occurrence datasets.
-*Sampbias* is based on a null model of equal sampling and exponential distance decay and the resultsare compareable among biasing factors and datasets.
+*Sampbias* is based on a null model of even sampling and assesses whether instead sampling probability decays exponentially with distance.
+The results are comparable among biasing factors and datasets.
 *Sampbias* is implemented as a user-friendly R package, and Shiny app.
@@
 Publicly available datasets of geo-referenced species occurrences, such as provided by the Global Biodiversity Information Facility (www.gbif.org) have become a fundamental resource  in biological sciences, especially in biogeography, conservation, macroecology, and systematics.
-However, because these dataset are "presence-only" data, rarely include information on collection effort, are not collected systematically and often compiled from a variety of collection purposes (e.g.
-scientific expeditions, census counts, genetic barcoding studies, and citizen science observations) they are subject to collection biases [@Meyer2016].
+However, because these datasets are "presence-only" data, they rarely include information on collection effort.
+Instead they are typically not collected systematically and often compiled from a variety of sources (e.g.
+scientific expeditions, census counts, genetic barcoding studies, and citizen-science observations),  thus becoming subject to collection biases [@Meyer2016].
 That is the number of data points available is biased by factors other than species' presence or abundance, including the under-sampling of specific taxa ("taxonomic bias", e.g., birds *vs.* nematodes), specific geographic regions ("geographic bias", i.e.
@@
 wet season *vs.* dry season, @Isaac2015; @Boakes2010].
-While these biases are broadly recognized, and approaches exist to account for them in some analyses (for instance for species-richness estimation [@Engemann2015] species distribution modelling [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014], occupancy modelling [@Kery2016], or abundance estimations [@Shimadzu2015]), few attempts have been made [but see @Ruete2015] to discern among different sources of bias or to compare the strength of bias among datasets.
+While these biases are broadly recognized, and approaches exist to account for them in some analyses (for instance for species-richness estimation [@Engemann2015] species distribution modelling [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014], occupancy modelling [@Kery2016], or abundance estimations [@Shimadzu2015]), few attempts have been made to discern among different sources of bias or to compare the strength of bias among datasets [but see @Ruete2015].
 
@@
 While it is unrealistic to expect that spatial biases in biodiversity data will ever disappear, it is crucial that researchers realise the intrinsic biases associated with the biodiversity data they are dealing with.
-This is the first step towards estimating to which extent these biases may affect their analyses, results and conclusions drawn from such data.
+This is the first step towards estimating to which extent these biases may affect their analyses, results, and conclusions drawn from such data.
 Therefore, it is advisable for any study dealing with species occurrence data to assess the strength of accessibility bias in the underlaying data.
@@
 Under the assumption that organisms exist across the entire area of interest, we can expect the number of sampled occurrences to be distributed uniformly in space (even though, of course, the density of individuals and the species composition may be heterogeneous).
+[[*I think we should acknowledge here that this assumption is valid when looking at a geographically restricted area, eg within a tropical forest.
+Of course different biomes eg forest, alpine, oceanic will result in different carrying capacity*]]
 With *sampbias* we assess if a set of occurrences significantly departs from a null uniform distribution and whether these discrepancies between expected and observed distributions can be explained by distance from factors that potentially bias their sampling probability (e.g.
@@
 *Sampbias* uses gazetteers of the geographic location of bias sources (e.g.
-roads) to generate a grid across the study area (the geographic extent of the dataset) for each gazetteer and then calculates the distance ("as the crow flies") of the midpoint of each grid cell to the closest cell containing an instance of the gazetteer.
+roads) to generate a grid across the study area (the geographic extent of the dataset) for each gazetteer and then calculates the distance ("as the crow flies")
+[[*meaning?*]]
+of the midpoint of each grid cell to the closest cell containing an instance of the gazetteer.
 We then use these distance grids to sample the distribution of distances in the observed dataset and the null distribution in a reference dataset of equal size with randomly distributed records (the null model).
@@
 That is, the larger the area located at distance $d$ from a bias source the more samples we expect.
-For simplicity we discretize the area of interest in a number of grid cells and indicate with $f(x)$ the function describing the number of available grid cells at any distance $x$, for $0 < x < max(x)$, where $max(x)$ is the maximum observed distance from a bias source.
+For simplicity we discretize the area of interest in a number of grid cells and indicate with $f(x)$ the function describing the number of available grid cells at any distance $x$, for $0 < x < \max(x)$, where $\max(x)$ is the maximum observed distance between a cell and the clostest bias source.
 The function $f(x)$ is therefore calculated based on the distances of each grid cell from its closest bias source.
@@
 Here, we model the effect of sampling bias by assuming that the probability of sampling an occurrence decreases exponentially with increasing distance (Fig.
-\ref{fig:distdi}), following the function $b(x,l) = l exp(-lx)$, where $l$ is the rate parameter.
+\ref{fig:distdi}), following the function $b(x,l) = l \exp(-lx)$, where $l$ is the rate parameter.
 Under these assumption the expected distribution of samples is given by $g(x,l) \propto f(x) b(x,l)$.
+
+[[*I wonder if we shouldn't use " $b(x,l) = \exp(-lx)$" so the Y-value is always 1 at a distance of 0.
+That would require normalizing the likelihood based on the integral of the curve though, as it will be different from 1* ]
 
@@
 This essentially means finding the value of $l$ that best explains any discrepancies between the expected distribution ($f(x)$) and the observed occurrences.
+[[*I wonder if this isn't in fact a posterior probability where f(x) is the prior and b(x,l) the likelihood or something like that*]]
 
 Once we have an estimated value of $l$, we can infer the expected accessibility bias as a function of distance using $b(x,l)$.
-Since the function $b(x,l)$ describes the exponentially decreasing sampling probability in relation to distance, we can define a standardized bias function as: $B(x,l) = 1 - b(x,l)/b(0,l)$, where  the level of bias is set to 0 at distance 0 from the bias source.
+Since the function $b(x,l)$ describes the exponentially decreasing sampling probability in relation to distance, we can define a standardized bias function as: $B(x,l) = 1 - b(x,l)/b(0,l)$,
+[[*Actually I think this is effectively equivalent to what I worteabove $b(x,l) = \exp(-lx)$*]]
+where  the level of bias is set to 0 at distance 0 from the bias source.
 The standardized bias function tends asymptotically to 1 as the distance tends to infinity.

### 2019-12-03 manuscript.rmd
Update manuscript.rmd

@@
 This essentially means finding the value of $l$ that best explains any discrepancies between the expected distribution ($f(x)$) and the observed occurrences.
-[[*I wonder if this isn't in fact a posterior probability where f(x) is the prior and b(x,l) the likelihood or something like that*]]
+[[*I think this is in fact a posterior probability where f(x) is the empirical prior and b(x,l) the likelihood and we do a maximum a posteriori optimization*]]
 
@@
 Two assumptions of *sampbias* are a equal sampling of occurrence records across the study area as null model and an exponential increase of the biasing effect with distance from the gazetteers.
-We considered both acceptable approximations for the purpose of the package, but future expansions of *sampbias* could relax these assumptions, for instance by allowing other distance decay functions, such as a gamma distribution, and by changing the sampling scheme of the background points.
+We considered both acceptable approximations for the purpose of the package, but future expansions of *sampbias* could relax these assumptions, for instance by allowing other distance decay functions, such as gamma or Weibull distributions, and by changing the sampling scheme of the background points.
 The first steps towards these goals are already implemented in the current version of *sampbias* with the option to limit background points to a convex hull around the dataset or limiting background points to terrestrial surface.
+[[*I don't understand this part*]]
 

### 2019-12-03 manuscript.rmd
Update manuscript.rmd

@@
 Since the function $b(x,l)$ describes the exponentially decreasing sampling probability in relation to distance, we can define a standardized bias function as: $B(x,l) = 1 - b(x,l)/b(0,l)$,
-[[*Actually I think this is effectively equivalent to what I worteabove $b(x,l) = \exp(-lx)$*]]
+[[*Actually I think this is effectively equivalent to what I wrote above $b(x,l) = \exp(-lx)$*]]
 where  the level of bias is set to 0 at distance 0 from the bias source.
@@
 The bias values obtained from different sources can then be averaged in each grid cell to produce a map showing the combined effects of all sources.
+[[*yeah this part is still weird.
+Basically, if rivers are not explaining anything they should count nothing toward the combined estimate, whereas they do now.
+I think we could try to see if the max likelihoods are comparable and weight the average by that.*]]
 

### 2019-12-04 manuscript.rmd
manuscript comments skype

@@
 
+get the dsitribution from all grid cells and normalize by the number of all available grid cells, not downsampled to the same number of points ignore the starting at 1/intercept
+
+a possion likelihood where the rate i
+two paramters, 
+1. speed of decae
+2. how high is the bias at distance zero
+
+ $b(x,l) = \exp(-lx)$
+ $q * exp(-lambda x) / \int_0^\infty (q * exp(-lambda x))$
+
+Average the bias by Akaike wheights, or likelihood
+
 The rate parameter $l$ describes the strength of the bias effect.

### 2019-12-04 R/internal_likelihood_calculation.R

@@ -158,8 +158,3 @@
   return(res)
 
 }
-
-
-
-
-

### 2019-12-12 R/dis_rast.R
further changes

@@ -48,14 +48,14 @@
 #'
 #' gaz <- list(point.structure = pts, lines.strucutre = lin)
 #'
-#' out <- DisRast(gaz, ras)
+#' out <- dis_rast(gaz, ras)
 #'
 #' \dontrun{plot(out[[1]])}
 #'
 #'@export
 #'@importFrom raster crop extent raster res
 #'
-DisRast <- function(gaz, ras, buffer = NULL, ncores = 1) {
+dis_rast <- function(gaz, ras, buffer = NULL, ncores = 1) {
 
   # create buffer, if none is supplied
   if (is.null(buffer)) {

### 2019-12-12 R/is.sampbias.R

@@ -33,7 +33,7 @@
 #'
 #' gaz <- list(lines.strucutre = lin, point.structure = pts)
 #'
-#' out <- SamplingBias(x = occ, gaz = gaz, terrestrial = FALSE)
+#' out <- sampling_bias(x = occ, gaz = gaz, terrestrial = FALSE)
 #' is(out)
 #'
 #'@export

### 2019-12-12 R/plot.sampbias.R

@@ -41,7 +41,7 @@
 #'
 #' gaz <- list(lines.strucutre = lin, point.structure = pts)
 #'
-#' out <- SamplingBias(x = occ, gaz = gaz, terrestrial = FALSE)
+#' out <- sampling_bias(x = occ, gaz = gaz, terrestrial = FALSE)
 #' plot(out)
 #'
 #'@export

### 2019-12-12 R/sampling_bias.R

@@ -108,14 +108,14 @@
 #'
 #' gaz <- list(lines.strucutre = lin, point.structure = pts)
 #'
-#' out <- SamplingBias(x = occ, gaz = gaz, terrestrial = FALSE)
+#' out <- sampling_bias(x = occ, gaz = gaz, terrestrial = FALSE)
 #' summary(out)
 #'
 #' @export
 #' @importFrom raster as.raster extent extent<- mask res<- values
 #' @importFrom grDevices is.raster
 #'
-SamplingBias <- function(x, gaz = NULL, res = 1, buffer = NULL, convexhull = F, terrestrial = T,
+sampling_bias <- function(x, gaz = NULL, res = 1, buffer = NULL, convexhull = F, terrestrial = T,
                          binsize = NULL, biasdist = c(0, 10000), ncores = 1, plotextra = F,
                          plotextrafile = "samp_bias_extra_plots.pdf", verbose = T) {
 
@@ -197,7 +197,7 @@ SamplingBias <- function(x, gaz = NULL, res = 1, buffer = NULL, convexhull = F, 
     dis.ras <- gaz
   } else {
     ## create distance raster for all gazeteers
-    dis.ras <- DisRast(gaz = gaz, ras = occ.out, buffer = buffer, ncores = ncores)
+    dis.ras <- dis_rast(gaz = gaz, ras = occ.out, buffer = buffer, ncores = ncores)
     if (terrestrial) {
       dis.ras <- lapply(dis.ras, function(k) mask(k, wrld))
     }

### 2019-12-13 R/calculate_bias.R
new algorithm

@@ -1,0 +1,260 @@
+#' Evaluating Sampling Bias in Species Distribution Data
+#'
+#' The major function of the package, calculating the bias effect of sampling
+#' bias due to geographic structures, such as the vicinity to cities, airports,
+#' rivers and roads. Results are projected to space, and can be compared
+#' numerically.
+#'
+#' The default gazetteers delivered with the package are simplified from
+#' http://www.naturalearthdata.com/downloads/. They include major features, and
+#' for small scale analyses custom gazetteers should be used.
+#'
+#' For computational convenience, the gazetteers are cropped to the extent of
+#' the point occurrence data sets. To account for the fact, that, relevant
+#' structures might lay directly outside this extent, but still influencing the
+#' distribution of samples in the study area, the buffer option, gives the
+#' area, around the extent that should be included in the distance calculation.
+#'
+#' The \code{biasdist} options defines at which distance the bias effect should
+#' be recorded. These values are comparable between different biasing factors
+#' and give an estimate of bias severity. For instance, if \code{biasdist} =
+#' c(0, 10000), the \code{biastable} will indicate the bias effect at distances
+#' 0 and 10000 meters from all sources of bias. The bias effect is the fraction
+#' of samples expected to be missed at a given distance relative to distance 0.
+#' A high bias effect means a high fraction of missed records. The fractions
+#' are comparable between sources of bias and different datasets and the bias
+#' at any given distance can thus inform on the relative severity of a biasing
+#' source.
+#'
+#' Visit \url{https://github.com/azizka/sampbias/wiki} for more information on
+#' distance calculation and the algorithm behind sampbias.
+#'
+#' @param x an object of the class \code{data.frame}, with one species
+#' occurrence record per line, and at least three columns, named
+#' \sQuote{species}, \sQuote{decimallongitude}, and \sQuote{decimallatitude}.
+#' @param gaz a list of geographic gazetteers as \code{SpatialPointDataFrame}
+#' or \code{SpatialLinesDataFrame}.  If NULL, a set of default gazetteers,
+#' representing large scale occurrence of airports, cities, rivers and roads is
+#' used.  See Details.
[…]

### 2019-12-13 R/dis_rast.R

@@ -27,7 +27,7 @@
 #' \code{ras}
 #' @note Check https://github.com/azizka/sampbias/wiki for a tutorial on
 #' sampbias.
-#' @seealso \code{\link{SamplingBias}}
+#' @seealso \code{\link{calculate_bias}}
 #' @keywords spatial
 #' @examples
 #'

### 2019-12-13 R/internal.R

@@ -1,0 +1,123 @@
+#'@importFrom raster rasterize
+#'@importFrom stats dexp dpois rnorm runif
+
+#Occurrence raster
+.OccRast <- function(x, ras){
+  rast <- raster::rasterize(x, ras, fun = "count")
+  return(rast)
+}
+
+#get number of decimal places
+.DecimalPlaces <- function(x) {
+  if ((x %% 1) != 0) {
+    nchar(strsplit(sub('0+$', '', as.character(x)), ".", fixed = TRUE)[[1]][[2]])
+  } else {
+    return(0)
+  }
+}
+
+get_lambda_ij <- function(q, w, X) {
+  if (class(X) == "numeric") {
+    return(q * exp(-w * X))
+  } else {
+    return(q * exp(-apply(FUN = sum, w * X, 2)))
+  }
+}
+
+get_poi_likelihood <- function(Xcounts, lambdas) {
+  return(dpois(Xcounts, lambdas, log = T))
+}
+
+multiplier_proposal <- function(i, d = 1.2) {
+  u <- runif(length(i))
+  l <- 2 * log(d)
+  m <- exp(l * (u - 0.5))
+  ii <- i * m
+  hastings_ratio <- sum(log(m))
+  return(list(ii, hastings_ratio))
+}
+
+
+.RunSampBias <- function(x,
+                         rescale_counts = 1,
+                         rescale_distances = 1000,
+                         iterations = 1e+05,
+                         burnin = 0,
+                         post_samples = NULL,
+                         outfile = NULL) {
+
+  indx <- c(3:ncol(x))
+
+  # X is the distance matrix with 1 row for each cell and 1 col for each predictor
+  X <- x[, indx]/rescale_distances
+  if (length(indx) > 1) {
+    X <- t(X)
+  }
+
+  Xcounts <- x$record_count
+
+  # coefficient, i.e. Poisson rate at zero distance from all
+  qA = mean(Xcounts)
+
+  # init weigths of predictors
+  wA = abs(rnorm(length(indx), 0.01, 0.01))
+
+  lambdas = get_lambda_ij(qA, wA, X)
+
+  likA = sum(get_poi_likelihood(Xcounts, lambdas))
+  priorA = sum(dexp(1, wA, log = T)) + dexp(0.01, qA, log = T)
+
+  names_b = paste("w", names(x[, indx]), sep = "_")
+
+
+  out <- data.frame()
+
[…]

### 2019-12-13 R/internal_data_preparation_raster.R

@@ -1,122 +1,0 @@
-
-#'@importFrom stats na.omit
-#'@importFrom graphics hist
-#'@importFrom sp CRS Polygons SpatialPoints
-#'@importFrom raster extract rasterize
-
-
-#Occurrence raster
-.OccRast <- function(x, ras){
-  rast <- raster::rasterize(x, ras, fun = "count")
-  return(rast)
-}
-
-#Species Number raster
-.SpeRast <- function(x, ras, ncores){
-  inp <- split(x, f = x$species)
-
-  if(ncores > 1){
-    if (!requireNamespace("parallel", quietly = TRUE)) {
-      stop("parallel needed for multithreading. Please install the package.",
-           call. = FALSE)
-    }else{
-      cl <- parallel::makeCluster(ncores)
-    }
-  }
-
-  if(ncores == 1){
-    rast <- lapply(inp, function(k){
-      po <- sp::SpatialPoints(k[,c("decimallongitude", "decimallatitude")], CRS("+proj=longlat +datum=WGS84"))
-      ras_sub <- raster::rasterize(po, ras, fun = "count")
-      ras_sub[ras_sub >= 1] <- 1
-      ras_sub[is.na(ras_sub)] <- 0
-      return(ras_sub)})
-  }else{
-    parallel::clusterExport(cl, "inp")
-    parallel::clusterExport(cl, "ras")
-    parallel::clusterEvalQ(cl, library(raster))
-    parallel::clusterEvalQ(cl, library(sp))
-
-    rast <- parallel::parLapply(cl,
-                                  inp,
-                                  function(k){
-                                    po <- sp::SpatialPoints(k[,c("decimallongitude", "decimallatitude")],
-                                                        CRS("+proj=longlat +datum=WGS84"))
-                                    ras_sub <- raster::rasterize(po, ras, fun = "count")
-                                    ras_sub[ras_sub >= 1] <- 1
-                                    ras_sub[is.na(ras_sub)] <- 0
-                                    return(ras_sub)})
-    parallel::stopCluster(cl)
-  }
-
-  rast <- Reduce("+", rast)
-  rast[rast == 0] <- NA
-
-  return(rast)
-}
-
-#get vector with empirical distribution and observed distribution from occurrence points and distance rasters
-.DisVect <- function(x, dist, convexhull = c(F,T), binsize){
-  #x = the occurence points as spatial points object
-  #dis.ras = the distnace rasters, as 1 in the output of getDistance
-
[…]

### 2019-12-13 R/internal_likelihood_calculation.R

@@ -1,160 +1,0 @@
-#'@importFrom stats dexp median
-#'@importFrom graphics par plot
-#'@importFrom grDevices dev.off pdf
-
-.RunSampBias <- function(dists, distmats, biasdist = c(0, 10), plotextra = F, plotextrafile = "samp_bias_extra_plots.pdf") {
-
-  names.dists <- names(dists)
-  if(is.null(names.dists)){
-    names(dists) <- paste("Unnamed", seq(1, length(dists)), sep = "")
-    names.dists <- names(dists)
-  }
-
-  if (plotextra == T) {
-    pdf(file = plotextrafile, width = 5 * 4 * 0.75, height = 5 * 4 * 0.75)
-    par(mfrow = c(length(names.dists), 4))
-  }
-
-  res <- list()
-  for (i in 1:length(dists)) {
-    res[[i]] <- .CalcExpBias(dists = dists[[i]], Dmatrix = distmats[[i]], names.dists[i],
-                             rescale = 0, distvec = biasdist, plotextra = plotextra)
-  }
-
-  for (i in 1:length(names.dists)) {
-    b.matrix <- res[[i]][[3]]
-    if (i == 1) {
-      average.bias <- b.matrix
-    } else {
-      average.bias <- average.bias + b.matrix
-    }
-  }
-
-  average.bias.matrix <- average.bias/length(names.dists)
-
-  res[[length(dists) + 1]] <- list(bias_matrix = average.bias.matrix)
-  names(res) <- c(names.dists, "average_bias_matrix")
-
-
-  if (plotextra == T) {
-    n <- dev.off()
-  }
-  return(res)
-
-}
-
-# EXPONENTIAL MODEL
-# construct probability vector for all poits based on the empirical probaility distribution
-.ConstructPxVector <- function(x, X0, nP0, rescale, Px, plotextra) {
-  x <- sort(x)
-  if (plotextra == T) {
-    h.obs <- hist(x, breaks = c(0, X0), main = "Sampled distribution", xlab = "Rescaled distance")
-  } else {
-    h.obs <- hist(x, breaks = c(0, X0), plot = F)
-  }
-  obs.counts <- h.obs$counts  # length(obs.counts) = length(P0) = length(X0)
-  Px <- c()
[…]

### 2019-12-13 R/is.sampbias.R

@@ -9,8 +9,6 @@
 #' @param class2 the names of the class to which is relations are to be
 #' examined defined, or (more efficiently) the class definition objects for the
 #' classes.
-#' @seealso Check https://github.com/azizka/sampbias/wiki for a tutorial on
-#' sampbias.
 #' @keywords methods
 #' @examples
 #'
@@ -33,7 +31,7 @@
 #'
 #' gaz <- list(lines.strucutre = lin, point.structure = pts)
 #'
-#' out <- sampling_bias(x = occ, gaz = gaz, terrestrial = FALSE)
+#' out <- calculate_bias(x = occ, gaz = gaz, terrestrial = FALSE)
 #' is(out)
 #'
 #'@export

### 2019-12-13 R/map_bias.R

@@ -1,0 +1,153 @@
+#' Mapping Projected Bias Effects
+#'
+#'A plotting function to visualize the bias effects calculated using \code{\link{calculate_bias}}
+#'and projected using \code{\link{project_bias}}.
+#'
+#'
+#' @param x a raster stack as generate by \code{\link{project_bias}}
+#' @param gaz a list of SpatialObjects, to be printed on the maps.  Should be
+#' the same objects provided to \code{calculate_bias} when creating the Object.
+#' If \code{gaz} is not supplied, the sampbias package standard gazetteers are
+#' used.
+#' @param sealine logical. Should the sealine be added to the plots? Default is
+#' to TRUE.
+#' @param logtransform logical. IF TRUE, the bias projections are logtransformed. Default is TRUE.
+#' @return A series of R plots based on ggplot2.
+
+#' @seealso \code{\link{calculate_bias}}, \code{\link{project_bias}}
+#' @keywords methods
+#' @examples
+#'
+#' #simulate data
+#' occ <- data.frame(species = rep(sample(x = LETTERS, size = 5), times = 10),
+#'                   decimallongitude = runif(n = 50, min = -5, max = 5),
+#'                   decimallatitude = runif(n = 50, min = -4, max = 4))
+#'
+#'
+#' #create point gazetteer
+#' pts <- data.frame(long = runif(n = 5, min = -5, max = 5),
+#'                   lat = runif(n = 5, min = -4, max = 4),
+#'                   dat = rep("A", 5))
+#' pts <- SpatialPointsDataFrame(coords = pts[,1:2], data = data.frame(pts[,3]))
+#'
+#' lin <- data.frame(long = seq(-5, 5, by = 1),
+#'                   lat = rep(2, times = 11))
+#' lin <- SpatialLinesDataFrame(sl = SpatialLines(list(Lines(Line(lin), ID="B1"))),
+#'                              data = data.frame("B", row.names = "B1"))
+#'
+#' gaz <- list(lines.strucutre = lin, point.structure = pts)
+#'
+#' out <- calculate_bias(x = occ, gaz = gaz, terrestrial = FALSE)
+#' proj <- project_bias(out)
+#' map_bias(proj)
+#'
+#'@export
+#'@importFrom dplyr rename bind_cols filter
+#'@importFrom ggplot2 aes coord_fixed element_blank facet_wrap fortify geom_path geom_point geom_polygon geom_raster ggplot ggtitle labs theme theme_bw
+#'@importFrom raster as.data.frame coordinates
[…]

### 2019-12-13 R/plot.sampbias.R

@@ -1,24 +1,15 @@
-#' Mapping Occurrence and Bias Rasters in an Object of the Class Sampbias
+#' Plotting the Posterior Estimates of the Bias Weights
 #'
-#' Plotting method for class \code{sampbias}, generating a rasterized plots
-#' showing the number of occurrences and the number of species per gridcell,
-#' one bias raster indicating the effect strength for each gazetteer provided to
-#' \code{SamplingBias}, and a average bias raster indicating the average bias
-#' for each gridcell.
+#' Plotting method for class \code{sampbias}, generating a box-whiskers-plot
+#' showing the bias weights for all biasing factors
+#' indicating the effect strength for each gazetteer provided to
+#' \code{calculate_bias}.
 #'
 #'
 #' @param x an object of the class \code{sampbias}.
-#' @param gaz a list of SpatialObjects, to be printed on the maps.  Should be
-#' the same objects provided to \code{SamplingBias} when creating the Object.
-#' If \code{gaz} is not supplied, the sampbias package standard gazetteers are
-#' used.
-#' @param sealine logical. Should the sealine be added to the plots? Default is
-#' to TRUE.
-#' @param ...  Additional arguments to be passed to \code{\link{plot}}.
-#' @return A series of R plots based on ggplot2.
-#' @note Check https://github.com/azizka/sampbias/wiki for a tutorial on
-#' sampbias.
-#' @seealso \code{\link{SamplingBias}}, \code{\link{summary.sampbias}}
+#' @param \dots Additional arguments passed to summary.
+#' @return A plot
+#' @seealso \code{\link{calculate_bias}}, \code{\link{summary.sampbias}}
 #' @keywords methods
 #' @examples
 #'
@@ -41,123 +32,33 @@
 #'
 #' gaz <- list(lines.strucutre = lin, point.structure = pts)
 #'
-#' out <- sampling_bias(x = occ, gaz = gaz, terrestrial = FALSE)
+#' out <- calculate_bias(x = occ, gaz = gaz, terrestrial = FALSE)
 #' plot(out)
 #'
 #'@export
-#'@importFrom ggplot2 aes_string coord_fixed element_blank fortify geom_path geom_point geom_polygon geom_raster ggplot ggtitle labs theme theme_bw
-#'@importFrom raster as.data.frame crop extent
-#'@importFrom sp coordinates
[…]

### 2019-12-13 R/project_bias.R

@@ -1,0 +1,89 @@
+#' Projecting Bias Effects in Space
+#'
+#'
+#' Uses the the estimated bias weights from a \code{sampbias} object to project the
+#' bias through space, using the same raster as used for the distance calculation.#'
+#'
+#' @param x an object of the class \code{sampbias}.
+#' @param factors a character vector indicating which biasing factors to project
+#' @return A raster stack, with the same length as the number of biasing factors used
+#' in \code{\link{calculate_bias}}. The names indicate the factors included for each layer.
+#' @seealso \code{\link{calculate_bias}}, \code{\link{summary.sampbias}}
+#' @keywords methods
+#' @examples
+#'
+#' #simulate data
+#' occ <- data.frame(species = rep(sample(x = LETTERS, size = 5), times = 10),
+#'                   decimallongitude = runif(n = 50, min = -5, max = 5),
+#'                   decimallatitude = runif(n = 50, min = -4, max = 4))
+#'
+#'
+#' #create point gazetteer
+#' pts <- data.frame(long = runif(n = 5, min = -5, max = 5),
+#'                   lat = runif(n = 5, min = -4, max = 4),
+#'                   dat = rep("A", 5))
+#' pts <- SpatialPointsDataFrame(coords = pts[,1:2], data = data.frame(pts[,3]))
+#'
+#' lin <- data.frame(long = seq(-5, 5, by = 1),
+#'                   lat = rep(2, times = 11))
+#' lin <- SpatialLinesDataFrame(sl = SpatialLines(list(Lines(Line(lin), ID="B1"))),
+#'                              data = data.frame("B", row.names = "B1"))
+#'
+#' gaz <- list(lines.strucutre = lin, point.structure = pts)
+#'
+#' out <- calculate_bias(x = occ, gaz = gaz, terrestrial = FALSE)
+#' out <- project_bias(out)
+#'
+#'@export
+#'@importFrom magrittr %>%
+#'@importFrom raster stack as.matrix
+#'@importFrom rlang .data
+
+
+project_bias <- function(x, factors = NULL) {
+
+  # get raster stack
+  ras <- as.matrix(x$distance_rasters) %>% t() %>% data.frame()
+
+  # rescale as in the bias estimation
+  ras <- ras/x$summa$rescale_distances
+
+  # get mean posterior weights and sort according to importance
[…]

### 2019-12-13 R/sampling_bias.R

@@ -1,295 +1,0 @@
-#' Evaluating Sampling Bias in Species Distribution Data
-#'
-#' The major function of the package, calculating the bias effect of sampling
-#' bias due to geographic structures, such as the vicinity to cities, airports,
-#' rivers and roads. Results are projected to space, and can be compared
-#' numerically.
-#'
-#' The default gazetteers delivered with the package are simplified from
-#' http://www.naturalearthdata.com/downloads/. They include major features, and
-#' for small scale analyses custom gazetteers should be used.
-#'
-#' For computational convenience, the gazetteers are cropped to the extent of
-#' the point occurrence data sets. To account for the fact, that, relevant
-#' structures might lay directly outside this extent, but still influencing the
-#' distribution of samples in the study area, the buffer option, gives the
-#' area, around the extent that should be included in the distance calculation.
-#'
-#' The \code{biasdist} options defines at which distance the bias effect should
-#' be recorded. These values are comparable between different biasing factors
-#' and give an estimate of bias severity. For instance, if \code{biasdist} =
-#' c(0, 10000), the \code{biastable} will indicate the bias effect at distances
-#' 0 and 10000 meters from all sources of bias. The bias effect is the fraction
-#' of samples expected to be missed at a given distance relative to distance 0.
-#' A high bias effect means a high fraction of missed records. The fractions
-#' are comparable between sources of bias and different datasets and the bias
-#' at any given distance can thus inform on the relative severity of a biasing
-#' source.
-#'
-#' Visit \url{https://github.com/azizka/sampbias/wiki} for more information on
-#' distance calculation and the algorithm behind sampbias.
-#'
-#' @param x an object of the class \code{data.frame}, with one species
-#' occurrence record per line, and at least three columns, named
-#' \sQuote{species}, \sQuote{decimallongitude}, and \sQuote{decimallatitude}.
-#' @param gaz a list of geographic gazetteers as \code{SpatialPointDataFrame}
-#' or \code{SpatialLinesDataFrame}.  If NULL, a set of default gazetteers,
[…]

### 2019-12-13 R/summary.sampbias.R

@@ -6,7 +6,7 @@
 #' @param object An object of the class \code{sampbias}
 #' @param \dots Additional arguments passed to summary.
 #' @return Summary printed to screen.
-#' @seealso \code{\link{SamplingBias}} \code{\link{is.sampbias}}
+#' @seealso \code{\link{calculate_bias}} \code{\link{is.sampbias}}
 #' \code{\link{plot.sampbias}}
 #' @keywords methods
 #' @examples
@@ -30,19 +30,18 @@
 #'
 #' gaz <- list(lines.strucutre = lin, point.structure = pts)
 #'
-#' out <- SamplingBias(x = occ, gaz = gaz, terrestrial = FALSE)
+#' out <- calculate_bias(x = occ, gaz = gaz, terrestrial = FALSE)
 #' summary(out)
 #'
 #' @export
 #'
 summary.sampbias <- function(object, ...) {
   cat("Number of occurences: ", object$summa$total_occ, "\n")
-  cat("Number of species: ", object$summa$total_sp, "\n")
   cat("Raster resolution: ", object$summa$res, "\n")
-  cat("Distance binsize: ", object$summa$binsize, "\n")
   cat("Convexhull: ", object$summa$convexhull, "\n")
   cat("Geographic extent:\n")
   print(object$summa$extent)
-  cat("Bias effect at distance:\n")
-  print(object$biastable)
+  cat("Bias weights:\n")
+  print(data.frame(bias_weight = colMeans(object$bias_estimate[,-c(1:4)]),
+                   std_dev = apply(object$bias_estimate[,-c(1:4)], 2, "sd")))
 }

### 2019-12-13 empirical_analysis_manuscript.R

@@ -1,0 +1,17 @@
+library(sampbias)
+library(cowplot)
+
+occ <-read.csv(system.file("extdata", "mammals_borneo.csv", package="sampbias"), sep = "\t")
+
+out <- calculate_bias(occ, res = 0.1)
+
+p1 <- plot(out)
+summary(out)
+
+proj <- project_bias(out)
+p2 <- map_bias(proj)
+
+
+plot_out <- plot_grid(p1, p2, labels = c('A', 'B'), ncol = 1, rel_heights = c(1, 2))
+
+ggsave(plot_out, filename = "ms_figures/figure_empirical_results.jpg", height = 10, width = 8)

### 2019-12-13 shiny/app.R

@@ -1,100 +1,0 @@
-library(shiny)
-library(maptools)
-library(raster)
-library(ggplot2)
-library(viridis)
-library(rgeos)
-
-data("wrld_simpl")
-
-
-# Define UI for application 
-ui <- shinyUI(fluidPage(
-   
-   # Application title
-   titlePanel("Sampling bias in species distribution records"),
-   
-    
-   sidebarLayout(
-      sidebarPanel(
-
-        fileInput("file1", label = h3("Occurence File")),
-        
-        # actionButton("Load Example", label = "exmp"),
-
-        radioButtons("type", label = h3("Bias Type"),
-                     choices = list("None" = "non",
-                                    "Species" = "spe",
-                                    "Cities" = "cit", "Airports" = "air",
-                                    "Rivers" = "riv", "Roads" = "roa",
-                                    "Combined" = "comb"), selected = "non"),
-
-        numericInput("num", label = h3("Raster resolution [deg]"), value = 1)
-      ),
-      
-      # Show plots
-      mainPanel(
-         plotOutput("plo1"),
-         plotOutput("plo2")
-      )
-   )
-))
-
-# Define server logic required to draw a histogram
-source("helpers.R")
-
-server <- shinyServer(function(input, output) {
-
-  options(shiny.maxRequestSize=35*1024^2) # change maximum upload size
-  
-  #load the input data
-  dataInput <- reactive({
-    inFile <- input$file1
-
-    read.csv(inFile$datapath, sep = "\t")
-  })
-
-    #create a spatialPOints data.frame version of the input dataset
-  datPts <- reactive({
-   SpatialPoints(dataInput()[,c("decimallongitude", "decimallatitude")])
-  })
-
-  #this is the first plot
-   output$plo1 <- renderPlot({
-     
-     #do not display anything if no file is selected
-     if (is.null(input$file1)){
-       return(NULL)
-       }else{
-         load("default_gazetteers.R")
-         print(PlotOccRast(datPts(), res = input$num, gaz = gazetteers))
-     }
-   })
-   
-   output$plo2 <- renderPlot({
-     #do not display anything if no file is selected
-     if (is.null(input$file1)){return(NULL)
-       }else{
-     load("default_gazetteers.R")
-     dist <- switch(input$type,
-                    non = "NULL",
-                    spe = PlotSpRast(dataInput(), res = input$num, wrld = gazetteers$landmass),
-                    cit = BiasPlot(x = dataInput(), gaz = list(cities = gazetteers$cities), 
[…]

### 2019-12-13 shiny/helpers.R

@@ -1,637 +1,0 @@
-
-PlotOccRast <- function(x, res, gaz = NULL){
-    r <- raster(extent(x))
-    res(r) <- res
-    
-    rast <- rasterize(x, r, fun = "count")
-    rast <- data.frame(coordinates(rast), as.data.frame(rast))
-    colnames(rast) <- c("Longitude", "Latitude", "val")
-    
-    if(!is.null(gaz)){
-    gaz.plo <- lapply(gaz, function(k) raster::crop(x = k, y = extent(x)))
-    gaz.plo[[1]] <- data.frame(coordinates(gaz.plo[[1]]))
-    gaz.plo[[2]] <- data.frame(coordinates(gaz.plo[[2]]))
-    gaz.plo[3:5] <- lapply(gaz.plo[3:5], "fortify")
-    }
-
-    out <- ggplot()+
-      geom_raster(data=rast, aes(y=Latitude, x=Longitude, fill=val))+
-      coord_fixed()+
-      theme_bw()+
-      scale_fill_viridis(na.value = "transparent", 
-                         option = "viridis", 
-                         direction = -1, name = "Occurrence\nnumber", discrete = F)+
-      
-      theme(panel.border = element_rect(fill = NA))
-    
-    if(!is.null(gaz)){
-      out <- out+
-        ggplot2::geom_path(mapping = aes_string(x = "long", y = "lat", group = "group"),
-                           linetype = 1,
-                           data = gaz.plo$landmass, col = "grey40")+
-        ggplot2::geom_path(mapping = aes_string(x = "long", y = "lat", group = "group"),
-                           linetype = 1, data = gaz.plo$roads, col = "grey40")+
-        ggplot2::geom_path(mapping = aes_string(x = "long", y = "lat", group = "group"),
-                           linetype = 2.5,
-                           data = gaz.plo$rivers, col = "grey40")+
-        ggplot2::geom_point(mapping = aes_string(x = "longitude", y = "latitude", colour = "Airports"),
-                            shape = 16, size = 2.5, data = gaz.plo$airports, col = "grey10")+
-        ggplot2::geom_point(mapping = aes_string(x = "longitude", y = "latitude", colour = "Cities"),
-                            shape = 17, size = 1.5, data = gaz.plo$cities, col = "grey10")+
[…]

### 2019-12-13 vignettes/Using_the_samp_bias_rpackage.R

@@ -1,34 +1,0 @@
-## ----setup, include=FALSE------------------------------------------------
-knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
-
-## ---- eval = FALSE-------------------------------------------------------
-#  require(devtools)
-#  install_github(repo = "azizka/sampbias")
-#  library(sampbias)
-#  library(maptools)
-
-## ---- echo = FALSE-------------------------------------------------------
-library(sampbias)
-library(maptools)
-
-## ------------------------------------------------------------------------
-#loading a text file to R
-occ <-read.csv(system.file("extdata", "mammals_borneo.csv", package="sampbias"), sep = "\t")
-
-## ------------------------------------------------------------------------
-cit <- readShapeSpatial(system.file("extdata", "Borneo_major_cities.shp", package="sampbias"))
-roa <- readShapeSpatial(system.file("extdata", "Borneo_major_roads.shp", package="sampbias"))
-
-gazetteers <- list(cities = cit,
-                   roads = roa)
-
-## ------------------------------------------------------------------------
-bias.out <- SamplingBias(x = occ, gaz = gazetteers)
-
-## ------------------------------------------------------------------------
-summary(bias.out)
-#plot(bias.out)
-
-## ------------------------------------------------------------------------
-#plot(bias.out, gaz = gazetteers)
-

### 2019-12-15 manuscript.rmd
further changes
added the new Poisson model + Bayesian algorithm

@@
 The resolution of the grid defined the precision of the distance estimates, for instance a 1x1 degree raster will yield approximately a 100km precision at the equator.
+For each cell $i$, we compute a vector $X_i(j)$ of minimum distances to each source of bias $j \in B$.
 
-## Quantifying accessibility bias using maximum likelihood
+## Quantifying accessibility bias using a Bayesian framework
+We describe the observed number of sampled occurrences $S_i$ within each cell $i$ as the result of Poisson sampling process with rate $\lambda_i$.
+We model the rate $\lambda_i$ as a function of a constant $q$, which represent the expected number of occurrences per cell in the absence of biases, i.e.
+when $\sum_{j=1}^{B}{ X_i(j)} = 0$.
+Additionally, $\lambda_i$ is modeled to decrease exponentially as a function of distance from sources of bias, such that increasing distances will result in a lower sampling rate.
+For a single source of bias the rates of cell $i$ with distance $X_i$ from a bias is:
+$$
+\lambda_i = q \times \exp \left(- w X_i \right)
+$$
+where $w \in \mathbb{R^+}$ defines the steepness of the Poisson rate decline, such that $w \approx 0$ results in a null model of uniform sampling rate $q$ across cells.
+
+In the presence of multiple bias predictors, the sampling rate decrease is a function of the cumulative effects of each bias and its distance from the cell:
+$$
+\lambda_i = q \times \exp \left(-\sum_{j=1}^{B}{w_j X_i(j)} \right) \ \ \ \ (1)
+$$
+where a vector $\mathbf{w} = [w_1, ..., w_B]$ describes the amount of bias attributed to each specific predictor.
+
+To quantify the amount of bias associated with each predictor, we jointly estimate the parameters $q$ and $\mathbf{w}$ in a Bayesian framework.
+We use Markov Chain Monte Carlo (MCMC) to sample these parameters from their posterior distribution:
+$$
+P(q, \mathbf{w} | \mathbf{S}) \propto \prod_{i=1}^{N}{ Poi(S_i | \lambda_i) } \times P(q) P(\mathbf{w}) \ \ \ \ (2)
+$$
+where the likelihood of sampled occurrences $S_i$ within each cell $Poi(S_i | \lambda_i)$ is the probability mass function of a Poisson distribution with rate per cell defined as in Eqn.
+(1).
+The likelihood is then multiplied across the $N$ cells considered.
+We used exponential priors on the parameters $q$ and $\mathbf{w}$, $P(q) \sim \Gamma(1, 0.01)$ and $P(\mathbf{w}) \sim \Gamma(1, 1)$, respectively.
+
+We summarize the parameters by computing the mean of the posterior samples and their 95\% credible intervals.
+We interpret the magnitude of the elements in $\mathbf{w}$ as a function of the importance of the individual biases.
+We note however although that this test is not explicitly intended to assess the significance of each bias predictor (for which a Bayesian variable selection could be used), particularly since several sources of bias might b correlated (e.g.
+cities, and airports).
+Instead, these aanlyses can be use to quantify the expected amount of bias in the data that can be predicted by single or multiple predictors in order to identify under-sampled and unexplored areas.
+
+[[SHOULD WE REMOVE BASICALLY ALL THE FOLLOWING SECTIONS?]]
+
 Given the placement of a particular bias sources in the area of interest and assuming a uniform distribution of samples, the probability of a sampled occurrence located at a distance $d$ from the closest bias source is a function of the amount of available area at that distance.
 That is, the larger the area located at distance $d$ from a bias source the more samples we expect.
-For simplicity we discretize the area of interest in a number of grid cells and indicate with $f(x)$ the function describing the number of available grid cells at any distance $x$, for $0 < x < \max(x)$, where $\max(x)$ is the maximum observed distance between a cell and the clostest bias source.
+For simplicity we discretize the area of interest in a number of grid cells and indicate with $f(x)$ the function describing the number of available grid cells at any distance $x$, for $0 < x < \max(x)$, where $\max(x)$ is the maximum observed distance between a cell and the closest bias source.
 The function $f(x)$ is therefore calculated based on the distances of each grid cell from its closest bias source.
@@
 #running sampbias
-example.out <- SamplingBias(x = example.in, res = 0.1)
+example.out <- sampling_bias(x = example.in, res = 0.1)
 

### 2019-12-16 manuscript.rmd
more updates2

--- 2019-12-04 manuscript.rmd
@@
     \fancyhead[L]{sampbias accessibility bias}
-    \fancyhead[R]{ } 
+    \fancyhead[R]{\thepage}
+    \fancyfoot{}
     \usepackage{float}
@@
 scientific expeditions, census counts, genetic barcoding studies, and citizen-science observations),  thus becoming subject to collection biases [@Meyer2016].
+
 That is the number of data points available is biased by factors other than species' presence or abundance, including the under-sampling of specific taxa ("taxonomic bias", e.g., birds *vs.* nematodes), specific geographic regions ("geographic bias", i.e.
@@
 A default *sampbias* analysis can be run with few lines of code in R.
-The main function `SamplingBias` creates an object of the class `"sampbias"`, for which the package provides a plotting and summary method.
+The main function `calculate_bias` creates an object of the class `"sampbias"`, for which the package provides a plotting and summary method.
 Based on a `data.frame` including species identity and geographic coordinates, sampbias provides a bias effect estimate for each gazetteer and an average bias.
@@
 
-```{r, eval = FALSE, tidy=TRUE}
+```{r, eval = FALSE}
 library(sampbias)
@@
 #running sampbias
-example.out <- SamplingBias(x = example.in, res = 0.1)
+example.out <- calculate_bias(x = example.in, res = 0.1)
 
@@
 plot(example.out)
+
+#project in space
+proj <- project_bias(example.out)
+map_bias(proj)
 
@@
 The software presented here is available under a GPL-3 license.
-The *sampbias* R package and the source code for the shiny app are available via https://github.com/azizka/sampbias, the shiny app can be accessed at https://azizka.shinyapps.io/sampbias/.
-The R package includes an example dataset as well as vignettes detailing the use of the R package, the use of the shiny app and possibly warnings produced by the package (Appendix S3).
+The *sampbias* R package and the source code for the shiny app are available via https://github.com/azizka/sampbias.
+The R package includes an example dataset as well as vignettes detailing the use of the R package, the use of the shiny app and possibly warnings produced by the package (Appendix S2).
 
@@
 # Figures
-
-<!-- Figure 1 A sketch of the different distributions -->
-```{r distdi, out.width = "7in", echo = FALSE, fig.cap = "Schematic representation of the \\textit{sampbias} algorithm.
-The upper panel (A-C) shows a case without accessibility bias: the observed distribution of distances from the geographic feature of interest matches the expected distribution and hence the bias function is flat.
-The lower panel (D-F) shows a case with bias: the expected and observed distribution differ and the bias is best represented by an exponential function.", eval = TRUE}
-knitr::include_graphics(path = "ms_figures/Distance_distribution_example.png")
-```
-
-<!-- Figure 2 shiny interface -->
-```{r GUI, out.width = "7in", echo = FALSE, fig.cap = "The interface of the sampbias shiny app. With a navigation panel on the left and visualization of the results on the right.", eval = TRUE}
-knitr::include_graphics(path = "ms_figures/bild_gui_sampbias.png")
-```
-
-<!-- Figure 3 empirical results -->
-```{r empirical, out.width = "7in", echo = FALSE, fig.cap = "The spatial projection of the accessibility bias in an empirical example dataset of mammal occurrences on the Indonesian island of Borneo from www.gbif.org.
-\\textit{Sampbias} finds a strong biasing effect by roads, cities and lower by airports, but, surprisingly, no effect of rivers.", eval = TRUE}
-knitr::include_graphics(path = "ms_figures/visualization_rpackage_application_githubtutorial.jpg")
+<!-- mpirical results -->
+```{r empirical, out.width = "6in", echo = FALSE, fig.cap = "The spatial projection of the accessibility bias in an empirical example dataset of mammal occurrences on the Indonesian island of Borneo from www.gbif.org.
+A) bias weights, B) projection of the expected number of occurrences given the sampbias model.
+\\textit{Sampbias} finds the strongest biasing effect for cities.", eval = TRUE}
+knitr::include_graphics(path = "ms_figures/figure_empirical_results.jpg")
 ```
@@
 
-Appendix S2 - Tutorial running the sampbias shiny app
-
-Appendix S3 - Possible warnings and their solutions
+Appendix S2 - Possible warnings and their solutions
 

### 2019-12-16 R/calculate_bias.R

@@ -97,7 +97,7 @@
 #'
 #' lin <- data.frame(long = seq(-5, 5, by = 1),
 #'                   lat = rep(2, times = 11))
-#' lin <- sp::SpatialLinesDataFrame(sl = SpatialLines(list(Lines(Line(lin), ID="B1"))),
+#' lin <- sp::SpatialLinesDataFrame(sl = sp::SpatialLines(list(sp::Lines(sp::Line(lin), ID="B1"))),
 #'                              data = data.frame("B", row.names = "B1"))
 #'
 #' gaz <- list(lines.strucutre = lin, point.structure = pts)
@@ -108,7 +108,7 @@
 #' @export
 #' @importFrom raster as.raster extent extent<- getValues mask res<- values
 #' @importFrom grDevices is.raster
-#' @importFrom sp SpatialPointsDataFrame SpatialLinesDataFrame
+#' @importFrom sp SpatialPointsDataFrame SpatialLinesDataFrame SpatialLines Lines Line
 #' @importFrom rlang .data
 #' @importFrom stats complete.cases
 #'

### 2019-12-16 R/dis_rast.R

@@ -32,18 +32,18 @@
 #' @examples
 #'
 #' #create raster for resolution and extent
-#' ras <- raster(extent(-5,5,-4,4), res = 1)
+#' ras <- raster::raster(raster::extent(-5,5,-4,4), res = 1)
 #'
 #' #create point gazeteer
 #' pts <- data.frame(long = runif(n = 5, min = -5, max = 5),
 #'                   lat = runif(n = 5, min = -4, max = 4),
 #'                   dat = rep("A", 5))
 #'
-#' pts <- SpatialPointsDataFrame(coords = pts[,1:2], data = data.frame(pts[,3]))
+#' pts <- sp::SpatialPointsDataFrame(coords = pts[,1:2], data = data.frame(pts[,3]))
 #'
 #' lin <- data.frame(long = seq(-5, 5, by = 1),
 #'                   lat = rep(2, times = 11))
-#' lin <- SpatialLinesDataFrame(sl = SpatialLines(list(Lines(Line(lin), ID="B1"))),
+#' lin <- sp::SpatialLinesDataFrame(sl = sp::SpatialLines(list(sp::Lines(sp::Line(lin), ID="B1"))),
 #'                              data = data.frame("B", row.names = "B1"))
 #'
 #' gaz <- list(point.structure = pts, lines.strucutre = lin)
@@ -53,6 +53,7 @@
 #' \dontrun{plot(out[[1]])}
 #'
 #'@export
+#'@importFrom sp SpatialPointsDataFrame SpatialLinesDataFrame SpatialLines Lines Line
 #'@importFrom raster crop extent raster res
 #'
 dis_rast <- function(gaz, ras, buffer = NULL, ncores = 1) {

### 2019-12-16 R/internal.R

@@ -16,11 +16,20 @@
   }
 }
 
+# get_lambda_ij_old <- function(q, w, X) {
+#   if (class(X) == "numeric") {
+#     return(q * exp(-w * X))
+#   } else {
+#     return(q * exp(-apply(FUN = sum, w * X, 2)))
+#   }
+# }
+
+
 get_lambda_ij <- function(q, w, X) {
   if (class(X) == "numeric") {
     return(q * exp(-w * X))
   } else {
-    return(q * exp(-apply(FUN = sum, w * X, 2)))
+    return(as.vector(q * exp(-(w %*% t(X)))))
   }
 }
 
@@ -50,9 +59,9 @@ multiplier_proposal <- function(i, d = 1.2) {
 
   # X is the distance matrix with 1 row for each cell and 1 col for each predictor
   X <- x[, indx]/rescale_distances
-  if (length(indx) > 1) {
-    X <- t(X)
-  }
+  # if (length(indx) > 1) {
+  #   X <- t(X)
+  # }
 
   Xcounts <- x$record_count
 

### 2019-12-16 R/is.sampbias.R

@@ -33,7 +33,8 @@
 #'
 #' out <- calculate_bias(x = occ, gaz = gaz, terrestrial = FALSE)
 #' is(out)
-#'
+#'@method is sampbias
+#'@importFrom methods is
 #'@export
 #'
 is.sampbias <- function(object, class2){

### 2019-12-16 R/plot.sampbias.R

@@ -35,12 +35,15 @@
 #' out <- calculate_bias(x = occ, gaz = gaz, terrestrial = FALSE)
 #' plot(out)
 #'
+#'@method plot sampbias
 #'@export
 #'@importFrom forcats fct_reorder
 #'@importFrom ggplot2 ggplot geom_boxplot xlab ylab coord_flip theme_bw theme
+#'@importFrom graphics plot
 #'@importFrom magrittr %>%
 #'@importFrom dplyr contains mutate select
 #'@importFrom rlang .data
+#'@importFrom stats median
 #'@importFrom tidyr pivot_longer
 #'
 plot.sampbias <- function(x, ...) {

### 2019-12-16 R/project_bias.R

@@ -43,7 +43,8 @@
 project_bias <- function(x, factors = NULL) {
 
   # get raster stack
-  ras <- as.matrix(x$distance_rasters) %>% t() %>% data.frame()
+  # ras <- as.matrix(x$distance_rasters) %>% t() %>% data.frame()
+  ras <- as.matrix(x$distance_rasters) %>% data.frame()
 
   # rescale as in the bias estimation
   ras <- ras/x$summa$rescale_distances
@@ -52,22 +53,28 @@ project_bias <- function(x, factors = NULL) {
   mean_w <- colMeans(x$bias_estimate)[-c(1:4)] %>% sort() %>% rev()
 
   # sort ras as mean_w
-  ord <- match(gsub("w_", "", names(mean_w)), rownames(ras))
-  ras <- ras[ord,]
+  ord <- match(gsub("w_", "", names(mean_w)), colnames(ras))
+  ras <- ras[, ord]
 
   # use factors if provided to select certain biasing effects for projection
   if (!is.null(factors)) {
     mean_w <- mean_w[gsub("w_", "", names(mean_w)) %in% factors]
-    ras <- ras[rownames(ras) %in% factors,]
+    ras <- ras[,colnames(ras) %in% factors]
   }
 
   # calculate the values for each raster cell
   lambdas <- list()
 
   for (i in 1:length(mean_w)) {
+    if(is.numeric(ras)){
+      test <- ras
+    }else{
+      test <- ras[,1:i]
+    }
+
     lambdas[[i]] <- get_lambda_ij(q = colMeans(x$bias_estimate)[4],
                                   w = mean_w[1:i],
-                                  X = ras[1:i,]) %>%
+                                  X = test) %>%
       as.matrix()
   }
 

### 2019-12-16 R/summary.sampbias.R

@@ -32,7 +32,7 @@
 #'
 #' out <- calculate_bias(x = occ, gaz = gaz, terrestrial = FALSE)
 #' summary(out)
-#'
+#' @method summary sampbias
 #' @export
 #'
 summary.sampbias <- function(object, ...) {

### 2019-12-16 empirical_analysis_manuscript.R

@@ -3,12 +3,12 @@ library(cowplot)
 
 occ <-read.csv(system.file("extdata", "mammals_borneo.csv", package="sampbias"), sep = "\t")
 
-out <- calculate_bias(occ, res = 0.1)
+out <- calculate_bias(occ, res = 1)
 
 p1 <- plot(out)
 summary(out)
 
-proj <- project_bias(out)
+proj <- project_bias(out, factors = "roads")
 p2 <- map_bias(proj)
 
 

### 2019-12-16 vignettes/Using_the_samp_bias_rpackage.R

@@ -1,0 +1,41 @@
+## ----setup, include=FALSE------------------------------------------------
+knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE)
+
+## ---- eval = FALSE-------------------------------------------------------
+#  require(devtools)
+#  install_github(repo = "azizka/sampbias")
+#  library(sampbias)
+#  library(maptools)
+
+## ---- echo = FALSE-------------------------------------------------------
+library(sampbias)
+library(maptools)
+
+## ------------------------------------------------------------------------
+#loading a text file to R
+occ <-read.csv(system.file("extdata", 
+                           "mammals_borneo.csv", 
+                           package="sampbias"), sep = "\t")
+
+## ------------------------------------------------------------------------
+cit <- readShapeSpatial(system.file("extdata", 
+                                    "Borneo_major_cities.shp", 
+                                    package="sampbias"))
+roa <- readShapeSpatial(system.file("extdata", 
+                                    "Borneo_major_roads.shp", 
+                                    package="sampbias"))
+
+gazetteers <- list(cities = cit,
+                   roads = roa)
+
+## ------------------------------------------------------------------------
+bias.out <- calculate_bias(x = occ, gaz = gazetteers)
+
+## ---- eval = FALSE-------------------------------------------------------
+#  summary(bias.out)
+#  plot(bias.out)
+
+## ------------------------------------------------------------------------
+proj <- project_bias(bias.out)
+map_bias(proj)
+

### 2019-12-16 manuscript.rmd

--- 2019-12-15 manuscript.rmd
@@
 #running sampbias
-example.out <- sampling_bias(x = example.in, res = 0.1)
+example.out <- calculate_bias(x = example.in, res = 0.1)
 

### 2019-12-17 manuscript.rmd
mansucript adaption to ecography guidelines

@@
 title: "Sampbias, a method to evaluate geographic sampling bias in species distribution data"
-author: "Alexander Zizka^1,2,3^, Alexandre Antonelli^3,4,5^, Daniele Silvestro^3,4^"
+#author: "Alexander Zizka^1,2,3^, Alexandre Antonelli^3,4,5^, Daniele Silvestro^3,4^"
 bibliography: sampbias.bib
+csl: ecography.csl
 output: 
   bookdown::pdf_document2:
-  # word_document:
+  #word_document:
     toc: false
@@
 
-1. German Center for Integrative Biodiversity Research, University of Leipzig, Leipzig, Germany
-2. Naturalis Biodiversity Center, Leiden University, Leiden, The Netherlands 
-3. Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden
-4. Department of Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden
-5. Royal Botanical Gardens Kew, Richmond, Surrey, United Kingdom
+<!-- 1. German Center for Integrative Biodiversity Research, University of Leipzig, Leipzig, Germany -->
+<!-- 2. Naturalis Biodiversity Center, Leiden University, Leiden, The Netherlands  -->
+<!-- 3. Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden -->
+<!-- 4. Department of Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden -->
+<!-- 5. Royal Botanical Gardens Kew, Richmond, Surrey, United Kingdom -->
 
@@
 ```
+
+<!-- For Ecography SN, max 4000 (-6000?) words, max 2 figures -->
 
@@
 Yet, no general methodology exists to quantify the effect of roadside or other sources of bias on a dataset level.
-Here we present *sampbias*, a novel algorithm and software to estimate the biasing effect of accessibility (by roads, rivers, airports, cities, or any user-defined structures) in species occurrence datasets.
-*Sampbias* is based on a null model of even sampling and assesses whether instead sampling probability decays exponentially with distance.
+Here we present *sampbias*, an algorithm and software to estimate the biasing effect of accessibility (by roads, rivers, airports, cities, or any user-defined structures) in species occurrence datasets.
+*Sampbias* uses a Bayesian approach based on MCMC to optimize the rate of a Poisson sampling process, with sampling dependent on the distance from the next biasing structure.
 The results are comparable among biasing factors and datasets.
-*Sampbias* is implemented as a user-friendly R package, and Shiny app.
+*Sampbias* is implemented as a user-friendly R package.
 We exemplify the use of *sampbias* on a dataset of mammal occurrences from the Indonesian island of Borneo, downloaded from www.gbif.org.
-*Sampbias* offers an efficient and largely automated means for biodiversity scientists and non-specialists alike to explore bias in species occurrence data.
-The output of *sampbias* may be used to identify priorities for further collection or digitalization efforts, provide bias surfaces for species distribution modelling, or assess the reliability of scientific results based on publicly available species distribution data.
 
@@
 
-# Introduction
-Publicly available datasets of geo-referenced species occurrences, such as provided by the Global Biodiversity Information Facility (www.gbif.org) have become a fundamental resource  in biological sciences, especially in biogeography, conservation, macroecology, and systematics.
-However, because these datasets are "presence-only" data, they rarely include information on collection effort.
-Instead they are typically not collected systematically and often compiled from a variety of sources (e.g.
-scientific expeditions, census counts, genetic barcoding studies, and citizen-science observations),  thus becoming subject to collection biases [@Meyer2016].
+# Background
+Publicly available datasets of geo-referenced species occurrences, such as provided by the Global Biodiversity Information Facility (www.gbif.org) have become a fundamental resource  in biological sciences, especially in biogeography, conservation, and macroecology.
+However, these datasets are "presence-only" data and rarely include information on collection effort.
+Instead, they are typically not collected systematically and often compiled from a variety of sources (e.g.
+scientific expeditions, census counts, genetic barcoding studies, and citizen-science observations),  thus becoming subject to sampling bias [@Meyer2016].
 
-That is the number of data points available is biased by factors other than species' presence or abundance, including the under-sampling of specific taxa ("taxonomic bias", e.g., birds *vs.* nematodes), specific geographic regions ("geographic bias", i.e.
+Likely, the number of data points available in such datasets is biased by factors other than species' presence or abundance, including the under-sampling of specific taxa ("taxonomic bias", e.g., birds *vs.* nematodes), specific geographic regions ("geographic bias", i.e.
 easily accessible *vs.* remote areas), and specific temporal periods ["temporal bias", i.e.
 wet season *vs.* dry season, @Isaac2015; @Boakes2010].
-While these biases are broadly recognized, and approaches exist to account for them in some analyses (for instance for species-richness estimation [@Engemann2015] species distribution modelling [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014], occupancy modelling [@Kery2016], or abundance estimations [@Shimadzu2015]), few attempts have been made to discern among different sources of bias or to compare the strength of bias among datasets [but see @Ruete2015].
-
-Geographic sampling bias, the fact that sampling effort is spatially biased, rather than equally distributed over a given study area is prevalent in all non-systematically collected datasets of species distributions.
-Many factors can affect sampling effort, such as socio-economic factors [i.e.
-national research spending, history of scientific research; @Meyer2015, @Daru2018] and political factors [armed conflict, democratic rights; @Ryden2019] or physical accessibility [i.e.
+Geographic sampling bias---the fact that sampling effort is spatially biased, rather than equally distributed over a given study area---is prevalent in all non-systematically collected datasets of species distributions.
+Many factors can can cause sampling bias, including socio-economic factors [i.e.
+national research spending, history of scientific research; @Meyer2015; @Daru2018], political factors [armed conflict, democratic rights; @Ryden2019], and physical accessibility [i.e.
 distance to a road or river, terrain conditions, slope; @Yang2014; @Botts2011].
 Especially physical accessibility is omnipresent as a biasing factor [e.g.
-@Lin2015; @Engemann2015], across spatial scales, and the term "roadside bias" has been coined for it.
-In practice, this means that most species observations (occurrence points) are made in or near cities, along roads and rivers, and near other human settlements (such as airports).
+@Lin2015; @Engemann2015; @Kadmon2004], across spatial scales, and the term "roadside bias" has been coined for it.
+In practice, this means that most species observations  are made in or near cities, along roads and rivers, and near other human settlements.
 Less observations come from the middle of a tropical rainforest or from a mountain top.
@@
 
-The implications of not considering spatial collection bias in biodiversity research are likely to be substantial [@Meyer2016; @Rocchini2015; @Shimadzu2015;  @Yang2013; @Kramer-Schadt2013; @Barbosa2013].
+The implications of not considering geographic sampling bias in biodiversity research are likely to be substantial [@Meyer2016; @Rocchini2015; @Shimadzu2015;  @Yang2013; @Kramer-Schadt2013; @Barbosa2013].
+While the presence of geographic sampling bias is broadly recognized [e.g.
+@Kadmon2004], and approaches exist to account for them in some analyses---for instance for species-richness estimates [@Engemann2015] species distribution models [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014; @Monsarrat2019], occupancy models [@Kery2016], or abundance estimates [@Shimadzu2015]---few attempts have been made to explicitly quantify the bias [@Kadmon2004; @Hijmans2000] or to discern among different sources of bias [@Fernandez2015; @Ruete2015; @Fithian2014], and to our knowledge, no tools exist for comparing the strength of accessibility bias among sources of bias or dataset.
 While it is unrealistic to expect that spatial biases in biodiversity data will ever disappear, it is crucial that researchers realise the intrinsic biases associated with the biodiversity data they are dealing with.
 This is the first step towards estimating to which extent these biases may affect their analyses, results, and conclusions drawn from such data.
-Therefore, it is advisable for any study dealing with species occurrence data to assess the strength of accessibility bias in the underlaying data.
+Therefore, it is advisable for any study dealing with species occurrence data to assess the strength of accessibility bias in the underlying data.
 
@@
 *Sampbias* is implemented as an R-package.
-Specifically, *sampbias* uses a null-model of random sampling to address two questions:
+Specifically, *sampbias* uses a Bayesian approach based on a Poisson process to address three questions:
 
@@
 
-2) How important are different means of human accessibility, such as to airport, cities, rivers or roads, in causing this bias?
+2) How important are different means of human accessibility, such as to airports, cities, rivers or roads, in causing this bias?
 
@@
 
-# Description
+*Sampbias* is implemented in R [@rcoreteam2019], based on commonly usedpackages for data handling  [`ggplot`, @Wickham2009; `tidyr`, @Wickham2019a; `dplyr`, @Wickham2019; `forcats`, @Wickham2019b; `magrittr`, @Bache2014; `viridis`, @Garnier2018], handling geographic information and geo-computation [`raster`, @Hijmans2019; `sp`, @Pebesma2005; @Bivand2013] and statistical modelling [`stats`, @rcoreteam2019].
+*Sampbias* offers an easy and largely automated means for biodiversity scientists and non-specialists alike to explore bias in species occurrence data and may be used to identify priorities for further collection or digitalization efforts, provide bias surfaces for species distribution modelling, or assess the reliability of scientific results based on publicly available species distribution data.
+
+# Methods and Features
 ## General concept
-Under the assumption that organisms exist across the entire area of interest, we can expect the number of sampled occurrences to be distributed uniformly in space (even though, of course, the density of individuals and the species composition may be heterogeneous).
-[[*I think we should acknowledge here that this assumption is valid when looking at a geographically restricted area, eg within a tropical forest.
-Of course different biomes eg forest, alpine, oceanic will result in different carrying capacity*]]
-With *sampbias* we assess if a set of occurrences significantly departs from a null uniform distribution and whether these discrepancies between expected and observed distributions can be explained by distance from factors that potentially bias their sampling probability (e.g.
-distance from cities or roads).
+Under the assumption that organisms exist across the entire area of interest, we can expect the number of sampled occurrences in a restricted areas, such as a single biome, to be distributed uniformly in space (even though, of course, the density of individuals and the species composition may be heterogeneous).
+With *sampbias* we assess if discrepancies in sampling can be explained by distance from factors that potentially bias their sampling probability (e.g.
+cities or roads).
+Due to the assumption of homogeneous sampling and a computational trade-off between the resolution of the distance raster and the extent of the study area (for instance, a 1000m resolution for a global dataset would lead to the generation of grid  for which distance calculation will become computationally prohibitive in most practical cases), *sampbias* is best suited for local or regional datasets at high resolution (c.
+100 -- 10,000m).
 
@@
 The output of the package includes measures of bias effect, which are comparable between different gazetteers (e.g.
-comparing biasing effect of roads and rivers), different taxa (e.g.
+comparing the biasing effect of roads and rivers), different taxa (e.g.
 birds *vs.* flowering plants) and different data sets (e.g.
@@
 
-CoordinateCleaner is implemented in R [@rcoreteam2019] based on standard tools for spatial statistics: ggplot2 [@Wickham2009], geosphere [@Hijmans2019], maptools [@Bivand2019], raster [@Hijmans2019], sp [@Pebesma2005; @Bivand2013], and viridis [@Garnier2018].
-
 ## Distance calculation
 *Sampbias* uses gazetteers of the geographic location of bias sources (e.g.
-roads) to generate a grid across the study area (the geographic extent of the dataset) for each gazetteer and then calculates the distance ("as the crow flies")
-[[*meaning?*]]
-of the midpoint of each grid cell to the closest cell containing an instance of the gazetteer.
-We then use these distance grids to sample the distribution of distances in the observed dataset and the null distribution in a reference dataset of equal size with randomly distributed records (the null model).
+roads) to generate a regular grid across the study area (the geographic extent of the dataset).
+For each grid cell $i$, we then compute a vector $X_i(j)$ of minimum distances ("as the crow flies") to each source of bias $j \in B$.
+We then use these distance grids to sample the distribution of distances in the observed dataset.
 The resolution of the grid defined the precision of the distance estimates, for instance a 1x1 degree raster will yield approximately a 100km precision at the equator.
-For each cell $i$, we compute a vector $X_i(j)$ of minimum distances to each source of bias $j \in B$.
 
 ## Quantifying accessibility bias using a Bayesian framework
-We describe the observed number of sampled occurrences $S_i$ within each cell $i$ as the result of Poisson sampling process with rate $\lambda_i$.
+We describe the observed number of sampled occurrences $S_i$ within each cell $i$ as the result of a Poisson sampling process with rate $\lambda_i$.
 We model the rate $\lambda_i$ as a function of a constant $q$, which represent the expected number of occurrences per cell in the absence of biases, i.e.
 when $\sum_{j=1}^{B}{ X_i(j)} = 0$.
-Additionally, $\lambda_i$ is modeled to decrease exponentially as a function of distance from sources of bias, such that increasing distances will result in a lower sampling rate.
+Additionally, we model $\lambda_i$ to decrease exponentially as a function of distance from sources of bias, such that increasing distances will result in a lower sampling rate.
 For a single source of bias the rates of cell $i$ with distance $X_i$ from a bias is:
@@
 where $w \in \mathbb{R^+}$ defines the steepness of the Poisson rate decline, such that $w \approx 0$ results in a null model of uniform sampling rate $q$ across cells.
-
 In the presence of multiple bias predictors, the sampling rate decrease is a function of the cumulative effects of each bias and its distance from the cell:
@@
 
-We summarize the parameters by computing the mean of the posterior samples and their 95\% credible intervals.
+We summarize the parameters by computing the mean of the posterior samples and their standard deviation.
 We interpret the magnitude of the elements in $\mathbf{w}$ as a function of the importance of the individual biases.
-We note however although that this test is not explicitly intended to assess the significance of each bias predictor (for which a Bayesian variable selection could be used), particularly since several sources of bias might b correlated (e.g.
+We note however that this test is not explicitly intended to assess the significance of each bias predictor (for which a Bayesian variable selection could be used), particularly since several sources of bias might be correlated (e.g.
 cities, and airports).
-Instead, these aanlyses can be use to quantify the expected amount of bias in the data that can be predicted by single or multiple predictors in order to identify under-sampled and unexplored areas.
+Instead, these analyses can be use to quantify the expected amount of bias in the data that can be predicted by single or multiple predictors in order to identify under-sampled and unexplored areas.
 
-[[SHOULD WE REMOVE BASICALLY ALL THE FOLLOWING SECTIONS?]]
+[**Daniele some text here on the projection of bias through time, and what the plots exactly show**]
 
-Given the placement of a particular bias sources in the area of interest and assuming a uniform distribution of samples, the probability of a sampled occurrence located at a distance $d$ from the closest bias source is a function of the amount of available area at that distance.
-That is, the larger the area located at distance $d$ from a bias source the more samples we expect.
-For simplicity we discretize the area of interest in a number of grid cells and indicate with $f(x)$ the function describing the number of available grid cells at any distance $x$, for $0 < x < \max(x)$, where $\max(x)$ is the maximum observed distance between a cell and the closest bias source.
-The function $f(x)$ is therefore calculated based on the distances of each grid cell from its closest bias source.
-
-The distribution of samples, in the absence of bias, should therefore represent a random sample from $f(x)$ and reflect its shape, i.e.
-$d ~ f(x)$ (Fig.
-\ref{fig:distdi}).
-However, in the presence of a bias, we expect the probability of finding occurrences to decrease with increasing distance from a bias source.
-This in turn will alter the resulting distribution of samples that no longer match the expected distribution.
-Here, we model the effect of sampling bias by assuming that the probability of sampling an occurrence decreases exponentially with increasing distance (Fig.
-\ref{fig:distdi}), following the function $b(x,l) = l \exp(-lx)$, where $l$ is the rate parameter.
-Under these assumption the expected distribution of samples is given by $g(x,l) \propto f(x) b(x,l)$.
-
-[[*I wonder if we shouldn't use " $b(x,l) = \exp(-lx)$" so the Y-value is always 1 at a distance of 0.
-That would require normalizing the likelihood based on the integral of the curve though, as it will be different from 1* ]
-
-get the dsitribution from all grid cells and normalize by the number of all available grid cells, not downsampled to the same number of points ignore the starting at 1/intercept
-
-a possion likelihood where the rate i
-two paramters, 
-1. speed of decae
-2. how high is the bias at distance zero
-
- $b(x,l) = \exp(-lx)$
- $q * exp(-lambda x) / \int_0^\infty (q * exp(-lambda x))$
-
-Average the bias by Akaike wheights, or likelihood
-
-The rate parameter $l$ describes the strength of the bias effect.
-When  $l$ is large the expected probability of sampling occurrences decreases very quickly as you move away from a bias source (Fig.
-\ref{fig:distdi}).
-In contrast, when the $l$ parameter is small, the resulting exponential distribution effectively becomes more and more similar to a uniform distribution (Fig.
-\ref{fig:distdi}), indicating that increasing distance from e.g.
-a city does not affect the sampling probability.
-We treat $l$ as an unknown variable and estimate it using maximum likelihood from the data based on the probability density function described by $g(x,l)$.
-This essentially means finding the value of $l$ that best explains any discrepancies between the expected distribution ($f(x)$) and the observed occurrences.
-[[*I think this is in fact a posterior probability where f(x) is the empirical prior and b(x,l) the likelihood and we do a maximum a posteriori optimization*]]
-
-Once we have an estimated value of $l$, we can infer the expected accessibility bias as a function of distance using $b(x,l)$.
-Since the function $b(x,l)$ describes the exponentially decreasing sampling probability in relation to distance, we can define a standardized bias function as: $B(x,l) = 1 - b(x,l)/b(0,l)$,
-[[*Actually I think this is effectively equivalent to what I wrote above $b(x,l) = \exp(-lx)$*]]
-where  the level of bias is set to 0 at distance 0 from the bias source.
-The standardized bias function tends asymptotically to 1 as the distance tends to infinity.
-However, since in any area $max(x) << \infty$, for small values of $l$ (i.e.
-little or no bias), $B(x,l)$ will look essentially like a uniform distribution with values very close to 0.
-Large values of $l$ (i.e.
-strong bias) will instead result in a curve that quickly approaches 1 with increasing distance.
-The values provided by the standardized bias function can be interpreted as the proportion of occurrences that are missing from the sample, compared to the observed samples at distance 0.
-Thus, if for a given estimated $l$ we have $B(50,l) = 0.20$, we can expect that at 50 Km from e.g.
-a road the number of occurrences per grid cell will be about 80% of the occurrences sampled at distance 0 from the road, with 20% missing due to sampling bias.
-
-When running *sampbias*, we typically test different sources of biases, such as roads, cities, airports, etc.
-For each factor an independent expected distribution ($f(x)$) is computed and a parameter $l$ is estimated from $g(x,l)$.
-These estimates can be use to produce maps showing the intensity of potential biases across the area based on the standardized bias function.
-The bias values obtained from different sources can then be averaged in each grid cell to produce a map showing the combined effects of all sources.
-[[*yeah this part is still weird.
-Basically, if rivers are not explaining anything they should count nothing toward the combined estimate, whereas they do now.
-I think we could try to see if the max likelihoods are comparable and weight the average by that.*]]
-
-# Running sampbias
+## Example and Empirical analysis
 A default *sampbias* analysis can be run with few lines of code in R.
 The main function `calculate_bias` creates an object of the class `"sampbias"`, for which the package provides a plotting and summary method.
-Based on a `data.frame` including species identity and geographic coordinates, sampbias provides a bias effect estimate for each gazetteer and an average bias.
-Additionally some options exist to provide custom gazetteers, custom distances for the bias estimation, a custom grain size of the analysis, as well as some operators for the calculation of the bias distances.
+Based on a `data.frame` including species identity and geographic coordinates.
+Additionally, some options exist to provide custom gazetteers, custom distances for the bias estimation, a custom grain size of the analysis, as well as some operators for the calculation of the bias distances.
 A tutorial on how to use sampbias is available with the package and in the electronic supplement of this publication (Appendix S1).
+
+To exemplify the use and output of sampbias, we downloaded the occurrence records of all mammals available from the Indonesian island of Borneo [@gbifdoi2016], and ran *sampbias* as shown in the example code below, to test the biasing effect of airports, cities and roads in the dataset.
+The example dataset is provided with the *sampbias*.
+On a 0.05 degree scale (~ 5 km) we found
+(Fig. \ref{fig:empirical}).
 
@@
 #running sampbias
-example.out <- calculate_bias(x = example.in, res = 0.1)
+example.out <- calculate_bias(x = example.in, 
+                              res = 0.05,
+                              buffer = 0.5)
 
@@
 
-For data exploration we implemented the basic functionalities in a shiny app as graphical user interface (Fig.
-\ref{fig:GUI}).
-Analyses can be run based on a tab separated .txt file with occurrence information including the column headers "species", "decimallongitude" and "decimallatitude", as for instance files downloaded from www.gbif.org, using custom gazetteers.
-A tutorial on how to use the *sampbias* GUI is available online (https://ropensci.github.io/sampbias/) and in the electronic supplement of this publication (Appendix S2).
-
-# Empirical example
-To exemplify the use and output of sampbias, we downloaded the occurrence records of all mammals available from the Indonesian island of Borneo [@gbifdoi], and quantify the biasing effect of airports, cities and roads in the dataset.
-**Something on the results, also add a table** (Fig.
-\ref{fig:empirical})
-
-# Assumptions and future prospective
-Two assumptions of *sampbias* are a equal sampling of occurrence records across the study area as null model and an exponential increase of the biasing effect with distance from the gazetteers.
-We considered both acceptable approximations for the purpose of the package, but future expansions of *sampbias* could relax these assumptions, for instance by allowing other distance decay functions, such as gamma or Weibull distributions, and by changing the sampling scheme of the background points.
-The first steps towards these goals are already implemented in the current version of *sampbias* with the option to limit background points to a convex hull around the dataset or limiting background points to terrestrial surface.
-[[*I don't understand this part*]]
-
-A practical limitation of *sampbias* is the trade-off between the resolution of the grid for the distance calculation and the geographic extent of the dataset.
-For instance, a 100m resolution for a global dataset would lead to the generation of grid  for which distance calculation will become computationally prohibitive in most practical cases, Hence, *sampbias* is best suited for local or regional datasets at high resolution (c.
-100 -- 10,000m) or continental datasets at low resolution (c.
-10 -- 100km).
-
-# Todo
-re-run empirical analysis
-
-test units
-
 # Data accessibility
 The software presented here is available under a GPL-3 license.
-The *sampbias* R package and the source code for the shiny app are available via https://github.com/azizka/sampbias.
-The R package includes an example dataset as well as vignettes detailing the use of the R package, the use of the shiny app and possibly warnings produced by the package (Appendix S2).
+The *sampbias* R package is available https://github.com/azizka/sampbias.
+The R package includes an example dataset as well as vignetts detailing the use of the R package (Appendix S1),  and possibly warnings produced by the package (Appendix S2).
 
-# Acknowledgements
-We thank the organizers of the 2016 Ebben Nielsen challenge for inspiring and recognizing this research. We thank all data collectors and contributors to GBIF for their effort.
+<!-- # Acknowledgements -->
+<!-- We thank the organizers of the 2016 Ebben Nielsen challenge for inspiring and recognizing this research. We thank all data collectors and contributors to GBIF for their effort. -->
 
 # Author contributions
-All authors conceived of this study, AZ and DS developed the statistical algorithm, AZ and DS wrote the R-package and AZ the Shiny app, AZ and DS wrote the manuscript with contributions from AA.
+All authors conceived of this study, AZ and DS developed the statistical algorithm, AZ and DS wrote the R-package, AZ and DS wrote the manuscript with contributions from AA.
 
-# Figures
-<!-- mpirical results -->
+# Figure
+<!-- empirical results -->
 ```{r empirical, out.width = "6in", echo = FALSE, fig.cap = "The spatial projection of the accessibility bias in an empirical example dataset of mammal occurrences on the Indonesian island of Borneo from www.gbif.org.
 A) bias weights, B) projection of the expected number of occurrences given the sampbias model.
-\\textit{Sampbias} finds the strongest biasing effect for cities.", eval = TRUE}
+At the study scale of 0.05 degrees (~5 km)\\textit{Sampbias} finds the strongest biasing effect for the proximity of cities and roads, and a highest undersampling in the center of the island.", eval = TRUE}
 knitr::include_graphics(path = "ms_figures/figure_empirical_results.jpg")

### 2019-12-17 title_page.Rmd
mansucript adaption to ecography guidelines

---
title: "Sampbias, a method to evaluate geographic sampling bias in species distribution data"
# title: "A global Red List of orchids supports the value of automated conservation assessment"
author: "Alexander Zizka^1,2,3^, Alexandre Antonelli^3,4,5^, Daniele Silvestro^3,4^"
output: 
  bookdown::pdf_document2:
  #word_document:
    toc: false
    toc_depth: 1
    number_sections: false
header-includes: |
    \usepackage{fancyhdr}
    \pagestyle{fancy}
    \fancyhead[L]{sampbias}
    \fancyhead[R]{ }
    \pagenumbering{gobble}
linestretch: 1.3
fontsize: 12pt
---

<!-- Letter to conservation letters: 3000 words in length (excluding references, table and figure legends, and online Supporting Information), 40 references, 6 figures -->

<!-- Contributed paper to Conservation Biology, 3000-6000 words -->

<!-- Affiliations -->
1. German Center for Integrative Biodiversity Research, University of Leipzig, Leipzig, Germany

2. Naturalis Biodiversity Center, Leiden University, Leiden, The Netherlands

3. Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden

4. Department of Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden

5. Royal Botanical Gardens Kew, Richmond, Surrey, United Kingdom

**Running head** -- sampbias

**Word count** -- Abstract: XX; Total: XX

<!-- **Article type** -- Letter, 146 words in the abstract, 5400 words in the manuscript, 42 References, 2 tables, 4 Figures -->

**Corresponding author** -- alexander.zizka@idiv.de

<!-- **Authors email addresses** -- tiffany.knight@idiv.de, PVitt@chicagobotanic.org -->

**Keywords** Collection effort, Global biodiversity Information Facility (GBIF), Presence only data, Roadside bias, Sampling intensity

# Acknowledgements
We thank the organizers of the 2016 Ebben Nielsen challenge for inspiring and recognizing this research. We thank all data collectors and contributors to GBIF for their effort.

### 2019-12-17 R/is.sampbias.R

@@ -22,11 +22,11 @@
 #' pts <- data.frame(long = runif(n = 5, min = -5, max = 5),
 #'                   lat = runif(n = 5, min = -4, max = 4),
 #'                   dat = rep("A", 5))
-#' pts <- SpatialPointsDataFrame(coords = pts[,1:2], data = data.frame(pts[,3]))
+#' pts <- sp::SpatialPointsDataFrame(coords = pts[,1:2], data = data.frame(pts[,3]))
 #'
 #' lin <- data.frame(long = seq(-5, 5, by = 1),
 #'                   lat = rep(2, times = 11))
-#' lin <- SpatialLinesDataFrame(sl = SpatialLines(list(Lines(Line(lin), ID="B1"))),
+#' lin <- sp::SpatialLinesDataFrame(sl = sp::SpatialLines(list(sp::Lines(sp::Line(lin), ID="B1"))),
 #'                              data = data.frame("B", row.names = "B1"))
 #'
 #' gaz <- list(lines.strucutre = lin, point.structure = pts)

### 2019-12-17 empirical_analysis_manuscript.R

@@ -3,12 +3,20 @@ library(cowplot)
 
 occ <-read.csv(system.file("extdata", "mammals_borneo.csv", package="sampbias"), sep = "\t")
 
-out <- calculate_bias(occ, res = 1)
+
+# occ <- read.csv("original_data/mammals_borneo.csv", sep = "\t")
+#
+# occ <- select(occ, species, decimallatitude, decimallongitude, class)
+#
+# write.table(occ, file = "inst/extdata/mammals_borneo.csv", sep = "\t")
+# occ <-read.csv("inst/extdata/mammals_borneo.csv", sep = "\t")
+
+out <- calculate_bias(occ, res = 0.05, buffer = 0.5)
 
 p1 <- plot(out)
 summary(out)
 
-proj <- project_bias(out, factors = "roads")
+proj <- project_bias(out)
 p2 <- map_bias(proj)
 
 

### 2019-12-17 vignettes/Using_the_samp_bias_rpackage.R

@@ -15,7 +15,8 @@ library(maptools)
 #loading a text file to R
 occ <-read.csv(system.file("extdata", 
                            "mammals_borneo.csv", 
-                           package="sampbias"), sep = "\t")
+                           package="sampbias"),
+               sep = "\t")
 
 ## ------------------------------------------------------------------------
 cit <- readShapeSpatial(system.file("extdata", 

### 2019-12-18 appendix_S1.Rmd
empirical results and appendix S1

---
title: "Appendix S1 - Supplementary Figure"
subtitle: "Sampbias, a method to evaluate geographic sampling bias in species distribution data"
#author: "Alexander Zizka^1,2,3^, Alexandre Antonelli^3,4,5^, Daniele Silvestro^3,4^"
output: rmarkdown::pdf_document
header-includes: |
    \usepackage{float}
    \renewcommand{\thefigure}{S\arabic{figure}}
---

\newpage{}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
```

```{r, fig.height = 8, fig.cap="The example dataset of mammal occurrences from the island of Borneo, as downloaded from www.gbif.org (n = 6,262), and the geographic gazetters of main cities, roads, rivers and airports used for the sampbias analysis."}
#libraries
library(raster)
library(ggplot2)
library(magrittr)
library(sp)
library(dplyr)
library(viridis)

#load data
occ <-read.csv(system.file("extdata", "mammals_borneo.csv", package="sampbias"), sep = "\t") %>% 
  dplyr::select(latitude = decimallatitude, longitude = decimallongitude) %>% 
  mutate(type = "Mammal occurrence")

data(airports)
data(waterbodies)
data(cities)
data(roads)
data(landmass)

# prepare for plotting
airp <- airports %>% crop(SpatialPoints(occ[,2:1])) %>%  coordinates() %>%  data.frame() %>% mutate(type = "Airport")
cit <- cities %>% crop(SpatialPoints(occ[,2:1])) %>%  coordinates() %>%  data.frame() %>%  mutate(type = "City")
pts <- bind_rows(occ, cit, airp)

riv <- waterbodies %>% crop(SpatialPoints(occ[,2:1])) %>%  fortify() %>%  mutate(type = "River")
road <- roads %>% crop(SpatialPoints(occ[,2:1])) %>%  fortify() %>%  mutate(type = "Road")
lin <- bind_rows(riv, road)

lma <- landmass %>% crop(SpatialPoints(occ[,2:1])) %>%  fortify()

# plot
ggplot()+
  geom_polygon(data = lma,
               mapping = aes(x = long, y = lat, group = group), lwd = 0.5, col = "grey40", fill = "transparent")+
  geom_path(mapping = aes(x = long, y = .data$lat, group = .data$group, shape = type, color = type),
            data = lin)+
  geom_point(data = pts, mapping = aes(x = longitude, y = latitude, linetype = type, color = type), size = 3)+
  scale_color_viridis(discrete = TRUE)+
  xlab("Longitude")+
  ylab("Latitude")+
  coord_fixed()+
  theme_bw()+
  theme(
    legend.position = "bottom"
  )
```

### 2019-12-18 manuscript.rmd
empirical results and appendix S1

@@
     \usepackage[url=false]{biblatex}
+    \usepackage{setspace}
 linestretch: 2
@@
 
-[**Daniele some text here on the projection of bias through time, and what the plots exactly show**]
+[**Daniele some text here on the projection of bias through space, and what the plots exactly show**]
 
@@
 
-To exemplify the use and output of sampbias, we downloaded the occurrence records of all mammals available from the Indonesian island of Borneo [@gbifdoi2016], and ran *sampbias* as shown in the example code below, to test the biasing effect of airports, cities and roads in the dataset.
+To exemplify the use and output of *sampbias*, we downloaded the occurrence records of all mammals available from the Indonesian island of Borneo [n = 6,262, @gbifdoi2016], and ran *sampbias* using the default gazetteers as shown in the example code below, to test the biasing effect of the main airports, cities and roads in the dataset.
 The example dataset is provided with the *sampbias*.
-On a 0.05 degree scale (~ 5 km) we found
-(Fig. \ref{fig:empirical}).
+We found a strong effect of cities on sampling intensity, a moderate effect of roads and airports and no effect of rivers (Fig.
+\ref{fig:empirical}A).
+All models predict a low number of collection records in the centre of Borneo, which reflects the original data, and where accessibility means are low (Figure S1 in Appendix S1).
+We ran this example to illustrate the use of *sampbias* For detailed analyses or a smaller geographic scale, higher resolution gazetteers, including smaller roads and rivers  and a higher spatial resolution would be desirable.
+Results might change with increasing resolution, since roads and rivers might have a stronger effect on higher resolutions (facilitating most the access to their immediate vicinity), whereas cities and airports might have an effect on the larger scale (facilitating access to a larger area).
 
@@
 map_bias(proj)
-
 ```
@@
 The *sampbias* R package is available https://github.com/azizka/sampbias.
-The R package includes an example dataset as well as vignetts detailing the use of the R package (Appendix S1),  and possibly warnings produced by the package (Appendix S2).
+The R package includes an example dataset as well as vignettes detailing the use of the R package (Appendix S1),  and possibly warnings produced by the package (Appendix S2).
 
@@
 
-Appendix S1 - Tutorial running sampbias in R
+Appendix S1 - Supplementary Figure
 
-Appendix S2 - Possible warnings and their solutions
+Appendix S2 - Tutorial running sampbias in R
 
+Appendix S3 - Possible warnings and their solutions
+
+\newpage{}
 # References

### 2019-12-18 empirical_analysis_manuscript.R

@@ -23,3 +23,6 @@ p2 <- map_bias(proj)
 plot_out <- plot_grid(p1, p2, labels = c('A', 'B'), ncol = 1, rel_heights = c(1, 2))
 
 ggsave(plot_out, filename = "ms_figures/figure_empirical_results.jpg", height = 10, width = 8)
+
+
+

### 2019-12-18 manuscript.rmd
release version

@@
 *Sampbias* is implemented as a user-friendly R package.
-We exemplify the use of *sampbias* on a dataset of mammal occurrences from the Indonesian island of Borneo, downloaded from www.gbif.org.
+We exemplify the use of *sampbias* on a dataset of mammal occurrences from the Indonesian island of Borneo, downloaded from www.gbif.org, showing a high biasing effect of cities and a moderate effect of roads.
 
@@
 Publicly available datasets of geo-referenced species occurrences, such as provided by the Global Biodiversity Information Facility (www.gbif.org) have become a fundamental resource in biological sciences, especially in biogeography, conservation, and macroecology.
-However, these datasets are "presence-only" data and rarely include information on collection effort.
-Instead, they are typically not collected systematically and often compiled from a variety of sources (e.g.
-scientific expeditions, census counts, genetic barcoding studies, and citizen-science observations),  thus becoming subject to sampling bias [@Meyer2016].
+However, these datasets are are typically not collected systematically and rarely include information on collection effort.
+Instead, they  are often compiled from a variety of sources (e.g.
+scientific expeditions, census counts, genetic barcoding studies, and citizen-science observations),  therefore often subject to sampling bias [@Meyer2016].
 
-Likely, the number of data points available in such datasets is biased by factors other than species' presence or abundance, including the under-sampling of specific taxa ("taxonomic bias", e.g., birds *vs.* nematodes), specific geographic regions ("geographic bias", i.e.
+Likely, the number of occurrence available in such datasets is biased by factors other than species' presence or abundance, including the under-sampling of specific taxa ("taxonomic bias", e.g., birds *vs.* nematodes), specific geographic regions ("geographic bias", i.e.
 easily accessible *vs.* remote areas), and specific temporal periods ["temporal bias", i.e.
 wet season *vs.* dry season, @Isaac2015; @Boakes2010].
-Geographic sampling bias---the fact that sampling effort is spatially biased, rather than equally distributed over a given study area---is prevalent in all non-systematically collected datasets of species distributions.
+Geographic sampling bias---the fact that sampling effort is spatially biased, rather than equally distributed over the study area---is prevalent in all non-systematically collected datasets of species distributions.
 Many factors can can cause sampling bias, including socio-economic factors [i.e.
@@
 
-The implications of not considering geographic sampling bias in biodiversity research are likely to be substantial [@Meyer2016; @Rocchini2015; @Shimadzu2015;  @Yang2013; @Kramer-Schadt2013; @Barbosa2013].
+The implications of not considering geographic sampling bias in biodiversity research are likely  substantial [@Meyer2016; @Rocchini2015; @Shimadzu2015;  @Yang2013; @Kramer-Schadt2013; @Barbosa2013].
 While the presence of geographic sampling bias is broadly recognized [e.g.
-@Kadmon2004], and approaches exist to account for them in some analyses---for instance for species-richness estimates [@Engemann2015] species distribution models [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014; @Monsarrat2019], occupancy models [@Kery2016], or abundance estimates [@Shimadzu2015]---few attempts have been made to explicitly quantify the bias [@Kadmon2004; @Hijmans2000] or to discern among different sources of bias [@Fernandez2015; @Ruete2015; @Fithian2014], and to our knowledge, no tools exist for comparing the strength of accessibility bias among sources of bias or dataset.
-While it is unrealistic to expect that spatial biases in biodiversity data will ever disappear, it is crucial that researchers realise the intrinsic biases associated with the biodiversity data they are dealing with.
+@Kadmon2004], and approaches exist to account for them in some analyses---for instance for species-richness estimates [@Engemann2015] species distribution models [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014; @Monsarrat2019], occupancy models [@Kery2016], or abundance estimates [@Shimadzu2015]---few attempts have been made to explicitly quantify the bias [@Kadmon2004; @Hijmans2000] or to discern among different sources of bias [@Fernandez2015; @Ruete2015; @Fithian2014], and to our knowledge, no tools exist for comparing the strength of accessibility bias among sources of bias or datasets.
+While it is unrealistic to expect that spatial biases in biodiversity data will ever disappear, it is crucial that researchers realise the intrinsic bias associated with the  data they are dealing with.
 This is the first step towards estimating to which extent these biases may affect their analyses, results, and conclusions drawn from such data.
@@
 
-Here, we present *sampbias*, a novel method to quantify accessibility bias in individual datasets of species occurrences, in a way that is comparable across datasets.
-*Sampbias* is implemented as an R-package.
+Here, we present *sampbias*, a method to quantify accessibility bias in individual datasets of species occurrences, in a way that is comparable across datasets.
+*Sampbias* is implemented as user-friendly R-package.
 Specifically, *sampbias* uses a Bayesian approach based on a Poisson process to address three questions:
@@
 
-*Sampbias* is implemented in R [@rcoreteam2019], based on commonly usedpackages for data handling  [`ggplot`, @Wickham2009; `tidyr`, @Wickham2019a; `dplyr`, @Wickham2019; `forcats`, @Wickham2019b; `magrittr`, @Bache2014; `viridis`, @Garnier2018], handling geographic information and geo-computation [`raster`, @Hijmans2019; `sp`, @Pebesma2005; @Bivand2013] and statistical modelling [`stats`, @rcoreteam2019].
+*Sampbias* is implemented in R [@rcoreteam2019], based on commonly used packages for data handling  [`ggplot`, @Wickham2009; `tidyr`, @Wickham2019a; `dplyr`, @Wickham2019; `forcats`, @Wickham2019b; `magrittr`, @Bache2014; `viridis`, @Garnier2018], handling geographic information and geo-computation [`raster`, @Hijmans2019; `sp`, @Pebesma2005; @Bivand2013] and statistical modelling [`stats`, @rcoreteam2019].
 *Sampbias* offers an easy and largely automated means for biodiversity scientists and non-specialists alike to explore bias in species occurrence data and may be used to identify priorities for further collection or digitalization efforts, provide bias surfaces for species distribution modelling, or assess the reliability of scientific results based on publicly available species distribution data.
@@
 cities or roads).
-Due to the assumption of homogeneous sampling and a computational trade-off between the resolution of the distance raster and the extent of the study area (for instance, a 1000m resolution for a global dataset would lead to the generation of grid  for which distance calculation will become computationally prohibitive in most practical cases), *sampbias* is best suited for local or regional datasets at high resolution (c.
-100 -- 10,000m).
 
@@
 The resolution of the grid defined the precision of the distance estimates, for instance a 1x1 degree raster will yield approximately a 100km precision at the equator.
+Due to the assumption of homogeneous sampling and a computational trade-off between the resolution of the distance raster and the extent of the study area (for instance, a 1000m resolution for a global dataset would lead to the generation of grid  for which distance calculation will become computationally prohibitive in most practical cases), *sampbias* is best suited for local or regional datasets at high resolution (c.
+100 -- 10,000m).
 
@@
 All models predict a low number of collection records in the centre of Borneo, which reflects the original data, and where accessibility means are low (Figure S1 in Appendix S1).
-We ran this example to illustrate the use of *sampbias* For detailed analyses or a smaller geographic scale, higher resolution gazetteers, including smaller roads and rivers  and a higher spatial resolution would be desirable.
-Results might change with increasing resolution, since roads and rivers might have a stronger effect on higher resolutions (facilitating most the access to their immediate vicinity), whereas cities and airports might have an effect on the larger scale (facilitating access to a larger area).
+The empirical example illustrates the use of *sampbias*, for detailed analyses or a smaller geographic scale, higher resolution gazetteers, including smaller roads and rivers and a higher spatial resolution would be desirable.
+Results might change with increasing resolution, since roads and rivers might have a stronger effect on higher resolutions (facilitating most the access to their immediate vicinity), whereas cities and airports might have a stronger effect on the larger scale (facilitating access to a larger area).
 
@@
 # Data accessibility
-The software presented here is available under a GPL-3 license.
-The *sampbias* R package is available https://github.com/azizka/sampbias.
-The R package includes an example dataset as well as vignettes detailing the use of the R package (Appendix S1),  and possibly warnings produced by the package (Appendix S2).
+*Sampbias* is available under a GPL-3 license from https://github.com/azizka/sampbias, and includes an example dataset as well as a tutorial to run *sampbias* (Appendix S2) and as summary of possibly warnings produced by the package (Appendix S3).
 

### 2019-12-18 R/calculate_bias.R

@@ -60,9 +60,6 @@
 #' @param mcmc_burnin numerical. the burnin for the MCMC, default is to o
 #' @param mcmc_outfile character string. the path on where to write
 #' the results of the MCMC, optional.
-#' @param ncores numerical.  The number of cores used for parallel computing.
-#' Must be lower than the available number of cores. Not finally implemented in
-#' version 0.1.0.
 #' @param verbose logical.  If TRUE, progress is reported.  Default = TRUE.
 #' @return An object of the S3-class \sQuote{sampbias}, which is a list
 #' including the following objects: \item{summa}{A list of summary statistics
@@ -125,7 +122,6 @@ calculate_bias <- function(x,
                           mcmc_iterations = 1e+05,
                           mcmc_burnin = 0,
                           mcmc_outfile = NULL,
-                          ncores = 1,
                           verbose = TRUE) {
 
   #convert x to SpatialPoints
@@ -200,8 +196,7 @@ calculate_bias <- function(x,
     ## create distance raster for all gazeteers
     dis.ras <- dis_rast(gaz = gaz,
                         ras = occ.out,
-                        buffer = buffer,
-                        ncores = ncores)
+                        buffer = buffer)
     if (terrestrial) {
       dis.ras <- lapply(dis.ras, function(k) mask(k, wrld))
     }

### 2019-12-18 R/dis_rast.R

@@ -53,7 +53,7 @@
 #'@importFrom sp SpatialPointsDataFrame SpatialLinesDataFrame SpatialLines Lines Line
 #'@importFrom raster crop extent raster res
 #'
-dis_rast <- function(gaz, ras, buffer = NULL, ncores = 1) {
+dis_rast <- function(gaz, ras, buffer = NULL) {
 
   # create buffer, if none is supplied
   if (is.null(buffer)) {

### 2019-12-19 manuscript.rmd
updated manuscript

@@
 differences in sampling intensity among localities caused by differences in accessibility for humans is ubiquitous and might differ in strength among taxonomic groups and datasets.
-Yet, no general methodology exists to quantify the effect of roadside or other sources of bias on a dataset level.
-Here we present *sampbias*, an algorithm and software to estimate the biasing effect of accessibility (by roads, rivers, airports, cities, or any user-defined structures) in species occurrence datasets.
-*Sampbias* uses a Bayesian approach based on MCMC to optimize the rate of a Poisson sampling process, with sampling dependent on the distance from the next biasing structure.
-The results are comparable among biasing factors and datasets.
+While several bias factors exist, here defined as anthropogenic or natural features that facilitate human accessibility (e.g.
+roads, rivers, airports, cities), quantifying their effect on occurrence data remains difficult.
+Here we present *sampbias*, an algorithm and software to quantify the effect of roadside  in species occurrence datasets.
+*Sampbias* uses a Bayesian approach to estimate how sampling rates vary as a function of proximity to one or multiple bias factors.
+The results are comparable among bias factors and datasets.
 *Sampbias* is implemented as a user-friendly R package.
-We exemplify the use of *sampbias* on a dataset of mammal occurrences from the Indonesian island of Borneo, downloaded from www.gbif.org, showing a high biasing effect of cities and a moderate effect of roads.
+We demonstrate the use of *sampbias* on a dataset of mammal occurrences from the Indonesian island of Borneo, showing a high biasing effect of cities and a moderate effect of roads and airports.
 
@@
 
-Likely, the number of occurrence available in such datasets is biased by factors other than species' presence or abundance, including the under-sampling of specific taxa ("taxonomic bias", e.g., birds *vs.* nematodes), specific geographic regions ("geographic bias", i.e.
+The number of occurrence available in such datasets is likely biased by factors other than species' presence or abundance, including the under-sampling of specific taxa ("taxonomic bias", e.g., birds *vs.* nematodes), specific geographic regions ("geographic bias", i.e.
 easily accessible *vs.* remote areas), and specific temporal periods ["temporal bias", i.e.
@@
 In practice, this means that most species observations  are made in or near cities, along roads and rivers, and near other human settlements.
-Less observations come from the middle of a tropical rainforest or from a mountain top.
+Relatively fewer observations are expected to be available from the middle of a tropical rainforest or from a mountain top.
 Interestingly, since the observation of different taxonomic groups has different challenges, geographic sampling bias and the effect of accessibility may differ among taxonomic groups [@Vale2012].
@@
 @Kadmon2004], and approaches exist to account for them in some analyses---for instance for species-richness estimates [@Engemann2015] species distribution models [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014; @Monsarrat2019], occupancy models [@Kery2016], or abundance estimates [@Shimadzu2015]---few attempts have been made to explicitly quantify the bias [@Kadmon2004; @Hijmans2000] or to discern among different sources of bias [@Fernandez2015; @Ruete2015; @Fithian2014], and to our knowledge, no tools exist for comparing the strength of accessibility bias among sources of bias or datasets.
+We define as *bias factors* any anthropogenic or natural features that facilitate human accessibility and sampling, such as roads, rivers, airports, and cities.
+
 While it is unrealistic to expect that spatial biases in biodiversity data will ever disappear, it is crucial that researchers realise the intrinsic bias associated with the  data they are dealing with.
@@
 Therefore, it is advisable for any study dealing with species occurrence data to assess the strength of accessibility bias in the underlying data.
+Finally, a quantification of spatial sampling biases can help researchers to target their sampling efforts.
 
-Here, we present *sampbias*, a method to quantify accessibility bias in individual datasets of species occurrences, in a way that is comparable across datasets.
-*Sampbias* is implemented as user-friendly R-package.
-Specifically, *sampbias* uses a Bayesian approach based on a Poisson process to address three questions:
+Here, we present *sampbias*, a probabilistic method to quantify accessibility bias in datasets of species occurrences, in a way that is comparable across datasets.
+*Sampbias* is implemented as user-friendly R-package and uses a Bayesian approach to address three questions:
 
@@
 
-2) How important are different means of human accessibility, such as to airports, cities, rivers or roads, in causing this bias?
+2) How important are different factors in causing this bias?
 
@@
 ## General concept
-Under the assumption that organisms exist across the entire area of interest, we can expect the number of sampled occurrences in a restricted areas, such as a single biome, to be distributed uniformly in space (even though, of course, the density of individuals and the species composition may be heterogeneous).
-With *sampbias* we assess if discrepancies in sampling can be explained by distance from factors that potentially bias their sampling probability (e.g.
+Under the assumption that organisms exist across the entire area of interest, we can expect the number of sampled occurrences in a restricted areas, such as a single biome, to be distributed uniformly in space (even though, of course, the density of individuals and the species diversity may be heterogeneous).
+With *sampbias* we assess to which extent variation in sampling rates can be explained by distance from factors that potentially bias their sampling probability (e.g.
 cities or roads).
@@
 Default large-scale gazetteers for airports, cities, rivers and roads are provided with *sampbias*.
-Species occurrence data as downloaded from the data portal of GBIF can be directly used as input data for sampbias.
-The output of the package includes measures of bias effect, which are comparable between different gazetteers (e.g.
+Species occurrence data as downloaded from the data portal of GBIF can be directly used as input data for *sampbias*.
+The output of the package includes measures of the sampling rates across space, which are comparable between different gazetteers (e.g.
 comparing the biasing effect of roads and rivers), different taxa (e.g.
@@
 ## Distance calculation
-*Sampbias* uses gazetteers of the geographic location of bias sources (e.g.
+*Sampbias* uses gazetteers of the geographic location of bias factors (e.g.
 roads) to generate a regular grid across the study area (the geographic extent of the dataset).
-For each grid cell $i$, we then compute a vector $X_i(j)$ of minimum distances ("as the crow flies") to each source of bias $j \in B$.
-We then use these distance grids to sample the distribution of distances in the observed dataset.
-The resolution of the grid defined the precision of the distance estimates, for instance a 1x1 degree raster will yield approximately a 100km precision at the equator.
-Due to the assumption of homogeneous sampling and a computational trade-off between the resolution of the distance raster and the extent of the study area (for instance, a 1000m resolution for a global dataset would lead to the generation of grid  for which distance calculation will become computationally prohibitive in most practical cases), *sampbias* is best suited for local or regional datasets at high resolution (c.
-100 -- 10,000m).
+For each grid cell $i$, we then compute a vector $X_i(j)$ of minimum distances ("as the crow flies") to each bias factor $j \in B$.
+The resolution of the grid defines the precision of the distance estimates, for instance a 1x1 degree raster will yield approximately a 100 km precision at the equator.
+Due to the assumption of homogeneous sampling and a computational trade-off between the resolution of the distance raster and the extent of the study area (for instance, a 1000 m resolution for a global dataset would lead to the generation of grid  for which distance calculation will become computationally prohibitive in most practical cases), *sampbias* is best suited for local or regional datasets at high resolution (c.
+100 -- 10,000 m).
 
@@
 We describe the observed number of sampled occurrences $S_i$ within each cell $i$ as the result of a Poisson sampling process with rate $\lambda_i$.
-We model the rate $\lambda_i$ as a function of a constant $q$, which represent the expected number of occurrences per cell in the absence of biases, i.e.
+We model the rate $\lambda_i$ as a function of a parameter $q$, which represents the expected number of occurrences per cell in the absence of biases, i.e.
 when $\sum_{j=1}^{B}{ X_i(j)} = 0$.
-Additionally, we model $\lambda_i$ to decrease exponentially as a function of distance from sources of bias, such that increasing distances will result in a lower sampling rate.
-For a single source of bias the rates of cell $i$ with distance $X_i$ from a bias is:
+Additionally, we model $\lambda_i$ to decrease exponentially as a function of distance from bias factors, such that increasing distances will result in a lower sampling rate.
+For a single bias factor the rates of cell $i$ with distance $X_i$ from a bias is:
 $$
@@
 where $w \in \mathbb{R^+}$ defines the steepness of the Poisson rate decline, such that $w \approx 0$ results in a null model of uniform sampling rate $q$ across cells.
-In the presence of multiple bias predictors, the sampling rate decrease is a function of the cumulative effects of each bias and its distance from the cell:
+In the presence of multiple bias factors (e.g.
+roads and rivers), the sampling rate decrease is a function of the cumulative effects of each bias and its distance from the cell:
 $$
@@
 $$
-where a vector $\mathbf{w} = [w_1, ..., w_B]$ describes the amount of bias attributed to each specific predictor.
+where a vector $\mathbf{w} = [w_1, ..., w_B]$ describes the amount of bias attributed to each specific factor.
 
-To quantify the amount of bias associated with each predictor, we jointly estimate the parameters $q$ and $\mathbf{w}$ in a Bayesian framework.
+To quantify the amount of bias associated with each factor, we jointly estimate the parameters $q$ and $\mathbf{w}$ in a Bayesian framework.
 We use Markov Chain Monte Carlo (MCMC) to sample these parameters from their posterior distribution:
@@
 We interpret the magnitude of the elements in $\mathbf{w}$ as a function of the importance of the individual biases.
-We note however that this test is not explicitly intended to assess the significance of each bias predictor (for which a Bayesian variable selection could be used), particularly since several sources of bias might be correlated (e.g.
+We note, however, that this test is not explicitly intended to assess the significance of each bias factor (for which a Bayesian variable selection method could be used), particularly since several sources of bias might be correlated (e.g.
 cities, and airports).
@@
 
-[**Daniele some text here on the projection of bias through space, and what the plots exactly show**]
+We summarize the results by mapping the estimated sampling rates ($\lambda_i$) across space.
+These rates represent the expected number of sampled occurrences for each grid cell and provide a graphical representation of the spatial variation of sampling rates.
+Provided that the cells are of comparable size, the estimated rates will be comparable across data sets, regions, and taxonomic groups. 
+Analyzing different regions, biomes, or taxa in separate analyses allows to account for differences in over sampling rates, which are not linked with bias factors.
+For instance, the unbiased sampling rate $q$ is expected to differ between a highly sampled clade like birds and under-sampled groups of invertebrates, but their sampling biases ($\mathbf{w}$) might be similar across the two groups.
 
@@
 The example dataset is provided with the *sampbias*.
-We found a strong effect of cities on sampling intensity, a moderate effect of roads and airports and no effect of rivers (Fig.
+We found a strong effect of cities on sampling intensity, a moderate effect of roads and airports and negligible effect of rivers (Fig.
 \ref{fig:empirical}A).
@@
 
+[**DO we need a Discussion paragraph?**
+
 # Author contributions
@@
 <!-- empirical results -->
-```{r empirical, out.width = "6in", echo = FALSE, fig.cap = "The spatial projection of the accessibility bias in an empirical example dataset of mammal occurrences on the Indonesian island of Borneo from www.gbif.org.
-A) bias weights, B) projection of the expected number of occurrences given the sampbias model.
+```{r empirical, out.width = "6in", echo = FALSE, fig.cap = "The spatial projection of the estimated sampling rates in an empirical example dataset of mammal occurrences on the Indonesian island of Borneo from www.gbif.org.
+A) bias weights ($w$) defining the effects of each bias factor, B) projection of the sampling rates (i.e.
+expected number of occurrences per cell) given the inferred *sampbias* model.
 At the study scale of 0.05 degrees (~5 km)\\textit{Sampbias} finds the strongest biasing effect for the proximity of cities and roads, and a highest undersampling in the center of the island.", eval = TRUE}

### 2019-12-20 manuscript.rmd
v1.0-2

@@
 Many factors can can cause sampling bias, including socio-economic factors [i.e.
-national research spending, history of scientific research; @Meyer2015; @Daru2018], political factors [armed conflict, democratic rights; @Ryden2019], and physical accessibility [i.e.
+national research spending, history of scientific research; www.bio-dem.surge.sh, @Meyer2015; @Daru2018], political factors [armed conflict, democratic rights; @Ryden2019], and physical accessibility [i.e.
 distance to a road or river, terrain conditions, slope; @Yang2014; @Botts2011].
@@
 *Sampbias* works on a user-defined scale, and any dataset of multi-species occurrence records can be tested against any geographic gazetteer (reliability increases with increasing dataset size).
-Default large-scale gazetteers for airports, cities, rivers and roads are provided with *sampbias*.
+Default global gazetteers for airports, cities, rivers and roads are provided with *sampbias*.
 Species occurrence data as downloaded from the data portal of GBIF can be directly used as input data for *sampbias*.
@@
 roads) to generate a regular grid across the study area (the geographic extent of the dataset).
-For each grid cell $i$, we then compute a vector $X_i(j)$ of minimum distances ("as the crow flies") to each bias factor $j \in B$.
+For each grid cell $i$, we then compute a vector $X_i(j)$ of minimum distances (straight aerial distance, "as the crow flies") to each bias factor $j \in B$.
 The resolution of the grid defines the precision of the distance estimates, for instance a 1x1 degree raster will yield approximately a 100 km precision at the equator.
@@
 
-[**DO we need a Discussion paragraph?**
-
 # Author contributions
@@
 
-# Figure
+# Figures
 <!-- empirical results -->
-```{r empirical, out.width = "6in", echo = FALSE, fig.cap = "The spatial projection of the estimated sampling rates in an empirical example dataset of mammal occurrences on the Indonesian island of Borneo from www.gbif.org.
-A) bias weights ($w$) defining the effects of each bias factor, B) projection of the sampling rates (i.e.
-expected number of occurrences per cell) given the inferred *sampbias* model.
-At the study scale of 0.05 degrees (~5 km)\\textit{Sampbias} finds the strongest biasing effect for the proximity of cities and roads, and a highest undersampling in the center of the island.", eval = TRUE}
+```{r empirical, out.width = "6in", echo = FALSE, fig.cap = "Results of a \textit{sampbias} analysis.
+A) bias weights ($w$) defining the effects of each bias factor, B) the sampling rates as function of distance to the colsest instance of each biasing factor (i.e.
+expected number of occurrences) given the inferred *sampbias* model.
+At the study scale of 0.05 degrees (~5 km)\\textit{sampbias} finds the strongest biasing effect for the proximity of cities and roads.", eval = TRUE}
+knitr::include_graphics(path = "ms_figures/figure_empirical_results.jpg")
+```
+
+```{r spatial, out.width = "6in", echo = FALSE, fig.cap = "The spatial projection of the estimated sampling rates in an empirical example dataset of mammal occurrences on the Indonesian island of Borneo from www.gbif.org.
+The colours show the projection of the sampling rates (i.e.
+expected number of occurrences per cell) given the inferred \textit{sampbias} model.
+The highest undersampling is in the center of the island.", eval = TRUE}
 knitr::include_graphics(path = "ms_figures/figure_empirical_results.jpg")

### 2019-12-20 R/calculate_bias.R

@@ -111,6 +111,7 @@
 #' @importFrom sp SpatialPointsDataFrame SpatialLinesDataFrame SpatialLines Lines Line
 #' @importFrom rlang .data
 #' @importFrom stats complete.cases
+#' @importFrom rgeos gBuffer
 #'
 calculate_bias <- function(x,
                           gaz = NULL,
@@ -120,7 +121,7 @@ calculate_bias <- function(x,
                           terrestrial = TRUE,
                           mcmc_rescale_distances = 1000,
                           mcmc_iterations = 1e+05,
-                          mcmc_burnin = 0,
+                          mcmc_burnin = 2e+04,
                           mcmc_outfile = NULL,
                           verbose = TRUE) {
 
@@ -232,8 +233,9 @@ calculate_bias <- function(x,
     out <- .RunSampBias(x = dis.vec,
                         rescale_distances = 1000,
                         iterations = 1e+05,
-                        burnin = 0,
+                        burnin = mcmc_burnin,
                         outfile = NULL)
+
 
     # create output file, a list of the class sampbias
     if (verbose) {

### 2019-12-20 R/plot.sampbias.R

@@ -38,30 +38,55 @@
 #'@method plot sampbias
 #'@export
 #'@importFrom forcats fct_reorder
-#'@importFrom ggplot2 ggplot geom_boxplot xlab ylab coord_flip theme_bw theme
+#'@importFrom cowplot plot_grid
+#'@importFrom ggplot2 ggplot geom_boxplot geom_point xlab ylab coord_flip theme_bw theme
 #'@importFrom graphics plot
 #'@importFrom magrittr %>%
 #'@importFrom dplyr contains mutate select
 #'@importFrom rlang .data
 #'@importFrom stats median
 #'@importFrom tidyr pivot_longer
+#'@importFrom viridis scale_color_viridis
 #'
 plot.sampbias <- function(x, ...) {
 
-  plo <-  x$bias_estimate %>%
+  plo1 <-  x$bias_estimate %>%
     pivot_longer(cols = contains("w_"), names_to = "bias",
                  values_to = "posterior_estimate") %>%
     mutate(bias = gsub("w_", "", .data$bias)) %>%
     mutate(bias = fct_reorder(.data$bias, .data$posterior_estimate, .fun = median, .desc = FALSE))
 
-  out <- ggplot(plo)+
-    geom_boxplot(aes(x = .data$bias, y = .data$posterior_estimate))+
+  plo2_w <-colMeans(x$bias_estimate)
+  plo2_dist <- seq(1,1000,length.out=1000)
+  plo2 <- data.frame(dist = plo2_dist,
+                     rate = plo2_w[4] * exp(-plo2_w[5:length(plo2_w)]*plo2_dist),
+                     id = names(x$bias_estimate)[-c(1:4)]) %>%
+    mutate(id = gsub("w_", "", .data$id)) %>%
+    mutate(id = factor(.data$id, levels = levels(plo1$bias)))
+
+  p1 <- ggplot(plo1)+
+    geom_boxplot(aes(x = .data$bias, y = .data$posterior_estimate, fill = .data$bias))+
+    scale_fill_viridis(discrete = TRUE)+
     xlab("Biasing factor")+
     ylab("Posterior weight")+
     coord_flip()+
     theme_bw()+
     theme(panel.grid.minor.x = element_blank(),
-          panel.grid.major.y = element_blank())
+          panel.grid.major.y = element_blank(),
+          legend.position = "none")
+
+  p2 <- ggplot(plo2)+
+    geom_point(aes(x = .data$dist, y = .data$rate, color = .data$id))+
[…]

### 2019-12-20 empirical_analysis_manuscript.R

@@ -16,13 +16,17 @@ out <- calculate_bias(occ, res = 0.05, buffer = 0.5)
 p1 <- plot(out)
 summary(out)
 
+ggsave(p1, filename = "ms_figures/figure_empirical_results.jpg", height = 8, width = 8)
+
 proj <- project_bias(out)
 p2 <- map_bias(proj)
 
 
-plot_out <- plot_grid(p1, p2, labels = c('A', 'B'), ncol = 1, rel_heights = c(1, 2))
+ggsave(p1, filename = "ms_figures/figure_empirical_results.jpg", height = 8, width = 8)
 
-ggsave(plot_out, filename = "ms_figures/figure_empirical_results.jpg", height = 10, width = 8)
+
+
+
 
 
 

### 2019-12-20 manuscript.rmd
manuscript to co-authors

@@
 ---
-title: "Sampbias, a method to evaluate geographic sampling bias in species distribution data"
+title: "Sampbias, a method to quantify geographic sampling bias in species distribution data"
 #author: "Alexander Zizka^1,2,3^, Alexandre Antonelli^3,4,5^, Daniele Silvestro^3,4^"
@@
 output: 
-  bookdown::pdf_document2:
-  #word_document:
+  #bookdown::pdf_document2:
+  word_document:
     toc: false
     toc_depth: 1
-    fig_caption: true
-    number_sections: false
+   # fig_caption: true
+    #number_sections: false
 header-includes: |
@@
 Geographically biased sampling is a widely recognized issue that might severely affect analyses.
-Especially "roadside bias", i.e.
+Especially "accessibility bias", i.e.
 differences in sampling intensity among localities caused by differences in accessibility for humans is ubiquitous and might differ in strength among taxonomic groups and datasets.
@@
 roads, rivers, airports, cities), quantifying their effect on occurrence data remains difficult.
-Here we present *sampbias*, an algorithm and software to quantify the effect of roadside  in species occurrence datasets.
+Here we present *sampbias*, an algorithm and software to quantify the effect of accessibility bias in species occurrence datasets.
 *Sampbias* uses a Bayesian approach to estimate how sampling rates vary as a function of proximity to one or multiple bias factors.
@@
 While the presence of geographic sampling bias is broadly recognized [e.g.
-@Kadmon2004], and approaches exist to account for them in some analyses---for instance for species-richness estimates [@Engemann2015] species distribution models [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014; @Monsarrat2019], occupancy models [@Kery2016], or abundance estimates [@Shimadzu2015]---few attempts have been made to explicitly quantify the bias [@Kadmon2004; @Hijmans2000] or to discern among different sources of bias [@Fernandez2015; @Ruete2015; @Fithian2014], and to our knowledge, no tools exist for comparing the strength of accessibility bias among sources of bias or datasets.
+@Kadmon2004], and approaches exist to account for it in some analyses---for instance for species-richness estimates [@Engemann2015] species distribution models [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014; @Monsarrat2019], occupancy models [@Kery2016], or abundance estimates [@Shimadzu2015]---few attempts have been made to explicitly quantify the bias [@Kadmon2004; @Hijmans2000] or to discern among different sources of bias [@Fernandez2015; @Ruete2015; @Fithian2014], and to our knowledge, no tools exist for comparing the strength of accessibility bias among bias factors or datasets.
 We define as *bias factors* any anthropogenic or natural features that facilitate human accessibility and sampling, such as roads, rivers, airports, and cities.
 
-While it is unrealistic to expect that spatial biases in biodiversity data will ever disappear, it is crucial that researchers realise the intrinsic bias associated with the  data they are dealing with.
+While it is unrealistic to expect that accessibility in biodiversity data will ever disappear, it is crucial that researchers realise the intrinsic bias associated with the  data they are dealing with.
 This is the first step towards estimating to which extent these biases may affect their analyses, results, and conclusions drawn from such data.
 Therefore, it is advisable for any study dealing with species occurrence data to assess the strength of accessibility bias in the underlying data.
-Finally, a quantification of spatial sampling biases can help researchers to target their sampling efforts.
+Finally, a quantification of accessibility bias can help researchers to target their sampling efforts.
 
@@
 
-2) How important are different factors in causing this bias?
+2) How important are different bias factors in causing this bias?
 
-3) How is sampling bias distributed in space, i.e. which areas are a priority for targeted sampling?
+3) How is accessibility bias distributed in space, i.e. which areas are a priority for targeted sampling?
 
@@
 Under the assumption that organisms exist across the entire area of interest, we can expect the number of sampled occurrences in a restricted areas, such as a single biome, to be distributed uniformly in space (even though, of course, the density of individuals and the species diversity may be heterogeneous).
-With *sampbias* we assess to which extent variation in sampling rates can be explained by distance from factors that potentially bias their sampling probability (e.g.
-cities or roads).
+With *sampbias* we assess to which extent variation in sampling rates can be explained by distance from bias factors.
 
@@
 ## Distance calculation
-*Sampbias* uses gazetteers of the geographic location of bias factors (e.g.
-roads) to generate a regular grid across the study area (the geographic extent of the dataset).
+*Sampbias* uses gazetteers of the geographic location of bias factors to generate a regular grid across the study area (the geographic extent of the dataset).
 For each grid cell $i$, we then compute a vector $X_i(j)$ of minimum distances (straight aerial distance, "as the crow flies") to each bias factor $j \in B$.
@@
 We interpret the magnitude of the elements in $\mathbf{w}$ as a function of the importance of the individual biases.
-We note, however, that this test is not explicitly intended to assess the significance of each bias factor (for which a Bayesian variable selection method could be used), particularly since several sources of bias might be correlated (e.g.
+We note, however, that this test is not explicitly intended to assess the significance of each bias factor (for which a Bayesian variable selection method could be used), particularly since several bias factors might be correlated (e.g.
 cities, and airports).
@@
 To exemplify the use and output of *sampbias*, we downloaded the occurrence records of all mammals available from the Indonesian island of Borneo [n = 6,262, @gbifdoi2016], and ran *sampbias* using the default gazetteers as shown in the example code below, to test the biasing effect of the main airports, cities and roads in the dataset.
-The example dataset is provided with the *sampbias*.
+The example dataset is provided with *sampbias*.
 We found a strong effect of cities on sampling intensity, a moderate effect of roads and airports and negligible effect of rivers (Fig.
-\ref{fig:empirical}A).
-All models predict a low number of collection records in the centre of Borneo, which reflects the original data, and where accessibility means are low (Figure S1 in Appendix S1).
+\ref{fig:empirical}).
+All models predict a low number of collection records in the centre of Borneo (Fig.
+\ref{fig:spatial}), which reflects the original data, and where accessibility means are low (Figure S1 in Appendix S1).
 The empirical example illustrates the use of *sampbias*, for detailed analyses or a smaller geographic scale, higher resolution gazetteers, including smaller roads and rivers and a higher spatial resolution would be desirable.
@@
 # Data accessibility
-*Sampbias* is available under a GPL-3 license from https://github.com/azizka/sampbias, and includes an example dataset as well as a tutorial to run *sampbias* (Appendix S2) and as summary of possibly warnings produced by the package (Appendix S3).
+*Sampbias* is available under a GPL-3 license from https://github.com/azizka/sampbias, and includes the example dataset as well as a tutorial (Appendix S2) and a summary of possibly warnings produced by the package (Appendix S3).
 
@@
 <!-- empirical results -->
-```{r empirical, out.width = "6in", echo = FALSE, fig.cap = "Results of a \textit{sampbias} analysis.
-A) bias weights ($w$) defining the effects of each bias factor, B) the sampling rates as function of distance to the colsest instance of each biasing factor (i.e.
-expected number of occurrences) given the inferred *sampbias* model.
-At the study scale of 0.05 degrees (~5 km)\\textit{sampbias} finds the strongest biasing effect for the proximity of cities and roads.", eval = TRUE}
+```{r empirical, out.width = "6in", echo = FALSE, fig.cap = "Results of a \\textit{sampbias} analysis.
+A) bias weights ($w$) defining the effects of each bias factor, B) the sampling rates as function of distance to the closest instance of each biasing factor (i.e.
+expected number of occurrences) given the inferred \\textit{sampbias} model.
+At the study scale of 0.05 degrees (~5 km) \\textit{sampbias} finds the strongest biasing effect for the proximity of cities and roads.", eval = TRUE}
 knitr::include_graphics(path = "ms_figures/figure_empirical_results.jpg")
@@
 The highest undersampling is in the center of the island.", eval = TRUE}
-knitr::include_graphics(path = "ms_figures/figure_empirical_results.jpg")
+knitr::include_graphics(path = "ms_figures/figure_empirical_results_spatial_projection.jpg")
 ```
+
+\newpage{}
 
@@
 
-Appendix S1 - Supplementary Figure
+Appendix S1 - Supplementary Figure S1
 

### 2019-12-20 title_page.Rmd
manuscript to co-authors

@@
 ---
-title: "Sampbias, a method to evaluate geographic sampling bias in species distribution data"
+title: "Sampbias, a method to quantify geographic sampling bias in species distribution data"
 # title: "A global Red List of orchids supports the value of automated conservation assessment"
@@
 
-5. Royal Botanical Gardens Kew, Richmond, Surrey, United Kingdom
+5. Royal Botanic Gardens Kew, Richmond, Surrey, United Kingdom
 
@@
 
-**Word count** -- Abstract: XX; Total: XX
+**Word count** -- Abstract: 179 words; Total: 3269 words
 
@@
 # Acknowledgements
-We thank the organizers of the 2016 Ebben Nielsen challenge for inspiring and recognizing this research. We thank all data collectors and contributors to GBIF for their effort.
+We thank the organizers of the 2016 Ebben Nielsen challenge for inspiring and recognizing this research.
+We thank all data collectors and contributors to GBIF for their effort.
+AZ is thankful for funding by iDiv via the German Research Foundation (DFG FZT 118), specifically through sDiv, the Synthesis Centre of iDiv.
+AA is supported by grants from the Swedish Research Council, the Knut and Alice Wallenberg Foundation, the Swedish Foundation for Strategic Research and the Royal Botanic Gardens, Kew.

### 2019-12-20 empirical_analysis_manuscript.R

@@ -1,5 +1,6 @@
 library(sampbias)
 library(cowplot)
+library(ggplot2)
 
 occ <-read.csv(system.file("extdata", "mammals_borneo.csv", package="sampbias"), sep = "\t")
 
@@ -22,7 +23,7 @@ proj <- project_bias(out)
 p2 <- map_bias(proj)
 
 
-ggsave(p1, filename = "ms_figures/figure_empirical_results.jpg", height = 8, width = 8)
+ggsave(p2, filename = "ms_figures/figure_empirical_results_spatial_projection.jpg", height = 8, width = 8)
 
 
 

### 2020-01-13 manuscript.rmd
release version 1.0.2

@@
 ---
-title: "Sampbias, a method to quantify geographic sampling bias in species distribution data"
-#author: "Alexander Zizka^1,2,3^, Alexandre Antonelli^3,4,5^, Daniele Silvestro^3,4^"
+title: "sampbias, a method for quantifying geographic sampling biases in species distribution data"
+author: "Alexander Zizka^1,2^, Alexandre Antonelli^3,4,5^, Daniele Silvestro^3,4,6^"
 bibliography: sampbias.bib
@@
 output: 
-  #bookdown::pdf_document2:
-  word_document:
+  bookdown::pdf_document2:
+  #word_document:
     toc: false
     toc_depth: 1
-   # fig_caption: true
-    #number_sections: false
+    fig_caption: true
+    number_sections: false
 header-includes: |
@@
 
-<!-- 1. German Center for Integrative Biodiversity Research, University of Leipzig, Leipzig, Germany -->
-<!-- 2. Naturalis Biodiversity Center, Leiden University, Leiden, The Netherlands  -->
-<!-- 3. Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden -->
-<!-- 4. Department of Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden -->
-<!-- 5. Royal Botanical Gardens Kew, Richmond, Surrey, United Kingdom -->
+1. German Centre for Integrative Biodiversity Research Halle-Jena-Leipzig (iDiv), University of Leipzig, Deutscher Platz 5e, 04103 Leipzig, Germany
+2. Naturalis Biodiversity Center, Leiden University, Leiden, Darwinweg 2, 2333 CR Leiden The Netherlands
+3. Gothenburg Global Biodiversity Centre, University of Gothenburg, Box 461, 405 30 Gothenburg, Sweden
+4. Department for Biological and Environmental Sciences, University of Gothenburg, Box 461, 405 30 Gothenburg, Sweden
+5. Royal Botanic Gardens Kew, TW9 3AE, Richmond, Surrey, United Kingdom
+6. Department of Biology, University of Fribourg, Ch. du Musée 10, 1700 Fribourg, Switzerland
 
@@
 # Abstract
-Georeferenced species occurrences from public databases have become essential to biodiversity research and conservation, but have limitations.
-Geographically biased sampling is a widely recognized issue that might severely affect analyses.
-Especially "accessibility bias", i.e.
-differences in sampling intensity among localities caused by differences in accessibility for humans is ubiquitous and might differ in strength among taxonomic groups and datasets.
-While several bias factors exist, here defined as anthropogenic or natural features that facilitate human accessibility (e.g.
-roads, rivers, airports, cities), quantifying their effect on occurrence data remains difficult.
-Here we present *sampbias*, an algorithm and software to quantify the effect of accessibility bias in species occurrence datasets.
+Geo-referenced species occurrences from public databases have become essential to biodiversity research and conservation.
+However, geographical biases are widely recognized as a limiting factor limiting the usefulness of such data for understanding species diversity and distribution.
+In particular, differences in sampling intensity across a landscape due to differences in human accessibility are ubiquitous but may differ in strength among taxonomic groups and datasets.
+Although several factors have been described to influence human access (such as presence of roads, rivers, airports and cities), quantifying their specific and combined effects on recorded occurrence data remains challenging.
+Here we present *sampbias*, an algorithm and software for quantifying the effect of accessibility biases in species occurrence datasets.
 *Sampbias* uses a Bayesian approach to estimate how sampling rates vary as a function of proximity to one or multiple bias factors.
 The results are comparable among bias factors and datasets.
-*Sampbias* is implemented as a user-friendly R package.
-We demonstrate the use of *sampbias* on a dataset of mammal occurrences from the Indonesian island of Borneo, showing a high biasing effect of cities and a moderate effect of roads and airports.
+We demonstrate the use of *sampbias* on a dataset of mammal occurrences from the island of Borneo, showing a high biasing effect of cities and a moderate effect of roads and airports.
+*Sampbias* is implemented as a well-documented, open-access and  user-friendly R package that we hope will become a standard tool for anyone working with species occurrences in ecology, evolution, conservation  and related fields.
 
@@
 Publicly available datasets of geo-referenced species occurrences, such as provided by the Global Biodiversity Information Facility (www.gbif.org) have become a fundamental resource in biological sciences, especially in biogeography, conservation, and macroecology.
-However, these datasets are are typically not collected systematically and rarely include information on collection effort.
+However, these datasets are typically not collected systematically and rarely include information on collection effort.
 Instead, they  are often compiled from a variety of sources (e.g.
-scientific expeditions, census counts, genetic barcoding studies, and citizen-science observations),  therefore often subject to sampling bias [@Meyer2016].
+scientific expeditions, census counts, genetic barcoding studies, and citizen-science observations).
+Species occurrences are therefore often subject to multiple sampling biases [@Meyer2016].
 
-The number of occurrence available in such datasets is likely biased by factors other than species' presence or abundance, including the under-sampling of specific taxa ("taxonomic bias", e.g., birds *vs.* nematodes), specific geographic regions ("geographic bias", i.e.
-easily accessible *vs.* remote areas), and specific temporal periods ["temporal bias", i.e.
-wet season *vs.* dry season, @Isaac2015; @Boakes2010].
-Geographic sampling bias---the fact that sampling effort is spatially biased, rather than equally distributed over the study area---is prevalent in all non-systematically collected datasets of species distributions.
-Many factors can can cause sampling bias, including socio-economic factors [i.e.
+Sampling biases that may affect the recording of species occurrences [presence, absence and abundance, @Isaac2015; @Boakes2010] include the under-sampling of specific taxa ("taxonomic bias", e.g., birds *vs.* nematodes), specific geographic regions ("geographic bias", i.e.
+easily accessible *vs.* remote areas), and specific temporal periods ("temporal bias", i.e.
+wet season *vs.* dry season).
+In particular geographic sampling bias---the fact that sampling effort is spatially biased, rather than equally distributed over the study area---is likely to be widespread in all non-systematically collected datasets of species distributions.
+
+Many aspects can lead to sampling biases, including socio-economic factors [i.e.
 national research spending, history of scientific research; www.bio-dem.surge.sh, @Meyer2015; @Daru2018], political factors [armed conflict, democratic rights; @Ryden2019], and physical accessibility [i.e.
 distance to a road or river, terrain conditions, slope; @Yang2014; @Botts2011].
-Especially physical accessibility is omnipresent as a biasing factor [e.g.
-@Lin2015; @Engemann2015; @Kadmon2004], across spatial scales, and the term "roadside bias" has been coined for it.
-In practice, this means that most species observations  are made in or near cities, along roads and rivers, and near other human settlements.
-Relatively fewer observations are expected to be available from the middle of a tropical rainforest or from a mountain top.
-Interestingly, since the observation of different taxonomic groups has different challenges, geographic sampling bias and the effect of accessibility may differ among taxonomic groups [@Vale2012].
+Especially physical accessibility by people is omnipresent as a bias factor [e.g.
+@Lin2015; @Engemann2015; @Kadmon2004], across spatial scales, as the commonly used term "roadside bias" testifies.
+In practice, this means that most species observations are made in or near cities, along roads, paths, and rivers, and near human settlements.
+Relatively fewer observations are expected to be available from inaccessible areas in e.g.
+a tropical rainforest or a mountain top.
+Since the recording of different taxonomic groups poses different challenges, geographic sampling bias and the effect of accessibility may differ among taxonomic groups [@Vale2012].
 
-The implications of not considering geographic sampling bias in biodiversity research are likely  substantial [@Meyer2016; @Rocchini2015; @Shimadzu2015;  @Yang2013; @Kramer-Schadt2013; @Barbosa2013].
-While the presence of geographic sampling bias is broadly recognized [e.g.
-@Kadmon2004], and approaches exist to account for it in some analyses---for instance for species-richness estimates [@Engemann2015] species distribution models [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014; @Monsarrat2019], occupancy models [@Kery2016], or abundance estimates [@Shimadzu2015]---few attempts have been made to explicitly quantify the bias [@Kadmon2004; @Hijmans2000] or to discern among different sources of bias [@Fernandez2015; @Ruete2015; @Fithian2014], and to our knowledge, no tools exist for comparing the strength of accessibility bias among bias factors or datasets.
-We define as *bias factors* any anthropogenic or natural features that facilitate human accessibility and sampling, such as roads, rivers, airports, and cities.
+The implications of not considering geographic sampling biases in biodiversity research are likely to be substantial [@Meyer2016; @Rocchini2015; @Shimadzu2015;  @Yang2013; @Kramer-Schadt2013; @Barbosa2013].
+The presence of geographic sampling biases is broadly recognized [e.g.
+@Kadmon2004], and approaches exist to account for it in some analyses---such as for species-richness estimates [@Engemann2015] species distribution models [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014; @Monsarrat2019], occupancy models [@Kery2016], and abundance estimates [@Shimadzu2015].
+In contrast, few attempts have been made to explicitly quantify the overall bias [@Kadmon2004; @Hijmans2000] or to discern and quantify different sources of bias [@Fernandez2015; @Ruete2015; @Fithian2014].
+To our knowledge, no tools exist for comparing the strength of bias factors or datasets.
+We define as *bias factors* any anthropogenic or natural features that facilitate human access and sampling, such as roads, rivers, airports, and cities.
 
-While it is unrealistic to expect that accessibility in biodiversity data will ever disappear, it is crucial that researchers realise the intrinsic bias associated with the  data they are dealing with.
-This is the first step towards estimating to which extent these biases may affect their analyses, results, and conclusions drawn from such data.
-Therefore, it is advisable for any study dealing with species occurrence data to assess the strength of accessibility bias in the underlying data.
-Finally, a quantification of accessibility bias can help researchers to target their sampling efforts.
+It is unrealistic to expect that accessibility bias in biodiversity data will ever disappear even after more automated observation technologies are developed.
+It is therefore crucial that researchers realise the intrinsic biases associated with the data they deal with.
+This is the first step towards estimating to which extent these biases may affect their analyses, results, and conclusions.
+Any study dealing with species occurrence data should arguably assess the strength of accessibility biases in the underlying data.
+Such a quantification can also help researchers to target further sampling efforts.
 
-Here, we present *sampbias*, a probabilistic method to quantify accessibility bias in datasets of species occurrences, in a way that is comparable across datasets.
-*Sampbias* is implemented as user-friendly R-package and uses a Bayesian approach to address three questions:
+Here, we present *sampbias*, a probabilistic method to quantify accessibility bias in datasets of species occurrences.
+*Sampbias* is implemented as a user-friendly R-package and uses a Bayesian approach to address three questions:
 
@@
 
-2) How important are different bias factors in causing this bias?
+2) How strong is the effect of different bias factors in causing the overall accessibility bias?
 
@@
 *Sampbias* is implemented in R [@rcoreteam2019], based on commonly used packages for data handling  [`ggplot`, @Wickham2009; `tidyr`, @Wickham2019a; `dplyr`, @Wickham2019; `forcats`, @Wickham2019b; `magrittr`, @Bache2014; `viridis`, @Garnier2018], handling geographic information and geo-computation [`raster`, @Hijmans2019; `sp`, @Pebesma2005; @Bivand2013] and statistical modelling [`stats`, @rcoreteam2019].
-*Sampbias* offers an easy and largely automated means for biodiversity scientists and non-specialists alike to explore bias in species occurrence data and may be used to identify priorities for further collection or digitalization efforts, provide bias surfaces for species distribution modelling, or assess the reliability of scientific results based on publicly available species distribution data.
+*Sampbias* offers an easy and largely automated means for biodiversity scientists and non-specialists alike to explore bias in species occurrence data, in a way that is comparable across datasets.
+The results may be used to identify priorities for further collection or digitalization efforts, improve species distribution models (by providing bias surfaces in the analyses), or assess the reliability of scientific results based on publicly available species distribution data.
 
@@
 ## General concept
-Under the assumption that organisms exist across the entire area of interest, we can expect the number of sampled occurrences in a restricted areas, such as a single biome, to be distributed uniformly in space (even though, of course, the density of individuals and the species diversity may be heterogeneous).
+Under the assumption that organisms exist across the entire area of interest, we can expect the number of sampled occurrences in a restricted area, such as a single biome, to be distributed uniformly in space (even though, of course, the density of individuals and the species diversity may be heterogeneous).
 With *sampbias* we assess to which extent variation in sampling rates can be explained by distance from bias factors.
 
-*Sampbias* works on a user-defined scale, and any dataset of multi-species occurrence records can be tested against any geographic gazetteer (reliability increases with increasing dataset size).
-Default global gazetteers for airports, cities, rivers and roads are provided with *sampbias*.
+*Sampbias* works at a user-defined spatial scale, and any dataset of multi-species occurrence records can be tested against any geographic gazetteer.
+Reliability increases with increasing dataset size.
+Default global gazetteers for airports, cities, rivers and roads are provided with *sampbias*, and user-defined gazetteers can be added easily.
 Species occurrence data as downloaded from the data portal of GBIF can be directly used as input data for *sampbias*.
@@
 ## Distance calculation
-*Sampbias* uses gazetteers of the geographic location of bias factors to generate a regular grid across the study area (the geographic extent of the dataset).
+*Sampbias* uses gazetteers of the geographic location of bias factors (hereafter indicated with B) to generate a regular grid across the study area (the geographic extent of the dataset).
 For each grid cell $i$, we then compute a vector $X_i(j)$ of minimum distances (straight aerial distance, "as the crow flies") to each bias factor $j \in B$.
-The resolution of the grid defines the precision of the distance estimates, for instance a 1x1 degree raster will yield approximately a 100 km precision at the equator.
-Due to the assumption of homogeneous sampling and a computational trade-off between the resolution of the distance raster and the extent of the study area (for instance, a 1000 m resolution for a global dataset would lead to the generation of grid  for which distance calculation will become computationally prohibitive in most practical cases), *sampbias* is best suited for local or regional datasets at high resolution (c.
+The resolution of the grid defines the precision of the distance estimates, for instance a 1x1 degree raster will yield approximately a 110 km precision at the equator.
+Due to the assumption of homogeneous sampling and a computational trade-off between the resolution of the regular grid and the extent of the study area (for instance, a 1 second resolution for a global dataset would become computationally prohibitive in most practical cases), *sampbias* is best suited for local or regional datasets at high resolution (c.
 100 -- 10,000 m).
@@
 cities, and airports).
-Instead, these analyses can be use to quantify the expected amount of bias in the data that can be predicted by single or multiple predictors in order to identify under-sampled and unexplored areas.
+Instead, these analyses can be used to quantify the expected amount of bias in the data that can be predicted by single or multiple predictors in order to identify under-sampled and unexplored areas.
 
@@
 These rates represent the expected number of sampled occurrences for each grid cell and provide a graphical representation of the spatial variation of sampling rates.
-Provided that the cells are of comparable size, the estimated rates will be comparable across data sets, regions, and taxonomic groups. 
-Analyzing different regions, biomes, or taxa in separate analyses allows to account for differences in over sampling rates, which are not linked with bias factors.
+Provided that the cells are of equal size, the estimated rates will be comparable across data sets, regions, and taxonomic groups.
+Analysing different regions, biomes, or taxa in separate analyses allows to account for differences in over sampling rates, which are not linked with bias factors.
 For instance, the unbiased sampling rate $q$ is expected to differ between a highly sampled clade like birds and under-sampled groups of invertebrates, but their sampling biases ($\mathbf{w}$) might be similar across the two groups.
 
-## Example and Empirical analysis
+## Example and Empirical validation
 A default *sampbias* analysis can be run with few lines of code in R.
@@
 Based on a `data.frame` including species identity and geographic coordinates.
-Additionally, some options exist to provide custom gazetteers, custom distances for the bias estimation, a custom grain size of the analysis, as well as some operators for the calculation of the bias distances.
-A tutorial on how to use sampbias is available with the package and in the electronic supplement of this publication (Appendix S1).
+Additional options exist to provide custom gazetteers, a custom grain size of the analysis, as well as some operators for the calculation of the bias distances.
+A tutorial on how to use *sampbias* is available with the package and in the electronic supplement of this publication (Appendix S1).
 
-To exemplify the use and output of *sampbias*, we downloaded the occurrence records of all mammals available from the Indonesian island of Borneo [n = 6,262, @gbifdoi2016], and ran *sampbias* using the default gazetteers as shown in the example code below, to test the biasing effect of the main airports, cities and roads in the dataset.
+To exemplify the use and output of *sampbias*, we downloaded the occurrence records of all mammals available from the island of Borneo [n = 6,262, @gbifdoi2016], and ran *sampbias* using the default gazetteers as shown in the example code below, to test the biasing effect of the main airports, cities and roads in the dataset.
 The example dataset is provided with *sampbias*.
+
 We found a strong effect of cities on sampling intensity, a moderate effect of roads and airports and negligible effect of rivers (Fig.
@@
 All models predict a low number of collection records in the centre of Borneo (Fig.
-\ref{fig:spatial}), which reflects the original data, and where accessibility means are low (Figure S1 in Appendix S1).
+\ref{fig:spatial}), which reflects the original data, and where accessibility means are low (Figure S1 in Appendix S2).
 The empirical example illustrates the use of *sampbias*, for detailed analyses or a smaller geographic scale, higher resolution gazetteers, including smaller roads and rivers and a higher spatial resolution would be desirable.
@@
 
-# summarizing the results
+#summary
 summary(example.out)
@@
 
-#project in space
+#projecting the bias effect in space
 proj <- project_bias(example.out)
@@
 # Data accessibility
-*Sampbias* is available under a GPL-3 license from https://github.com/azizka/sampbias, and includes the example dataset as well as a tutorial (Appendix S2) and a summary of possibly warnings produced by the package (Appendix S3).
+*Sampbias* is available under a GNU General Public license v3 from https://github.com/azizka/sampbias, and includes the example dataset as well as a tutorial (Appendix S1) and a summary of possible warnings produced by the package (Appendix S3).
 
-<!-- # Acknowledgements -->
-<!-- We thank the organizers of the 2016 Ebben Nielsen challenge for inspiring and recognizing this research. We thank all data collectors and contributors to GBIF for their effort. -->
+# Acknowledgements
+We thank the organizers of the 2016 Ebben Nielsen challenge for inspiring and recognizing this research.
+We thank all data collectors and contributors to GBIF for their effort.
+AZ is thankful for funding by iDiv via the German Research Foundation (DFG FZT 118), specifically through sDiv, the Synthesis Centre of iDiv.
+AA is supported by grants from the Swedish Research Council, the Knut and Alice Wallenberg Foundation, the Swedish Foundation for Strategic Research and the Royal Botanic Gardens, Kew.
+DS received funding from the Swedish Research Council (2015-04748) and from the Swiss National Science Foundation (PCEFP3_187012)
 
 # Author contributions
-All authors conceived of this study, AZ and DS developed the statistical algorithm, AZ and DS wrote the R-package, AZ and DS wrote the manuscript with contributions from AA.
+All authors conceived this study, AZ and DS developed the statistical algorithm and wrote the R-package, AZ and DS wrote the manuscript with contributions from AA.
 
@@
 <!-- empirical results -->
-```{r empirical, out.width = "6in", echo = FALSE, fig.cap = "Results of a \\textit{sampbias} analysis.
-A) bias weights ($w$) defining the effects of each bias factor, B) the sampling rates as function of distance to the closest instance of each biasing factor (i.e.
+```{r empirical, out.width = "6in", echo = FALSE, fig.cap = "Results of the empirical validation analysis, estimating the accessibility bias in mammal occurrences from Borneo).
+A) bias weights ($w$) defining the effects of each bias factor, B) sampling rate as function of distance to the closest instance of each bias factor (i.e.
 expected number of occurrences) given the inferred \\textit{sampbias} model.
-At the study scale of 0.05 degrees (~5 km) \\textit{sampbias} finds the strongest biasing effect for the proximity of cities and roads.", eval = TRUE}
+At the study scale of 0.05 degrees (c.
+5km) \\textit{sampbias} finds the strongest biasing effect for the proximity of cities and roads.", eval = TRUE}
 knitr::include_graphics(path = "ms_figures/figure_empirical_results.jpg")
@@
 
-```{r spatial, out.width = "6in", echo = FALSE, fig.cap = "The spatial projection of the estimated sampling rates in an empirical example dataset of mammal occurrences on the Indonesian island of Borneo from www.gbif.org.
+```{r spatial, out.width = "6in", echo = FALSE, fig.cap = "Spatial projection of the estimated sampling rates in an empirical example dataset of mammal occurrences on the Indonesian island of Borneo (downloaded from www.gbif.org.
+GBIF.org, 2016).
 The colours show the projection of the sampling rates (i.e.
 expected number of occurrences per cell) given the inferred \textit{sampbias} model.
-The highest undersampling is in the center of the island.", eval = TRUE}
-knitr::include_graphics(path = "ms_figures/figure_empirical_results_spatial_projection.jpg")
+The highest undersampling is in the centre of the island.", eval = TRUE}
+knitr::include_graphics(path = "ms_figures/figure_empirical_results_spatial_projection_adapted.jpg")
 ```
@@
 
-Appendix S1 - Supplementary Figure S1
+Appendix S1 - Tutorial running *sampbias* in R
 
-Appendix S2 - Tutorial running sampbias in R
+Appendix S2 - Supplementary Figure S1
 

### 2020-01-13 title_page.Rmd
release version 1.0.2

@@
 ---
-title: "Sampbias, a method to quantify geographic sampling bias in species distribution data"
+title: "sampbias, a method for quantifying geographic sampling biases in species distribution data"
 # title: "A global Red List of orchids supports the value of automated conservation assessment"
-author: "Alexander Zizka^1,2,3^, Alexandre Antonelli^3,4,5^, Daniele Silvestro^3,4^"
+author: "Alexander Zizka^1,2^, Alexandre Antonelli^3,4,5^, Daniele Silvestro^3,4,6^"
 output: 
@@
 
-<!-- Letter to conservation letters: 3000 words in length (excluding references, table and figure legends, and online Supporting Information), 40 references, 6 figures -->
-
-<!-- Contributed paper to Conservation Biology, 3000-6000 words -->
-
 <!-- Affiliations -->
-1. German Center for Integrative Biodiversity Research, University of Leipzig, Leipzig, Germany
-
-2. Naturalis Biodiversity Center, Leiden University, Leiden, The Netherlands
-
-3. Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden
-
-4. Department of Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden
-
-5. Royal Botanic Gardens Kew, Richmond, Surrey, United Kingdom
+1. German Centre for Integrative Biodiversity Research Halle-Jena-Leipzig (iDiv), University of Leipzig, Deutscher Platz 5e, 04103 Leipzig, Germany
+2. Naturalis Biodiversity Center, Leiden University, Leiden, Darwinweg 2, 2333 CR Leiden The Netherlands
+3. Gothenburg Global Biodiversity Centre, University of Gothenburg, Box 461, 405 30 Gothenburg, Sweden
+4. Department for Biological and Environmental Sciences, University of Gothenburg, Box 461, 405 30 Gothenburg, Sweden
+5. Royal Botanic Gardens Kew, TW9 3AE, Richmond, Surrey, United Kingdom
+6. Department of Biology, University of Fribourg, Ch. du Musée 10, 1700 Fribourg, Switzerland
 
@@
 
-**Word count** -- Abstract: 179 words; Total: 3269 words
+**Word count** -- Abstract:  words; Total:  words
 
@@
 AA is supported by grants from the Swedish Research Council, the Knut and Alice Wallenberg Foundation, the Swedish Foundation for Strategic Research and the Royal Botanic Gardens, Kew.
+DS received funding from the Swedish Research Council (2015-04748) and from the Swiss National Science Foundation (PCEFP3_187012)

### 2020-01-13 R/dis_rast.R

@@ -110,7 +110,7 @@ dis_rast <- function(gaz, ras, buffer = NULL) {
   dist.d <- lapply(dist.r, function(k) suppressWarnings(raster::distance(k)))
 
   ## crop resulting distance raster to study area
-    if (buffer%%as.numeric(as.character(res(ras)[1])) == 0) {
+    if ( .DecimalPlaces(buffer/as.numeric(as.character(res(ras)[1]))) == 0) {
       dist.out <- lapply(dist.d, function(k) raster::crop(k, extent(ras)))
     } else {
       stop("'Buffer' is not a multiple of res. Set 'buffer' to a multiple of res")

### 2020-01-13 R/map_bias.R

@@ -112,7 +112,7 @@ map_bias <- function(x,
     scale_fill_viridis(na.value = "transparent",
                        option = "viridis",
                        direction = 1,
-                       name = "Bias effect [log]",
+                       name = "Bias effect\nsize [log]",
                        discrete = FALSE)+
     theme(axis.title = element_blank())+
     facet_wrap(split~ .)

### 2020-01-13 empirical_analysis_manuscript.R

@@ -23,7 +23,7 @@ proj <- project_bias(out)
 p2 <- map_bias(proj)
 
 
-ggsave(p2, filename = "ms_figures/figure_empirical_results_spatial_projection.jpg", height = 8, width = 8)
+ggsave(p2, filename = "ms_figures/figure_empirical_results_spatial_projection.pdf", height = 16, width = 16)
 
 
 

### 2020-03-30 manuscript.rmd
v1.0.3

@@
 Geo-referenced species occurrences from public databases have become essential to biodiversity research and conservation.
-However, geographical biases are widely recognized as a limiting factor limiting the usefulness of such data for understanding species diversity and distribution.
+However, geographical biases are widely recognized as a factor limiting the usefulness of such data for understanding species diversity and distribution.
 In particular, differences in sampling intensity across a landscape due to differences in human accessibility are ubiquitous but may differ in strength among taxonomic groups and datasets.

### 2020-03-30 title_page.Rmd
v1.0.3

@@
 
-**Running head** -- sampbias
+**Running head** -- sampbias accessibility bias
 
-**Word count** -- Abstract:  words; Total:  words
-
-<!-- **Article type** -- Letter, 146 words in the abstract, 5400 words in the manuscript, 42 References, 2 tables, 4 Figures -->
+**Word count** -- Abstract: 3,051 words; Total: 208 words, 2 figures
 
 **Corresponding author** -- alexander.zizka@idiv.de
-
-<!-- **Authors email addresses** -- tiffany.knight@idiv.de, PVitt@chicagobotanic.org -->
-
-**Keywords** Collection effort, Global biodiversity Information Facility (GBIF), Presence only data, Roadside bias, Sampling intensity
 
@@
 DS received funding from the Swedish Research Council (2015-04748) and from the Swiss National Science Foundation (PCEFP3_187012)
+
+# Author contributions
+All authors conceived this study, AZ and DS developed the statistical algorithm and wrote the R-package, AZ and DS wrote the manuscript with contributions from AA.

### 2020-03-30 R/calculate_bias.R

@@ -224,7 +224,7 @@ calculate_bias <- function(x,
     dis.vec <- as.data.frame(do.call(cbind, dis.vec))
     names(dis.vec) <- names(dis.ras)
 
-    dis.vec <- data.frame(cell_id = 1:nrow(dis.vec),
+    dis.vec <- data.frame(cell_id = seq_len(nrow(dis.vec)),
                           record_count = getValues(occ.out),
                           dis.vec)
     dis.vec <- dis.vec[complete.cases(dis.vec),]

### 2020-03-30 R/dis_rast.R

@@ -66,9 +66,15 @@ dis_rast <- function(gaz, ras, buffer = NULL) {
     warning(sprintf("Adapting buffer precision to resolution. Buffer set to %s", buffer))
   }
 
-  #make buffer even in case it is odd
+  #make buffer even in case it is odd, by setting to the closes even multiple of the resolution
   if((buffer * 10^decs) %%2 != 0  ){
-    buffer  <- buffer + 1/10^decs
+
+    ref <- seq(0, res(ras)[1]* 100, by = res(ras)[1])
+    ref <- ref[which((ref* 10^decs) %%2 == 0 )]
+
+    buffer <- ref[which.min(abs(ref - buffer))]
+
+    # buffer  <- buffer + 1/10^decs
     warning(sprintf("Evening buffer. Buffer set to %s", buffer))
   }
 
@@ -113,7 +119,14 @@ dis_rast <- function(gaz, ras, buffer = NULL) {
     if ( .DecimalPlaces(buffer/as.numeric(as.character(res(ras)[1]))) == 0) {
       dist.out <- lapply(dist.d, function(k) raster::crop(k, extent(ras)))
     } else {
-      stop("'Buffer' is not a multiple of res. Set 'buffer' to a multiple of res")
+      ref <- seq(0, res(ras)[1]* 100, by = res(ras)[1])
+      ref <- ref[which((ref* 10^decs) %%2 == 0 )]
+
+      buffer <- ref[which.min(abs(ref - buffer))]
+
+      warning(sprintf("'Buffer' was not a multiple of res. Set to %s", buffer))
+
+      dist.out <- lapply(dist.d, function(k) raster::crop(k, extent(ras)))
     }
 
   return(dist.out)

### 2020-03-30 R/internal.R

@@ -120,7 +120,7 @@ multiplier_proposal <- function(i, d = 1.2) {
       out <- rbind(out, c(it, likA, priorA, qA, wA))
 
       if (!is.null(outfile)){
-        cat(c(it, likA, priorA, qA, wA, "\n"), file = outfile, sep = "\t", append = T)
+        cat(c(it, likA, priorA, qA, wA, "\n"), file = outfile, sep = "\t", append = TRUE)
       }
     }
   }

### 2020-03-30 R/map_bias.R

@@ -63,7 +63,7 @@ map_bias <- function(x,
     point.gaz <- gaz.plo[list.condition]
     point.gaz <- lapply(point.gaz, "data.frame")
 
-    for (i in 1:length(point.gaz)) {
+    for (i in seq_along(point.gaz)) {
       point.gaz[[i]] <- data.frame(long = point.gaz[[i]]$longitude,
                                    lat = point.gaz[[i]]$latitude,
                                    bias = rep(names(point.gaz)[i],
@@ -75,7 +75,7 @@ map_bias <- function(x,
     line.gaz <- gaz.plo[!list.condition]
     line.gaz <- lapply(line.gaz, "fortify")
 
-    for (i in 1:length(line.gaz)) {
+    for (i in seq_along(line.gaz)) {
       line.gaz[[i]] <- data.frame(line.gaz[[i]], bias = rep(names(line.gaz)[i],
                                                             nrow(line.gaz[[i]])))
     }

### 2020-03-30 R/project_bias.R

@@ -65,7 +65,7 @@ project_bias <- function(x, factors = NULL) {
   # calculate the values for each raster cell
   lambdas <- list()
 
-  for (i in 1:length(mean_w)) {
+  for (i in seq_along(mean_w)) {
     if(is.numeric(ras)){
       test <- ras
     }else{
@@ -84,7 +84,7 @@ project_bias <- function(x, factors = NULL) {
 
   # define the names
   nam <- vector()
-  for (i in 1:length(mean_w)) {
+  for (i in seq_along(mean_w)) {
     nam <- c(nam, paste(gsub("w_", "",
                              names(mean_w)[1:i]),
                         collapse = "+"))

### 2020-03-30 R/sampbias.R

@@ -1,10 +1,10 @@
 
 
 #' Global Airport Coordinates
-#' 
+#'
 #' A gazetteer of major global airport locations.
-#' 
-#' 
+#'
+#'
 #' @name airports
 #' @docType data
 #' @format A data frame with 891 observations on the following 4 variables.
@@ -14,10 +14,10 @@
 #' @source \url{http://www.naturalearthdata.com/}
 #' @keywords datasets
 #' @examples
-#' 
+#'
 #' data(airports)
 #' str(airports)
-#' 
+#'
 NULL
 
 
@@ -25,11 +25,11 @@ NULL
 
 
 #' Global Major Populated Places
-#' 
+#'
 #' A gazetteer of global country and province capitals, as well as major cities
 #' and smaller cities in less populated areas.
-#' 
-#' 
+#'
+#'
 #' @name cities
 #' @docType data
 #' @format A data frame with 7322 observations on the following 4 variables.
@@ -41,10 +41,10 @@ NULL
 #' \url{http://www.naturalearthdata.com/downloads/10m-cultural-vectors/}
 #' @keywords datasets
 #' @examples
-#' 
+#'
 #' data(cities)
 #' str(cities)
-#' 
+#'
 NULL
 
 
@@ -52,20 +52,20 @@ NULL
 
 
 #' Global Coastlines
-#' 
+#'
 #' A \code{SpatialPolygonsDataFrame} with global coastlines.
-#' 
-#' 
+#'
+#'
 #' @name landmass
 #' @docType data
 #' @source
 #' \url{http://www.naturalearthdata.com/downloads/10m-physical-vectors/}
 #' @keywords datasets
 #' @examples
-#' 
+#'
 #' data("landmass")
-#' plot(landmass)
-#' 
+#'
+#'
 NULL
 
 
@@ -73,19 +73,19 @@ NULL
 
 
 #' Global Network of Major Roads
-#' 
+#'
 #' A gazetteer of major global roads as \code{SpatialPolygonsDataFrame}.
-#' 
-#' 
+#'
+#'
 #' @name roads
 #' @docType data
 #' @source
 #' \url{http://www.naturalearthdata.com/downloads/10m-cultural-vectors/}
 #' @keywords datasets
 #' @examples
-#' 
+#'
 #' data(roads)
-#' 
+#'
 NULL
 
 
@@ -93,21 +93,20 @@ NULL
 
 
 #' Global Major Waterbodies
-#' 
[…]

### 2020-03-30 pre_submission_tests.R

@@ -1,7 +1,6 @@
 
 pkgdown::build_site()
 
-source("https://install-github.me/MangoTheCat/goodpractice")
 library(goodpractice)
 gp("C:/Users/az64mycy/Dropbox (iDiv)/research_projects/16_spatial_bias_package/sampbias")
 

### 2020-07-11 manuscript.rmd
reviewer comments

@@
 We used exponential priors on the parameters $q$ and $\mathbf{w}$, $P(q) \sim \Gamma(1, 0.01)$ and $P(\mathbf{w}) \sim \Gamma(1, 1)$, respectively.
+We chose exponential priors because they represent the standard choice for rate parameters such as $q$ and the weights $\mathbf{w}$, all of which must be positive and have support [0, +inf].
+We designed the priors to be informative (i.e.
+not allowing negative values) and yet vague enough to encompass a much wider range of parameter space than that poster range of values observed in empirical tests.
+For instance, the posterior weights estimated in our empirical study ranged from 0 to 0.013, while the prior distribution applied on these parameters, Exp(1) has a 95% of it density in range 0 – 3.
+Custom priors, within the flexible family of gamma distributions are possible via the `prior_q` and `prior_w` arguments of the `calcualte_bias` function.
 
@@
 We interpret the magnitude of the elements in $\mathbf{w}$ as a function of the importance of the individual biases.
-We note, however, that this test is not explicitly intended to assess the significance of each bias factor (for which a Bayesian variable selection method could be used), particularly since several bias factors might be correlated (e.g.
-cities, and airports).
+We note, however, although that this test is not explicitly intended to assess the significance of each bias factor, for which a Bayesian variable selection method could be used.
 Instead, these analyses can be used to quantify the expected amount of bias in the data that can be predicted by single or multiple predictors in order to identify under-sampled and unexplored areas.
+Because particularly since several bias factors might be correlated (e.g.
+cities, and airports), cumulating their biases.
+estimated from independent analyses of each factor would result in an overestimation of the bias.
+For this reason, it is important to jointly estimate the effects of correlated factors, as this is based on the likelihood of the data given the combined effects of all biasing factors.
 
@@
 #running sampbias
+## res defines the resolution of the spatial raster 
+## for distance calculation in degrees latitude and longitude
+## buffer defines the buffer around the study area to account for biasing
+## structure adjecent to the study area, in degrees latitude and longitude
+## All other options at default, see ?calculate_bias for a description
 example.out <- calculate_bias(x = example.in, 
@@
 GBIF.org, 2016).
-The colours show the projection of the sampling rates (i.e.
+The colours show the projection of the log-transformedsampling rates (i.e.
 expected number of occurrences per cell) given the inferred \textit{sampbias} model.

### 2020-07-11 R/calculate_bias.R

@@ -60,6 +60,13 @@
 #' @param mcmc_burnin numerical. the burnin for the MCMC, default is to o
 #' @param mcmc_outfile character string. the path on where to write
 #' the results of the MCMC, optional.
+#' @param prior_q the gamma prior for the sampling rate $q$,
+#' which represents the expected number of occurrences per cell
+#' in the absence of biases. In the format c(shape,scale).
+#' @param prior_w the gamma prior for the steepness of the Poisson rate decline,
+#' such that $w \approx 0$ results in a
+#' null model of uniform sampling rate $q$ across cells.
+#' In the format c(shape,scale).
 #' @param verbose logical.  If TRUE, progress is reported.  Default = TRUE.
 #' @return An object of the S3-class \sQuote{sampbias}, which is a list
 #' including the following objects: \item{summa}{A list of summary statistics
@@ -124,7 +131,7 @@ calculate_bias <- function(x,
                           mcmc_burnin = 2e+04,
                           mcmc_outfile = NULL,
                           prior_q = 0.01,
-                          prior_w = 1, 
+                          prior_w = 1,
                           verbose = TRUE) {
 
   #convert x to SpatialPoints
@@ -236,7 +243,7 @@ calculate_bias <- function(x,
                         rescale_distances = 1000,
                         iterations = 1e+05,
                         burnin = mcmc_burnin,
-                        prior_q = prior_q, 
+                        prior_q = prior_q,
                         prior_w = prior_w,
                         outfile = NULL)
 

### 2020-07-11 R/map_bias.R

@@ -112,7 +112,7 @@ map_bias <- function(x,
     scale_fill_viridis(na.value = "transparent",
                        option = "viridis",
                        direction = 1,
-                       name = "Bias effect\nsize [log]",
+                       name = "Log Sampling rate",
                        discrete = FALSE)+
     theme(axis.title = element_blank())+
     facet_wrap(split~ .)

### 2020-07-11 empirical_analyses_rarefaction.R

@@ -1,0 +1,68 @@
+library(sampbias)
+library(cowplot)
+library(ggplot2)
+
+
+# rarefaction steps:
+rar <- c(1, 0.5, 0.25, 0.1, 0.01, 0.001)
+ID <- 1
+res <- 0.05
+
+
+# rarefaction of the empirical data
+occ <-read.csv(system.file("extdata", "mammals_borneo.csv", package="sampbias"), sep = "\t")
+
+# rarefaction and bias calculation loop, write out a map and the bias estiamtes
+
+for(i in 1:length(rar)){
+  print(i)
+  sub <- occ[base::sample(x = 1:nrow(occ), size = round(nrow(occ) * rar[i], 0)), ]
+  out <- calculate_bias(occ, res = res, buffer = 0.5)
+  out$ID <- ID
+  out$rar <- rar[i]
+  out$res <- res
+  out$type <- "empirical"
+
+  if(ID == 1 & i == 1){
+    write_csv(out, "empirical_analyses/weight_estimates.csv", append = FALSE)
+  }else{
+    write_csv(out, "empirical_analyses/weight_estimates.csv", append = TRUE)
+  }
+
+  proj <- project_bias(out)
+  p2 <- map_bias(proj)
+
+  ggsave(p2, filename = paste("empirical_analyses/figure_empirical_results_spatial_projection_empirical_", ID, "_", rar[i], ".pdf", sep = ""), height = 16, width = 16)
+
+  write
+}
+
+
+#rarefaction of a randomly sampled dataset across the study area without bias
+
+## get a polygon of borneo
+
+## randomly sample points across teh island
+
+# rarefaction loop for the bias calculation, write out the resulting paramter estimates and a map
+
+for(i in 1:length(rar)){
+  print(i)
+  sub <- occ[base::sample(x = 1:nrow(occ), size = round(nrow(occ) * rar[i], 0)), ]
+  out <- calculate_bias(occ, res = res, buffer = 0.5)
+  out$ID <- ID
+  out$rar <- rar[i]
+  out$res <- res
+  out$type <- "empirical"
+
+  if(ID == 1 & i == 1){
+    write_csv(out, "empirical_analyses/weight_estimates.csv", append = FALSE)
+  }else{
+    write_csv(out, "empirical_analyses/weight_estimates.csv", append = TRUE)
+  }
+
+  proj <- project_bias(out)
+  p2 <- map_bias(proj)
+
[…]

### 2020-07-21 R/calculate_bias.R
version 1.0.4 equal area, user-defined extent

@@ -46,13 +46,16 @@
 #' for geographic structures neighbouring the study area (such as a road right
 #' outside the study area). Should be a multiple of res.  Default is to res.
 #' See Details.
-#' @param convexhull logical.  If TRUE, the empirical distribution (and the
-#' output maps) is restricted to cells within a convex hull polygon around
-#' \code{x}. If FALSE a rectangle around \code{x} is used.  Default = FALSE.
+#' @param restrict_sample a SpatialPolygons or SpatialPolygonDataframe.
+#' If provided the area for the bias test will be restricted to raster cells
+#' within these polygons (and the extent of the sampled points in x).
+#' Make sure to use adequate values for res. Default = NULL.
 #' @param terrestrial logical.  If TRUE, the empirical distribution (and the
 #' output maps) are restricted to terrestrial areas.  Uses the
 #' \code{\link{landmass}} to define what is terrestrial.  Default = TRUE.
-#'
+#' @param inp_raster an object of class raster. A template raster for the
+#' counts and distance calculation. Can be used to provide a special resolution, or
+#' for different coordiante reference systems. See vignette.
 #' @param mcmc_rescale_distances numerical. rescaling factor for the
 #' distance calculation
 #' @param mcmc_iterations numerical. the number of iterations for the MCMC,
@@ -62,11 +65,13 @@
 #' the results of the MCMC, optional.
 #' @param prior_q the gamma prior for the sampling rate $q$,
 #' which represents the expected number of occurrences per cell
-#' in the absence of biases. In the format c(shape,scale).
+#' in the absence of biases. In the format c(shape,rate).
 #' @param prior_w the gamma prior for the steepness of the Poisson rate decline,
-#' such that $w \approx 0$ results in a
-#' null model of uniform sampling rate $q$ across cells.
-#' In the format c(shape,scale).
+#' such that w approximating 0 results in a
+#' null model of uniform sampling rate q across cells.
+#' In the format c(shape,rate).
+#' @param plot_raster logical. If TRUE, a plot of the occurrence raster is shown
+#' for diagnostic purposes. Dwfault = FALSE
 #' @param verbose logical.  If TRUE, progress is reported.  Default = TRUE.
 #' @return An object of the S3-class \sQuote{sampbias}, which is a list
[…]

### 2020-07-21 empirical_analyses_rarefaction.R

@@ -1,6 +1,8 @@
 library(sampbias)
 library(cowplot)
 library(ggplot2)
+library(sp)
+library(rgdal)
 
 
 # rarefaction steps:
@@ -41,8 +43,12 @@ for(i in 1:length(rar)){
 #rarefaction of a randomly sampled dataset across the study area without bias
 
 ## get a polygon of borneo
+born <- readOGR(dsn = "empirical_analyses", layer = "ne_110m_land")
+
+STILL NEED TO CROP
 
 ## randomly sample points across teh island
+occ <- sp::spsample(x = born, n = 100000)
 
 # rarefaction loop for the bias calculation, write out the resulting paramter estimates and a map
 
@@ -53,7 +59,7 @@ for(i in 1:length(rar)){
   out$ID <- ID
   out$rar <- rar[i]
   out$res <- res
-  out$type <- "empirical"
+  out$type <- "simulated"
 
   if(ID == 1 & i == 1){
     write_csv(out, "empirical_analyses/weight_estimates.csv", append = FALSE)
@@ -66,3 +72,5 @@ for(i in 1:length(rar)){
 
   ggsave(p2, filename = paste("empirical_analyses/figure_empirical_results_spatial_projection_empirical_", ID, "_", rar[i], ".pdf", sep = ""), height = 16, width = 16)
 }
+
+

### 2020-07-21 vignettes/Using_the_samp_bias_rpackage.R

@@ -1,42 +1,94 @@
-## ----setup, include=FALSE------------------------------------------------
+## ----setup, include=FALSE-----------------------------------------------------
 knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE)
 
-## ---- eval = FALSE-------------------------------------------------------
+## ---- eval = FALSE------------------------------------------------------------
 #  require(devtools)
 #  install_github(repo = "azizka/sampbias")
 #  library(sampbias)
 #  library(maptools)
+#  library(rgdal)
+#  library(raster)
 
-## ---- echo = FALSE-------------------------------------------------------
+## ---- echo = FALSE------------------------------------------------------------
 library(sampbias)
 library(maptools)
+library(rgdal)
+library(raster)
 
-## ------------------------------------------------------------------------
+## -----------------------------------------------------------------------------
 #loading a text file to R
 occ <-read.csv(system.file("extdata", 
                            "mammals_borneo.csv", 
                            package="sampbias"),
                sep = "\t")
 
-## ------------------------------------------------------------------------
-cit <- readShapeSpatial(system.file("extdata", 
-                                    "Borneo_major_cities.shp", 
-                                    package="sampbias"))
-roa <- readShapeSpatial(system.file("extdata", 
-                                    "Borneo_major_roads.shp", 
-                                    package="sampbias"))
+## -----------------------------------------------------------------------------
+cit <- readOGR(dsn = system.file("extdata", package="sampbias"), 
+               layer = "Borneo_major_cities")
+roa <- readOGR(dsn = system.file("extdata", package="sampbias"), 
+               layer = "Borneo_major_roads")
 
 gazetteers <- list(cities = cit,
                    roads = roa)
 
-## ------------------------------------------------------------------------
+## -----------------------------------------------------------------------------
 bias.out <- calculate_bias(x = occ, gaz = gazetteers)
 
-## ---- eval = FALSE-------------------------------------------------------
+## ---- eval = FALSE------------------------------------------------------------
 #  summary(bias.out)
 #  plot(bias.out)
 
-## ------------------------------------------------------------------------
+## -----------------------------------------------------------------------------
 proj <- project_bias(bias.out)
[…]

### 2020-07-23 manuscript.rmd
adapt package description and documentation

@@
 The presence of geographic sampling biases is broadly recognized [e.g.
-@Kadmon2004], and approaches exist to account for it in some analyses---such as for species-richness estimates [@Engemann2015] species distribution models [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014; @Monsarrat2019], occupancy models [@Kery2016], and abundance estimates [@Shimadzu2015].
-In contrast, few attempts have been made to explicitly quantify the overall bias [@Kadmon2004; @Hijmans2000] or to discern and quantify different sources of bias [@Fernandez2015; @Ruete2015; @Fithian2014].
+@Kadmon2004], and approaches exist to account for it in some analyses---such as for species-richness estimates [@Engemann2015], occupancy models [@Kery2016], and abundance estimates [@Shimadzu2015].
+In the case of species distribution modelling---the statistical estimation of species geographic distributions based on known occurrences and environmental conditions--- geographically baised sampling is problematic if it a cause a biased sampling in the environmantal space.
+Multiple appraoches exist to adress biased sampling in the context of sdms [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014], for instance spatial thinning [@Aiello-Lammens2015] and selecting background points using a bias surface (so that they are equally biased as the presence records) [REFERENCES].
+In contrast, few attempts have been made to explicitly quantify the overall bias [@Kadmon2004; @Hijmans2000] or to discern and quantify different sources of bias [@Fernandez2015; @Ruete2015; @Fithian2014; @Monsarrat2019].
 To our knowledge, no tools exist for comparing the strength of bias factors or datasets.
 We define as *bias factors* any anthropogenic or natural features that facilitate human access and sampling, such as roads, rivers, airports, and cities.
+
+Add references above and remove all those cited again from the first citation. add integration with other methods in as a future goal?!?
 
@@
 *Sampbias* offers an easy and largely automated means for biodiversity scientists and non-specialists alike to explore bias in species occurrence data, in a way that is comparable across datasets.
-The results may be used to identify priorities for further collection or digitalization efforts, improve species distribution models (by providing bias surfaces in the analyses), or assess the reliability of scientific results based on publicly available species distribution data.
+The results may be used to identify priorities for further collection or digitalization efforts, assess the reliability of scientific results based on publicly available species distribution data, and potentially to improve species distirbution models (by providing bias surfaces in the analyses).
 
@@
 ## Distance calculation
-*Sampbias* uses gazetteers of the geographic location of bias factors (hereafter indicated with B) to generate a regular grid across the study area (the geographic extent of the dataset).
+*Sampbias* uses gazetteers of the geographic location of bias factors (hereafter indicated with B) to generate a regular grid across the study area.
+By default the study area is defined by the geographic extent of the study dataset, but it can also be customized via user-defined polygons, for instance to limit the analyses to an environmentally homogenous region (e.g.
+rainforest) or an area of special interest (e.g.
+a national park).
 For each grid cell $i$, we then compute a vector $X_i(j)$ of minimum distances (straight aerial distance, "as the crow flies") to each bias factor $j \in B$.
@@
 100 -- 10,000 m).
+Since the differences in grid cell size are neglible on the local and regional scale, sampbias uses a latitude/longitude grid by default, but a custom grid in any projection and coordinate reference system---for instance of equal area grid cells---may be provided by the user.
 
@@
 Based on a `data.frame` including species identity and geographic coordinates.
-Additional options exist to provide custom gazetteers, a custom grain size of the analysis, as well as some operators for the calculation of the bias distances.
+Additional options exist to provide custom gazetteers, study area, spatial grid and grain size of the analysis, as well as some operators for the calculation of the bias distances, including  priors for $q$ and $\mathbf{w}$.
 A tutorial on how to use *sampbias* is available with the package and in the electronic supplement of this publication (Appendix S1).
@@
 #running sampbias
-## res defines the resolution of the spatial raster 
+## res defines the resolution of the spatial grid
 ## for distance calculation in degrees latitude and longitude

### 2020-07-23 R/calculate_bias.R

@@ -120,7 +120,7 @@
 #' @export
 #' @importFrom raster as.raster extent extent<- getValues mask res<- values rasterize trim stack plot
 #' @importFrom grDevices is.raster
-#' @importFrom sp SpatialPointsDataFrame SpatialLinesDataFrame SpatialLines Lines Line
+#' @importFrom sp SpatialPointsDataFrame SpatialLinesDataFrame SpatialLines Lines Line spTransform proj4string
 #' @importFrom rlang .data
 #' @importFrom stats complete.cases
 #' @importFrom rgeos gBuffer
@@ -220,7 +220,7 @@ calculate_bias <- function(x,
     dis.ras <- dis_rast(gaz = gaz,
                         ras = occ.out,
                         buffer = buffer)
-
+    dis.ras <- lapply(dis.ras, function(k) raster::crop(k, occ.out))
     dis.ras <- lapply(dis.ras, function(k) raster::mask(k, occ.out))
 
     ##  check if there are values in the distance raster

### 2020-07-23 R/dis_rast.R

@@ -59,6 +59,15 @@ dis_rast <- function(gaz, ras, buffer = NULL) {
   if (is.null(buffer)) {
     buffer <- res(ras)[1]
   }
+
+  # if x and y resolution are different, set buffer to 0 and return warning
+  # In a later version this should also be enabled for resolutions with uneven
+
+  if(res(ras)[1] != res(ras)[2]){
+    buffer <- 0
+    warning("Buffer not supported for CRS with unequal resolution. Setting buffer to 0. Mind edge effects!")
+  }
+
   #adapt buffer to resolution, buffer always has to be a multiple of resolution
   decs <- .DecimalPlaces(raster::res(ras)[1])
   if(.DecimalPlaces(buffer) > decs){

### 2020-07-23 R/sampbias.R

@@ -111,3 +111,45 @@ NULL
 
 
 
+#' Example Dataset for a Custom Study Area
+#'
+#' An example of the format needed to provide custom areas for \code{\link{calculate_bias}}
+#'
+#' @name area_example
+#' @docType data
+#' @keywords datasets
+#' @examples
+#'
+#' data(area_example)
+#'
+"area_example"
+
+#' Detailed Example for a Custom Study Area
+#'
+#' An example of the format needed to provide custom areas for \code{\link{calculate_bias}} based on a publicly available set of global ecoregions.
+#'
+#' @name ecoregion_example
+#' @docType data
+#' @source
+#' \url{https://www.worldwildlife.org/publications/terrestrial-ecoregions-of-the-world}
+#' @keywords datasets
+#' @examples
+#'
+#' data(ecoregion_example)
+#'
+"ecoregion_example"
+
+#' Equal Area Raster
+#'
+#' An example for an global equal area raster (in Behrmann projection) for the format needed for a custom grid provided to \code{\link{calculate_bias}}.
+#'
+#' @name ea_raster
+#' @docType data
+#' @keywords datasets
+#' @examples
+#'
+#' data(ea_raster)
+#'
+"ea_raster"
+
+

### 2020-07-23 empirical_analyses_rarefaction.R

@@ -3,6 +3,7 @@ library(cowplot)
 library(ggplot2)
 library(sp)
 library(rgdal)
+library(readr)
 
 
 # rarefaction steps:
@@ -20,57 +21,27 @@ for(i in 1:length(rar)){
   print(i)
   sub <- occ[base::sample(x = 1:nrow(occ), size = round(nrow(occ) * rar[i], 0)), ]
   out <- calculate_bias(occ, res = res, buffer = 0.5)
+
+  # plot of bias projection in space
+  proj <- project_bias(out)
+  p2 <- map_bias(proj)
+
+  ggsave(p2, filename = paste("empirical_analyses/simulations/figure_empirical_results_spatial_projection_empirical_",
+                              ID, "_", rar[i], ".pdf", sep = ""),
+         height = 16, width = 16)
+
+
+  # prepare bias estimate output
+  out <- out$bias_estimate
   out$ID <- ID
   out$rar <- rar[i]
   out$res <- res
   out$type <- "empirical"
 
+  # write to disk
   if(ID == 1 & i == 1){
-    write_csv(out, "empirical_analyses/weight_estimates.csv", append = FALSE)
+    write_csv(out, "empirical_analyses/simulations/weight_estimates.csv", append = FALSE)
   }else{
-    write_csv(out, "empirical_analyses/weight_estimates.csv", append = TRUE)
+    write_csv(out, "empirical_analyses/simulations/weight_estimates.csv", append = TRUE)
   }
-
-  proj <- project_bias(out)
-  p2 <- map_bias(proj)
-
-  ggsave(p2, filename = paste("empirical_analyses/figure_empirical_results_spatial_projection_empirical_", ID, "_", rar[i], ".pdf", sep = ""), height = 16, width = 16)
-
-  write
 }
-
-
-#rarefaction of a randomly sampled dataset across the study area without bias
-
-## get a polygon of borneo
-born <- readOGR(dsn = "empirical_analyses", layer = "ne_110m_land")
-
-STILL NEED TO CROP
-
-## randomly sample points across teh island
-occ <- sp::spsample(x = born, n = 100000)
-
-# rarefaction loop for the bias calculation, write out the resulting paramter estimates and a map
-
-for(i in 1:length(rar)){
-  print(i)
-  sub <- occ[base::sample(x = 1:nrow(occ), size = round(nrow(occ) * rar[i], 0)), ]
[…]

### 2020-07-23 vignettes/Using_the_samp_bias_rpackage.R

@@ -91,4 +91,3 @@ ea_bias <- calculate_bias(x = ea_occ,
                            inp_raster = ea_raster)
 summary(ea_bias)
 
-

### 2020-07-24 empirical_analyses_simulation.R
manuscript and appendices

@@ -21,7 +21,7 @@ data(landmass)
 born2 <- st_read("empirical_analyses/Borneo.kml")
 born2 <- sf:::st_zm(born2$geom)
 born2 <- as(born2, 'Spatial')
-born <- intersect(born, born2)
+born <- intersect(landmass, born2)
 
 
 ## randomly sample points across teh island

### 2020-07-27 appendix_S1.Rmd

@@
 subtitle: "Sampbias, a method to evaluate geographic sampling bias in species distribution data"
-#author: "Alexander Zizka^1,2,3^, Alexandre Antonelli^3,4,5^, Daniele Silvestro^3,4^"
+author: "Alexander Zizka, Alexandre Antonelli, Daniele Silvestro"
 output: rmarkdown::pdf_document
@@
     \usepackage{float}
-    \renewcommand{\thefigure}{S\arabic{figure}}
+    \renewcommand{\thefigure}{S1.\arabic{figure}}
 ---
@@
 occ <-read.csv(system.file("extdata", "mammals_borneo.csv", package="sampbias"), sep = "\t") %>% 
-  dplyr::select(latitude = decimallatitude, longitude = decimallongitude) %>% 
+  dplyr::select(latitude = decimalLatitude, longitude = decimalLongitude) %>% 
   mutate(type = "Mammal occurrence")

### 2020-07-27 appendix_S3.Rmd

---
title: "Appendix S3 - The effect of dataset size on parameter estimates"
subtitle: "Sampbias, a method to evaluate geographic sampling bias in species distribution data"
author: "Alexander Zizka, Alexandre Antonelli, Daniele Silvestro"
output: rmarkdown::pdf_document
header-includes: |
    \usepackage{float}
    \renewcommand{\thefigure}{S3.\arabic{figure}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

library(readr)
library(ggplot2)
library(magrittr)
library(sf)
library(geosphere)
library(dplyr)
library(tidyr)
library(forcats)
library(viridis)

```

The density of records might affect to estimate the sampling rate and bias weights (i.e. "How many records are needed to get reliable estimates?". To test the sensitivity of sampbias to the number of records in the input dataset, we ran two simulation studies:

1) As an example for a biased data set, we used the empirical example dataset with occurrence records of mammals from the island of Borneo. We down sampled the records by the factors 0.5, 0.25, 0.1, 0.01, 0.001 times the original number of records (i.e. 3131, 1566, 626, 62, and 6 records for all of Borneo) and ran a separate sampbias analyses on each of these down sampled datasets and compared the estimated posterior weights for each biasing factor. We replicated this analyses 5 times (A total of 25 analyses). The estimates of the bias weights were similar across analyses, suggesting that sampbias can robustly quantify sampling bias even with low record densities (Fig. 1).

2) We simulated seven data sets with 100,000, 50,000, 10,000, 1,000, 100, and 10 records randomly distributed across Borneo (no accessibility bias) and ran a separate sampbias analyses for each of these data sets, and replicated the analyses five times (a total of 35 analyses). The estimates of the bias weights were similar across analyses, suggesting that sampbias is robust against false positives in bias detection even in datasets with low record density (Fig. 2).

```{r}
occ <-read.csv(system.file("extdata", "mammals_borneo.csv", package="sampbias"), sep = "\t")

data(landmass)

born2 <- st_read("empirical_analyses/Borneo.kml", quiet = TRUE)
born2 <- sf:::st_zm(born2$geom)
born2 <- as(born2, 'Spatial')
born <- raster::intersect(landmass, born2)

area <- round(areaPolygon(born) / 1000000, 0)

dat <- read_csv("empirical_analyses/simulations/weight_estimates.csv") %>% 
  mutate(records = rar * nrow(occ)) %>% 
  mutate(records_area = round(records / area * 10000, 2)) %>% 
  select(contains("w"), ID, type, records_area) %>% 
    pivot_longer(cols = contains("w_"), names_to = "bias",
                 values_to = "posterior_estimate") %>%
    mutate(bias = gsub("w_", "", .data$bias)) %>%
    mutate(bias = fct_reorder(.data$bias, .data$posterior_estimate, .fun = median, .desc = FALSE))
```

```{r, echo = FALSE, fig.cap="The bias weights ($w$) defining the effects of each bias factor estimated from data sets with differing density of occurrence records across the study area. Datasets generated by down sampling the empirical example dataset of mammal occurrences on Borneo."}
plo <- dat %>%  filter(type == "empirical")
# ggplot(data = plo)+
#   geom_point(aes(x = bias, 
#                  y = posterior_estimate, 
#                  group = as.factor(records_area), 
#                  color = as.factor(ID)), 
#              position = position_jitterdodge())+
#   xlab("Posterior weight")+
#   xlab("Bias factor")+
#   scale_color_viridis(discrete = TRUE)+
#   theme_bw()

ggplot(data = plo)+
  geom_boxplot(aes(x = bias, 
                   y = posterior_estimate,
                 fill = as.factor(records_area)))+
  ylab("Posterior weight")+
  xlab("Bias factor")+
  scale_fill_viridis(discrete = TRUE, name = "Records per\n10,000 sqkm")+
  theme_bw()
```

```{r, echo = FALSE, fig.cap="The bias weights ($w$) defining the effects of each bias factor estimated from data sets with differing density of occurrence records across the study area. Datasets generated by random simulation of records across Borneo (no bias)."}
plo <- dat %>%  filter(type == "simulated")

ggplot(data = plo)+
  geom_boxplot(aes(x = bias, 
                   y = posterior_estimate,
                 fill = as.factor(records_area)))+
  ylab("Posterior weight")+
  xlab("Bias factor")+
  scale_fill_viridis(discrete = TRUE, name = "Records per\n10,000 sqkm")+
  theme_bw()
```

### 2020-07-27 manuscript.rmd

--- 2020-03-30 manuscript.rmd
@@
 
-The implications of not considering geographic sampling biases in biodiversity research are likely to be substantial [@Meyer2016; @Rocchini2015; @Shimadzu2015;  @Yang2013; @Kramer-Schadt2013; @Barbosa2013].
-The presence of geographic sampling biases is broadly recognized [e.g.
-@Kadmon2004], and approaches exist to account for it in some analyses---such as for species-richness estimates [@Engemann2015] species distribution models [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014; @Monsarrat2019], occupancy models [@Kery2016], and abundance estimates [@Shimadzu2015].
-In contrast, few attempts have been made to explicitly quantify the overall bias [@Kadmon2004; @Hijmans2000] or to discern and quantify different sources of bias [@Fernandez2015; @Ruete2015; @Fithian2014].
-To our knowledge, no tools exist for comparing the strength of bias factors or datasets.
-We define as *bias factors* any anthropogenic or natural features that facilitate human access and sampling, such as roads, rivers, airports, and cities.
+\textcolor{blue}{The implications of not considering geographic sampling biases in biodiversity research are likely substantial} [@Meyer2016; @Yang2013; @Barbosa2013].
+\textcolor{blue}{The presence of geographic sampling biases is broadly recognized} [e.g.
+@Kadmon2004], \textcolor{blue}{and approaches exist to account for it in some analyses---such as species-richness estimates} [@Engemann2015], \textcolor{blue}{occupancy models} [@Kery2016], \textcolor{blue}{and abundance estimates} [@Shimadzu2015].
+\textcolor{blue}{In the case of species distribution modelling---the statistical estimation of species geographic distributions based on known occurrences and environmental conditions---geographically biased sampling is problematic because it often causes environmentally biased sampling which decreases model performance} [@Kadmon2004; @Lobo2011; @Bystriakova2012; @Kramer-Schadt2013; @Varela2014].
+\textcolor{blue}{Many approaches exist to remedy the effect of biased sampling on species distribution models} [@Fourcade2014], \textcolor{blue}{including rarefaction to reduce clumped sampling in geographic} [@Aiello-Lammens2015; @Beck2014; @Boria2014] \textcolor{blue}{or environmental space} [@Varela2014]\textcolor{blue}{, collecting background points for presence-only models to reflect the same sampling bias as the presence records} [@Phillips2009]\textcolor{blue}{, and explicitly including modelling sampling bias} [@Komori2020; @Fithian2014; @Stolar2015]\textcolor{blue}{.
+In contrast, few attempts have been made to compare the geographic sampling bias among datasets} [@Fernandez2015; @Ruete2015; @Monsarrat2019] \textcolor{blue}{and to our knowledge, no tools exist to quantify the effect size of specific bias factors and compare it among them.
+We define as *bias factors* any anthropogenic or natural features that facilitate human access and sampling, such as roads, rivers, airports, and cities.}
 
 It is unrealistic to expect that accessibility bias in biodiversity data will ever disappear even after more automated observation technologies are developed.
-It is therefore crucial that researchers realise the intrinsic biases associated with the data they deal with.
+It is therefore crucial that researchers realise the intrinsic biases associated with the data they deal with, \textcolor{blue}{especially in cross-taxonomic studies, since occurrence datasets from different taxa are likely differently affected by sampling biases due to differences in specimen collection and transportation}.
 This is the first step towards estimating to which extent these biases may affect their analyses, results, and conclusions.
-Any study dealing with species occurrence data should arguably assess the strength of accessibility biases in the underlying data.
-Such a quantification can also help researchers to target further sampling efforts.
+Any study dealing with species occurrence data should arguably assess the strength of accessibility biases in the underlying data Such a quantification can also help researchers to target further sampling efforts.
 
@@
 *Sampbias* offers an easy and largely automated means for biodiversity scientists and non-specialists alike to explore bias in species occurrence data, in a way that is comparable across datasets.
-The results may be used to identify priorities for further collection or digitalization efforts, improve species distribution models (by providing bias surfaces in the analyses), or assess the reliability of scientific results based on publicly available species distribution data.
+\textcolor{blue}{The results may be used to identify priorities for further collection or digitalization efforts, assess the reliability of scientific results based on publicly available species distribution data, and potentially to improve species distribution models (by providing bias surfaces in the analyses).}
 
@@
 ## Distance calculation
-*Sampbias* uses gazetteers of the geographic location of bias factors (hereafter indicated with B) to generate a regular grid across the study area (the geographic extent of the dataset).
+*Sampbias* uses gazetteers of the geographic location of bias factors (hereafter indicated with B) to generate a regular grid across the study area.
+\textcolor{blue}{By default the study area is defined by the geographic extent of the study dataset, but it can also be customized via user-defined polygons, for instance to limit the analyses to an environmentally homogeneous region (e.g.
+rainforest) or an area of special interest (e.g.
+a national park)}.
 For each grid cell $i$, we then compute a vector $X_i(j)$ of minimum distances (straight aerial distance, "as the crow flies") to each bias factor $j \in B$.
@@
 100 -- 10,000 m).
+\textcolor{blue}{Since the differences in grid cell size are negligible on the local and regional scale, sampbias uses a latitude/longitude grid by default, but a custom grid in any projection and coordinate reference system---for instance of equal area grid cells---may be provided by the user.}
 
@@
 We used exponential priors on the parameters $q$ and $\mathbf{w}$, $P(q) \sim \Gamma(1, 0.01)$ and $P(\mathbf{w}) \sim \Gamma(1, 1)$, respectively.
+\textcolor{blue}{We chose exponential priors because they represent the standard choice for rate parameters such as} $q$ \textcolor{blue}{and the weights} $\mathbf{w}$ \textcolor{blue}{, all of which must be positive and have support [0, +inf].
+We designed the priors to be informative (i.e.
+not allowing negative values) and yet vague enough to encompass a much wider range of parameter space than that poster range of values observed in empirical tests.
+For instance, the posterior weights estimated in our empirical study ranged from 0 to 0.013, while the prior distribution applied on these parameters, Exp(1) has a} 95% \textcolor{blue}{of it density in range 0 – 3.
+Custom priors, within the flexible family of gamma distributions are possible via the} `prior_q` and `prior_w` \textcolor{blue}{arguments of the} `calculate_bias` \textcolor{blue}{function.}
 
@@
 We interpret the magnitude of the elements in $\mathbf{w}$ as a function of the importance of the individual biases.
-We note, however, that this test is not explicitly intended to assess the significance of each bias factor (for which a Bayesian variable selection method could be used), particularly since several bias factors might be correlated (e.g.
-cities, and airports).
-Instead, these analyses can be used to quantify the expected amount of bias in the data that can be predicted by single or multiple predictors in order to identify under-sampled and unexplored areas.
+\textcolor{blue}{We note, however, although this test is not explicitly intended to assess the significance of each bias factor, for which a Bayesian variable selection method could be used these analyses can be used to quantify the expected amount of bias in the data that can be predicted by single or multiple predictors in order to identify under-sampled and unexplored areas.
+Because particularly since several bias factors might be correlated (e.g.
+cities, and airports), cumulating their biases.
+estimated from independent analyses of each factor would result in an overestimation of the bias.
+For this reason, it is important to jointly estimate the effects of correlated factors, as this is based on the likelihood of the data given the combined effects of all biasing factors.}
 
@@
 Based on a `data.frame` including species identity and geographic coordinates.
-Additional options exist to provide custom gazetteers, a custom grain size of the analysis, as well as some operators for the calculation of the bias distances.
+\textcolor{blue}{Additional options exist to provide custom gazetteers, study area, spatial grid and grain size of the analysis, as well as some operators for the calculation of the bias distances, including  priors for $q$ and $\mathbf{w}$}.
 A tutorial on how to use *sampbias* is available with the package and in the electronic supplement of this publication (Appendix S1).
@@
 #running sampbias
+## 'res' defines the resolution of the spatial grid
+## for distance calculation in degrees latitude and longitude
+## 'buffer' defines the buffer around the study area to account for biasing
+## structure adjecent to the study area, in degrees latitude and longitude
+## All other options at default, see ?calculate_bias for a description
 example.out <- calculate_bias(x = example.in, 
@@
 # Data accessibility
-*Sampbias* is available under a GNU General Public license v3 from https://github.com/azizka/sampbias, and includes the example dataset as well as a tutorial (Appendix S1) and a summary of possible warnings produced by the package (Appendix S3).
+*Sampbias* is available under a GNU General Public license v3 from https://github.com/azizka/sampbias, and includes the example dataset as well as a tutorial (Appendix S1) and a summary of possible warnings produced by the package \textcolor{blue}{(Appendix S4)}.
 
@@
 GBIF.org, 2016).
-The colours show the projection of the sampling rates (i.e.
+The colours show the projection of the \\textcolor{blue}{log-transformed} sampling rates (i.e.
 expected number of occurrences per cell) given the inferred \textit{sampbias} model.
@@
 
-Appendix S3 - Possible warnings and their solutions
+\textcolor{blue}{Appendix S3 - The effect of dataset size on parameter estimates}
+
+\textcolor{blue}{Appendix S4 - Possible warnings and their solutions}
 

### 2020-07-27 appendix_S1.Rmd
manuscript changes

@@
 ---
-title: "Appendix S1 - Supplementary Figure"
-subtitle: "Sampbias, a method to evaluate geographic sampling bias in species distribution data"
+title: "sampbias, a method for quantifying geographic sampling biases in species distribution data"
+subtitle: "Appendix S1 - Supplementary Figure"
 author: "Alexander Zizka, Alexandre Antonelli, Daniele Silvestro"
@@
     \usepackage{float}
-    \renewcommand{\thefigure}{S1.\arabic{figure}}
+    \renewcommand{\thefigure}{S\arabic{figure}}
 ---
@@
 
-```{r, fig.height = 8, fig.cap="The example dataset of mammal occurrences from the island of Borneo, as downloaded from www.gbif.org (n = 6,262), and the geographic gazetters of main cities, roads, rivers and airports used for the sampbias analysis."}
+```{r, fig.height = 8, fig.cap="The example dataset of mammal occurrences from the island of Borneo, as downloaded from www.gbif.org (n = 6,262, 2016, https://doi.org/10.15468/dl.7fg4zx), and the geographic gazetters of main cities, roads, rivers and airports used for the sampbias analysis."}
 #libraries
@@
 library(viridis)
+library(readr)
+library(sf)
+library(geosphere)
+library(tidyr)
+library(forcats)
 
@@
 ```
+
+```{r}
+occ <-read.csv(system.file("extdata", "mammals_borneo.csv", package="sampbias"), sep = "\t")
+
+data(landmass)
+
+born2 <- st_read("empirical_analyses/Borneo.kml", quiet = TRUE)
+born2 <- sf:::st_zm(born2$geom)
+born2 <- as(born2, 'Spatial')
+born <- raster::intersect(landmass, born2)
+
+area <- round(areaPolygon(born) / 1000000, 0)
+
+dat <- read_csv("empirical_analyses/simulations/weight_estimates.csv") %>% 
+  mutate(records = ifelse(type=="empirical", rar * nrow(occ), rar * 100000)) %>% 
+  mutate(records_area = round(records / area * 10000, 2)) %>% 
+  select(contains("w"), ID, type, records_area) %>% 
+    pivot_longer(cols = contains("w_"), names_to = "bias",
+                 values_to = "posterior_estimate") %>%
+    mutate(bias = gsub("w_", "", .data$bias))
+```
+
+```{r, echo = FALSE, fig.cap="The bias weights ($w$) quantifying the effects of each bias factor estimated from data sets with differing density of occurrence records across the study area.
+Datasets generated by down-sampling the empirical example dataset of mammal occurrences on Borneo (www.gbif.org,  2016, https://doi.org/10.15468/dl.7fg4zx)."}
+plo <- dat %>%  filter(type == "empirical") %>%
+    mutate(bias = fct_reorder(.data$bias, .data$posterior_estimate, .fun = median, .desc = FALSE))
+# ggplot(data = plo)+
+#   geom_point(aes(x = bias, 
+#                  y = posterior_estimate, 
+#                  group = as.factor(records_area), 
+#                  color = as.factor(ID)), 
+#              position = position_jitterdodge())+
+#   xlab("Posterior weight")+
+#   xlab("Bias factor")+
+#   scale_color_viridis(discrete = TRUE)+
+#   theme_bw()
+
+ggplot(data = plo)+
+  geom_boxplot(aes(x = bias, 
+                   y = posterior_estimate,
+                 fill = as.factor(records_area),
+                 color = as.factor(records_area)))+
+  ylab("Posterior weight")+
+  xlab("Bias factor")+
+  scale_fill_viridis(discrete = TRUE, name = "Records per\n10,000 sqkm")+
+  scale_color_viridis(discrete = TRUE, name = "Records per\n10,000 sqkm")+
+  theme_bw()
+```
+
+```{r, out.width = "6in", fig.cap = "Projected bias surface for the distribution data of mammals from Borneo (www.gbif.org, 2016, https://doi.org/10.15468/dl.7fg4zx), at different levels of downsampling, from a random replicate."}
+
+knitr::include_graphics(path = "ms_figures/figure_projection_empirical_rarefaction.jpg")
+
+```
+
+```{r, echo = FALSE, fig.cap="The bias weights ($w$) defining the effects of each bias factor estimated from data sets with differing density of occurrence records across the study area.
+Datasets generated by random simulation of records across Borneo (no bias)."}
+plo <- dat %>%  filter(type == "simulated") %>%
+    mutate(bias = fct_reorder(.data$bias, .data$posterior_estimate, .fun = median, .desc = FALSE))
+
+ggplot(data = plo)+
+  geom_boxplot(aes(x = bias, 
+                   y = posterior_estimate,
+                 fill = as.factor(records_area)))+
+  ylab("Posterior weight")+
+  xlab("Bias factor")+
+  scale_fill_viridis(discrete = TRUE, name = "Records per\n10,000 sqkm")+
+  theme_bw()
+```
+
+```{r, out.width = "6in", fig.cap = "Projected bias surface for randomly sampled points across Borneo, from a random replicate."}
+
+knitr::include_graphics(path = "ms_figures/figure_projection_empirical_simulation.jpg")
+
+```

### 2020-07-27 manuscript.rmd
manuscript changes

@@
 # Keywords
-Collection effort, Global biodiversity Information Facility (GBIF), Presence only data, Roadside bias, Sampling intensity
+Collection effort, Global Biodiversity Information Facility (GBIF), Presence only data, Roadside bias, Sampling intensity
 
@@
 Many aspects can lead to sampling biases, including socio-economic factors [i.e.
-national research spending, history of scientific research; www.bio-dem.surge.sh, @Meyer2015; @Daru2018], political factors [armed conflict, democratic rights; @Ryden2019], and physical accessibility [i.e.
+national research spending, history of scientific research; @Zizka2020c; @Meyer2015; @Daru2018], political factors [armed conflict, democratic rights; @Ryden2019], and physical accessibility [i.e.
 distance to a road or river, terrain conditions, slope; @Yang2014; @Botts2011].
@@
 \textcolor{blue}{In the case of species distribution modelling---the statistical estimation of species geographic distributions based on known occurrences and environmental conditions---geographically biased sampling is problematic because it often causes environmentally biased sampling which decreases model performance} [@Kadmon2004; @Lobo2011; @Bystriakova2012; @Kramer-Schadt2013; @Varela2014].
-\textcolor{blue}{Many approaches exist to remedy the effect of biased sampling on species distribution models} [@Fourcade2014], \textcolor{blue}{including rarefaction to reduce clumped sampling in geographic} [@Aiello-Lammens2015; @Beck2014; @Boria2014] \textcolor{blue}{or environmental space} [@Varela2014]\textcolor{blue}{, collecting background points for presence-only models to reflect the same sampling bias as the presence records} [@Phillips2009]\textcolor{blue}{, and explicitly including modelling sampling bias} [@Komori2020; @Fithian2014; @Stolar2015]\textcolor{blue}{.
+\textcolor{blue}{Many approaches exist to remedy the effect of biased sampling on species distribution models} [@Fourcade2014], \textcolor{blue}{including rarefaction to reduce clumped sampling in geographic} [@Aiello-Lammens2015; @Beck2014; @Boria2014] \textcolor{blue}{or environmental space} [@Varela2014]\textcolor{blue}{, collecting background points for presence-only models to reflect the same sampling bias as the presence records} [@Phillips2009]\textcolor{blue}{, and explicitly modelling sampling bias} [@Komori2020; @Fithian2014; @Stolar2015]\textcolor{blue}{.
 In contrast, few attempts have been made to compare the geographic sampling bias among datasets} [@Fernandez2015; @Ruete2015; @Monsarrat2019] \textcolor{blue}{and to our knowledge, no tools exist to quantify the effect size of specific bias factors and compare it among them.
-We define as *bias factors* any anthropogenic or natural features that facilitate human access and sampling, such as roads, rivers, airports, and cities.}
+We define as \textit{bias factors} any anthropogenic or natural features that facilitate human access and sampling, such as roads, rivers, airports, and cities.}
 
@@
 This is the first step towards estimating to which extent these biases may affect their analyses, results, and conclusions.
-Any study dealing with species occurrence data should arguably assess the strength of accessibility biases in the underlying data Such a quantification can also help researchers to target further sampling efforts.
+Any study dealing with species occurrence data should arguably assess the strength of accessibility biases in the underlying data.
+Such a quantification can also help researchers to target further sampling efforts.
 
-Here, we present *sampbias*, a probabilistic method to quantify accessibility bias in datasets of species occurrences.
+Here, we present *sampbias* 1, a probabilistic method to quantify accessibility bias in datasets of species occurrences.
 *Sampbias* is implemented as a user-friendly R-package and uses a Bayesian approach to address three questions:
@@
 *Sampbias* offers an easy and largely automated means for biodiversity scientists and non-specialists alike to explore bias in species occurrence data, in a way that is comparable across datasets.
-\textcolor{blue}{The results may be used to identify priorities for further collection or digitalization efforts, assess the reliability of scientific results based on publicly available species distribution data, and potentially to improve species distribution models (by providing bias surfaces in the analyses).}
+\textcolor{blue}{The results may be used to identify priorities for further collection or digitalization efforts, assess the reliability of scientific results based on publicly available species distribution data.}
 
@@
 100 -- 10,000 m).
-\textcolor{blue}{Since the differences in grid cell size are negligible on the local and regional scale, sampbias uses a latitude/longitude grid by default, but a custom grid in any projection and coordinate reference system---for instance of equal area grid cells---may be provided by the user.}
+\textcolor{blue}{Since the differences in grid cell size are negligible on the local and regional scale, sampbias uses a latitude/longitude grid by default, but a custom grid in any projection and coordinate reference system---for instance an equal area grid---may be provided by the user.}
 
@@
 We used exponential priors on the parameters $q$ and $\mathbf{w}$, $P(q) \sim \Gamma(1, 0.01)$ and $P(\mathbf{w}) \sim \Gamma(1, 1)$, respectively.
-\textcolor{blue}{We chose exponential priors because they represent the standard choice for rate parameters such as} $q$ \textcolor{blue}{and the weights} $\mathbf{w}$ \textcolor{blue}{, all of which must be positive and have support [0, +inf].
+\textcolor{blue}{We chose exponential priors because they represent the standard choice for rate parameters such as} $q$ \textcolor{blue}{and the weights} $\mathbf{w}$\textcolor{blue}{, all of which must be positive and have support [0, +inf].
 We designed the priors to be informative (i.e.
-not allowing negative values) and yet vague enough to encompass a much wider range of parameter space than that poster range of values observed in empirical tests.
+not allowing negative values) and yet vague enough to encompass a much wider range of parameter space than the range of values observed in empirical tests.
 For instance, the posterior weights estimated in our empirical study ranged from 0 to 0.013, while the prior distribution applied on these parameters, Exp(1) has a} 95% \textcolor{blue}{of it density in range 0 – 3.
@@
 We interpret the magnitude of the elements in $\mathbf{w}$ as a function of the importance of the individual biases.
-\textcolor{blue}{We note, however, although this test is not explicitly intended to assess the significance of each bias factor, for which a Bayesian variable selection method could be used these analyses can be used to quantify the expected amount of bias in the data that can be predicted by single or multiple predictors in order to identify under-sampled and unexplored areas.
-Because particularly since several bias factors might be correlated (e.g.
-cities, and airports), cumulating their biases.
-estimated from independent analyses of each factor would result in an overestimation of the bias.
+\textcolor{blue}{We note that although this test is not explicitly intended to assess the significance of each bias factor, for which a Bayesian variable selection method could be used, it can quantify the expected amount of bias in the data predicted by single or multiple predictors in order to identify under-sampled and unexplored areas.
+Because several bias factors might be correlated (e.g.
+cities, and airports), cumulating their bias estimates from independent analyses of each factor would result in an overestimation of the bias.
 For this reason, it is important to jointly estimate the effects of correlated factors, as this is based on the likelihood of the data given the combined effects of all biasing factors.}
@@
 Provided that the cells are of equal size, the estimated rates will be comparable across data sets, regions, and taxonomic groups.
-Analysing different regions, biomes, or taxa in separate analyses allows to account for differences in over sampling rates, which are not linked with bias factors.
+Analysing different regions, biomes, or taxa in separate analyses allows to account for differences in sampling rates, which are not linked with bias factors.
 For instance, the unbiased sampling rate $q$ is expected to differ between a highly sampled clade like birds and under-sampled groups of invertebrates, but their sampling biases ($\mathbf{w}$) might be similar across the two groups.
@@
 
-To exemplify the use and output of *sampbias*, we downloaded the occurrence records of all mammals available from the island of Borneo [n = 6,262, @gbifdoi2016], and ran *sampbias* using the default gazetteers as shown in the example code below, to test the biasing effect of the main airports, cities and roads in the dataset.
+To exemplify the use of *sampbias*, we downloaded the occurrence records of all mammals available from the island of Borneo [n = 6,262, @gbifdoi2016], and ran *sampbias* using the default gazetteers as shown in the example code below, to test the biasing effect of the main airports, cities and roads in the dataset.
 The example dataset is provided with *sampbias*.
 
-We found a strong effect of cities on sampling intensity, a moderate effect of roads and airports and negligible effect of rivers (Fig.
+We found a strong effect of cities on sampling intensity, a moderate effect of roads and airports and a negligible effect of rivers (Fig.
 \ref{fig:empirical}).
@@
 
-#a data table with species identify, longitude, and latitude
+# a data table with species identity, longitude, and latitude
 example.in <- read.csv(system.file("extdata", 
@@
 
-#running sampbias
+# running sampbias
 ## 'res' defines the resolution of the spatial grid
@@
 
-#summary
+# summary
 summary(example.out)
@@
 
-#projecting the bias effect in space
+# projecting the bias effect in space
 proj <- project_bias(example.out)
@@
 
+*Sampbias* \textcolor{blue}{is designed to work with sparsely sampled datasets, and to estimate bias effects from datasets with low coverage.
+To test the performence of} *sampbias* \textcolor{blue}{on small datasets we ran two simulation experiments: 1) we ran the analyses as presented above on the distribution of mammals on Borneo, sub-sampling of the initial dataset (to 6,262, 3,131, 1,566, 626, 62 records for all of Borneo respectively, 5 replicates each).
+The results showed that estimates of} $\mathbf{w}$ \textcolor{blue}{ and the projection of the bias effect in space in a biased dataset were robust to decreases in data, although uncertainty increased (Figs.
+S2nd S3, in Appendix S2).
+2) in a second simulation we  randomly sampled occurrence records across Borneo as an example of an unbiased dataset.
+We sampled 100,000, 50,000, 25,000, and 10,000 records and ran a} *sampbias* \textcolor{blue}{analysis with the same settings as for the empirical dataset}.
+\textcolor{blue}{The results show that} *sampbias* \textcolor{blue}{reliably rejected a biasing effect even for small datasets (Figs.
+S4 and S5, in Appendix S2), however, we note that the analyses took a very long time to converge for datasets with no sampling bias and datasets with less than 10,000 points.
+In general, the more records, the more precise the parameter estimates will be.}
+
 # Data accessibility
-*Sampbias* is available under a GNU General Public license v3 from https://github.com/azizka/sampbias, and includes the example dataset as well as a tutorial (Appendix S1) and a summary of possible warnings produced by the package \textcolor{blue}{(Appendix S4)}.
+*Sampbias* is available under a GNU General Public license v3 from https://github.com/azizka/sampbias, and includes the example dataset as well as a tutorial (Appendix S1) and a summary of possible warnings produced by the package (Appendix S3).
 
-# Acknowledgements
-We thank the organizers of the 2016 Ebben Nielsen challenge for inspiring and recognizing this research.
+<!-- # Acknowledgements -->
+<!-- We thank the organizers of the 2016 Ebben Nielsen challenge for inspiring and recognizing this research.
 We thank all data collectors and contributors to GBIF for their effort.
@@
 AA is supported by grants from the Swedish Research Council, the Knut and Alice Wallenberg Foundation, the Swedish Foundation for Strategic Research and the Royal Botanic Gardens, Kew.
-DS received funding from the Swedish Research Council (2015-04748) and from the Swiss National Science Foundation (PCEFP3_187012)
+DS received funding from the Swedish Research Council (2015-04748) and from the Swiss National Science Foundation (PCEFP3_187012) -->
 
-# Author contributions
-All authors conceived this study, AZ and DS developed the statistical algorithm and wrote the R-package, AZ and DS wrote the manuscript with contributions from AA.
+<!-- # Author contributions -->
+<!-- All authors conceived this study, AZ and DS developed the statistical algorithm and wrote the R-package, AZ and DS wrote the manuscript with contributions from AA. -->
 
@@
 
-Appendix S2 - Supplementary Figure S1
+Appendix S2 - Supplementary Figures
 
-\textcolor{blue}{Appendix S3 - The effect of dataset size on parameter estimates}
-
-\textcolor{blue}{Appendix S4 - Possible warnings and their solutions}
+Appendix S3 - Possible warnings and their solutions
 

### 2020-07-27 appendix_S1.Rmd
focus on Borneo appendix 1

@@
 
+## get a polygon of Borneo
+data(landmass)
+
+born2 <- st_read("empirical_analyses/Borneo.kml")
+born2 <- sf:::st_zm(born2$geom)
+born2 <- as(born2, 'Spatial')
+
 #load data
@@
 # prepare for plotting
-airp <- airports %>% crop(SpatialPoints(occ[,2:1])) %>%  coordinates() %>%  data.frame() %>% mutate(type = "Airport")
-cit <- cities %>% crop(SpatialPoints(occ[,2:1])) %>%  coordinates() %>%  data.frame() %>%  mutate(type = "City")
+airp <- airports %>% crop(SpatialPoints(occ[,2:1])) %>% raster::intersect(born2) %>%  coordinates() %>%  data.frame() %>% mutate(type = "Airport")
+cit <- cities %>% crop(SpatialPoints(occ[,2:1])) %>% raster::intersect(born2) %>%  coordinates() %>%  data.frame() %>%  mutate(type = "City")
 pts <- bind_rows(occ, cit, airp)
@@
 riv <- waterbodies %>% crop(SpatialPoints(occ[,2:1])) %>%  fortify() %>%  mutate(type = "River")
-road <- roads %>% crop(SpatialPoints(occ[,2:1])) %>%  fortify() %>%  mutate(type = "Road")
+road <- roads %>% crop(SpatialPoints(occ[,2:1])) %>% raster::intersect(born2) %>%  fortify() %>%  mutate(type = "Road")
 lin <- bind_rows(riv, road)
 
-lma <- landmass %>% crop(SpatialPoints(occ[,2:1])) %>%  fortify()
+lma <- landmass %>% crop(SpatialPoints(occ[,2:1])) %>% raster::intersect(born2) %>%  fortify()
 

### 2020-07-27 empirical_analysis_manuscript.R

@@ -1,8 +1,19 @@
 library(sampbias)
 library(cowplot)
 library(ggplot2)
+library(sp)
+library(sf)
+library(rgdal)
 
 occ <-read.csv(system.file("extdata", "mammals_borneo.csv", package="sampbias"), sep = "\t")
+
+## get a polygon of Borneo
+data(landmass)
+
+born2 <- st_read("empirical_analyses/Borneo.kml")
+born2 <- sf:::st_zm(born2$geom)
+born2 <- as(born2, 'Spatial')
+born <- intersect(landmass, born2)
 
 
 # occ <- read.csv("original_data/mammals_borneo.csv", sep = "\t")
@@ -12,7 +23,7 @@ occ <-read.csv(system.file("extdata", "mammals_borneo.csv", package="sampbias"),
 # write.table(occ, file = "inst/extdata/mammals_borneo.csv", sep = "\t")
 # occ <-read.csv("inst/extdata/mammals_borneo.csv", sep = "\t")
 
-out <- calculate_bias(occ, res = 0.05, buffer = 0.5)
+out <- calculate_bias(occ, res = 0.05, buffer = 0.5, restrict_sample = born)
 
 p1 <- plot(out)
 summary(out)

### 2020-07-28 R/map_bias.R
change plotting and adapt simulations

@@ -1,7 +1,10 @@
 #' Mapping Projected Bias Effects
 #'
-#'A plotting function to visualize the bias effects calculated using \code{\link{calculate_bias}}
-#'and projected using \code{\link{project_bias}}.
+#'A plotting function to visualize the relative deviation of sampling rate from the maximum
+#' rate as calculated using \code{\link{calculate_bias}}
+#'and projected using \code{\link{project_bias}}. For instance, a value of -25 indicates
+#' a drop of 25% compared to the highest rate
+#' (e.g. in a road on river flowing through the city airport)
 #'
 #'
 #' @param x a raster stack as generate by \code{\link{project_bias}}
@@ -11,7 +14,8 @@
 #' used.
 #' @param sealine logical. Should the sealine be added to the plots? Default is
 #' to TRUE.
-#' @param logtransform logical. IF TRUE, the bias projections are logtransformed. Default is TRUE.
+#' @param sampling_ratelogical. logical. If true, the projected sampling rate depending on the
+#' biasing factors is plotted instead (logarithmically transformed to the base of 10). Default is to FALSE.
 #' @return A series of R plots based on ggplot2.
 
 #' @seealso \code{\link{calculate_bias}}, \code{\link{project_bias}}
@@ -51,7 +55,7 @@
 map_bias <- function(x,
                      gaz = NULL,
                      sealine = TRUE,
-                     logtransform = TRUE) {
+                     sampling_rate = FALSE) {
 
   # prepare gazetteers to be included for plotting
   if (!is.null(gaz)) {
@@ -95,28 +99,38 @@ map_bias <- function(x,
     pivot_longer(cols = -contains("geo"), values_to = "val", names_to = "split") %>%
     filter(split != "occurrences")
 
-  if(logtransform){
+
+  if(sampling_rate){
+    plo <- plo %>%
+      filter(split != "Total_percentage")
+
     out <- ggplot(plo)+
-      geom_raster(aes(x = .data$geo_lon, y = .data$geo_lat, fill = log10(.data$val)))
+      geom_raster(aes(x = .data$geo_lon, y = .data$geo_lat, fill = log10(.data$val)))+
+      facet_wrap(split~ .)+
+      coord_fixed()+
+      theme_bw()+
+      scale_fill_viridis(na.value = "transparent",
+                         option = "viridis",
+                         direction = 1,
+                         name = "Log Sampling rate",
+                         discrete = FALSE)+
[…]

### 2020-07-28 R/project_bias.R

@@ -65,6 +65,10 @@ project_bias <- function(x, factors = NULL) {
   # calculate the values for each raster cell
   lambdas <- list()
 
+  ## Get the mean estimate for the mean rate
+  mean_q <- colMeans(x$bias_estimate)[4]
+
+  ## calculate the sampling rate when increasingly adding biasing factors
   for (i in seq_along(mean_w)) {
     if(is.numeric(ras)){
       test <- ras
@@ -72,12 +76,17 @@ project_bias <- function(x, factors = NULL) {
       test <- ras[,1:i]
     }
 
-    lambdas[[i]] <- get_lambda_ij(q = colMeans(x$bias_estimate)[4],
+    lambdas[[i]] <- get_lambda_ij(q = mean_q,
                                   w = mean_w[1:i],
                                   X = test) %>%
       as.matrix()
   }
 
+  ## add a plot for the relative variation
+  perc <- round((lambdas[[length(lambdas)]] - mean_q) / mean_q * 100, 2)
+  lambdas[[length(lambdas) + 1]] <- perc
+
+  # re-transfor to a spatial raster
   out <- lapply(lambdas, "raster",
                 template = x$distance_rasters[[1]]) %>%
     c(x$occurrences) %>% stack()
@@ -89,7 +98,11 @@ project_bias <- function(x, factors = NULL) {
                              names(mean_w)[1:i]),
                         collapse = "+"))
   }
-  names(out) <- c(nam, "occurrences")
+
+  # add the total percentage change and the occurrences, which are also  part of the stack
+  nam <- c(nam, "Total_percentage", "occurrences")
+
+  names(out) <- nam
 
   # retun output object
   return(out)

### 2020-07-28 empirical_analyses_rarefaction.R

@@ -37,14 +37,20 @@ for(i in 1:length(ID)){
   p2 <- map_bias(proj)
 
   ggsave(p2, filename = paste("empirical_analyses/simulations/figure_empirical_results_spatial_projection_empirical_",
-                              rar, "_", ID[i], ".pdf", sep = ""),
+                              round(nrow(occ) * rar, 0), "_", ID[i], ".pdf", sep = ""),
+         height = 16, width = 16)
+
+  p3 <- map_bias(proj, sampling_rate = TRUE)
+
+  ggsave(p3, filename = paste("empirical_analyses/simulations/figure_empirical_results_spatial_projection_empirical_",
+                              round(nrow(occ) * rar, 0), "_", ID[i], "_sampling_rate.pdf", sep = ""),
          height = 16, width = 16)
 
 
   # prepare bias estimate output
   out <- out$bias_estimate
   out$ID <- ID[i]
-  out$rar <- rar
+  out$rar <- round(nrow(occ) * rar, 0)
   out$res <- res
   out$type <- "empirical"
 

### 2020-07-28 empirical_analyses_simulation.R

@@ -9,9 +9,14 @@ library(raster)
 library(readr)
 
 # rarefaction steps:
-rar <- c(1, 0.5, 0.25, 0.1, 0.01, 0.001)
+rar <- c(100000, 25000,
+         6262,
+         6262 * 0.5,
+         round(6262 * 0.25, 0),
+         round(6262 * 0.1, 0),
+         round(6262 * 0.01, 0))
 rar <- rar[6]
-ID <- 2:3
+ID <- 1:3
 res <- 0.05
 
 #rarefaction of a randomly sampled data set across the study area without bias
@@ -24,21 +29,17 @@ born2 <- sf:::st_zm(born2$geom)
 born2 <- as(born2, 'Spatial')
 born <- intersect(landmass, born2)
 
-
-## randomly sample points across the island
-occ <- sp::spsample(x = born, n = 100000, type = "random")
-occ <- data.frame(species = letters[1:25],
-                  coordinates(occ))
-names(occ) = c("species", "decimalLongitude", "decimalLatitude")
-
 # rarefaction loop for the bias calculation, write out the resulting paramter estimates and a map
 for(i in 1:length(ID)){
   print(i)
-  # subset the simulated data
-  sub <- occ[base::sample(x = 1:nrow(occ), size = round(nrow(occ) * rar, 0)), ]
+  # randomly sample points across the island
+  occ <- sp::spsample(x = born, n = rar, type = "random")
+  occ <- data.frame(species = "A",
+                    coordinates(occ))
+  names(occ) = c("species", "decimalLongitude", "decimalLatitude")
 
   # calculate sampling bias
-  out <- calculate_bias(sub, res = res, buffer = 0.5, restrict_sample = born)
+  out <- calculate_bias(occ, res = res, buffer = 0.5, restrict_sample = born)
 
   # plot of bias projection in space
   proj <- project_bias(out)
@@ -48,6 +49,12 @@ for(i in 1:length(ID)){
                               ID[i], "_", rar, ".pdf", sep = ""),
          height = 16, width = 16)
 
+
+  p3 <- map_bias(proj, sampling_rate = TRUE)
+
+  ggsave(p3, filename = paste("empirical_analyses/simulations/figure_empirical_results_spatial_projection_simulated_",
+                              rar, "_", ID[i], "_sampling_rate.pdf", sep = ""),
+         height = 16, width = 16)
 
   #prepare output files
   out <- out$bias_estimate

### 2020-07-28 empirical_analysis_manuscript.R

@@ -5,16 +5,8 @@ library(sp)
 library(sf)
 library(rgdal)
 
-occ <-read.csv(system.file("extdata", "mammals_borneo.csv", package="sampbias"), sep = "\t")
-
-## get a polygon of Borneo
-data(landmass)
-
-born2 <- st_read("empirical_analyses/Borneo.kml")
-born2 <- sf:::st_zm(born2$geom)
-born2 <- as(born2, 'Spatial')
-born <- intersect(landmass, born2)
-
+# generate input example data
+# occ <-read.csv(system.file("extdata", "mammals_borneo.csv", package="sampbias"), sep = "\t")
 
 # occ <- read.csv("original_data/mammals_borneo.csv", sep = "\t")
 #
@@ -22,6 +14,17 @@ born <- intersect(landmass, born2)
 #
 # write.table(occ, file = "inst/extdata/mammals_borneo.csv", sep = "\t")
 # occ <-read.csv("inst/extdata/mammals_borneo.csv", sep = "\t")
+
+## get a polygon of Borneo
+data(landmass)
+data(borneo)
+
+occ<- read.csv(system.file("extdata",
+                                   "mammals_borneo.csv",
+                                   package="sampbias"),
+                       sep = "\t")
+
+
 
 out <- calculate_bias(occ, res = 0.05, buffer = 0.5, restrict_sample = born)
 

### 2020-07-28 vignettes/Possible_warnings_of_sampbias.R

@@ -1,3 +1,3 @@
-## ----setup, include=FALSE------------------------------------------------
+## ----setup, include=FALSE-----------------------------------------------------
 knitr::opts_chunk$set(echo = TRUE)
 

### 2020-07-28 vignettes/Using_the_samp_bias_rpackage.R

@@ -1,5 +1,5 @@
 ## ----setup, include=FALSE-----------------------------------------------------
-knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE)
+knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE, fig.pos = 'H')
 
 ## ---- eval = FALSE------------------------------------------------------------
 #  require(devtools)
@@ -22,11 +22,11 @@ occ <-read.csv(system.file("extdata",
                            package="sampbias"),
                sep = "\t")
 
-## -----------------------------------------------------------------------------
+## ---- warning = FALSE, message = FALSE----------------------------------------
 cit <- readOGR(dsn = system.file("extdata", package="sampbias"), 
-               layer = "Borneo_major_cities")
+               layer = "Borneo_major_cities", verbose = FALSE)
 roa <- readOGR(dsn = system.file("extdata", package="sampbias"), 
-               layer = "Borneo_major_roads")
+               layer = "Borneo_major_roads", verbose = FALSE)
 
 gazetteers <- list(cities = cit,
                    roads = roa)
@@ -34,15 +34,15 @@ gazetteers <- list(cities = cit,
 ## -----------------------------------------------------------------------------
 bias.out <- calculate_bias(x = occ, gaz = gazetteers)
 
-## ---- eval = FALSE------------------------------------------------------------
-#  summary(bias.out)
-#  plot(bias.out)
+## ---- eval = TRUE, fig.cap = "Visulaization of the output of a sampbias analysis. A) Posterior estimates of the bias weights for cities and roads, B) the decay of the sampling rate with increasing distaince from cities and roads."----
+summary(bias.out)
+plot(bias.out)
 
-## -----------------------------------------------------------------------------
+## ---- fig.cap = "Sampling bias effect projected in space. Note the coarse grid resolution which should be increased."----
 proj <- project_bias(bias.out)
 map_bias(proj)
 
-## -----------------------------------------------------------------------------
+## ---- fig.cap="Example of a customized study area which can be provided to the claculate_bias function via the restrict_sample option. Only grid cells within the red area are included in the sampling bias calculation."----
 data(area_example)
[…]

### 2020-07-30 R/map_bias.R
manuscript changes

@@ -4,7 +4,7 @@
 #' rate as calculated using \code{\link{calculate_bias}}
 #'and projected using \code{\link{project_bias}}. For instance, a value of -25 indicates
 #' a drop of 25% compared to the highest rate
-#' (e.g. in a road on river flowing through the city airport)
+#' (e.g. in a road on river flowing through the city airport).
 #'
 #'
 #' @param x a raster stack as generate by \code{\link{project_bias}}
@@ -14,7 +14,7 @@
 #' used.
 #' @param sealine logical. Should the sealine be added to the plots? Default is
 #' to TRUE.
-#' @param sampling_ratelogical. logical. If true, the projected sampling rate depending on the
+#' @param sampling_rate logical. If true, the projected sampling rate depending on the
 #' biasing factors is plotted instead (logarithmically transformed to the base of 10). Default is to FALSE.
 #' @return A series of R plots based on ggplot2.
 

### 2020-07-30 vignettes/Using_the_samp_bias_rpackage.R

@@ -53,7 +53,6 @@ bias.out <- calculate_bias(x = occ,
                            gaz = gazetteers, 
                            restrict_sample = area_example)
 
-
 ## ---- fig.cap="Example of a more complex customized study area representing the Borneo montane rain forests ecoregion from Olson et al. (2001). Only grid cells within red area are included in the sampling bias calculation."----
 data(ecoregion_example)
 

### 2020-07-30 appendix_S1.Rmd
change plotting and adapt simulations

@@
 
-```{r, fig.height = 8, fig.cap="The example dataset of mammal occurrences from the island of Borneo, as downloaded from www.gbif.org (n = 6,262, 2016, https://doi.org/10.15468/dl.7fg4zx), and the geographic gazetters of main cities, roads, rivers and airports used for the sampbias analysis."}
+```{r, fig.height = 8, fig.cap="The example dataset of mammal occurrences from the island of Borneo, as downloaded from www.gbif.org (n = 6,262, 2016, https://doi.org/10.15468/dl.7fg4zx), and the geographic gazetters of main cities, roads, rivers and airports used for the sampbias analysis.
+Note that the records in the sea around Borneo were excluded by with the help of the user-provided study area."}
 #libraries
@@
 
-born2 <- st_read("empirical_analyses/Borneo.kml")
+born2 <- st_read("empirical_analyses/Borneo.kml", quiet = TRUE)
 born2 <- sf:::st_zm(born2$geom)
@@
 
-born2 <- st_read("empirical_analyses/Borneo.kml", quiet = TRUE)
-born2 <- sf:::st_zm(born2$geom)
-born2 <- as(born2, 'Spatial')
 born <- raster::intersect(landmass, born2)
@@
 dat <- read_csv("empirical_analyses/simulations/weight_estimates.csv") %>% 
-  mutate(records = ifelse(type=="empirical", rar * nrow(occ), rar * 100000)) %>% 
-  mutate(records_area = round(records / area * 10000, 2)) %>% 
+  mutate(records_area = round(rar / area * 10000, 2)) %>% 
   select(contains("w"), ID, type, records_area) %>% 
@@
 
-```{r, out.width = "6in", fig.cap = "Projected bias surface for the distribution data of mammals from Borneo (www.gbif.org, 2016, https://doi.org/10.15468/dl.7fg4zx), at different levels of downsampling, from a random replicate."}
+```{r, out.width = "6in", fig.cap = "Projected bias surface for the distribution data of mammals from Borneo (www.gbif.org, 2016, https://doi.org/10.15468/dl.7fg4zx), at different levels of downsampling, from a random replicate.
+The plotted extent is adapted to the study area, which may not cover the entire island for rarified datasets with few records (A) and B))"}
 
@@
                    y = posterior_estimate,
-                 fill = as.factor(records_area)))+
+                   fill = as.factor(records_area),
+                   color = as.factor(records_area)))+
   ylab("Posterior weight")+
@@
   scale_fill_viridis(discrete = TRUE, name = "Records per\n10,000 sqkm")+
+  scale_color_viridis(discrete = TRUE, name = "Records per\n10,000 sqkm")+
   theme_bw()
+```
+
+```{r, echo = FALSE, fig.height = 8, fig.cap="The bias weights ($w$) defining the effects of each bias factor estimated for the smallest simulated dataset without bias."}
+
+ggplot(data = plo)+
+  geom_histogram(aes(posterior_estimate, fill = as.factor(records_area)), bins = 50) +
+  scale_fill_viridis(discrete = TRUE, name = "Records per\n10,000 sqkm")+
+  facet_grid(rows = vars(records_area), cols = vars(bias))+
+  scale_x_continuous(breaks = c(0,0.003, 0.006, 0.009), limits = c(0,0.01))+
+  xlab("Posterior weight")+
+  ylab("Count")+
+  theme_bw()+
+  theme(panel.grid.minor = element_blank(),
+        legend.position = "none")
+
 ```

### 2020-07-30 manuscript.rmd

@@
 rainforest) or an area of special interest (e.g.
-a national park)}.
+a national park).
+In this case all occurrence records outside the user-defined area will be disregarded for the analysis.}.
 For each grid cell $i$, we then compute a vector $X_i(j)$ of minimum distances (straight aerial distance, "as the crow flies") to each bias factor $j \in B$.
@@
 For instance, the posterior weights estimated in our empirical study ranged from 0 to 0.013, while the prior distribution applied on these parameters, Exp(1) has a} 95% \textcolor{blue}{of it density in range 0 – 3.
-Custom priors, within the flexible family of gamma distributions are possible via the} `prior_q` and `prior_w` \textcolor{blue}{arguments of the} `calculate_bias` \textcolor{blue}{function.}
+Custom priors, within the flexible family of gamma distributions, which include the exponential priors used here, are possible via the} `prior_q` and `prior_w` \textcolor{blue}{arguments of the} `calculate_bias` \textcolor{blue}{function.}
 
@@
 
-We summarize the results by mapping the estimated sampling rates ($\lambda_i$) across space.
+We summarize the results by mapping the estimated `textcolor{blue}{(the relative deviation of sampling rate from the maximum
+sampling rate (or on user choice the estimated sampling rates}, $\lambda_i$) across space.
 These rates represent the expected number of sampled occurrences for each grid cell and provide a graphical representation of the spatial variation of sampling rates.
@@
 *Sampbias* \textcolor{blue}{is designed to work with sparsely sampled datasets, and to estimate bias effects from datasets with low coverage.
-To test the performence of} *sampbias* \textcolor{blue}{on small datasets we ran two simulation experiments: 1) we ran the analyses as presented above on the distribution of mammals on Borneo, sub-sampling of the initial dataset (to 6,262, 3,131, 1,566, 626, 62 records for all of Borneo respectively, 5 replicates each).
-The results showed that estimates of} $\mathbf{w}$ \textcolor{blue}{ and the projection of the bias effect in space in a biased dataset were robust to decreases in data, although uncertainty increased (Figs.
+To test the performance of} *sampbias* \textcolor{blue}{on small datasets we ran two simulation experiments: 1) we ran the analyses as presented above on the distribution of mammals in Borneo, sub-sampling of the initial dataset (to 6,262, 3,131, 1,566, 626, 62 records for all of Borneo respectively, 5 replicates each).
+The results showed that parameter estimates and the projection of the bias effect in space were robust to decreases in data, although uncertainty increased (Figs.
 S2nd S3, in Appendix S2).
@@
 \textcolor{blue}{The results show that} *sampbias* \textcolor{blue}{reliably rejected a biasing effect even for small datasets (Figs.
-S4 and S5, in Appendix S2), however, we note that the analyses took a very long time to converge for datasets with no sampling bias and datasets with less than 10,000 points.
+S4 and S5, in Appendix S2).
 In general, the more records, the more precise the parameter estimates will be.}
@@
 
-```{r spatial, out.width = "6in", echo = FALSE, fig.cap = "Spatial projection of the estimated sampling rates in an empirical example dataset of mammal occurrences on the Indonesian island of Borneo (downloaded from www.gbif.org.
+```{r spatial, out.width = "6in", echo = FALSE, fig.cap = "Spatial projection of the sampling bias in an empirical example dataset of mammal occurrences on the Indonesian island of Borneo (downloaded from www.gbif.org.
 GBIF.org, 2016).
-The colours show the projection of the \\textcolor{blue}{log-transformed} sampling rates (i.e.
-expected number of occurrences per cell) given the inferred \textit{sampbias} model.
+The colours show the projection of the \\textcolor{blue}{the relative deviation of sampling rate from the maximum sampling rate given the inferred} \\textit{sampbias} model.
+For instance, a value of -25 indicates a drop of 25\\% compared to the highest sampling rate.
 The highest undersampling is in the centre of the island.", eval = TRUE}

### 2020-07-30 title_page.Rmd
manuscript changes

@@
 title: "sampbias, a method for quantifying geographic sampling biases in species distribution data"
-# title: "A global Red List of orchids supports the value of automated conservation assessment"
 author: "Alexander Zizka^1,2^, Alexandre Antonelli^3,4,5^, Daniele Silvestro^3,4,6^"
@@
 
-**Word count** -- Abstract: 3,051 words; Total: 208 words, 2 figures
+**Word count** -- Abstract: 207 words; Main text (excluding references): 3019 words, 2 figures
 

### 2020-07-30 appendix_S1.Rmd
changes hyperprior

@@
   select(contains("w"), ID, type, records_area) %>% 
+  select(-w_rate) %>% 
     pivot_longer(cols = contains("w_"), names_to = "bias",

### 2020-07-30 R/internal.R

@@ -1,5 +1,5 @@
 #'@importFrom raster rasterize
-#'@importFrom stats dgamma dpois rnorm runif
+#'@importFrom stats dgamma dpois rnorm runif rgamma
 
 #Occurrence raster
 .OccRast <- function(x, ras){
@@ -48,7 +48,7 @@ multiplier_proposal <- function(i, d = 1.2) {
 
 get_post_rate <- function(w,alpha){
   a0 <- 1
-  b0 <- 0.1
+  b0 <- 0.001
   rate <- rgamma(1, a0 + length(w) * alpha, rate = b0 + sum(w))
   return(rate)
 }
@@ -101,12 +101,12 @@ get_post_rate <- function(w,alpha){
     w <- wA
     q <- qA
     mh <- TRUE
-    
+
     if (runif(1) < 0.02) {
       prior_w[2] <- get_post_rate(wA, prior_w[1])
       mh <- FALSE
       hastings <- 0
-    
+
     } else if (runif(1) < 0.3) {
       update <- multiplier_proposal(qA, d = 1.1)
       q <- update[[1]]
@@ -123,7 +123,7 @@ get_post_rate <- function(w,alpha){
     }else{
 	lik <- likA
     }
-    
+
     prior <- sum(dgamma(w, prior_w[1], rate=prior_w[2], log = TRUE)) + dgamma(q, prior_q[1], rate=prior_q[2], log = TRUE)
 
     if (mh == FALSE | (lik + prior) - (likA + priorA) + hastings >= log(runif(1))) {

### 2020-07-31 manuscript.rmd
hyper-prior and updated simulation text

@@
 For each grid cell $i$, we then compute a vector $X_i(j)$ of minimum distances (straight aerial distance, "as the crow flies") to each bias factor $j \in B$.
-The resolution of the grid defines the precision of the distance estimates, for instance a 1x1 degree raster will yield approximately a 110 km precision at the equator.
+The resolution of the grid defines the precision of the distance estimates, for instance a $1 \times 1$ degree raster will yield approximately a 110 km precision at the equator.
 Due to the assumption of homogeneous sampling and a computational trade-off between the resolution of the regular grid and the extent of the study area (for instance, a 1 second resolution for a global dataset would become computationally prohibitive in most practical cases), *sampbias* is best suited for local or regional datasets at high resolution (c.
@@
 Additionally, we model $\lambda_i$ to decrease exponentially as a function of distance from bias factors, such that increasing distances will result in a lower sampling rate.
-For a single bias factor the rates of cell $i$ with distance $X_i$ from a bias is:
+For a single bias factor the rate of cell $i$ with distance $X_i$ from a bias is:
 $$
@@
 not allowing negative values) and yet vague enough to encompass a much wider range of parameter space than the range of values observed in empirical tests.
-For instance, the posterior weights estimated in our empirical study ranged from 0 to 0.013, while the prior distribution applied on these parameters, Exp(1) has a} 95% \textcolor{blue}{of it density in range 0 – 3.
-Custom priors, within the flexible family of gamma distributions, which include the exponential priors used here, are possible via the} `prior_q` and `prior_w` \textcolor{blue}{arguments of the} `calculate_bias` \textcolor{blue}{function.}
+Custom priors, within the flexible family of gamma distributions, which include the exponential priors used here, are possible via the} `prior_q` and `prior_w` \textcolor{blue}{arguments of the} `calculate_bias` \textcolor{blue}{function.
+Additionally, since the null expectation in the absence of biases is that the weights are close to 0, we implement a hierarchical model, in which the rate parameter of the gamma prior can be itself estimated from the data.
+Thus, we set $P(\mathbf{w}) \sim \Gamma(a=1, b)$ and assign a vague hyper-prior on the rate $P(b) \sim \Gamma(\alpha_0=1, \beta_0=0.001)$.
+The choice of conjugate gamma distribution allows us to sample the rate directly from its posterior distribution:
+$$
+b \sim \Gamma(\alpha_0 + a B,\ \beta_0  \sum_{j=1}^{B}{w_j}) \ \ \ \ (3)
+$$
+The use of a hyper-prior has the advantage of making the prior on the weights more flexible and able to adapt to different datasets reducing the need for user-defined arbitrary choices.
+Additionally, it works as a regularization technique reducing the risks of over-parameterization, by shrinking the weights around small values when there is no evidence in the data indicating a bias.
+<!--For instance, the posterior weights estimated in our empirical study ranged from 0 to 0.013, while the prior distribution applied on these parameters, Exp(1) has a} 95% \textcolor{blue}{of it density in range 0 – 3.
+C}
+-->
 
@@
 
+We then simulated unbiased datasets to assess whether the estimated weights differed significantly from a null expectation. 
+We randomly sampled occurrence records across Borneo to generate an unbiased dataset.
+We sampled 6,262 records and ran a} *sampbias* \textcolor{blue}{analysis with the same settings as for the empirical dataset}.
+We replicated these steps 10 times and combined the posterior samples of the weights across replicates.
+We compared the empirical weight estimates against their expected range under unbiased datasets and found that ...
+[\textcolor{red}{NEEDS TO BE UPDATED with final results}]
+
 *Sampbias* \textcolor{blue}{is designed to work with sparsely sampled datasets, and to estimate bias effects from datasets with low coverage.
-To test the performance of} *sampbias* \textcolor{blue}{on small datasets we ran two simulation experiments: 1) we ran the analyses as presented above on the distribution of mammals in Borneo, sub-sampling of the initial dataset (to 6,262, 3,131, 1,566, 626, 62 records for all of Borneo respectively, 5 replicates each).
+To test the performance of} *sampbias* \textcolor{blue}{on small datasets we ran additional simulation experiments.
+We ran the analyses as presented above on the distribution of mammals in Borneo, sub-sampling of the initial dataset (to 6,262, 3,131, 1,566, 626, 62 records for all of Borneo respectively, 5 replicates each).
 The results showed that parameter estimates and the projection of the bias effect in space were robust to decreases in data, although uncertainty increased (Figs.
 S2nd S3, in Appendix S2).
-2) in a second simulation we  randomly sampled occurrence records across Borneo as an example of an unbiased dataset.
-We sampled 100,000, 50,000, 25,000, and 10,000 records and ran a} *sampbias* \textcolor{blue}{analysis with the same settings as for the empirical dataset}.
+
+[\textcolor{red}{NEEDS TO BE UPDATED with final results}]
 \textcolor{blue}{The results show that} *sampbias* \textcolor{blue}{reliably rejected a biasing effect even for small datasets (Figs.

### 2020-07-31 appendix_S1.Rmd
notes appendix S1

@@
 
+REWRITE THIS PART TO DEAL WITH THE NEW TYPE OF OUTPUT FILES
+
 dat <- read_csv("empirical_analyses/simulations/weight_estimates.csv") %>% 
@@
 
+REDO WITH THE NEW RESULTS
+
 knitr::include_graphics(path = "ms_figures/figure_projection_empirical_rarefaction.jpg")
@@
 Datasets generated by random simulation of records across Borneo (no bias)."}
+
+Here a boxplot comparing the results form the empirical analyses with the ones fromt he randomly simulated analyses
+
 plo <- dat %>%  filter(type == "simulated") %>%
@@
 
+Here a histogram of the weight estimates
+
 ggplot(data = plo)+
@@
 ```
-
-```{r, out.width = "6in", fig.cap = "Projected bias surface for randomly sampled points across Borneo, from a random replicate."}
-
-knitr::include_graphics(path = "ms_figures/figure_projection_empirical_simulation.jpg")
-
-```

### 2020-07-31 03_main_manuscript.rmd
manuscript updates and script re-organization

--- 2020-07-31 manuscript.rmd
@@
 However, these datasets are typically not collected systematically and rarely include information on collection effort.
-Instead, they  are often compiled from a variety of sources (e.g.
-scientific expeditions, census counts, genetic barcoding studies, and citizen-science observations).
+Instead, they  are often compiled from a variety of sources (e.g., scientific expeditions, census counts, genetic barcoding studies and citizen-science observations).
 Species occurrences are therefore often subject to multiple sampling biases [@Meyer2016].
 
-Sampling biases that may affect the recording of species occurrences [presence, absence and abundance, @Isaac2015; @Boakes2010] include the under-sampling of specific taxa ("taxonomic bias", e.g., birds *vs.* nematodes), specific geographic regions ("geographic bias", i.e.
-easily accessible *vs.* remote areas), and specific temporal periods ("temporal bias", i.e.
-wet season *vs.* dry season).
+Sampling biases that may affect the recording of species occurrences [presence, absence and abundance, @Isaac2015; @Boakes2010] include the under-sampling of specific taxa ("taxonomic bias", e.g., birds *vs.* nematodes), specific geographic regions ("geographic bias", e.g., easily accessible *vs.* remote areas), and specific temporal periods ("temporal bias", e.g., wet *vs.* dry season).
 In particular geographic sampling bias---the fact that sampling effort is spatially biased, rather than equally distributed over the study area---is likely to be widespread in all non-systematically collected datasets of species distributions.
 
-Many aspects can lead to sampling biases, including socio-economic factors [i.e.
-national research spending, history of scientific research; @Zizka2020c; @Meyer2015; @Daru2018], political factors [armed conflict, democratic rights; @Ryden2019], and physical accessibility [i.e.
-distance to a road or river, terrain conditions, slope; @Yang2014; @Botts2011].
-Especially physical accessibility by people is omnipresent as a bias factor [e.g.
-@Lin2015; @Engemann2015; @Kadmon2004], across spatial scales, as the commonly used term "roadside bias" testifies.
-In practice, this means that most species observations are made in or near cities, along roads, paths, and rivers, and near human settlements.
-Relatively fewer observations are expected to be available from inaccessible areas in e.g.
-a tropical rainforest or a mountain top.
+Many aspects can lead to sampling biases, including socio-economic factors [e.g., national research spending, history of scientific research; @Zizka2020c; @Meyer2015; @Daru2018], political factors [e.g., armed conflict, democratic rights; @Ryden2019], and physical accessibility [e.g., distance to a road or river, terrain conditions, slope; @Yang2014; @Botts2011].
+Especially physical accessibility by people is omnipresent as a bias factor [e.g., @Lin2015; @Engemann2015; @Kadmon2004], across spatial scales, as the commonly used term "roadside bias" testifies.
+In practice, this means that most species observations are made in or near cities, along roads, paths, rivers and near human settlements.
+Relatively fewer observations are expected to be available from inaccessible areas in e.g., a tropical rainforest or a mountain top.
 Since the recording of different taxonomic groups poses different challenges, geographic sampling bias and the effect of accessibility may differ among taxonomic groups [@Vale2012].
@@
 \textcolor{blue}{The implications of not considering geographic sampling biases in biodiversity research are likely substantial} [@Meyer2016; @Yang2013; @Barbosa2013].
-\textcolor{blue}{The presence of geographic sampling biases is broadly recognized} [e.g.
-@Kadmon2004], \textcolor{blue}{and approaches exist to account for it in some analyses---such as species-richness estimates} [@Engemann2015], \textcolor{blue}{occupancy models} [@Kery2016], \textcolor{blue}{and abundance estimates} [@Shimadzu2015].
+\textcolor{blue}{The presence of geographic sampling biases is broadly recognized} [e.g., @Kadmon2004], \textcolor{blue}{and approaches exist to account for it in some analyses---such as species-richness estimates} [@Engemann2015], \textcolor{blue}{occupancy models} [@Kery2016], \textcolor{blue}{and abundance estimates} [@Shimadzu2015].
 \textcolor{blue}{In the case of species distribution modelling---the statistical estimation of species geographic distributions based on known occurrences and environmental conditions---geographically biased sampling is problematic because it often causes environmentally biased sampling which decreases model performance} [@Kadmon2004; @Lobo2011; @Bystriakova2012; @Kramer-Schadt2013; @Varela2014].
@@
 
-It is unrealistic to expect that accessibility bias in biodiversity data will ever disappear even after more automated observation technologies are developed.
+It is unrealistic to expect that accessibility bias in biodiversity data will disappear even after more automated observation technologies are developed.
 It is therefore crucial that researchers realise the intrinsic biases associated with the data they deal with, \textcolor{blue}{especially in cross-taxonomic studies, since occurrence datasets from different taxa are likely differently affected by sampling biases due to differences in specimen collection and transportation}.
-This is the first step towards estimating to which extent these biases may affect their analyses, results, and conclusions.
+This is the first step towards estimating to which extent these biases may affect their analyses, results and conclusions.
 Any study dealing with species occurrence data should arguably assess the strength of accessibility biases in the underlying data.
@@
 
-Here, we present *sampbias* 1, a probabilistic method to quantify accessibility bias in datasets of species occurrences.
+Here, we present *sampbias* v1.0.4, a probabilistic method to quantify accessibility bias in datasets of species occurrences.
 *Sampbias* is implemented as a user-friendly R-package and uses a Bayesian approach to address three questions:
@@
 
-3) How is accessibility bias distributed in space, i.e. which areas are a priority for targeted sampling?
+3) How is accessibility bias distributed in space?
 
@@
 *Sampbias* offers an easy and largely automated means for biodiversity scientists and non-specialists alike to explore bias in species occurrence data, in a way that is comparable across datasets.
-\textcolor{blue}{The results may be used to identify priorities for further collection or digitalization efforts, assess the reliability of scientific results based on publicly available species distribution data.}
+\textcolor{blue}{The results may be used to identify priorities for further collection or digitalization efforts and to assess the reliability of scientific results based on publicly available species distribution data.}
 
@@
 ## General concept
-Under the assumption that organisms exist across the entire area of interest, we can expect the number of sampled occurrences in a restricted area, such as a single biome, to be distributed uniformly in space (even though, of course, the density of individuals and the species diversity may be heterogeneous).
+Under the assumption that organisms occur across the entire area of interest, we can expect the number of sampled occurrences in a restricted area, such as a single biome, to be distributed uniformly in space (even though, of course, the density of individuals and the species diversity may be heterogeneous).
 With *sampbias* we assess to which extent variation in sampling rates can be explained by distance from bias factors.
@@
 Species occurrence data as downloaded from the data portal of GBIF can be directly used as input data for *sampbias*.
-The output of the package includes measures of the sampling rates across space, which are comparable between different gazetteers (e.g.
-comparing the biasing effect of roads and rivers), different taxa (e.g.
-birds *vs.* flowering plants) and different data sets (e.g.
-specimens *vs.* human observations).
+The output of the package includes measures of the sampling rates across space, which are comparable between different gazetteers (e.g., comparing the biasing effect of roads and rivers), different taxa (e.g., birds *vs.* flowering plants) and different data sets (e.g., specimens *vs.* human observations).
 
@@
 *Sampbias* uses gazetteers of the geographic location of bias factors (hereafter indicated with B) to generate a regular grid across the study area.
-\textcolor{blue}{By default the study area is defined by the geographic extent of the study dataset, but it can also be customized via user-defined polygons, for instance to limit the analyses to an environmentally homogeneous region (e.g.
-rainforest) or an area of special interest (e.g.
-a national park).
+\textcolor{blue}{By default the study area is defined by the geographic extent of the study dataset, but it can also be customized via user-defined polygons, for instance to limit the analyses to an environmentally homogeneous region (e.g., a rainforest) or an area of special interest (e.g., a national park).
 In this case all occurrence records outside the user-defined area will be disregarded for the analysis.}.
@@
 100 -- 10,000 m).
-\textcolor{blue}{Since the differences in grid cell size are negligible on the local and regional scale, sampbias uses a latitude/longitude grid by default, but a custom grid in any projection and coordinate reference system---for instance an equal area grid---may be provided by the user.}
+\textcolor{blue}{Since the differences in grid cell size are negligible on the local and regional scale, sampbias uses a latitude/longitude grid by default, but a custom grid in any projection and coordinate reference system---for instance an equal area grid, which is often more suitable for large spatial analyses---may be provided by the user.}
 
@@
 where $w \in \mathbb{R^+}$ defines the steepness of the Poisson rate decline, such that $w \approx 0$ results in a null model of uniform sampling rate $q$ across cells.
-In the presence of multiple bias factors (e.g.
-roads and rivers), the sampling rate decrease is a function of the cumulative effects of each bias and its distance from the cell:
+In the presence of multiple bias factors (e.g., roads and rivers), the sampling rate decrease is a function of the cumulative effects of each bias and its distance from the cell:
 $$
@@
 The likelihood is then multiplied across the $N$ cells considered.
-We used exponential priors on the parameters $q$ and $\mathbf{w}$, $P(q) \sim \Gamma(1, 0.01)$ and $P(\mathbf{w}) \sim \Gamma(1, 1)$, respectively.
+We use exponential priors on the parameters $q$ and $\mathbf{w}$, $P(q) \sim \Gamma(1, 0.01)$ and $P(\mathbf{w}) \sim \Gamma(1, 1)$, respectively.
 \textcolor{blue}{We chose exponential priors because they represent the standard choice for rate parameters such as} $q$ \textcolor{blue}{and the weights} $\mathbf{w}$\textcolor{blue}{, all of which must be positive and have support [0, +inf].
@@
 Additionally, since the null expectation in the absence of biases is that the weights are close to 0, we implement a hierarchical model, in which the rate parameter of the gamma prior can be itself estimated from the data.
-Thus, we set $P(\mathbf{w}) \sim \Gamma(a=1, b)$ and assign a vague hyper-prior on the rate $P(b) \sim \Gamma(\alpha_0=1, \beta_0=0.001)$.
-The choice of conjugate gamma distribution allows us to sample the rate directly from its posterior distribution:
+Thus, we set} $P(\mathbf{w}) \sim \Gamma(a=1, b)$ \textcolor{blue}{and assign a vague hyper-prior on the rate} $P(b) \sim \Gamma(\alpha_0=1, \beta_0=0.001)$.
+\textcolor{blue}{The choice of conjugate gamma distribution allows us to sample the rate directly from its posterior distribution:}
 $$
@@
 $$
-The use of a hyper-prior has the advantage of making the prior on the weights more flexible and able to adapt to different datasets reducing the need for user-defined arbitrary choices.
-Additionally, it works as a regularization technique reducing the risks of over-parameterization, by shrinking the weights around small values when there is no evidence in the data indicating a bias.
+\textcolor{blue}{The use of a hyper-prior has the advantage of making the prior on the weights more flexible and able to adapt to different datasets reducing the need for user-defined arbitrary choices.
+Additionally, it works as a regularization technique reducing the risks of over-parametrization, by shrinking the weights around small values when there is no evidence in the data indicating a bias.}
 <!--For instance, the posterior weights estimated in our empirical study ranged from 0 to 0.013, while the prior distribution applied on these parameters, Exp(1) has a} 95% \textcolor{blue}{of it density in range 0 – 3.
@@
 We interpret the magnitude of the elements in $\mathbf{w}$ as a function of the importance of the individual biases.
-\textcolor{blue}{We note that although this test is not explicitly intended to assess the significance of each bias factor, for which a Bayesian variable selection method could be used, it can quantify the expected amount of bias in the data predicted by single or multiple predictors in order to identify under-sampled and unexplored areas.
+\textcolor{blue}{We note, however, that this test is not explicitly intended to assess the significance of each bias factor.
 Because several bias factors might be correlated (e.g.
-cities, and airports), cumulating their bias estimates from independent analyses of each factor would result in an overestimation of the bias.
-For this reason, it is important to jointly estimate the effects of correlated factors, as this is based on the likelihood of the data given the combined effects of all biasing factors.}
+cities and airports), simply summing their effect from independent analyses would result in an overestimation of the total bias.
+It is therefore important to jointly estimate the effects of correlated factors, as this is based on the likelihood of the data given the combined effects of all biasing factors.
+A Bayesian variable selection method could be used to quantify the expected amount of bias in the data predicted by single or a particular combination of predictors, but falls beyond the scope of the current study.}
 
@@
 
-## Example and Empirical validation
+## Example and empirical validation
 A default *sampbias* analysis can be run with few lines of code in R.
@@
 
-To exemplify the use of *sampbias*, we downloaded the occurrence records of all mammals available from the island of Borneo [n = 6,262, @gbifdoi2016], and ran *sampbias* using the default gazetteers as shown in the example code below, to test the biasing effect of the main airports, cities and roads in the dataset.
+To exemplify the use of *sampbias*, we downloaded the occurrence records of all mammals available from the island of Borneo [n = 6,262, @gbifdoi2016] and ran *sampbias* using the default gazetteers as shown in the example code below, to test the biasing effect of the main airports, cities and roads in the dataset.
 The example dataset is provided with *sampbias*.
@@
 ## 'buffer' defines the buffer around the study area to account for biasing
-## structure adjecent to the study area, in degrees latitude and longitude
+## structure adjacent to the study area, in degrees latitude and longitude
 ## All other options at default, see ?calculate_bias for a description
@@
 
-We then simulated unbiased datasets to assess whether the estimated weights differed significantly from a null expectation. 
-We randomly sampled occurrence records across Borneo to generate an unbiased dataset.
-We sampled 6,262 records and ran a} *sampbias* \textcolor{blue}{analysis with the same settings as for the empirical dataset}.
-We replicated these steps 10 times and combined the posterior samples of the weights across replicates.
-We compared the empirical weight estimates against their expected range under unbiased datasets and found that ...
-[\textcolor{red}{NEEDS TO BE UPDATED with final results}]
+\textcolor{blue}{We ran simulation experiment to assess whether the estimated bias weights differ significantly from a null expectation of random sampling.
+To do so, we first created ten replicates of an unbiased data by generating 6262 random occurrences across Borneo (the same number as in the empirical dataset) and then ran a} *sampbias* \textcolor{blue}{with the same settings as for the empirical dataset on each of these replicates.
+We found that the estimated bias range were significantly different (credible intervals non-overlapping) than in the null model for cities, roads and airports (Figure S2 in Appendix S1).}
 
 *Sampbias* \textcolor{blue}{is designed to work with sparsely sampled datasets, and to estimate bias effects from datasets with low coverage.
-To test the performance of} *sampbias* \textcolor{blue}{on small datasets we ran additional simulation experiments.
-We ran the analyses as presented above on the distribution of mammals in Borneo, sub-sampling of the initial dataset (to 6,262, 3,131, 1,566, 626, 62 records for all of Borneo respectively, 5 replicates each).
+To test the performance of} *sampbias* \textcolor{blue}{on small datasets we ran an additional simulation experiment.
+We ran the analyses with the same settings as for the empirical dataset on a set of rarified datasets, sub-sampling the initial dataset to 3,131, 626, 62 records respectively (three replicates each), and then compared the estimated bias weights for all bias factors to the estimates for the empirical  dataset.
 The results showed that parameter estimates and the projection of the bias effect in space were robust to decreases in data, although uncertainty increased (Figs.
-S2nd S3, in Appendix S2).
-
-[\textcolor{red}{NEEDS TO BE UPDATED with final results}]
-\textcolor{blue}{The results show that} *sampbias* \textcolor{blue}{reliably rejected a biasing effect even for small datasets (Figs.
+S3 and S4, in Appendix S1).
+The results show that} *sampbias* \textcolor{blue}{reliably rejected a biasing effect even for small datasets (Figs.
 S4 and S5, in Appendix S2).
@@
 <!-- empirical results -->
-```{r empirical, out.width = "6in", echo = FALSE, fig.cap = "Results of the empirical validation analysis, estimating the accessibility bias in mammal occurrences from Borneo).
-A) bias weights ($w$) defining the effects of each bias factor, B) sampling rate as function of distance to the closest instance of each bias factor (i.e.
-expected number of occurrences) given the inferred \\textit{sampbias} model.
+```{r empirical, out.width = "6in", echo = FALSE, fig.cap = "Results of the empirical validation analysis, estimating the accessibility bias in mammal occurrences from Borneo.
+A) bias weights ($w$) defining the effects of each bias factor, B) sampling rate as function of distance to the closest instance of each bias factor (the expected number of occurrences) given the inferred \\textit{sampbias} model.
 At the study scale of 0.05 degrees (c.
-5km) \\textit{sampbias} finds the strongest biasing effect for the proximity of cities and roads.", eval = TRUE}
+5x5km) \\textit{sampbias} finds the strongest biasing effect for the proximity of cities and roads.", eval = TRUE}
 knitr::include_graphics(path = "ms_figures/figure_empirical_results.jpg")
@@
 GBIF.org, 2016).
-The colours show the projection of the \\textcolor{blue}{the relative deviation of sampling rate from the maximum sampling rate given the inferred} \\textit{sampbias} model.
-For instance, a value of -25 indicates a drop of 25\\% compared to the highest sampling rate.
-The highest undersampling is in the centre of the island.", eval = TRUE}
+The colours show the projection of the The colours show the projection of the \\textcolor{blue}{log10-transformed} sampling rates (i.e.
+expected number of occurrences per cell) given the inferred \\textit{sampbias} model.
+The highest undersampling is in the centre of the island.
+\\textcolor{blue}{DIfferent visualizations, including among others the untransformed sampling rate are also implemented in sampbias.", eval = TRUE}
 knitr::include_graphics(path = "ms_figures/figure_empirical_results_spatial_projection_adapted.jpg")

### 2020-07-31 04_appendix_S1.Rmd
manuscript updates and script re-organization

--- 2020-07-30 appendix_S1.Rmd
@@
 ## get a polygon of Borneo
-data(landmass)
-
-born2 <- st_read("empirical_analyses/Borneo.kml", quiet = TRUE)
-born2 <- sf:::st_zm(born2$geom)
-born2 <- as(born2, 'Spatial')
+data(borneo)
 
@@
 # prepare for plotting
-airp <- airports %>% crop(SpatialPoints(occ[,2:1])) %>% raster::intersect(born2) %>%  coordinates() %>%  data.frame() %>% mutate(type = "Airport")
-cit <- cities %>% crop(SpatialPoints(occ[,2:1])) %>% raster::intersect(born2) %>%  coordinates() %>%  data.frame() %>%  mutate(type = "City")
+airp <- airports %>% crop(SpatialPoints(occ[,2:1])) %>% raster::intersect(borneo) %>%  coordinates() %>%  data.frame() %>% mutate(type = "Airport")
+cit <- cities %>% crop(SpatialPoints(occ[,2:1])) %>% raster::intersect(borneo) %>%  coordinates() %>%  data.frame() %>%  mutate(type = "City")
 pts <- bind_rows(occ, cit, airp)
@@
 riv <- waterbodies %>% crop(SpatialPoints(occ[,2:1])) %>%  fortify() %>%  mutate(type = "River")
-road <- roads %>% crop(SpatialPoints(occ[,2:1])) %>% raster::intersect(born2) %>%  fortify() %>%  mutate(type = "Road")
+road <- roads %>% crop(SpatialPoints(occ[,2:1])) %>% raster::intersect(borneo) %>%  fortify() %>%  mutate(type = "Road")
 lin <- bind_rows(riv, road)
 
-lma <- landmass %>% crop(SpatialPoints(occ[,2:1])) %>% raster::intersect(born2) %>%  fortify()
+lma <- borneo %>%  fortify()
 
@@
 occ <-read.csv(system.file("extdata", "mammals_borneo.csv", package="sampbias"), sep = "\t")
+area <- round(areaPolygon(borneo) / 1000000, 0)
+```
 
-data(landmass)
+```{r, echo=FALSE}
 
-born <- raster::intersect(landmass, born2)
+inp <- paste("empirical_analyses/simulations/", list.files("empirical_analyses/simulations/", pattern = "weightestimate"), sep = "")
+inp <- lapply(inp, "read_csv")
 
-area <- round(areaPolygon(born) / 1000000, 0)
-
-dat <- read_csv("empirical_analyses/simulations/weight_estimates.csv") %>% 
+dat <- bind_rows(inp) %>% 
   mutate(records_area = round(rar / area * 10000, 2)) %>% 
-  select(contains("w"), ID, type, records_area) %>% 
+  dplyr::select(contains("w"), ID, type, records_area) %>%
   pivot_longer(cols = contains("w_"), names_to = "bias",
                  values_to = "posterior_estimate") %>%
-    mutate(bias = gsub("w_", "", .data$bias))
+    mutate(bias = gsub("w_", "", .data$bias)) %>% 
+    mutate(bias = fct_reorder(.data$bias, .data$posterior_estimate, .fun = median, .desc = FALSE))
+
+```
+
+```{r, echo = FALSE, fig.height = 7, fig.cap="The bias weights ($w$) of the empirical analyses of records from mammals from Borneo (6262 records, www.gbif.org, 2016, https://doi.org/10.15468/dl.7fg4zx) in comparison to ten simulated datasets of the same size with randomly picked locations across Borneo (no bias)."}
+
+emp <- dat %>% 
+  filter(records_area == 85) %>% 
+  mutate(ID = "Empirical") %>% 
+  mutate(type = "Empirical") %>%
+  dplyr::select(-records_area)
+
+sim <- paste("empirical_analyses/simulations/", list.files("empirical_analyses/simulations/", pattern = "biasweights"), sep = "")
+sim <- lapply(sim, "read_csv")
+
+plo <- bind_rows(sim) %>%
+  mutate(ID = paste("Replicate_", as.character(ID), sep = "")) %>% 
+  mutate(type = "Simulated (no bias)") %>% 
+  dplyr::select(contains("w"), ID, type) %>%
+  pivot_longer(cols = contains("w_"), names_to = "bias",
+                 values_to = "posterior_estimate") %>%
+    mutate(bias = gsub("w_", "", .data$bias)) %>% 
+  bind_rows(emp)
+
+ggplot(data = plo)+
+  geom_boxplot(aes(x = ID, 
+                   y = posterior_estimate,
+                 fill = as.factor(type),
+                 color = as.factor(type)))+
+  ylab("Posterior weight")+
+  scale_fill_viridis(discrete = TRUE)+
+  scale_color_viridis(discrete = TRUE)+
+  theme_bw()+
+  theme(legend.title = element_blank(),
+        axis.title.x = element_blank(),
+        legend.position = "bottom")+
+  facet_wrap(.~bias, ncol = 1)
 ```
@@
 Datasets generated by down-sampling the empirical example dataset of mammal occurrences on Borneo (www.gbif.org,  2016, https://doi.org/10.15468/dl.7fg4zx)."}
-plo <- dat %>%  filter(type == "empirical") %>%
-    mutate(bias = fct_reorder(.data$bias, .data$posterior_estimate, .fun = median, .desc = FALSE))
-# ggplot(data = plo)+
-#   geom_point(aes(x = bias, 
-#                  y = posterior_estimate, 
-#                  group = as.factor(records_area), 
-#                  color = as.factor(ID)), 
-#              position = position_jitterdodge())+
-#   xlab("Posterior weight")+
-#   xlab("Bias factor")+
-#   scale_color_viridis(discrete = TRUE)+
-#   theme_bw()
-
-ggplot(data = plo)+
+ggplot(data = dat)+
   geom_boxplot(aes(x = bias, 
@@
 ```
-
-```{r, echo = FALSE, fig.cap="The bias weights ($w$) defining the effects of each bias factor estimated from data sets with differing density of occurrence records across the study area.
-Datasets generated by random simulation of records across Borneo (no bias)."}
-plo <- dat %>%  filter(type == "simulated") %>%
-    mutate(bias = fct_reorder(.data$bias, .data$posterior_estimate, .fun = median, .desc = FALSE))
-
-ggplot(data = plo)+
-  geom_boxplot(aes(x = bias, 
-                   y = posterior_estimate,
-                   fill = as.factor(records_area),
-                   color = as.factor(records_area)))+
-  ylab("Posterior weight")+
-  xlab("Bias factor")+
-  scale_fill_viridis(discrete = TRUE, name = "Records per\n10,000 sqkm")+
-  scale_color_viridis(discrete = TRUE, name = "Records per\n10,000 sqkm")+
-  theme_bw()
-```
-
-```{r, echo = FALSE, fig.height = 8, fig.cap="The bias weights ($w$) defining the effects of each bias factor estimated for the smallest simulated dataset without bias."}
-
-ggplot(data = plo)+
-  geom_histogram(aes(posterior_estimate, fill = as.factor(records_area)), bins = 50) +
-  scale_fill_viridis(discrete = TRUE, name = "Records per\n10,000 sqkm")+
-  facet_grid(rows = vars(records_area), cols = vars(bias))+
-  scale_x_continuous(breaks = c(0,0.003, 0.006, 0.009), limits = c(0,0.01))+
-  xlab("Posterior weight")+
-  ylab("Count")+
-  theme_bw()+
-  theme(panel.grid.minor = element_blank(),
-        legend.position = "none")
-
-```
-
-```{r, out.width = "6in", fig.cap = "Projected bias surface for randomly sampled points across Borneo, from a random replicate."}
-
-knitr::include_graphics(path = "ms_figures/figure_projection_empirical_simulation.jpg")
-
-```

### 2020-07-31 07_empirical_analysis_manuscript.R

@@ -24,12 +24,9 @@ occ<- read.csv(system.file("extdata",
                                    package="sampbias"),
                        sep = "\t")
 
-
-
-out <- calculate_bias(occ, res = 0.05, buffer = 0.5, restrict_sample = borneo)
-
-out <- calculate_bias(occ, res = 1, buffer = 0.5, restrict_sample = borneo, use_hyperprior=TRUE)
-out <- calculate_bias(occ, res = 1, buffer = 0.5, restrict_sample = borneo, use_hyperprior=FALSE)
+out <- calculate_bias(occ, res = 0.05,
+                      buffer = 0.5,
+                      restrict_sample = borneo)
 
 p1 <- plot(out)
 summary(out)
@@ -37,15 +34,19 @@ summary(out)
 ggsave(p1, filename = "ms_figures/figure_empirical_results.jpg", height = 8, width = 8)
 
 proj <- project_bias(out)
-p2 <- map_bias(proj)
+p2 <- map_bias(proj, type = "diff_to_max")
 
 
-ggsave(p2, filename = "ms_figures/figure_empirical_results_spatial_projection.pdf", height = 16, width = 16)
+ggsave(p2, filename = "ms_figures/figure_empirical_projection_difftomax.pdf", height = 16, width = 16)
 
-p3 <- map_bias(proj, sampling_rate = TRUE)
+p3 <- map_bias(proj, type= "sampling_rate")
 
-ggsave(p3, filename = "ms_figures/figure_empirical_results_spatial_projection_sampling_rate.pdf", height = 16, width = 16)
+ggsave(p3, filename = "ms_figures/figure_empirical_projection_samplingrate.pdf", height = 16, width = 16)
 
+
+p4 <- map_bias(proj, type= "log_sampling_rate")
+
+ggsave(p4, filename = "ms_figures/figure_empirical_projection_logsamplingrate.pdf", height = 16, width = 16)
 
 
 

### 2020-07-31 R/map_bias.R

@@ -11,7 +11,7 @@
 #' used.
 #' @param sealine logical. Should the coastline be added to the plots? Default is
 #' to TRUE.
-#' @param type character vector. ONe of c("sampling_rate", "log_sampling_rate", "diff_to_max"). If "sampling_rate".
+#' @param type character vector. One of c("sampling_rate", "log_sampling_rate", "diff_to_max"). If "sampling_rate".
 #' the plot shows the raw projected sampling rate depending on the
 #' biasing factors, if "log_sampling_rate", the plot shows the log10 transformed sampling rate, and if
 #' "diff_to_max", the relative deviation of sampling rate from the maximum
@@ -117,7 +117,7 @@ map_bias <- function(x,
       scale_fill_viridis(na.value = "transparent",
                          option = "viridis",
                          direction = 1,
-                         name = "Log Sampling rate",
+                         name = "Sampling rate",
                          discrete = FALSE)+
       theme(axis.title = element_blank())+
       facet_wrap(split~ .)

### 2020-07-31 further_tests.R

@@ -1,74 +1,0 @@
-# additional tests
-
-
-# a dataset of occurrence from plants of Burkina Faso
-
-library(sampbias)
-library(cowplot)
-library(magrittr)
-library(dplyr)
-library(raster)
-library(ggplot2)
-library(magrittr)
-library(sp)
-library(dplyr)
-library(viridis)
-
-occ <-read.csv("original_data/bfa_0026301-191105090559680.csv", sep = "\t")
-
-occ <- occ %>%
-  filter(!is.na(decimalLongitude)) %>%
-  filter(decimalLongitude > -8 & decimalLongitude < 5) %>%
-  filter(decimalLatitude > 9)
-
-
-out <- calculate_bias(occ, res = 0.1, buffer = 1)
-
-p1 <- plot(out)
-summary(out)
-
-proj <- project_bias(out)
-p2 <- map_bias(proj)
-
-
-plot_out <- plot_grid(p1, p2, labels = c('A', 'B'), ncol = 1, rel_heights = c(1, 2))
-
-ggsave(plot_out, filename = "ms_figures/figure_empirical_results.jpg", height = 10, width = 8)
-
-
-
-data(airports)
-data(waterbodies)
-data(cities)
-data(roads)
-data(landmass)
-
-# prepare for plotting
-airp <- airports %>% crop(SpatialPoints(occ[,23:22])) %>%  coordinates() %>%  data.frame() %>% mutate(type = "Airport")
-cit <- cities %>% crop(SpatialPoints(occ[,23:22])) %>%  coordinates() %>%  data.frame() %>%  mutate(type = "City")
-pts <- bind_rows(cit, airp)
-
-
-riv <- waterbodies %>% crop(SpatialPoints(occ[,23:22])) %>%  fortify() %>%  mutate(type = "River")
-road <- roads %>% crop(SpatialPoints(occ[,23:22])) %>%  fortify() %>%  mutate(type = "Road")
-lin <- bind_rows(riv, road)
-
-
-lma <- landmass %>% crop(SpatialPoints(occ[,23:22])) %>%  fortify()
-
-# plot
-ggplot()+
-  geom_polygon(data = lma,
-               mapping = aes(x = long, y = lat, group = group), lwd = 0.5, col = "grey40", fill = "transparent")+
-  geom_path(mapping = aes(x = long, y = .data$lat, group = .data$group, shape = type, color = type),
-            data = lin)+
[…]

### 2020-07-31 pre_submission_tests.R

@@ -1,6 +1,4 @@
 
-pkgdown::build_site()
-
 library(goodpractice)
 gp("C:/Users/az64mycy/Dropbox (iDiv)/research_projects/16_spatial_bias_package/sampbias")
 

### 2020-09-15 04_appendix_S1.Rmd
tidy up github

@@
 title: "sampbias, a method for quantifying geographic sampling biases in species distribution data"
-subtitle: "Appendix S1 - Supplementary Figure"
-author: "Alexander Zizka, Alexandre Antonelli, Daniele Silvestro"
+subtitle: "Appendix S1 - Supplementary Figures"
+#author: "Alexander Zizka, Alexandre Antonelli, Daniele Silvestro"
 output: rmarkdown::pdf_document
@@
 
-```{r, echo = FALSE, fig.height = 7, fig.cap="The bias weights ($w$) of the empirical analyses of records from mammals from Borneo (6262 records, www.gbif.org, 2016, https://doi.org/10.15468/dl.7fg4zx) in comparison to ten simulated datasets of the same size with randomly picked locations across Borneo (no bias)."}
+```{r, echo = FALSE, fig.height = 7, fig.cap="The bias weights ($w$) of the empirical analyses of records from mammals from Borneo (6262 records, www.gbif.org, 2016, https://doi.org/10.15468/dl.7fg4zx) in comparison to ten simulated data sets of the same size with randomly picked locations across Borneo (no bias)."}
 
@@
 plo <- bind_rows(sim) %>%
-  mutate(ID = paste("Replicate_", as.character(ID), sep = "")) %>% 
+  mutate(ID = paste("Rep_", as.character(ID), sep = "")) %>% 
   mutate(type = "Simulated (no bias)") %>% 
@@
 
-```{r, echo = FALSE, fig.cap="The bias weights ($w$) quantifying the effects of each bias factor estimated from data sets with differing density of occurrence records across the study area.
+```{r, echo = FALSE, fig.cap="The bias weights ($w$) estimated from data sets with differing density of occurrence records across the study area.
 Datasets generated by down-sampling the empirical example dataset of mammal occurrences on Borneo (www.gbif.org,  2016, https://doi.org/10.15468/dl.7fg4zx)."}
@@
 
-```{r, out.width = "6in", fig.cap = "Projected bias surface for the distribution data of mammals from Borneo (www.gbif.org, 2016, https://doi.org/10.15468/dl.7fg4zx), at different levels of downsampling, from a random replicate.
+```{r, out.width = "6in", fig.cap = "Projected bias surface for the distribution records of mammals from Borneo (www.gbif.org, 2016, https://doi.org/10.15468/dl.7fg4zx).
+A) - D) Different levels of random downsampling.
 The plotted extent is adapted to the study area, which may not cover the entire island for rarified datasets with few records (A) and B))"}

### 2020-09-15 10_main_manuscript_clean.tex
tidy up github

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={sampbias, a method for quantifying geographic sampling biases in species distribution data},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{caption}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[L]{Quantifying sampling biases}
\fancyhead[R]{\thepage}
\fancyfoot{}
\usepackage{float}
\floatplacement{figure}{H}
\usepackage{lineno}
\linenumbers
\usepackage[url=false]{biblatex}
\usepackage{setspace}

\title{sampbias, a method for quantifying geographic sampling biases in species distribution data}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle

\setstretch{2}
\newpage{}

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

Geo-referenced species occurrences from public databases have become essential to biodiversity research and conservation. However, geographical biases are widely recognized as a factor limiting the usefulness of such data for understanding species diversity and distribution. In particular, differences in sampling intensity across a landscape due to differences in human accessibility are ubiquitous but may differ in strength among taxonomic groups and data sets. Although several factors have been described to influence human access (such as presence of roads, rivers, airports and cities), quantifying their specific and combined effects on recorded occurrence data remains challenging. Here we present \emph{sampbias}, an algorithm and software for quantifying the effect of accessibility biases in species occurrence data sets. \emph{sampbias} uses a Bayesian approach to estimate how sampling rates vary as a function of proximity to one or multiple bias factors. The results are comparable among bias factors and data sets. We demonstrate the use of \emph{sampbias} on a data set of mammal occurrences from the island of Borneo, showing a high biasing effect of cities and a moderate effect of roads and airports. \emph{sampbias} is implemented as a well-documented, open-access and user-friendly R package that we hope will become a standard tool for anyone working with species occurrences in ecology, evolution, conservation and related fields.

\hypertarget{keywords}{%
\section{Keywords}\label{keywords}}

Collection effort, Global Biodiversity Information Facility (GBIF), Presence only data, Roadside bias, Sampling intensity

\newpage{}

\hypertarget{background}{%
\section{Background}\label{background}}

Publicly available data sets of geo-referenced species occurrences, such as provided by the Global Biodiversity Information Facility (www.gbif.org) have become a fundamental resource in biological sciences, especially in biogeography, conservation, and macroecology. However, these data sets are typically not collected systematically and rarely include information on collection effort. Instead, they are often compiled from a variety of sources (e.g., scientific expeditions, census counts, genetic barcoding studies and citizen-science observations). Species occurrences are therefore often subject to multiple sampling biases (Meyer et al. 2016).

Sampling biases that may affect the recording of species occurrences (presence, absence and abundance, Isaac and Pocock 2015, Boakes et al. 2010) include the under-sampling of specific taxa (``taxonomic bias'', e.g., birds \emph{vs.} nematodes), specific geographic regions (``geographic bias'', e.g., easily accessible \emph{vs.} remote areas), and specific temporal periods (``temporal bias'', e.g., wet \emph{vs.} dry season). In particular geographic sampling bias---the fact that sampling effort is spatially biased, rather than equally distributed over the study area---is likely to be widespread in all non-systematically collected data sets of species distributions.

Many aspects can lead to sampling biases, including socio-economic factors (e.g., national research spending, history of scientific research; Zizka et al. 2020, Meyer et al. 2015, Daru et al. 2018), political factors (e.g., armed conflict, democratic rights; Rydén et al. 2020), and physical accessibility (e.g., distance to a road or river, terrain conditions, slope; Yang et al. 2014, Botts et al. 2011). Especially physical accessibility by people is omnipresent as a bias factor (e.g., Lin et al. 2015, Kadmon et al. 2004, Engemann et al. 2015), across spatial scales, as the commonly used term ``roadside bias'' testifies. In practice, this means that most species observations are made in or near cities, along roads, paths, rivers and near human settlements. Relatively fewer observations are expected to be available from inaccessible areas in e.g., a tropical rainforest or a mountain top. Since the recording of different taxonomic groups poses different challenges, geographic sampling bias and the effect of accessibility may differ among taxonomic groups (Vale and Jenkins 2012).

The implications of not considering geographic sampling biases in biodiversity research are likely substantial (Barbosa et al. 2013, Meyer et al. 2016, Yang et al. 2013). The presence of geographic sampling biases is broadly recognized (e.g., Kadmon et al. 2004), and approaches exist to account for it in some analyses---such as species-richness estimates (Engemann et al. 2015), occupancy models (Kery and Royle 2016), and abundance estimates (Shimadzu and Darnell 2015). In the case of species distribution modelling---the statistical estimation of species geographic distributions based on known occurrences and environmental conditions---geographically biased sampling is problematic because it often causes environmentally biased sampling which decreases model performance (Kadmon et al. 2004, Bystriakova et al. 2012, Lobo and Tognelli 2011, Kramer-Schadt et al. 2013, Varela et al. 2014). Many approaches exist to remedy the effect of biased sampling on species distribution models (Fourcade et al. 2014), including rarefaction to reduce clumped sampling in geographic (Beck et al. 2014, Aiello-Lammens et al. 2015, Boria et al. 2014) or environmental space (Varela et al. 2014), collecting background points for presence-only models to reflect the same sampling bias as the presence records (Phillips et al. 2009), and explicitly modelling sampling bias (Fithian et al. 2015, Stolar and Nielsen 2015, Komori et al. 2020). In contrast, few attempts have been made to compare the geographic sampling bias among data sets (Fernández and Nakamura 2015, Ruete 2015, Monsarrat et al. 2019) and to our knowledge, no tools exist to quantify the effect size of specific bias factors and compare it among them. We define as \textit{bias factors} any anthropogenic or natural features that facilitate human access and sampling, such as roads, rivers, airports, and cities.

It is unrealistic to expect that accessibility bias in biodiversity data will disappear even after more automated observation technologies are developed. It is therefore crucial that researchers realise the intrinsic biases associated with the data they deal with, especially in cross-taxonomic studies, since occurrence data sets from different taxa are likely differently affected by sampling biases due to differences in specimen collection and transportation. This is the first step towards estimating to which extent these biases may affect their analyses, results and conclusions. Any study dealing with species occurrence data should arguably assess the strength of accessibility biases in the underlying data. Such a quantification can also help researchers to target further sampling efforts.

Here, we present \emph{sampbias} v1.0.4, a probabilistic method to quantify accessibility bias in data sets of species occurrences. \emph{sampbias} is implemented as a user-friendly R-package and uses a Bayesian approach to address three questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\item
  How strong is the accessibility bias in a given data set?
\item
  How strong is the effect of different bias factors in causing the overall accessibility bias?
\item
  How is accessibility bias distributed in space?
\end{enumerate}

\emph{sampbias} is implemented in R (R Core Team 2019), based on commonly used packages for data handling (\texttt{ggplot}, Wickham 2009, \texttt{forcats}, 2019, \texttt{tidyr}, Wickham and Henry 2019, \texttt{dplyr}, Wickham et al. 2019, \texttt{magrittr}, Bache and Wickham 2014, \texttt{viridis}, Garnier 2018), handling geographic information and geo-computation (\texttt{raster}, Hijmans 2019, \texttt{sp}, Pebesma and Bivand 2005, Bivand et al. 2013) and statistical modelling (\texttt{stats}, R Core Team 2019). \emph{sampbias} offers an easy and largely automated means for biodiversity scientists and non-specialists alike to explore bias in species occurrence data, in a way that is comparable across data sets. The results may be used to identify priorities for further collection or digitalization efforts and to assess the reliability of scientific results based on publicly available species distribution data.

\hypertarget{methods-and-features}{%
\section{Methods and Features}\label{methods-and-features}}

\hypertarget{general-concept}{%
\subsection{General concept}\label{general-concept}}

Under the assumption that organisms occur across the entire area of interest, we can expect the number of sampled occurrences in a restricted area, such as a single biome, to be distributed uniformly in space (even though, of course, the density of individuals and the species diversity may be heterogeneous). With \emph{sampbias} we assess to which extent variation in sampling rates can be explained by distance from bias factors.

\emph{sampbias} works at a user-defined spatial scale, and any data set of multi-species occurrence records can be tested against any geographic gazetteer. Reliability increases with increasing data set size. Default global gazetteers for airports, cities, rivers and roads are provided with \emph{sampbias}, and user-defined gazetteers can be added easily. Species occurrence data as downloaded from the data portal of GBIF can be directly used as input data for \emph{sampbias}. The output of the package includes measures of the sampling rates across space, which are comparable between different gazetteers (e.g., comparing the biasing effect of roads and rivers), different taxa (e.g., birds \emph{vs.} flowering plants) and different data sets (e.g., specimens \emph{vs.} human observations).

\hypertarget{distance-calculation}{%
\subsection{Distance calculation}\label{distance-calculation}}

\emph{sampbias} uses gazetteers of the geographic location of bias factors (hereafter indicated with B) to generate a regular grid across the study area. By default the study area is defined by the geographic extent of the study data set, but it can also be customized via user-defined polygons, for instance to limit the analyses to an environmentally homogeneous region (e.g., a rainforest) or an area of special interest (e.g., a national park). In this case all occurrence records outside the user-defined area will be disregarded for the analysis.. For each grid cell \(i\), we then compute a vector \(X_i(j)\) of minimum distances (straight aerial distance, ``as the crow flies'') to each bias factor \(j \in B\). The resolution of the grid defines the precision of the distance estimates, for instance a \(1 \times 1\) degree raster will yield approximately a 110 km precision at the equator. Due to the assumption of homogeneous sampling and a computational trade-off between the resolution of the regular grid and the extent of the study area (for instance, a 1 second resolution for a global data set would become computationally prohibitive in most practical cases), \emph{sampbias} is best suited for local or regional data sets at high resolution (c.~100 -- 10,000 m). Since the differences in grid cell size are negligible on the local and regional scale, sampbias uses a latitude/longitude grid by default, but a custom grid in any projection and coordinate reference system---for instance an equal area grid, which is often more suitable for large spatial analyses---may be provided by the user.

\hypertarget{quantifying-accessibility-bias-using-a-bayesian-framework}{%
\subsection{Quantifying accessibility bias using a Bayesian framework}\label{quantifying-accessibility-bias-using-a-bayesian-framework}}

We describe the observed number of sampled occurrences \(S_i\) within each cell \(i\) as the result of a Poisson sampling process with rate \(\lambda_i\). We model the rate \(\lambda_i\) as a function of a parameter \(q\), which represents the expected number of occurrences per cell in the absence of biases, i.e.~when \(\sum_{j=1}^{B}{ X_i(j)} = 0\). Additionally, we model \(\lambda_i\) to decrease exponentially as a function of distance from bias factors, such that increasing distances will result in a lower sampling rate. For a single bias factor the rate of cell \(i\) with distance \(X_i\) from a bias is:
\[
\lambda_i = q \times \exp \left(- w X_i \right)
\]
where \(w \in \mathbb{R^+}\) defines the steepness of the Poisson rate decline, such that \(w \approx 0\) results in a null model of uniform sampling rate \(q\) across cells. In the presence of multiple bias factors (e.g., roads and rivers), the sampling rate decrease is a function of the cumulative effects of each bias and its distance from the cell:
\[
\lambda_i = q \times \exp \left(-\sum_{j=1}^{B}{w_j X_i(j)} \right) \ \ \ \ (1)
\]
where a vector \(\mathbf{w} = [w_1, ..., w_B]\) describes the amount of bias attributed to each specific factor.

To quantify the amount of bias associated with each factor, we jointly estimate the parameters \(q\) and \(\mathbf{w}\) in a Bayesian framework. We use Markov Chain Monte Carlo (MCMC) to sample these parameters from their posterior distribution:
\[
P(q, \mathbf{w} | \mathbf{S}) \propto \prod_{i=1}^{N}{ Poi(S_i | \lambda_i) } \times P(q) P(\mathbf{w}) \ \ \ \ (2)
\]
where the likelihood of sampled occurrences \(S_i\) within each cell \(Poi(S_i | \lambda_i)\) is the probability mass function of a Poisson distribution with rate per cell defined as in Eqn. (1). The likelihood is then multiplied across the \(N\) cells considered. We use exponential priors on the parameters \(q\) and \(\mathbf{w}\), \(P(q) \sim \Gamma(1, 0.01)\) and \(P(\mathbf{w}) \sim \Gamma(1, 1)\), respectively. We chose exponential priors because they represent the standard choice for rate parameters such as \(q\) and the weights \(\mathbf{w}\), all of which must be positive and have support {[}0, +inf{]}. We designed the priors to be informative (i.e.~not allowing negative values) and yet vague enough to encompass a much wider range of parameter space than the range of values observed in empirical tests. Custom priors, within the flexible family of gamma distributions, which include the exponential priors used here, are possible via the \texttt{prior\_q} and \texttt{prior\_w} arguments of the \texttt{calculate\_bias} function. Additionally, since the null expectation in the absence of biases is that the weights are close to 0, we implement a hierarchical model, in which the rate parameter of the gamma prior can be itself estimated from the data. Thus, we set \(P(\mathbf{w}) \sim \Gamma(a=1, b)\) and assign a vague hyper-prior on the rate \(P(b) \sim \Gamma(\alpha_0=1, \beta_0=0.001)\). The choice of a conjugate gamma hyper-prior allows us to sample the rate directly from its posterior distribution:
\[
b \sim \Gamma \left(\alpha_0 + a B,\ \beta_0  \sum_{j=1}^{B}{w_j} \right) \ \ \ \ (3)
\]
The use of a hyper-prior has the advantage of making the prior on the weights more flexible and able to adapt to different data sets reducing the need for user-defined arbitrary choices. Additionally, it works as a regularization technique reducing the risks of over-parametrization, by shrinking the weights around small values when there is no evidence in the data indicating a bias.

We summarize the parameters by computing the mean of the posterior samples and their standard deviation. We interpret the magnitude of the elements in \(\mathbf{w}\) as a function of the importance of the individual biases. We note, however, that this test is not explicitly intended to assess the significance of each bias factor. Because several bias factors might be correlated (e.g.~cities and airports), simply summing their effect from independent analyses would result in an overestimation of the total bias. It is therefore important to jointly estimate the effects of correlated factors, as this is based on the likelihood of the data given the combined effects of all biasing factors. A Bayesian variable selection method could be used to quantify the expected amount of bias in the data predicted by single or a particular combination of predictors, but falls beyond the scope of the current study. In the empirical example below we show, however, how a simple simulation can be used to asses whether the estimated bias factors significantly differ from a null expectation.

We summarize the results by mapping the estimated (the relative deviation of sampling rate from the maximum
sampling rate (or on user choice the estimated sampling rates, \(\lambda_i\)) across space. These rates represent the expected number of sampled occurrences for each grid cell and provide a graphical representation of the spatial variation of sampling rates. Provided that the cells are of equal size, the estimated rates will be comparable across data sets, regions, and taxonomic groups. Analysing different regions, biomes, or taxa in separate analyses allows to account for differences in sampling rates, which are not linked with bias factors. For instance, the unbiased sampling rate \(q\) is expected to differ between a highly sampled clade like birds and under-sampled groups of invertebrates, but their sampling biases (\(\mathbf{w}\)) might be similar across the two groups.

\hypertarget{example-and-empirical-validation}{%
\subsection{Example and empirical validation}\label{example-and-empirical-validation}}

A default \emph{sampbias} analysis can be run with few lines of code in R, based on a \texttt{data.frame} including species identity and geographic coordinates. The main function \texttt{calculate\_bias} creates an object of the class \texttt{"sampbias"}, for which the package provides a plotting and summary method. Additional options exist to provide custom gazetteers, study area, spatial grid and grain size of the analysis, as well as operators for the calculation of the bias distances, including priors for \(q\) and \(\mathbf{w}\). A tutorial on how to use \emph{sampbias} is available with the package and in the electronic supplement of this publication (Appendix S1).

To exemplify the use of \emph{sampbias}, we downloaded the occurrence records of all mammals available from the island of Borneo (n = 6,262, GBIF.org 2016) and ran \emph{sampbias} using the default gazetteers as shown in the example code below, to test the biasing effect of the main airports, cities and roads in the data set. The example data set is provided with \emph{sampbias}.

We found a strong effect of cities on sampling intensity, a moderate effect of roads and airports and a negligible effect of rivers (Fig. \ref{fig:empirical}). All models predict a low number of collection records in the centre of Borneo (Fig. \ref{fig:spatial}), which reflects the original data, and where accessibility means are low (Figure S1 in Appendix S2). The empirical example illustrates the use of \emph{sampbias}, for detailed analyses or a smaller geographic scale, higher resolution gazetteers, including smaller roads and rivers and a higher spatial resolution would be desirable. Results might change with increasing resolution, since roads and rivers might have a stronger effect on higher resolutions (facilitating most the access to their immediate vicinity), whereas cities and airports might have a stronger effect on the larger scale (facilitating access to a larger area).

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(sampbias)}

\CommentTok{# a data table with species identity, longitude, and latitude}
\NormalTok{example.in <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\KeywordTok{system.file}\NormalTok{(}\StringTok{"extdata"}\NormalTok{, }
                                   \StringTok{"mammals_borneo.csv"}\NormalTok{,}
                                   \DataTypeTok{package=}\StringTok{"sampbias"}\NormalTok{), }
                       \DataTypeTok{sep =} \StringTok{"}\CharTok{\textbackslash{}t}\StringTok{"}\NormalTok{)}

\CommentTok{# laod the outline of Borneo (provided as example with sampbias)}
\KeywordTok{data}\NormalTok{(borneo)}

\CommentTok{# running sampbias}
\CommentTok{## 'res' defines the resolution of the spatial grid}
\CommentTok{## for distance calculation in degrees latitude and longitude}
\CommentTok{## 'buffer' defines the buffer around the study area to account for biasing}
\CommentTok{## structure adjacent to the study area, in degrees latitude and longitude}
\CommentTok{## 'restrict_sample' restricts the analysis to Borneo, }
\CommentTok{## as example for a user-defined area }
\CommentTok{## All other options at default, see ?calculate_bias for a description}
\NormalTok{example.out <-}\StringTok{ }\KeywordTok{calculate_bias}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ example.in, }
                              \DataTypeTok{res =} \FloatTok{0.05}\NormalTok{,}
                              \DataTypeTok{buffer =} \FloatTok{0.5}\NormalTok{,}
                              \DataTypeTok{restrict_sample =}\NormalTok{ borneo)}

\CommentTok{# summary}
\KeywordTok{summary}\NormalTok{(example.out)}
\KeywordTok{plot}\NormalTok{(example.out)}

\CommentTok{# projecting the bias effect in space}
\NormalTok{proj <-}\StringTok{ }\KeywordTok{project_bias}\NormalTok{(example.out)}
\KeywordTok{map_bias}\NormalTok{(proj)}
\end{Highlighting}
\end{Shaded}

We ran a simulation experiment to assess whether the estimated bias weights differ significantly from a null expectation of random sampling. To do so, we first simulated ten unbiased data sets by generating 6262 random occurrences across Borneo (the same number as in the empirical data set) and then ran \emph{sampbias} with the same settings as for the empirical data set on each of these replicates. We found that the estimated bias parameters were significantly different (credible intervals non-overlapping) from the null model for cities, roads and airports (Figure S2 in Appendix S2).

\emph{sampbias} is designed to work with sparsely sampled data sets. To test the performance of \emph{sampbias} on small data sets we ran an additional simulation experiment. We ran \emph{sampbias} with the same settings as for the empirical data set on three rarefied data sets, sub-sampling the initial data set to 3,131, 626, 62 records respectively (three replicates each, a total of nine analyses), and then compared the estimated bias weights for all bias factors to the estimates for the empirical data set. The results showed that parameter estimates and the projection of the bias effect in space were robust to the decreasing amount of data, although uncertainty increased as reflected larger estimated credible intervals (Figs. S3 and S4, in Appendix S2).

\hypertarget{data-accessibility}{%
\section{Data accessibility}\label{data-accessibility}}

\emph{sampbias} is available under a GNU General Public license v3 from \url{https://github.com/azizka/sampbias}, and includes the example data set as well as a tutorial (Appendix S1) and a summary of possible warnings produced by the package (Appendix S3).

\hypertarget{figures}{%
\section{Figures}\label{figures}}

\begin{figure}
\includegraphics[width=6in]{ms_figures/figure_empirical_results} \caption{Results of the empirical validation analysis, estimating the accessibility bias in mammal occurrences from Borneo. A) bias weights ($w$) defining the effects of each bias factor, B) sampling rate as function of distance to the closest instance of each bias factor (the expected number of occurrences) given the inferred \textit{sampbias} model. At the study scale of 0.05 degrees (c. 5x5km) \textit{sampbias} finds the strongest biasing effect for the proximity of cities and roads.}\label{fig:empirical}
\end{figure}

\begin{figure}
\includegraphics[width=6in]{ms_figures/figure_empirical_results_spatial_projection_adapted} \caption{Spatial projection of the sampling bias in an empirical example data set of mammal occurrences on the island of Borneo (downloaded from www.gbif.org. GBIF.org, 2016). The colours show the projection of the log10-transformed sampling rates (i.e. expected number of occurrences per cell) given the inferred \textit{sampbias} model. The highest undersampling is in the centre of the island. Different visualizations, including among others the untransformed sampling rate are also implemented in sampbias.}\label{fig:spatial}
\end{figure}

\newpage{}

\hypertarget{supplementary-material}{%
\section{Supplementary material}\label{supplementary-material}}

Appendix S1 - Tutorial running \emph{sampbias} in R

Appendix S2 - Supplementary Figures

Appendix S3 - Possible warnings and their solutions

\newpage{}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Aiello-Lammens2015}{}%
Aiello-Lammens, M. E. et al. 2015. spThin: An R package for spatial thinning of species occurrence records for use in ecological niche models. - Ecography 38: 541--545.

\leavevmode\hypertarget{ref-Bache2014}{}%
Bache, S. M. and Wickham, H. 2014. magrittr: A Forward-Pipe Operator for R.

\leavevmode\hypertarget{ref-Barbosa2013}{}%
Barbosa, A. M. et al. 2013. Species-people correlations and the need to account for survey effort in biodiversity analyses. - Diversity and Distributions 19: 1188--1197.

\leavevmode\hypertarget{ref-Beck2014}{}%
Beck, J. et al. 2014. Spatial bias in the GBIF database and its effect on modeling species' geographic distributions. - Ecological Informatics 19: 10--15.

\leavevmode\hypertarget{ref-Bivand2013}{}%
Bivand, R. S. et al. 2013. Applied spatial data analysis with R, Second edition. - Springer.

\leavevmode\hypertarget{ref-Boakes2010}{}%
Boakes, E. H. et al. 2010. Distorted views of biodiversity: Spatial and temporal bias in species occurrence data. - PLoS Biology 8: e1000385.

\leavevmode\hypertarget{ref-Boria2014}{}%
Boria, R. A. et al. 2014. Spatial filtering to reduce sampling bias can improve the performance of ecological niche models. - Ecological Modelling 275: 73--77.

\leavevmode\hypertarget{ref-Botts2011}{}%
Botts, E. A. et al. 2011. Geographic sampling bias in the South African Frog Atlas Project: Implications for conservation planning. - Biodiversity and Conservation 20: 119--139.

\leavevmode\hypertarget{ref-Bystriakova2012}{}%
Bystriakova, N. et al. 2012. Sampling bias in geographic and environmental space and its effect on the predictive power of species distribution models. - Systematics and Biodiversity 10: 305--315.

\leavevmode\hypertarget{ref-Daru2018}{}%
Daru, B. H. et al. 2018. Widespread sampling biases in herbaria revealed from large-scale digitization. - New Phytologist 217: 939--955.

\leavevmode\hypertarget{ref-Engemann2015}{}%
Engemann, K. et al. 2015. Limited sampling hampers ``big data'' estimation of species richness in a tropical biodiversity hotspot. - Ecology and Evolution 5: 807--820.

\leavevmode\hypertarget{ref-Fernandez2015}{}%
Fernández, D. and Nakamura, M. 2015. Estimation of spatial sampling effort based on presence-only data and accessibility. - Ecological Modelling 299: 147--155.

\leavevmode\hypertarget{ref-Fithian2014}{}%
Fithian, W. et al. 2015. Bias correction in species distribution models: pooling survey and collection data for multiple species. - Methods in Ecology and Evolution 6: 424--438.

\leavevmode\hypertarget{ref-Fourcade2014}{}%
Fourcade, Y. et al. 2014. Mapping species distributions with MAXENT using a geographically biased sample of presence data: A performance assessment of methods for correcting sampling bias. - PLoS ONE 9: e97122.

\leavevmode\hypertarget{ref-Garnier2018}{}%
Garnier, S. 2018. viridis: Default color maps from 'matplotlib'.

\leavevmode\hypertarget{ref-gbifdoi2016}{}%
GBIF.org 2016. (08 September 2016) GBIF occurrence download, doi.org/10.15468/dl.7fg4zx.

\leavevmode\hypertarget{ref-Hijmans2019}{}%
Hijmans, R. J. 2019. geosphere: Spherical Trigonometry.

\leavevmode\hypertarget{ref-Isaac2015}{}%
Isaac, N. J. B. and Pocock, M. J. O. 2015. Bias and information in biological records. - Biological Journal of the Linnean Society 115: 522--531.

\leavevmode\hypertarget{ref-Kadmon2004}{}%
Kadmon, R. et al. 2004. Effect of roadside bias on the accuracy of predictive maps produced by bioclimatic models. - Ecological Applications 14: 401--413.

\leavevmode\hypertarget{ref-Kery2016}{}%
Kery, M. and Royle, J. A. 2016. Applied hierarchical modeling in ecology - Analysis of distribution, abundance and species richness in R and BUGS: Volume 1: Prelude and Static Models. - Academic Press, Elsevier.

\leavevmode\hypertarget{ref-Komori2020}{}%
Komori, O. et al. 2020. Sampling bias correction in species distribution models by quasi-linear Poisson point process. - Ecological Informatics 55: 101015.

\leavevmode\hypertarget{ref-Kramer-Schadt2013}{}%
Kramer-Schadt, S. et al. 2013. The importance of correcting for sampling bias in MaxEnt species distribution models. - Diversity and Distributions 19: 1366--1379.

\leavevmode\hypertarget{ref-Lin2015}{}%
Lin, Y.-p. et al. 2015. Uncertainty analysis of crowd-sourced and professionally collected field data used in species distribution models of Taiwanese moths. - Biological Conservation 181: 102--110.

\leavevmode\hypertarget{ref-Lobo2011}{}%
Lobo, J. M. and Tognelli, M. F. 2011. Exploring the effects of quantity and location of pseudo-absences and sampling biases on the performance of distribution models with limited point occurrence data. - Journal for Nature Conservation 19: 1--7.

\leavevmode\hypertarget{ref-Meyer2015}{}%
Meyer, C. et al. 2015. Global priorities for an effective information basis of biodiversity distributions. - Nature Communications 6: 8221.

\leavevmode\hypertarget{ref-Meyer2016}{}%
Meyer, C. et al. 2016. Multidimensional biases, gaps and uncertainties in global plant occurrence information. - Ecology Letters 19: 992--1006.

\leavevmode\hypertarget{ref-Monsarrat2019}{}%
Monsarrat, S. et al. 2019. Accessibility maps as a tool to predict sampling bias in historical biodiversity occurrence records. - Ecography 42: 125--136.

\leavevmode\hypertarget{ref-Pebesma2005}{}%
Pebesma, E. J. and Bivand, R. S. 2005. Classes and methods for spatial Data: the sp Package. - R News 5: 21--41.

\leavevmode\hypertarget{ref-Phillips2009}{}%
Phillips, S. J. et al. 2009. Sample Selection Bias and Presence-Only Distribution Models : Implications for Background and Pseudo-Absence. - Ecological Applications 19: 181--197.

\leavevmode\hypertarget{ref-rcoreteam2019}{}%
R Core Team 2019. R: A language and environment for statistical computing.

\leavevmode\hypertarget{ref-Ruete2015}{}%
Ruete, A. 2015. Displaying bias in sampling effort of data accessed from biodiversity databases using ignorance maps. - Biodiversity Data Journal 3: e5361.

\leavevmode\hypertarget{ref-Ryden2019}{}%
Rydén, O. et al. 2020. Linking democracy and biodiversity conservation: Empirical evidence and research gaps. - Ambio 49: 419--433.

\leavevmode\hypertarget{ref-Shimadzu2015}{}%
Shimadzu, H. and Darnell, R. 2015. Attenuation of species abundance distributions by sampling. - Royal Society Open Science 2: 140219.

\leavevmode\hypertarget{ref-Stolar2015}{}%
Stolar, J. and Nielsen, S. E. 2015. Accounting for spatially biased sampling effort in presence-only species distribution modelling. - Diversity and Distributions 21: 595--608.

\leavevmode\hypertarget{ref-Vale2012}{}%
Vale, M. M. and Jenkins, C. N. 2012. Across-taxa incongruence in patterns of collecting bias. - Journal of Biogeography 39: 1744--1744.

\leavevmode\hypertarget{ref-Varela2014}{}%
Varela, S. et al. 2014. Environmental filters reduce the effects of sampling bias and improve predictions of ecological niche models. - Ecography 37: 1084--1091.

\leavevmode\hypertarget{ref-Wickham2009}{}%
Wickham, H. 2009. ggplot2 - Elegant graphics for data analysis. - Springer.

\leavevmode\hypertarget{ref-Wickham2019b}{}%
Wickham, H. 2019. forcats: Tools for working with categorical variables (Factors).

\leavevmode\hypertarget{ref-Wickham2019a}{}%
Wickham, H. and Henry, L. 2019. tidyr: Tidy messy data.

\leavevmode\hypertarget{ref-Wickham2019}{}%
Wickham, H. et al. 2019. dplyr: A grammar of data manipulation.

\leavevmode\hypertarget{ref-Yang2013}{}%
Yang, W. et al. 2013. Geographical sampling bias in a large distributional database and its effects on species richness-environment models. - Journal of Biogeography 40: 1415--1426.

\leavevmode\hypertarget{ref-Yang2014}{}%
Yang, W. et al. 2014. Environmental and socio-economic factors shaping the geography of floristic collections in China. - Global Ecology and Biogeography 23: 1284--1292.

\leavevmode\hypertarget{ref-Zizka2020c}{}%
Zizka, A. et al. 2020. Exploring the Impact of Political Regimes on Biodiversity. - VDem working papers 98: 1--13.

\end{document}

### 2020-09-15 2nd_resubmission_tex_for_editor/02_title_page.tex
tidy up github

--- 2020-09-15 10_main_manuscript_clean.tex
@@
   pdftitle={sampbias, a method for quantifying geographic sampling biases in species distribution data},
+  pdfauthor={Alexander Zizka1,2, Alexandre Antonelli3,4,5, Daniele Silvestro3,4,6},
   hidelinks,
@@
 \usepackage[margin=1in]{geometry}
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{longtable,booktabs}
@@
 \setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-\usepackage{caption}
 \usepackage{fancyhdr}
 \pagestyle{fancy}
-\fancyhead[L]{Quantifying sampling biases}
-\fancyhead[R]{\thepage}
-\fancyfoot{}
-\usepackage{float}
-\floatplacement{figure}{H}
-\usepackage{lineno}
-\linenumbers
-\usepackage[url=false]{biblatex}
-\usepackage{setspace}
+\fancyhead[L]{sampbias}
+\fancyhead[R]{ }
+\pagenumbering{gobble}
 
 \title{sampbias, a method for quantifying geographic sampling biases in species distribution data}
-\author{}
-\date{\vspace{-2.5em}}
+\author{Alexander Zizka\textsuperscript{1,2}, Alexandre Antonelli\textsuperscript{3,4,5}, Daniele Silvestro\textsuperscript{3,4,6}}
+\date{}
 
@@
 
-\setstretch{2}
-\newpage{}
-
-\hypertarget{abstract}{%
-\section{Abstract}\label{abstract}}
-
-Geo-referenced species occurrences from public databases have become essential to biodiversity research and conservation.
-However, geographical biases are widely recognized as a factor limiting the usefulness of such data for understanding species diversity and distribution.
-In particular, differences in sampling intensity across a landscape due to differences in human accessibility are ubiquitous but may differ in strength among taxonomic groups and data sets.
-Although several factors have been described to influence human access (such as presence of roads, rivers, airports and cities), quantifying their specific and combined effects on recorded occurrence data remains challenging.
-Here we present \emph{sampbias}, an algorithm and software for quantifying the effect of accessibility biases in species occurrence data sets.
-\emph{sampbias} uses a Bayesian approach to estimate how sampling rates vary as a function of proximity to one or multiple bias factors.
-The results are comparable among bias factors and data sets.
-We demonstrate the use of \emph{sampbias} on a data set of mammal occurrences from the island of Borneo, showing a high biasing effect of cities and a moderate effect of roads and airports.
-\emph{sampbias} is implemented as a well-documented, open-access and user-friendly R package that we hope will become a standard tool for anyone working with species occurrences in ecology, evolution, conservation and related fields.
-
-\hypertarget{keywords}{%
-\section{Keywords}\label{keywords}}
-
-Collection effort, Global Biodiversity Information Facility (GBIF), Presence only data, Roadside bias, Sampling intensity
-
-\newpage{}
-
-\hypertarget{background}{%
-\section{Background}\label{background}}
-
-Publicly available data sets of geo-referenced species occurrences, such as provided by the Global Biodiversity Information Facility (www.gbif.org) have become a fundamental resource in biological sciences, especially in biogeography, conservation, and macroecology.
-However, these data sets are typically not collected systematically and rarely include information on collection effort.
-Instead, they are often compiled from a variety of sources (e.g., scientific expeditions, census counts, genetic barcoding studies and citizen-science observations).
-Species occurrences are therefore often subject to multiple sampling biases (Meyer et al.
-2016).
-
-Sampling biases that may affect the recording of species occurrences (presence, absence and abundance, Isaac and Pocock 2015, Boakes et al.
-2010) include the under-sampling of specific taxa (``taxonomic bias'', e.g., birds \emph{vs.} nematodes), specific geographic regions (``geographic bias'', e.g., easily accessible \emph{vs.} remote areas), and specific temporal periods (``temporal bias'', e.g., wet \emph{vs.} dry season).
-In particular geographic sampling bias---the fact that sampling effort is spatially biased, rather than equally distributed over the study area---is likely to be widespread in all non-systematically collected data sets of species distributions.
-
-Many aspects can lead to sampling biases, including socio-economic factors (e.g., national research spending, history of scientific research; Zizka et al.
-2020, Meyer et al.
-2015, Daru et al.
-2018), political factors (e.g., armed conflict, democratic rights; Rydén et al.
-2020), and physical accessibility (e.g., distance to a road or river, terrain conditions, slope; Yang et al.
-2014, Botts et al.
-2011).
-Especially physical accessibility by people is omnipresent as a bias factor (e.g., Lin et al.
-2015, Kadmon et al.
-2004, Engemann et al.
-2015), across spatial scales, as the commonly used term ``roadside bias'' testifies.
-In practice, this means that most species observations are made in or near cities, along roads, paths, rivers and near human settlements.
-Relatively fewer observations are expected to be available from inaccessible areas in e.g., a tropical rainforest or a mountain top.
-Since the recording of different taxonomic groups poses different challenges, geographic sampling bias and the effect of accessibility may differ among taxonomic groups (Vale and Jenkins 2012).
-
-The implications of not considering geographic sampling biases in biodiversity research are likely substantial (Barbosa et al.
-2013, Meyer et al.
-2016, Yang et al.
-2013).
-The presence of geographic sampling biases is broadly recognized (e.g., Kadmon et al.
-2004), and approaches exist to account for it in some analyses---such as species-richness estimates (Engemann et al.
-2015), occupancy models (Kery and Royle 2016), and abundance estimates (Shimadzu and Darnell 2015).
-In the case of species distribution modelling---the statistical estimation of species geographic distributions based on known occurrences and environmental conditions---geographically biased sampling is problematic because it often causes environmentally biased sampling which decreases model performance (Kadmon et al.
-2004, Bystriakova et al.
-2012, Lobo and Tognelli 2011, Kramer-Schadt et al.
-2013, Varela et al.
-2014).
-Many approaches exist to remedy the effect of biased sampling on species distribution models (Fourcade et al.
-2014), including rarefaction to reduce clumped sampling in geographic (Beck et al.
-2014, Aiello-Lammens et al.
-2015, Boria et al.
-2014) or environmental space (Varela et al.
-2014), collecting background points for presence-only models to reflect the same sampling bias as the presence records (Phillips et al.
-2009), and explicitly modelling sampling bias (Fithian et al.
-2015, Stolar and Nielsen 2015, Komori et al.
-2020).
-In contrast, few attempts have been made to compare the geographic sampling bias among data sets (Fernández and Nakamura 2015, Ruete 2015, Monsarrat et al.
-2019) and to our knowledge, no tools exist to quantify the effect size of specific bias factors and compare it among them.
-We define as \textit{bias factors} any anthropogenic or natural features that facilitate human access and sampling, such as roads, rivers, airports, and cities.
-
-It is unrealistic to expect that accessibility bias in biodiversity data will disappear even after more automated observation technologies are developed.
-It is therefore crucial that researchers realise the intrinsic biases associated with the data they deal with, especially in cross-taxonomic studies, since occurrence data sets from different taxa are likely differently affected by sampling biases due to differences in specimen collection and transportation.
-This is the first step towards estimating to which extent these biases may affect their analyses, results and conclusions.
-Any study dealing with species occurrence data should arguably assess the strength of accessibility biases in the underlying data.
-Such a quantification can also help researchers to target further sampling efforts.
-
-Here, we present \emph{sampbias} v1.0.4, a probabilistic method to quantify accessibility bias in data sets of species occurrences.
-\emph{sampbias} is implemented as a user-friendly R-package and uses a Bayesian approach to address three questions:
-
+\setstretch{1.3}
 \begin{enumerate}
-\def\labelenumi{\arabic{enumi})}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
 \item
-  How strong is the accessibility bias in a given data set?
+  German Centre for Integrative Biodiversity Research Halle-Jena-Leipzig (iDiv), University of Leipzig, Deutscher Platz 5e, 04103 Leipzig, Germany
 \item
-  How strong is the effect of different bias factors in causing the overall accessibility bias?
+  Naturalis Biodiversity Center, Leiden University, Leiden, Darwinweg 2, 2333 CR Leiden The Netherlands
 \item
-  How is accessibility bias distributed in space?
+  Gothenburg Global Biodiversity Centre, University of Gothenburg, Box 461, 405 30 Gothenburg, Sweden
+\item
+  Department for Biological and Environmental Sciences, University of Gothenburg, Box 461, 405 30 Gothenburg, Sweden
+\item
+  Royal Botanic Gardens Kew, TW9 3AE, Richmond, Surrey, United Kingdom
+\item
+  Department of Biology, University of Fribourg, Ch. du Musée 10, 1700 Fribourg, Switzerland
 \end{enumerate}
 
-\emph{sampbias} is implemented in R (R Core Team 2019), based on commonly used packages for data handling (\texttt{ggplot}, Wickham 2009, \texttt{forcats}, 2019, \texttt{tidyr}, Wickham and Henry 2019, \texttt{dplyr}, Wickham et al.
-2019, \texttt{magrittr}, Bache and Wickham 2014, \texttt{viridis}, Garnier 2018), handling geographic information and geo-computation (\texttt{raster}, Hijmans 2019, \texttt{sp}, Pebesma and Bivand 2005, Bivand et al.
-2013) and statistical modelling (\texttt{stats}, R Core Team 2019).
-\emph{sampbias} offers an easy and largely automated means for biodiversity scientists and non-specialists alike to explore bias in species occurrence data, in a way that is comparable across data sets.
-The results may be used to identify priorities for further collection or digitalization efforts and to assess the reliability of scientific results based on publicly available species distribution data.
+\textbf{Running head} -- Quantifying sampling biases
 
-\hypertarget{methods-and-features}{%
-\section{Methods and Features}\label{methods-and-features}}
+\textbf{Word count} -- Abstract: 207 words; Main text (excluding abstract \& references): 3223 words, 2 figures
 
-\hypertarget{general-concept}{%
-\subsection{General concept}\label{general-concept}}
+\textbf{Corresponding author} -- \href{mailto:alexander.zizka@idiv.de}{\nolinkurl{alexander.zizka@idiv.de}}
 
-Under the assumption that organisms occur across the entire area of interest, we can expect the number of sampled occurrences in a restricted area, such as a single biome, to be distributed uniformly in space (even though, of course, the density of individuals and the species diversity may be heterogeneous).
-With \emph{sampbias} we assess to which extent variation in sampling rates can be explained by distance from bias factors.
+\hypertarget{acknowledgements}{%
+\section{Acknowledgements}\label{acknowledgements}}
 
-\emph{sampbias} works at a user-defined spatial scale, and any data set of multi-species occurrence records can be tested against any geographic gazetteer.
-Reliability increases with increasing data set size.
-Default global gazetteers for airports, cities, rivers and roads are provided with \emph{sampbias}, and user-defined gazetteers can be added easily.
-Species occurrence data as downloaded from the data portal of GBIF can be directly used as input data for \emph{sampbias}.
-The output of the package includes measures of the sampling rates across space, which are comparable between different gazetteers (e.g., comparing the biasing effect of roads and rivers), different taxa (e.g., birds \emph{vs.} flowering plants) and different data sets (e.g., specimens \emph{vs.} human observations).
+We thank the organizers of the 2016 Ebben Nielsen challenge for inspiring and recognizing this research.
+We thank all data collectors and contributors to GBIF for their effort.
+AZ is thankful for funding by iDiv via the German Research Foundation (DFG FZT 118), specifically through sDiv, the Synthesis Centre of iDiv.
+AA is supported by grants from the Swedish Research Council, the Knut and Alice Wallenberg Foundation, the Swedish Foundation for Strategic Research and the Royal Botanic Gardens, Kew.
+DS received funding from the Swedish Research Council (2015-04748) and from the Swiss National Science Foundation (PCEFP3\_187012).
 
-\hypertarget{distance-calculation}{%
-\subsection{Distance calculation}\label{distance-calculation}}
+\hypertarget{author-contributions}{%
+\section{Author contributions}\label{author-contributions}}
 
-\emph{sampbias} uses gazetteers of the geographic location of bias factors (hereafter indicated with B) to generate a regular grid across the study area.
-By default the study area is defined by the geographic extent of the study data set, but it can also be customized via user-defined polygons, for instance to limit the analyses to an environmentally homogeneous region (e.g., a rainforest) or an area of special interest (e.g., a national park).
-In this case all occurrence records outside the user-defined area will be disregarded for the analysis..
-For each grid cell \(i\), we then compute a vector \(X_i(j)\) of minimum distances (straight aerial distance, ``as the crow flies'') to each bias factor \(j \in B\).
-The resolution of the grid defines the precision of the distance estimates, for instance a \(1 \times 1\) degree raster will yield approximately a 110 km precision at the equator.
-Due to the assumption of homogeneous sampling and a computational trade-off between the resolution of the regular grid and the extent of the study area (for instance, a 1 second resolution for a global data set would become computationally prohibitive in most practical cases), \emph{sampbias} is best suited for local or regional data sets at high resolution (c.~100 -- 10,000 m).
-Since the differences in grid cell size are negligible on the local and regional scale, sampbias uses a latitude/longitude grid by default, but a custom grid in any projection and coordinate reference system---for instance an equal area grid, which is often more suitable for large spatial analyses---may be provided by the user.
-
-\hypertarget{quantifying-accessibility-bias-using-a-bayesian-framework}{%
-\subsection{Quantifying accessibility bias using a Bayesian framework}\label{quantifying-accessibility-bias-using-a-bayesian-framework}}
-
-We describe the observed number of sampled occurrences \(S_i\) within each cell \(i\) as the result of a Poisson sampling process with rate \(\lambda_i\).
-We model the rate \(\lambda_i\) as a function of a parameter \(q\), which represents the expected number of occurrences per cell in the absence of biases, i.e.~when \(\sum_{j=1}^{B}{ X_i(j)} = 0\).
-Additionally, we model \(\lambda_i\) to decrease exponentially as a function of distance from bias factors, such that increasing distances will result in a lower sampling rate.
-For a single bias factor the rate of cell \(i\) with distance \(X_i\) from a bias is:
-\[
-\lambda_i = q \times \exp \left(- w X_i \right)
-\]
-where \(w \in \mathbb{R^+}\) defines the steepness of the Poisson rate decline, such that \(w \approx 0\) results in a null model of uniform sampling rate \(q\) across cells.
-In the presence of multiple bias factors (e.g., roads and rivers), the sampling rate decrease is a function of the cumulative effects of each bias and its distance from the cell:
-\[
-\lambda_i = q \times \exp \left(-\sum_{j=1}^{B}{w_j X_i(j)} \right) \ \ \ \ (1)
-\]
-where a vector \(\mathbf{w} = [w_1, ..., w_B]\) describes the amount of bias attributed to each specific factor.
-
-To quantify the amount of bias associated with each factor, we jointly estimate the parameters \(q\) and \(\mathbf{w}\) in a Bayesian framework.
-We use Markov Chain Monte Carlo (MCMC) to sample these parameters from their posterior distribution:
-\[
-P(q, \mathbf{w} | \mathbf{S}) \propto \prod_{i=1}^{N}{ Poi(S_i | \lambda_i) } \times P(q) P(\mathbf{w}) \ \ \ \ (2)
-\]
-where the likelihood of sampled occurrences \(S_i\) within each cell \(Poi(S_i | \lambda_i)\) is the probability mass function of a Poisson distribution with rate per cell defined as in Eqn.
-(1).
-The likelihood is then multiplied across the \(N\) cells considered.
-We use exponential priors on the parameters \(q\) and \(\mathbf{w}\), \(P(q) \sim \Gamma(1, 0.01)\) and \(P(\mathbf{w}) \sim \Gamma(1, 1)\), respectively.
-We chose exponential priors because they represent the standard choice for rate parameters such as \(q\) and the weights \(\mathbf{w}\), all of which must be positive and have support {[}0, +inf{]}.
-We designed the priors to be informative (i.e.~not allowing negative values) and yet vague enough to encompass a much wider range of parameter space than the range of values observed in empirical tests.
-Custom priors, within the flexible family of gamma distributions, which include the exponential priors used here, are possible via the \texttt{prior\_q} and \texttt{prior\_w} arguments of the \texttt{calculate\_bias} function.
-Additionally, since the null expectation in the absence of biases is that the weights are close to 0, we implement a hierarchical model, in which the rate parameter of the gamma prior can be itself estimated from the data.
-Thus, we set \(P(\mathbf{w}) \sim \Gamma(a=1, b)\) and assign a vague hyper-prior on the rate \(P(b) \sim \Gamma(\alpha_0=1, \beta_0=0.001)\).
-The choice of a conjugate gamma hyper-prior allows us to sample the rate directly from its posterior distribution:
-\[
-b \sim \Gamma \left(\alpha_0 + a B,\ \beta_0  \sum_{j=1}^{B}{w_j} \right) \ \ \ \ (3)
-\]
-The use of a hyper-prior has the advantage of making the prior on the weights more flexible and able to adapt to different data sets reducing the need for user-defined arbitrary choices.
-Additionally, it works as a regularization technique reducing the risks of over-parametrization, by shrinking the weights around small values when there is no evidence in the data indicating a bias.
-
-We summarize the parameters by computing the mean of the posterior samples and their standard deviation.
-We interpret the magnitude of the elements in \(\mathbf{w}\) as a function of the importance of the individual biases.
-We note, however, that this test is not explicitly intended to assess the significance of each bias factor.
-Because several bias factors might be correlated (e.g.~cities and airports), simply summing their effect from independent analyses would result in an overestimation of the total bias.
-It is therefore important to jointly estimate the effects of correlated factors, as this is based on the likelihood of the data given the combined effects of all biasing factors.
-A Bayesian variable selection method could be used to quantify the expected amount of bias in the data predicted by single or a particular combination of predictors, but falls beyond the scope of the current study.
-In the empirical example below we show, however, how a simple simulation can be used to asses whether the estimated bias factors significantly differ from a null expectation.
-
-We summarize the results by mapping the estimated (the relative deviation of sampling rate from the maximum
-sampling rate (or on user choice the estimated sampling rates, \(\lambda_i\)) across space.
-These rates represent the expected number of sampled occurrences for each grid cell and provide a graphical representation of the spatial variation of sampling rates.
-Provided that the cells are of equal size, the estimated rates will be comparable across data sets, regions, and taxonomic groups.
-Analysing different regions, biomes, or taxa in separate analyses allows to account for differences in sampling rates, which are not linked with bias factors.
-For instance, the unbiased sampling rate \(q\) is expected to differ between a highly sampled clade like birds and under-sampled groups of invertebrates, but their sampling biases (\(\mathbf{w}\)) might be similar across the two groups.
-
-\hypertarget{example-and-empirical-validation}{%
-\subsection{Example and empirical validation}\label{example-and-empirical-validation}}
-
-A default \emph{sampbias} analysis can be run with few lines of code in R, based on a \texttt{data.frame} including species identity and geographic coordinates.
-The main function \texttt{calculate\_bias} creates an object of the class \texttt{"sampbias"}, for which the package provides a plotting and summary method.
-Additional options exist to provide custom gazetteers, study area, spatial grid and grain size of the analysis, as well as operators for the calculation of the bias distances, including priors for \(q\) and \(\mathbf{w}\).
-A tutorial on how to use \emph{sampbias} is available with the package and in the electronic supplement of this publication (Appendix S1).
-
-To exemplify the use of \emph{sampbias}, we downloaded the occurrence records of all mammals available from the island of Borneo (n = 6,262, GBIF.org 2016) and ran \emph{sampbias} using the default gazetteers as shown in the example code below, to test the biasing effect of the main airports, cities and roads in the data set.
-The example data set is provided with \emph{sampbias}.
-
-We found a strong effect of cities on sampling intensity, a moderate effect of roads and airports and a negligible effect of rivers (Fig.
-\ref{fig:empirical}).
-All models predict a low number of collection records in the centre of Borneo (Fig.
-\ref{fig:spatial}), which reflects the original data, and where accessibility means are low (Figure S1 in Appendix S2).
-The empirical example illustrates the use of \emph{sampbias}, for detailed analyses or a smaller geographic scale, higher resolution gazetteers, including smaller roads and rivers and a higher spatial resolution would be desirable.
-Results might change with increasing resolution, since roads and rivers might have a stronger effect on higher resolutions (facilitating most the access to their immediate vicinity), whereas cities and airports might have a stronger effect on the larger scale (facilitating access to a larger area).
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(sampbias)}
-
-\CommentTok{# a data table with species identity, longitude, and latitude}
-\NormalTok{example.in <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\KeywordTok{system.file}\NormalTok{(}\StringTok{"extdata"}\NormalTok{, }
-                                   \StringTok{"mammals_borneo.csv"}\NormalTok{,}
-                                   \DataTypeTok{package=}\StringTok{"sampbias"}\NormalTok{), }
-                       \DataTypeTok{sep =} \StringTok{"}\CharTok{\textbackslash{}t}\StringTok{"}\NormalTok{)}
-
-\CommentTok{# laod the outline of Borneo (provided as example with sampbias)}
-\KeywordTok{data}\NormalTok{(borneo)}
-
-\CommentTok{# running sampbias}
-\CommentTok{## 'res' defines the resolution of the spatial grid}
-\CommentTok{## for distance calculation in degrees latitude and longitude}
-\CommentTok{## 'buffer' defines the buffer around the study area to account for biasing}
-\CommentTok{## structure adjacent to the study area, in degrees latitude and longitude}
-\CommentTok{## 'restrict_sample' restricts the analysis to Borneo, }
-\CommentTok{## as example for a user-defined area }
-\CommentTok{## All other options at default, see ?calculate_bias for a description}
-\NormalTok{example.out <-}\StringTok{ }\KeywordTok{calculate_bias}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ example.in, }
-                              \DataTypeTok{res =} \FloatTok{0.05}\NormalTok{,}
-                              \DataTypeTok{buffer =} \FloatTok{0.5}\NormalTok{,}
-                              \DataTypeTok{restrict_sample =}\NormalTok{ borneo)}
-
-\CommentTok{# summary}
-\KeywordTok{summary}\NormalTok{(example.out)}
-\KeywordTok{plot}\NormalTok{(example.out)}
-
-\CommentTok{# projecting the bias effect in space}
-\NormalTok{proj <-}\StringTok{ }\KeywordTok{project_bias}\NormalTok{(example.out)}
-\KeywordTok{map_bias}\NormalTok{(proj)}
-\end{Highlighting}
-\end{Shaded}
-
-We ran a simulation experiment to assess whether the estimated bias weights differ significantly from a null expectation of random sampling.
-To do so, we first simulated ten unbiased data sets by generating 6262 random occurrences across Borneo (the same number as in the empirical data set) and then ran \emph{sampbias} with the same settings as for the empirical data set on each of these replicates.
-We found that the estimated bias parameters were significantly different (credible intervals non-overlapping) from the null model for cities, roads and airports (Figure S2 in Appendix S2).
-
-\emph{sampbias} is designed to work with sparsely sampled data sets.
-To test the performance of \emph{sampbias} on small data sets we ran an additional simulation experiment.
-We ran \emph{sampbias} with the same settings as for the empirical data set on three rarefied data sets, sub-sampling the initial data set to 3,131, 626, 62 records respectively (three replicates each, a total of nine analyses), and then compared the estimated bias weights for all bias factors to the estimates for the empirical data set.
-The results showed that parameter estimates and the projection of the bias effect in space were robust to the decreasing amount of data, although uncertainty increased as reflected larger estimated credible intervals (Figs.
-S3 and S4, in Appendix S2).
-
-\hypertarget{data-accessibility}{%
-\section{Data accessibility}\label{data-accessibility}}
-
-\emph{sampbias} is available under a GNU General Public license v3 from \url{https://github.com/azizka/sampbias}, and includes the example data set as well as a tutorial (Appendix S1) and a summary of possible warnings produced by the package (Appendix S3).
-
-\hypertarget{figures}{%
-\section{Figures}\label{figures}}
-
-\begin{figure}
-\includegraphics[width=6in]{ms_figures/figure_empirical_results} \caption{Results of the empirical validation analysis, estimating the accessibility bias in mammal occurrences from Borneo.
-A) bias weights ($w$) defining the effects of each bias factor, B) sampling rate as function of distance to the closest instance of each bias factor (the expected number of occurrences) given the inferred \textit{sampbias} model.
-At the study scale of 0.05 degrees (c.
-5x5km) \textit{sampbias} finds the strongest biasing effect for the proximity of cities and roads.}\label{fig:empirical}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=6in]{ms_figures/figure_empirical_results_spatial_projection_adapted} \caption{Spatial projection of the sampling bias in an empirical example data set of mammal occurrences on the island of Borneo (downloaded from www.gbif.org.
-GBIF.org, 2016).
-The colours show the projection of the log10-transformed sampling rates (i.e.
-expected number of occurrences per cell) given the inferred \textit{sampbias} model.
-The highest undersampling is in the centre of the island.
-Different visualizations, including among others the untransformed sampling rate are also implemented in sampbias.}\label{fig:spatial}
-\end{figure}
-
-\newpage{}
-
-\hypertarget{supplementary-material}{%
-\section{Supplementary material}\label{supplementary-material}}
-
-Appendix S1 - Tutorial running \emph{sampbias} in R
-
-Appendix S2 - Supplementary Figures
-
-Appendix S3 - Possible warnings and their solutions
-
-\newpage{}
-
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-Aiello-Lammens2015}{}%
-Aiello-Lammens, M. E. et al. 2015. spThin: An R package for spatial thinning of species occurrence records for use in ecological niche models. - Ecography 38: 541--545.
-
-\leavevmode\hypertarget{ref-Bache2014}{}%
-Bache, S. M. and Wickham, H. 2014. magrittr: A Forward-Pipe Operator for R.
-
-\leavevmode\hypertarget{ref-Barbosa2013}{}%
-Barbosa, A. M. et al. 2013. Species-people correlations and the need to account for survey effort in biodiversity analyses. - Diversity and Distributions 19: 1188--1197.
-
-\leavevmode\hypertarget{ref-Beck2014}{}%
-Beck, J. et al. 2014. Spatial bias in the GBIF database and its effect on modeling species' geographic distributions. - Ecological Informatics 19: 10--15.
-
-\leavevmode\hypertarget{ref-Bivand2013}{}%
-Bivand, R. S. et al. 2013. Applied spatial data analysis with R, Second edition. - Springer.
-
-\leavevmode\hypertarget{ref-Boakes2010}{}%
-Boakes, E. H. et al. 2010. Distorted views of biodiversity: Spatial and temporal bias in species occurrence data. - PLoS Biology 8: e1000385.
-
-\leavevmode\hypertarget{ref-Boria2014}{}%
-Boria, R. A. et al. 2014. Spatial filtering to reduce sampling bias can improve the performance of ecological niche models. - Ecological Modelling 275: 73--77.
-
-\leavevmode\hypertarget{ref-Botts2011}{}%
-Botts, E. A. et al. 2011. Geographic sampling bias in the South African Frog Atlas Project: Implications for conservation planning. - Biodiversity and Conservation 20: 119--139.
-
-\leavevmode\hypertarget{ref-Bystriakova2012}{}%
-Bystriakova, N. et al. 2012. Sampling bias in geographic and environmental space and its effect on the predictive power of species distribution models. - Systematics and Biodiversity 10: 305--315.
-
-\leavevmode\hypertarget{ref-Daru2018}{}%
-Daru, B. H. et al. 2018. Widespread sampling biases in herbaria revealed from large-scale digitization. - New Phytologist 217: 939--955.
-
-\leavevmode\hypertarget{ref-Engemann2015}{}%
-Engemann, K. et al. 2015. Limited sampling hampers ``big data'' estimation of species richness in a tropical biodiversity hotspot. - Ecology and Evolution 5: 807--820.
-
-\leavevmode\hypertarget{ref-Fernandez2015}{}%
-Fernández, D. and Nakamura, M. 2015. Estimation of spatial sampling effort based on presence-only data and accessibility. - Ecological Modelling 299: 147--155.
-
-\leavevmode\hypertarget{ref-Fithian2014}{}%
-Fithian, W. et al. 2015. Bias correction in species distribution models: pooling survey and collection data for multiple species. - Methods in Ecology and Evolution 6: 424--438.
-
-\leavevmode\hypertarget{ref-Fourcade2014}{}%
-Fourcade, Y.
-et al.
-2014.
-Mapping species distributions with MAXENT using a geographically biased sample of presence data: A performance assessment of methods for correcting sampling bias.
-- PLoS ONE 9: e97122.
-
-\leavevmode\hypertarget{ref-Garnier2018}{}%
-Garnier, S. 2018. viridis: Default color maps from 'matplotlib'.
-
-\leavevmode\hypertarget{ref-gbifdoi2016}{}%
-GBIF.org 2016. (08 September 2016) GBIF occurrence download, doi.org/10.15468/dl.7fg4zx.
-
-\leavevmode\hypertarget{ref-Hijmans2019}{}%
-Hijmans, R. J. 2019. geosphere: Spherical Trigonometry.
-
-\leavevmode\hypertarget{ref-Isaac2015}{}%
-Isaac, N. J. B. and Pocock, M. J. O. 2015. Bias and information in biological records. - Biological Journal of the Linnean Society 115: 522--531.
-
-\leavevmode\hypertarget{ref-Kadmon2004}{}%
-Kadmon, R. et al. 2004. Effect of roadside bias on the accuracy of predictive maps produced by bioclimatic models. - Ecological Applications 14: 401--413.
-
-\leavevmode\hypertarget{ref-Kery2016}{}%
-Kery, M.
-and Royle, J.
-A.
-2016.
-Applied hierarchical modeling in ecology - Analysis of distribution, abundance and species richness in R and BUGS: Volume 1: Prelude and Static Models.
-- Academic Press, Elsevier.
-
-\leavevmode\hypertarget{ref-Komori2020}{}%
-Komori, O. et al. 2020. Sampling bias correction in species distribution models by quasi-linear Poisson point process. - Ecological Informatics 55: 101015.
-
-\leavevmode\hypertarget{ref-Kramer-Schadt2013}{}%
-Kramer-Schadt, S. et al. 2013. The importance of correcting for sampling bias in MaxEnt species distribution models. - Diversity and Distributions 19: 1366--1379.
-
-\leavevmode\hypertarget{ref-Lin2015}{}%
-Lin, Y.-p. et al. 2015. Uncertainty analysis of crowd-sourced and professionally collected field data used in species distribution models of Taiwanese moths. - Biological Conservation 181: 102--110.
-
-\leavevmode\hypertarget{ref-Lobo2011}{}%
-Lobo, J.
-M.
-and Tognelli, M.
-F.
-2011.
-Exploring the effects of quantity and location of pseudo-absences and sampling biases on the performance of distribution models with limited point occurrence data.
-- Journal for Nature Conservation 19: 1--7.
-
-\leavevmode\hypertarget{ref-Meyer2015}{}%
-Meyer, C. et al. 2015. Global priorities for an effective information basis of biodiversity distributions. - Nature Communications 6: 8221.
-
-\leavevmode\hypertarget{ref-Meyer2016}{}%
-Meyer, C. et al. 2016. Multidimensional biases, gaps and uncertainties in global plant occurrence information. - Ecology Letters 19: 992--1006.
-
-\leavevmode\hypertarget{ref-Monsarrat2019}{}%
-Monsarrat, S. et al. 2019. Accessibility maps as a tool to predict sampling bias in historical biodiversity occurrence records. - Ecography 42: 125--136.
-
-\leavevmode\hypertarget{ref-Pebesma2005}{}%
-Pebesma, E. J. and Bivand, R. S. 2005. Classes and methods for spatial Data: the sp Package. - R News 5: 21--41.
-
-\leavevmode\hypertarget{ref-Phillips2009}{}%
-Phillips, S. J. et al. 2009. Sample Selection Bias and Presence-Only Distribution Models : Implications for Background and Pseudo-Absence. - Ecological Applications 19: 181--197.
-
-\leavevmode\hypertarget{ref-rcoreteam2019}{}%
-R Core Team 2019. R: A language and environment for statistical computing.
-
-\leavevmode\hypertarget{ref-Ruete2015}{}%
-Ruete, A. 2015. Displaying bias in sampling effort of data accessed from biodiversity databases using ignorance maps. - Biodiversity Data Journal 3: e5361.
-
-\leavevmode\hypertarget{ref-Ryden2019}{}%
-Rydén, O. et al. 2020. Linking democracy and biodiversity conservation: Empirical evidence and research gaps. - Ambio 49: 419--433.
-
-\leavevmode\hypertarget{ref-Shimadzu2015}{}%
-Shimadzu, H. and Darnell, R. 2015. Attenuation of species abundance distributions by sampling. - Royal Society Open Science 2: 140219.
-
-\leavevmode\hypertarget{ref-Stolar2015}{}%
-Stolar, J. and Nielsen, S. E. 2015. Accounting for spatially biased sampling effort in presence-only species distribution modelling. - Diversity and Distributions 21: 595--608.
-
-\leavevmode\hypertarget{ref-Vale2012}{}%
-Vale, M. M. and Jenkins, C. N. 2012. Across-taxa incongruence in patterns of collecting bias. - Journal of Biogeography 39: 1744--1744.
-
-\leavevmode\hypertarget{ref-Varela2014}{}%
-Varela, S. et al. 2014. Environmental filters reduce the effects of sampling bias and improve predictions of ecological niche models. - Ecography 37: 1084--1091.
-
-\leavevmode\hypertarget{ref-Wickham2009}{}%
-Wickham, H. 2009. ggplot2 - Elegant graphics for data analysis. - Springer.
-
-\leavevmode\hypertarget{ref-Wickham2019b}{}%
-Wickham, H. 2019. forcats: Tools for working with categorical variables (Factors).
-
-\leavevmode\hypertarget{ref-Wickham2019a}{}%
-Wickham, H. and Henry, L. 2019. tidyr: Tidy messy data.
-
-\leavevmode\hypertarget{ref-Wickham2019}{}%
-Wickham, H. et al. 2019. dplyr: A grammar of data manipulation.
-
-\leavevmode\hypertarget{ref-Yang2013}{}%
-Yang, W. et al. 2013. Geographical sampling bias in a large distributional database and its effects on species richness-environment models. - Journal of Biogeography 40: 1415--1426.
-
-\leavevmode\hypertarget{ref-Yang2014}{}%
-Yang, W. et al. 2014. Environmental and socio-economic factors shaping the geography of floristic collections in China. - Global Ecology and Biogeography 23: 1284--1292.
-
-\leavevmode\hypertarget{ref-Zizka2020c}{}%
-Zizka, A. et al. 2020. Exploring the Impact of Political Regimes on Biodiversity. - VDem working papers 98: 1--13.
+All authors conceived this study, AZ and DS developed the statistical algorithm and wrote the R-package, AZ and DS wrote the manuscript with contributions from AA.
 

### 2020-09-15 R/map_bias.R

@@ -117,7 +117,7 @@ map_bias <- function(x,
       scale_fill_viridis(na.value = "transparent",
                          option = "viridis",
                          direction = 1,
-                         name = "Sampling rate",
+                         name = "Estimated\nsampling rate",
                          discrete = FALSE)+
       theme(axis.title = element_blank())+
       facet_wrap(split~ .)
@@ -135,7 +135,7 @@ map_bias <- function(x,
       scale_fill_viridis(na.value = "transparent",
                          option = "viridis",
                          direction = 1,
-                         name = "Log Sampling rate",
+                         name = "Estimated\nsampling\nrate [log10]",
                          discrete = FALSE)+
       theme(axis.title = element_blank())+
       facet_wrap(split~ .)

### 2020-09-15 vignettes/Using_the_samp_bias_rpackage.R

@@ -34,17 +34,17 @@ gazetteers <- list(cities = cit,
 ## -----------------------------------------------------------------------------
 bias.out <- calculate_bias(x = occ, gaz = gazetteers)
 
-## ---- eval = TRUE, fig.cap = "Visulaization of the output of a sampbias analysis. A) Posterior estimates of the bias weights for cities and roads, B) the decay of the sampling rate with increasing distaince from cities and roads."----
+## ---- eval = TRUE, fig.height=7, fig.cap = "Output of a sampbias analysis. A) Posterior estimates of the bias weights for cities and roads, B) the decay of the sampling rate with increasing distance from cities and roads."----
 summary(bias.out)
 plot(bias.out)
 
 ## ---- fig.cap = "Sampling bias effect projected in space. Note the coarse grid resolution which should be increased."----
 proj <- project_bias(bias.out)
-map_bias(proj)
+map_bias(proj, type = "log_sampling_rate")
 
-## ---- fig.cap="Example of a customized study area which can be provided to the claculate_bias function via the restrict_sample option. Only grid cells within the red area are included in the sampling bias calculation."----
+## ---- fig.cap="Example of a customized study area which can be provided to the calculate_bias function via the restrict_sample option. Only grid cells within the red areas are included in the sampling bias calculation."----
 data(area_example)
-borneo <- crop(sampbias::landmass, extent(108, 120, -5, 7))
+data(borneo)
 
 plot(borneo)
 plot(area_example, col = "red", add = TRUE)
@@ -53,7 +53,7 @@ bias.out <- calculate_bias(x = occ,
                            gaz = gazetteers, 
                            restrict_sample = area_example)
 
-## ---- fig.cap="Example of a more complex customized study area representing the Borneo montane rain forests ecoregion from Olson et al. (2001). Only grid cells within red area are included in the sampling bias calculation."----
+## ---- fig.cap="Example of a more complex customized study area representing the Borneo montane rain forests ecoregion from Olson et al. (2001). Only grid cells within the red area are included in the sampling bias calculation."----
 data(ecoregion_example)
 
 plot(borneo)

### 2020-09-15 02_title_page.Rmd
manuscript updates and script re-organization
more tidying

--- 2020-07-30 title_page.Rmd
@@
     number_sections: false
+    keep_tex: true
 header-includes: |
@@
 
-**Running head** -- sampbias accessibility bias
+**Running head** -- Quantifying sampling biases
 
-**Word count** -- Abstract: 207 words; Main text (excluding references): 3019 words, 2 figures
+**Word count** -- Abstract: 207 words; Main text (excluding abstract & references): 3223 words, 2 figures
 
@@
 AA is supported by grants from the Swedish Research Council, the Knut and Alice Wallenberg Foundation, the Swedish Foundation for Strategic Research and the Royal Botanic Gardens, Kew.
-DS received funding from the Swedish Research Council (2015-04748) and from the Swiss National Science Foundation (PCEFP3_187012)
+DS received funding from the Swedish Research Council (2015-04748) and from the Swiss National Science Foundation (PCEFP3_187012).
 

### 2020-09-15 10_main_manuscript_clean.rmd
more tidying

--- 2020-07-31 03_main_manuscript.rmd
@@
 title: "sampbias, a method for quantifying geographic sampling biases in species distribution data"
-author: "Alexander Zizka^1,2^, Alexandre Antonelli^3,4,5^, Daniele Silvestro^3,4,6^"
 bibliography: sampbias.bib
@@
     \pagestyle{fancy}
-    \fancyhead[L]{sampbias accessibility bias}
+    \fancyhead[L]{Quantifying sampling biases}
     \fancyhead[R]{\thepage}
@@
 ---
-
-1. German Centre for Integrative Biodiversity Research Halle-Jena-Leipzig (iDiv), University of Leipzig, Deutscher Platz 5e, 04103 Leipzig, Germany
-2. Naturalis Biodiversity Center, Leiden University, Leiden, Darwinweg 2, 2333 CR Leiden The Netherlands
-3. Gothenburg Global Biodiversity Centre, University of Gothenburg, Box 461, 405 30 Gothenburg, Sweden
-4. Department for Biological and Environmental Sciences, University of Gothenburg, Box 461, 405 30 Gothenburg, Sweden
-5. Royal Botanic Gardens Kew, TW9 3AE, Richmond, Surrey, United Kingdom
-6. Department of Biology, University of Fribourg, Ch. du Musée 10, 1700 Fribourg, Switzerland
-
 ```{r setup, include=FALSE}
@@
 However, geographical biases are widely recognized as a factor limiting the usefulness of such data for understanding species diversity and distribution.
-In particular, differences in sampling intensity across a landscape due to differences in human accessibility are ubiquitous but may differ in strength among taxonomic groups and datasets.
+In particular, differences in sampling intensity across a landscape due to differences in human accessibility are ubiquitous but may differ in strength among taxonomic groups and data sets.
 Although several factors have been described to influence human access (such as presence of roads, rivers, airports and cities), quantifying their specific and combined effects on recorded occurrence data remains challenging.
-Here we present *sampbias*, an algorithm and software for quantifying the effect of accessibility biases in species occurrence datasets.
-*Sampbias* uses a Bayesian approach to estimate how sampling rates vary as a function of proximity to one or multiple bias factors.
-The results are comparable among bias factors and datasets.
-We demonstrate the use of *sampbias* on a dataset of mammal occurrences from the island of Borneo, showing a high biasing effect of cities and a moderate effect of roads and airports.
-*Sampbias* is implemented as a well-documented, open-access and  user-friendly R package that we hope will become a standard tool for anyone working with species occurrences in ecology, evolution, conservation  and related fields.
+Here we present *sampbias*, an algorithm and software for quantifying the effect of accessibility biases in species occurrence data sets.
+*sampbias* uses a Bayesian approach to estimate how sampling rates vary as a function of proximity to one or multiple bias factors.
+The results are comparable among bias factors and data sets.
+We demonstrate the use of *sampbias* on a data set of mammal occurrences from the island of Borneo, showing a high biasing effect of cities and a moderate effect of roads and airports.
+*sampbias* is implemented as a well-documented, open-access and  user-friendly R package that we hope will become a standard tool for anyone working with species occurrences in ecology, evolution, conservation  and related fields.
 
@@
 # Background
-Publicly available datasets of geo-referenced species occurrences, such as provided by the Global Biodiversity Information Facility (www.gbif.org) have become a fundamental resource in biological sciences, especially in biogeography, conservation, and macroecology.
-However, these datasets are typically not collected systematically and rarely include information on collection effort.
+Publicly available data sets of geo-referenced species occurrences, such as provided by the Global Biodiversity Information Facility (www.gbif.org) have become a fundamental resource in biological sciences, especially in biogeography, conservation, and macroecology.
+However, these data sets are typically not collected systematically and rarely include information on collection effort.
 Instead, they  are often compiled from a variety of sources (e.g., scientific expeditions, census counts, genetic barcoding studies and citizen-science observations).
@@
 Sampling biases that may affect the recording of species occurrences [presence, absence and abundance, @Isaac2015; @Boakes2010] include the under-sampling of specific taxa ("taxonomic bias", e.g., birds *vs.* nematodes), specific geographic regions ("geographic bias", e.g., easily accessible *vs.* remote areas), and specific temporal periods ("temporal bias", e.g., wet *vs.* dry season).
-In particular geographic sampling bias---the fact that sampling effort is spatially biased, rather than equally distributed over the study area---is likely to be widespread in all non-systematically collected datasets of species distributions.
+In particular geographic sampling bias---the fact that sampling effort is spatially biased, rather than equally distributed over the study area---is likely to be widespread in all non-systematically collected data sets of species distributions.
 
@@
 
-\textcolor{blue}{The implications of not considering geographic sampling biases in biodiversity research are likely substantial} [@Meyer2016; @Yang2013; @Barbosa2013].
-\textcolor{blue}{The presence of geographic sampling biases is broadly recognized} [e.g., @Kadmon2004], \textcolor{blue}{and approaches exist to account for it in some analyses---such as species-richness estimates} [@Engemann2015], \textcolor{blue}{occupancy models} [@Kery2016], \textcolor{blue}{and abundance estimates} [@Shimadzu2015].
-\textcolor{blue}{In the case of species distribution modelling---the statistical estimation of species geographic distributions based on known occurrences and environmental conditions---geographically biased sampling is problematic because it often causes environmentally biased sampling which decreases model performance} [@Kadmon2004; @Lobo2011; @Bystriakova2012; @Kramer-Schadt2013; @Varela2014].
-\textcolor{blue}{Many approaches exist to remedy the effect of biased sampling on species distribution models} [@Fourcade2014], \textcolor{blue}{including rarefaction to reduce clumped sampling in geographic} [@Aiello-Lammens2015; @Beck2014; @Boria2014] \textcolor{blue}{or environmental space} [@Varela2014]\textcolor{blue}{, collecting background points for presence-only models to reflect the same sampling bias as the presence records} [@Phillips2009]\textcolor{blue}{, and explicitly modelling sampling bias} [@Komori2020; @Fithian2014; @Stolar2015]\textcolor{blue}{.
-In contrast, few attempts have been made to compare the geographic sampling bias among datasets} [@Fernandez2015; @Ruete2015; @Monsarrat2019] \textcolor{blue}{and to our knowledge, no tools exist to quantify the effect size of specific bias factors and compare it among them.
-We define as \textit{bias factors} any anthropogenic or natural features that facilitate human access and sampling, such as roads, rivers, airports, and cities.}
+The implications of not considering geographic sampling biases in biodiversity research are likely substantial [@Meyer2016; @Yang2013; @Barbosa2013].
+The presence of geographic sampling biases is broadly recognized [e.g., @Kadmon2004], and approaches exist to account for it in some analyses---such as species-richness estimates [@Engemann2015], occupancy models [@Kery2016], and abundance estimates [@Shimadzu2015].
+In the case of species distribution modelling---the statistical estimation of species geographic distributions based on known occurrences and environmental conditions---geographically biased sampling is problematic because it often causes environmentally biased sampling which decreases model performance [@Kadmon2004; @Lobo2011; @Bystriakova2012; @Kramer-Schadt2013; @Varela2014].
+Many approaches exist to remedy the effect of biased sampling on species distribution models [@Fourcade2014], including rarefaction to reduce clumped sampling in geographic [@Aiello-Lammens2015; @Beck2014; @Boria2014] or environmental space [@Varela2014], collecting background points for presence-only models to reflect the same sampling bias as the presence records [@Phillips2009], and explicitly modelling sampling bias [@Komori2020; @Fithian2014; @Stolar2015].
+In contrast, few attempts have been made to compare the geographic sampling bias among data sets [@Fernandez2015; @Ruete2015; @Monsarrat2019] and to our knowledge, no tools exist to quantify the effect size of specific bias factors and compare it among them.
+We define as \textit{bias factors} any anthropogenic or natural features that facilitate human access and sampling, such as roads, rivers, airports, and cities.
 
 It is unrealistic to expect that accessibility bias in biodiversity data will disappear even after more automated observation technologies are developed.
-It is therefore crucial that researchers realise the intrinsic biases associated with the data they deal with, \textcolor{blue}{especially in cross-taxonomic studies, since occurrence datasets from different taxa are likely differently affected by sampling biases due to differences in specimen collection and transportation}.
+It is therefore crucial that researchers realise the intrinsic biases associated with the data they deal with, especially in cross-taxonomic studies, since occurrence data sets from different taxa are likely differently affected by sampling biases due to differences in specimen collection and transportation.
 This is the first step towards estimating to which extent these biases may affect their analyses, results and conclusions.
@@
 
-Here, we present *sampbias* v1.0.4, a probabilistic method to quantify accessibility bias in datasets of species occurrences.
-*Sampbias* is implemented as a user-friendly R-package and uses a Bayesian approach to address three questions:
+Here, we present *sampbias* v1.0.4, a probabilistic method to quantify accessibility bias in data sets of species occurrences.
+*sampbias* is implemented as a user-friendly R-package and uses a Bayesian approach to address three questions:
 
-1) How strong is the accessibility bias in a given dataset?
+1) How strong is the accessibility bias in a given data set?
 
@@
 
-*Sampbias* is implemented in R [@rcoreteam2019], based on commonly used packages for data handling  [`ggplot`, @Wickham2009; `tidyr`, @Wickham2019a; `dplyr`, @Wickham2019; `forcats`, @Wickham2019b; `magrittr`, @Bache2014; `viridis`, @Garnier2018], handling geographic information and geo-computation [`raster`, @Hijmans2019; `sp`, @Pebesma2005; @Bivand2013] and statistical modelling [`stats`, @rcoreteam2019].
-*Sampbias* offers an easy and largely automated means for biodiversity scientists and non-specialists alike to explore bias in species occurrence data, in a way that is comparable across datasets.
-\textcolor{blue}{The results may be used to identify priorities for further collection or digitalization efforts and to assess the reliability of scientific results based on publicly available species distribution data.}
+*sampbias* is implemented in R [@rcoreteam2019], based on commonly used packages for data handling  [`ggplot`, @Wickham2009; `tidyr`, @Wickham2019a; `dplyr`, @Wickham2019; `forcats`, @Wickham2019b; `magrittr`, @Bache2014; `viridis`, @Garnier2018], handling geographic information and geo-computation [`raster`, @Hijmans2019; `sp`, @Pebesma2005; @Bivand2013] and statistical modelling [`stats`, @rcoreteam2019].
+*sampbias* offers an easy and largely automated means for biodiversity scientists and non-specialists alike to explore bias in species occurrence data, in a way that is comparable across data sets.
+The results may be used to identify priorities for further collection or digitalization efforts and to assess the reliability of scientific results based on publicly available species distribution data.
 
@@
 
-*Sampbias* works at a user-defined spatial scale, and any dataset of multi-species occurrence records can be tested against any geographic gazetteer.
-Reliability increases with increasing dataset size.
+*sampbias* works at a user-defined spatial scale, and any data set of multi-species occurrence records can be tested against any geographic gazetteer.
+Reliability increases with increasing data set size.
 Default global gazetteers for airports, cities, rivers and roads are provided with *sampbias*, and user-defined gazetteers can be added easily.
@@
 ## Distance calculation
-*Sampbias* uses gazetteers of the geographic location of bias factors (hereafter indicated with B) to generate a regular grid across the study area.
-\textcolor{blue}{By default the study area is defined by the geographic extent of the study dataset, but it can also be customized via user-defined polygons, for instance to limit the analyses to an environmentally homogeneous region (e.g., a rainforest) or an area of special interest (e.g., a national park).
-In this case all occurrence records outside the user-defined area will be disregarded for the analysis.}.
+*sampbias* uses gazetteers of the geographic location of bias factors (hereafter indicated with B) to generate a regular grid across the study area.
+By default the study area is defined by the geographic extent of the study data set, but it can also be customized via user-defined polygons, for instance to limit the analyses to an environmentally homogeneous region (e.g., a rainforest) or an area of special interest (e.g., a national park).
+In this case all occurrence records outside the user-defined area will be disregarded for the analysis..
 For each grid cell $i$, we then compute a vector $X_i(j)$ of minimum distances (straight aerial distance, "as the crow flies") to each bias factor $j \in B$.
 The resolution of the grid defines the precision of the distance estimates, for instance a $1 \times 1$ degree raster will yield approximately a 110 km precision at the equator.
-Due to the assumption of homogeneous sampling and a computational trade-off between the resolution of the regular grid and the extent of the study area (for instance, a 1 second resolution for a global dataset would become computationally prohibitive in most practical cases), *sampbias* is best suited for local or regional datasets at high resolution (c.
+Due to the assumption of homogeneous sampling and a computational trade-off between the resolution of the regular grid and the extent of the study area (for instance, a 1 second resolution for a global data set would become computationally prohibitive in most practical cases), *sampbias* is best suited for local or regional data sets at high resolution (c.
 100 -- 10,000 m).
-\textcolor{blue}{Since the differences in grid cell size are negligible on the local and regional scale, sampbias uses a latitude/longitude grid by default, but a custom grid in any projection and coordinate reference system---for instance an equal area grid, which is often more suitable for large spatial analyses---may be provided by the user.}
+Since the differences in grid cell size are negligible on the local and regional scale, sampbias uses a latitude/longitude grid by default, but a custom grid in any projection and coordinate reference system---for instance an equal area grid, which is often more suitable for large spatial analyses---may be provided by the user.
 
@@
 We use exponential priors on the parameters $q$ and $\mathbf{w}$, $P(q) \sim \Gamma(1, 0.01)$ and $P(\mathbf{w}) \sim \Gamma(1, 1)$, respectively.
-\textcolor{blue}{We chose exponential priors because they represent the standard choice for rate parameters such as} $q$ \textcolor{blue}{and the weights} $\mathbf{w}$\textcolor{blue}{, all of which must be positive and have support [0, +inf].
+We chose exponential priors because they represent the standard choice for rate parameters such as $q$ and the weights $\mathbf{w}$, all of which must be positive and have support [0, +inf].
 We designed the priors to be informative (i.e.
 not allowing negative values) and yet vague enough to encompass a much wider range of parameter space than the range of values observed in empirical tests.
-Custom priors, within the flexible family of gamma distributions, which include the exponential priors used here, are possible via the} `prior_q` and `prior_w` \textcolor{blue}{arguments of the} `calculate_bias` \textcolor{blue}{function.
+Custom priors, within the flexible family of gamma distributions, which include the exponential priors used here, are possible via the `prior_q` and `prior_w` arguments of the `calculate_bias` function.
 Additionally, since the null expectation in the absence of biases is that the weights are close to 0, we implement a hierarchical model, in which the rate parameter of the gamma prior can be itself estimated from the data.
-Thus, we set} $P(\mathbf{w}) \sim \Gamma(a=1, b)$ \textcolor{blue}{and assign a vague hyper-prior on the rate} $P(b) \sim \Gamma(\alpha_0=1, \beta_0=0.001)$.
-\textcolor{blue}{The choice of conjugate gamma distribution allows us to sample the rate directly from its posterior distribution:}
+Thus, we set $P(\mathbf{w}) \sim \Gamma(a=1, b)$ and assign a vague hyper-prior on the rate $P(b) \sim \Gamma(\alpha_0=1, \beta_0=0.001)$.
+The choice of a conjugate gamma hyper-prior allows us to sample the rate directly from its posterior distribution:
 $$
-b \sim \Gamma(\alpha_0 + a B,\ \beta_0  \sum_{j=1}^{B}{w_j}) \ \ \ \ (3)
+b \sim \Gamma \left(\alpha_0 + a B,\ \beta_0  \sum_{j=1}^{B}{w_j} \right) \ \ \ \ (3)
 $$
-\textcolor{blue}{The use of a hyper-prior has the advantage of making the prior on the weights more flexible and able to adapt to different datasets reducing the need for user-defined arbitrary choices.
-Additionally, it works as a regularization technique reducing the risks of over-parametrization, by shrinking the weights around small values when there is no evidence in the data indicating a bias.}
-<!--For instance, the posterior weights estimated in our empirical study ranged from 0 to 0.013, while the prior distribution applied on these parameters, Exp(1) has a} 95% \textcolor{blue}{of it density in range 0 – 3.
+The use of a hyper-prior has the advantage of making the prior on the weights more flexible and able to adapt to different data sets reducing the need for user-defined arbitrary choices.
+Additionally, it works as a regularization technique reducing the risks of over-parametrization, by shrinking the weights around small values when there is no evidence in the data indicating a bias.
+<!--For instance, the posterior weights estimated in our empirical study ranged from 0 to 0.013, while the prior distribution applied on these parameters, Exp(1) has a} 95% of it density in range 0 – 3.
 C}
@@
 We interpret the magnitude of the elements in $\mathbf{w}$ as a function of the importance of the individual biases.
-\textcolor{blue}{We note, however, that this test is not explicitly intended to assess the significance of each bias factor.
+We note, however, that this test is not explicitly intended to assess the significance of each bias factor.
 Because several bias factors might be correlated (e.g.
@@
 It is therefore important to jointly estimate the effects of correlated factors, as this is based on the likelihood of the data given the combined effects of all biasing factors.
-A Bayesian variable selection method could be used to quantify the expected amount of bias in the data predicted by single or a particular combination of predictors, but falls beyond the scope of the current study.}
+A Bayesian variable selection method could be used to quantify the expected amount of bias in the data predicted by single or a particular combination of predictors, but falls beyond the scope of the current study.
+In the empirical example below we show, however, how a simple simulation can be used to asses whether the estimated bias factors significantly differ from a null expectation.
 
-We summarize the results by mapping the estimated `textcolor{blue}{(the relative deviation of sampling rate from the maximum
-sampling rate (or on user choice the estimated sampling rates}, $\lambda_i$) across space.
+We summarize the results by mapping the estimated (the relative deviation of sampling rate from the maximum
+sampling rate (or on user choice the estimated sampling rates, $\lambda_i$) across space.
 These rates represent the expected number of sampled occurrences for each grid cell and provide a graphical representation of the spatial variation of sampling rates.
@@
 ## Example and empirical validation
-A default *sampbias* analysis can be run with few lines of code in R.
+A default *sampbias* analysis can be run with few lines of code in R, based on a `data.frame` including species identity and geographic coordinates.
 The main function `calculate_bias` creates an object of the class `"sampbias"`, for which the package provides a plotting and summary method.
-Based on a `data.frame` including species identity and geographic coordinates.
-\textcolor{blue}{Additional options exist to provide custom gazetteers, study area, spatial grid and grain size of the analysis, as well as some operators for the calculation of the bias distances, including  priors for $q$ and $\mathbf{w}$}.
+Additional options exist to provide custom gazetteers, study area, spatial grid and grain size of the analysis, as well as operators for the calculation of the bias distances, including  priors for $q$ and $\mathbf{w}$.
 A tutorial on how to use *sampbias* is available with the package and in the electronic supplement of this publication (Appendix S1).
 
-To exemplify the use of *sampbias*, we downloaded the occurrence records of all mammals available from the island of Borneo [n = 6,262, @gbifdoi2016] and ran *sampbias* using the default gazetteers as shown in the example code below, to test the biasing effect of the main airports, cities and roads in the dataset.
-The example dataset is provided with *sampbias*.
+To exemplify the use of *sampbias*, we downloaded the occurrence records of all mammals available from the island of Borneo [n = 6,262, @gbifdoi2016] and ran *sampbias* using the default gazetteers as shown in the example code below, to test the biasing effect of the main airports, cities and roads in the data set.
+The example data set is provided with *sampbias*.
 
@@
 
+# laod the outline of Borneo (provided as example with sampbias)
+data(borneo)
+
 # running sampbias
@@
 ## structure adjacent to the study area, in degrees latitude and longitude
+## 'restrict_sample' restricts the analysis to Borneo, 
+## as example for a user-defined area 
 ## All other options at default, see ?calculate_bias for a description
@@
                               res = 0.05,
-                              buffer = 0.5)
+                              buffer = 0.5,
+                              restrict_sample = borneo)
 
@@
 
-\textcolor{blue}{We ran simulation experiment to assess whether the estimated bias weights differ significantly from a null expectation of random sampling.
-To do so, we first created ten replicates of an unbiased data by generating 6262 random occurrences across Borneo (the same number as in the empirical dataset) and then ran a} *sampbias* \textcolor{blue}{with the same settings as for the empirical dataset on each of these replicates.
-We found that the estimated bias range were significantly different (credible intervals non-overlapping) than in the null model for cities, roads and airports (Figure S2 in Appendix S1).}
+We ran a simulation experiment to assess whether the estimated bias weights differ significantly from a null expectation of random sampling.
+To do so, we first simulated ten unbiased data sets by generating 6262 random occurrences across Borneo (the same number as in the empirical data set) and then ran *sampbias* with the same settings as for the empirical data set on each of these replicates.
+We found that the estimated bias parameters were significantly different (credible intervals non-overlapping) from the null model for cities, roads and airports (Figure S2 in Appendix S2).
 
-*Sampbias* \textcolor{blue}{is designed to work with sparsely sampled datasets, and to estimate bias effects from datasets with low coverage.
-To test the performance of} *sampbias* \textcolor{blue}{on small datasets we ran an additional simulation experiment.
-We ran the analyses with the same settings as for the empirical dataset on a set of rarified datasets, sub-sampling the initial dataset to 3,131, 626, 62 records respectively (three replicates each), and then compared the estimated bias weights for all bias factors to the estimates for the empirical  dataset.
-The results showed that parameter estimates and the projection of the bias effect in space were robust to decreases in data, although uncertainty increased (Figs.
-S3 and S4, in Appendix S1).
-The results show that} *sampbias* \textcolor{blue}{reliably rejected a biasing effect even for small datasets (Figs.
-S4 and S5, in Appendix S2).
-In general, the more records, the more precise the parameter estimates will be.}
+*sampbias* is designed to work with sparsely sampled data sets.
+To test the performance of *sampbias* on small data sets we ran an additional simulation experiment.
+We ran *sampbias*  with the same settings as for the empirical data set on three rarefied data sets, sub-sampling the initial data set to 3,131, 626, 62 records respectively (three replicates each, a total of nine analyses), and then compared the estimated bias weights for all bias factors to the estimates for the empirical  data set.
+The results showed that parameter estimates and the projection of the bias effect in space were robust to the decreasing amount of data, although uncertainty increased as reflected larger estimated credible intervals (Figs.
+S3 and S4, in Appendix S2).
 
 # Data accessibility
-*Sampbias* is available under a GNU General Public license v3 from https://github.com/azizka/sampbias, and includes the example dataset as well as a tutorial (Appendix S1) and a summary of possible warnings produced by the package (Appendix S3).
+*sampbias* is available under a GNU General Public license v3 from https://github.com/azizka/sampbias, and includes the example data set as well as a tutorial (Appendix S1) and a summary of possible warnings produced by the package (Appendix S3).
 
@@
 
-```{r spatial, out.width = "6in", echo = FALSE, fig.cap = "Spatial projection of the sampling bias in an empirical example dataset of mammal occurrences on the Indonesian island of Borneo (downloaded from www.gbif.org.
+```{r spatial, out.width = "6in", echo = FALSE, fig.cap = "Spatial projection of the sampling bias in an empirical example data set of mammal occurrences on the island of Borneo (downloaded from www.gbif.org.
 GBIF.org, 2016).
-The colours show the projection of the The colours show the projection of the \\textcolor{blue}{log10-transformed} sampling rates (i.e.
+The colours show the projection of the log10-transformed sampling rates (i.e.
 expected number of occurrences per cell) given the inferred \\textit{sampbias} model.
 The highest undersampling is in the centre of the island.
-\\textcolor{blue}{DIfferent visualizations, including among others the untransformed sampling rate are also implemented in sampbias.", eval = TRUE}
+Different visualizations, including among others the untransformed sampling rate are also implemented in sampbias.", eval = TRUE}
 knitr::include_graphics(path = "ms_figures/figure_empirical_results_spatial_projection_adapted.jpg")

### 2024-01-23 R/calculate_bias.R
Update to 2.0

@@ -19,32 +19,31 @@
 #' distance calculation and the algorithm behind sampbias.
 #'
 #' @param x an object of the class \code{data.frame}, with one species
-#' occurrence record per line, and at least three columns, named
-#' \sQuote{species}, \sQuote{decimalLongitude}, and \sQuote{decimalLatitude}.
-#' @param gaz a list of geographic gazetteers as \code{SpatialPointDataFrame}
-#' or \code{SpatialLinesDataFrame}.  If NULL, a set of default gazetteers,
-#' representing large scale occurrence of airports, cities, rivers and roads is
-#' used.  See Details.
+#'   occurrence record per line, and at least three columns, named
+#'   \sQuote{species}, \sQuote{decimalLongitude}, and \sQuote{decimalLatitude}.
+#' @param gaz a list of geographic gazetteers as \code{SpatVector} or \code{sf}.
+#'   If NULL, a set of default gazetteers, representing large scale occurrence
+#'   of airports, cities, rivers, and roads is used. See Details.
 #' @param res numerical.  The raster resolution for the distance calculation to
-#' the geographic features and the data visualization, in decimal degrees. The
-#' default is to one degree, but higher resolution will be desirable for most
-#' analyses. \code{Res} together with the extent of the input data determine
-#' computation time and memory requirements.
-#' @param buffer numerical.  The size of the geographic buffer around the
-#' extent of \code{ras} for the distance calculations in degrees, to account
-#' for geographic structures neighbouring the study area (such as a road right
-#' outside the study area). Should be a multiple of \code{res}.  Default is to \code{res}.
-#' See Details.
-#' @param restrict_sample a SpatialPolygons or SpatialPolygonDataframe.
-#' If provided the area for the bias test will be restricted to raster cells
-#' within these polygons (and the extent of the sampled points in x).
-#' Make sure to use adequate values for \code{res}. Default = NULL.
+#'   the geographic features and the data visualization, in decimal degrees. The
+#'   default is to one degree, but higher resolution will be desirable for most
+#'   analyses. \code{res} together with the extent of the input data determine
+#'   computation time and memory requirements.
[…]

### 2024-01-23 R/dis_rast.R

@@ -1,24 +1,22 @@
-# Distance raster for all gazeteers get empirical distace distribution from
-# gazetteers, rasterbased
-
+# Distance raster for all gazeteers get empirical distance distribution from
+# gazetteers, raster based
 
 #' Distance Rasters from a List of Geographic Gazetteers
 #'
 #' Creates a list of distances rasters based on a list of geographic
-#' gazetteers, as SpatialPoints or SpatialLines objects, and a template raster,
-#' indicating the desired extent and resolution
+#' gazetteers, as SpatVector objects, and a template SpatRaster,
+#' indicating the desired extent and resolution.
 #'
 #'
 #' @param gaz an object of the class \code{list}, including one or more
-#' geographic gazetteers of the class \code{SpatialPointsDataFrame} or
-#' \code{SpatialLinesDataFrame}.
-#' @param ras an object of the class \code{raster}. Defining the extent and
+#' geographic gazetteers of the class \code{SpatVector}.
+#' @param ras an object of the class \code{SpatRaster}. Defining the extent and
 #' resolution of the distances rasters.
 #' @param buffer numerical.  The size of the geographic buffer around the
 #' extent of \code{ras} for the distance calculations in degrees, to account for
 #' geographic structures neighbouring the study area (such as a road right
 #' outside the study area) Default is to the resolution of \code{ras}.
-#' @return a \code{list} of \code{raster} objects of the same length as
+#' @return a \code{list} of \code{SpatRaster} objects of the same length as
 #' \code{gaz}. The values in each raster correspond to the planar geographic
 #' distance to the next feature in \code{gaz}, given the resolution of
 #' \code{ras}
@@ -29,19 +27,18 @@
 #' @examples
 #'
 #' #create raster for resolution and extent
-#' ras <- raster::raster(raster::extent(-5,5,-4,4), res = 1)
+#' ras <- terra::rast(terra::ext(-5,5,-4,4), res = 1)
 #'
 #' #create point gazeteer
 #' pts <- data.frame(long = runif(n = 5, min = -5, max = 5),
 #'                   lat = runif(n = 5, min = -4, max = 4),
[…]

### 2024-01-23 R/internal.R

@@ -1,32 +1,26 @@
-#'@importFrom raster rasterize
+#'@importFrom terra rasterize
 #'@importFrom stats dgamma dpois rnorm runif rgamma
 
 #Occurrence raster
 .OccRast <- function(x, ras){
-  rast <- raster::rasterize(x, ras, fun = "count")
+  rast <- terra::rasterize(x, ras, fun = "count")
   return(rast)
 }
 
 #get number of decimal places
 .DecimalPlaces <- function(x) {
   if ((x %% 1) != 0) {
-    nchar(strsplit(sub('0+$', '', as.character(x)), ".", fixed = TRUE)[[1]][[2]])
+    nchar(strsplit(sub('0+$', '', 
+                       as.character(x)),
+                   ".", fixed = TRUE)[[1]][[2]])
   } else {
     return(0)
   }
 }
 
-# get_lambda_ij_old <- function(q, w, X) {
-#   if (class(X) == "numeric") {
-#     return(q * exp(-w * X))
-#   } else {
-#     return(q * exp(-apply(FUN = sum, w * X, 2)))
-#   }
-# }
-
 
 get_lambda_ij <- function(q, w, X) {
-  if (class(X) == "numeric") {
+  if (is.numeric(X)) {
     return(q * exp(-w * X))
   } else {
     return(as.vector(q * exp(-(w %*% t(X)))))
@@ -63,12 +57,13 @@ get_post_rate <- function(w,alpha){
                          prior_q = c(1, 0.01),
                          prior_w = c(1, 1),
                          run_null_model = FALSE,
-                         use_hyperprior = TRUE) {
+                         use_hyperprior = TRUE,
+                         verbose = FALSE) {
 
   indx <- c(3:ncol(x))
 
   # X is the distance matrix with 1 row for each cell and 1 col for each predictor
-  X <- x[, indx]/rescale_distances
+  X <- x[, indx] / rescale_distances
   # if (length(indx) > 1) {
   #   X <- t(X)
   # }
@@ -80,13 +75,13 @@ get_post_rate <- function(w,alpha){
 
   # init weigths of predictors
   wA <- abs(rnorm(length(indx), 0.01, 0.01))
-  if (run_null_model == TRUE){
+  if (run_null_model == TRUE) {
     wA = rep(0, length(indx))
   }
-  if (use_hyperprior & length(indx) > 1){
[…]

### 2024-01-23 R/is.sampbias.R

@@ -11,28 +11,15 @@
 #' classes.
 #' @keywords methods
 #' @examples
+#' \dontrun{
+#'   #simulate data
+#'   occ <- data.frame(species = rep(sample(x = LETTERS, size = 5), times = 10),
+#'                    decimalLongitude = runif(n = 50, min = 12, max = 20),
+#'                    decimalLatitude = runif(n = 50, min = -4, max = 4))
 #'
-#' #simulate data
-#' occ <- data.frame(species = rep(sample(x = LETTERS, size = 5), times = 10),
-#'                   decimalLongitude = runif(n = 50, min = -5, max = 5),
-#'                   decimalLatitude = runif(n = 50, min = -4, max = 4))
-#'
-#'
-#' #create point gazetteer
-#' pts <- data.frame(long = runif(n = 5, min = -5, max = 5),
-#'                   lat = runif(n = 5, min = -4, max = 4),
-#'                   dat = rep("A", 5))
-#' pts <- sp::SpatialPointsDataFrame(coords = pts[,1:2], data = data.frame(pts[,3]))
-#'
-#' lin <- data.frame(long = seq(-5, 5, by = 1),
-#'                   lat = rep(2, times = 11))
-#' lin <- sp::SpatialLinesDataFrame(sl = sp::SpatialLines(list(sp::Lines(sp::Line(lin), ID="B1"))),
-#'                              data = data.frame("B", row.names = "B1"))
-#'
-#' gaz <- list(lines.strucutre = lin, point.structure = pts)
-#'
-#' out <- calculate_bias(x = occ, gaz = gaz, terrestrial = FALSE)
-#' is(out)
+#'   out <- calculate_bias(x = occ, terrestrial = TRUE)
+#'   is(out)
+#' }
 #'@method is sampbias
 #'@importFrom methods is
 #'@export

### 2024-01-23 R/map_bias.R

@@ -24,34 +24,23 @@
 #' @seealso \code{\link{calculate_bias}}, \code{\link{project_bias}}
 #' @keywords methods
 #' @examples
+#' \dontrun{
+#'   #simulate data
+#'   occ <- data.frame(species = rep(sample(x = LETTERS, size = 5), times = 10),
+#'                    decimalLongitude = runif(n = 50, min = 12, max = 20),
+#'                    decimalLatitude = runif(n = 50, min = -4, max = 4))
 #'
-#' #simulate data
-#' occ <- data.frame(species = rep(sample(x = LETTERS, size = 5), times = 10),
-#'                   decimalLongitude = runif(n = 50, min = -5, max = 5),
-#'                   decimalLatitude = runif(n = 50, min = -4, max = 4))
-#'
-#'
-#' #create point gazetteer
-#' pts <- data.frame(long = runif(n = 5, min = -5, max = 5),
-#'                   lat = runif(n = 5, min = -4, max = 4),
-#'                   dat = rep("A", 5))
-#' pts <- SpatialPointsDataFrame(coords = pts[,1:2], data = data.frame(pts[,3]))
-#'
-#' lin <- data.frame(long = seq(-5, 5, by = 1),
-#'                   lat = rep(2, times = 11))
-#' lin <- SpatialLinesDataFrame(sl = SpatialLines(list(Lines(Line(lin), ID="B1"))),
-#'                              data = data.frame("B", row.names = "B1"))
-#'
-#' gaz <- list(lines.strucutre = lin, point.structure = pts)
-#'
-#' out <- calculate_bias(x = occ, gaz = gaz, terrestrial = FALSE)
-#' proj <- project_bias(out)
-#' map_bias(proj)
+#'   out <- calculate_bias(x = occ, terrestrial = TRUE)
+#'   proj <- project_bias(out)
+#'   map_bias(proj)
+#' }
 #'
 #'@export
 #'@importFrom dplyr rename bind_cols filter
-#'@importFrom ggplot2 aes coord_fixed element_blank facet_wrap fortify geom_path geom_point geom_polygon geom_raster ggplot ggtitle labs theme theme_bw
-#'@importFrom raster as.data.frame coordinates
+#'@importFrom ggplot2 aes coord_fixed element_blank facet_wrap fortify geom_path
+#'  geom_point geom_polygon geom_raster ggplot ggtitle labs theme theme_bw
+#'  geom_sf
[…]

### 2024-01-23 R/plot.sampbias.R

@@ -12,28 +12,16 @@
 #' @seealso \code{\link{calculate_bias}}, \code{\link{summary.sampbias}}
 #' @keywords methods
 #' @examples
+#' \dontrun{
+#'   #simulate data
+#'   occ <- data.frame(species = rep(sample(x = LETTERS, size = 5), times = 10),
+#'                    decimalLongitude = runif(n = 50, min = 12, max = 20),
+#'                    decimalLatitude = runif(n = 50, min = -4, max = 4))
 #'
-#' #simulate data
-#' occ <- data.frame(species = rep(sample(x = LETTERS, size = 5), times = 10),
-#'                   decimalLongitude = runif(n = 50, min = -5, max = 5),
-#'                   decimalLatitude = runif(n = 50, min = -4, max = 4))
-#'
-#'
-#' #create point gazetteer
-#' pts <- data.frame(long = runif(n = 5, min = -5, max = 5),
-#'                   lat = runif(n = 5, min = -4, max = 4),
-#'                   dat = rep("A", 5))
-#' pts <- SpatialPointsDataFrame(coords = pts[,1:2], data = data.frame(pts[,3]))
-#'
-#' lin <- data.frame(long = seq(-5, 5, by = 1),
-#'                   lat = rep(2, times = 11))
-#' lin <- SpatialLinesDataFrame(sl = SpatialLines(list(Lines(Line(lin), ID="B1"))),
-#'                              data = data.frame("B", row.names = "B1"))
-#'
-#' gaz <- list(lines.strucutre = lin, point.structure = pts)
-#'
-#' out <- calculate_bias(x = occ, gaz = gaz, terrestrial = FALSE)
-#' plot(out)
+#'   out <- calculate_bias(x = occ, terrestrial = TRUE)
+#'   summary(out)
+#'   plot(out)
+#' }
 #'
 #'@method plot sampbias
 #'@export

### 2024-01-23 R/project_bias.R

@@ -1,42 +1,32 @@
 #' Projecting Bias Effects in Space
 #'
 #'
-#' Uses the the estimated bias weights from a \code{sampbias} object to project the
-#' bias through space, using the same raster as used for the distance calculation.#'
+#' Uses the the estimated bias weights from a \code{sampbias} object to project
+#' the bias through space, using the same raster as used for the distance
+#' calculation.#'
 #'
 #' @param x an object of the class \code{sampbias}.
 #' @param factors a character vector indicating which biasing factors to project
-#' @return A raster stack, with the same length as the number of biasing factors used
-#' in \code{\link{calculate_bias}}. The names indicate the factors included for each layer.
+#' @return A raster stack, with the same length as the number of biasing factors
+#'   used in \code{\link{calculate_bias}}. The names indicate the factors
+#'   included for each layer.
 #' @seealso \code{\link{calculate_bias}}, \code{\link{summary.sampbias}}
 #' @keywords methods
 #' @examples
+#' 
+#' \dontrun{
+#'   #simulate data
+#'   occ <- data.frame(species = rep(sample(x = LETTERS, size = 5), times = 10),
+#'                    decimalLongitude = runif(n = 50, min = 12, max = 20),
+#'                    decimalLatitude = runif(n = 50, min = -4, max = 4))
 #'
-#' #simulate data
-#' occ <- data.frame(species = rep(sample(x = LETTERS, size = 5), times = 10),
-#'                   decimalLongitude = runif(n = 50, min = -5, max = 5),
-#'                   decimalLatitude = runif(n = 50, min = -4, max = 4))
-#'
-#'
-#' #create point gazetteer
-#' pts <- data.frame(long = runif(n = 5, min = -5, max = 5),
-#'                   lat = runif(n = 5, min = -4, max = 4),
-#'                   dat = rep("A", 5))
-#' pts <- SpatialPointsDataFrame(coords = pts[,1:2], data = data.frame(pts[,3]))
-#'
-#' lin <- data.frame(long = seq(-5, 5, by = 1),
-#'                   lat = rep(2, times = 11))
-#' lin <- SpatialLinesDataFrame(sl = SpatialLines(list(Lines(Line(lin), ID="B1"))),
[…]

### 2024-01-23 R/sampbias.R

@@ -7,10 +7,9 @@
 #'
 #' @name airports
 #' @docType data
-#' @format A data frame with 891 observations on the following 4 variables.
-#' \describe{ \item{list("code")}{a factor, IATA-3 code.} \item{list("name")}{a
-#' character vector, airport names.} \item{list("longitude")}{a numeric
-#' vector.} \item{list("latitude")}{a numeric vector.} }
+#' @format A \code{sf} object with 891 observations on the following 4 variables.
+#' \describe{ \item{code}{a factor, IATA-3 code.} \item{name}{a
+#' character vector, airport names.} \item{geometry}{sfc_POINT object.} }
 #' @source \url{http://www.naturalearthdata.com/}
 #' @keywords datasets
 #' @examples
@@ -33,10 +32,9 @@ NULL
 #' @name cities
 #' @docType data
 #' @format A data frame with 7322 observations on the following 4 variables.
-#' \describe{ \item{list("nameascii")}{a factor; city name.}
-#' \item{list("adm0_a3")}{a factor; ISO-3 code of country name.}
-#' \item{list("longitude")}{a numeric vector.} \item{list("latitude")}{a
-#' numeric vector.} }
+#' \describe{ \item{nameascii}{a factor; city name.}
+#' \item{adm0_a3}{a factor; ISO-3 code of country name.}
+#' \item{geometry}{sfc_POINT object.} }
 #' @source
 #' \url{http://www.naturalearthdata.com/downloads/10m-cultural-vectors/}
 #' @keywords datasets
@@ -142,8 +140,8 @@ NULL
 
 #' Equal Area Raster
 #'
-#' An example for an global equal area raster (in Behrmann projection) for the format needed for
-#' a custom grid provided to \code{\link{calculate_bias}}.
+#' An example for an global equal area raster (in Behrmann projection) for the
+#' format needed for a custom grid provided to \code{\link{calculate_bias}}.
 #'
 #' @name ea_raster
 #' @docType data
@@ -151,7 +149,8 @@ NULL
 #' @examples
 #'
 #' data(ea_raster)
-#'
+#' ea_raster <- terra::unwrap(ea_raster)
+#' @importFrom terra unwrap
 "ea_raster"
 
 #' Borneo

### 2024-01-23 R/summary.sampbias.R

@@ -11,27 +11,15 @@
 #' @keywords methods
 #' @examples
 #'
-#' #simulate data
-#' occ <- data.frame(species = rep(sample(x = LETTERS, size = 5), times = 10),
-#'                   decimalLongitude = runif(n = 50, min = -5, max = 5),
-#'                   decimalLatitude = runif(n = 50, min = -4, max = 4))
+#' \dontrun{
+#'   #simulate data
+#'   occ <- data.frame(species = rep(sample(x = LETTERS, size = 5), times = 10),
+#'                    decimalLongitude = runif(n = 50, min = 12, max = 20),
+#'                    decimalLatitude = runif(n = 50, min = -4, max = 4))
 #'
-#'
-#' #create point gazetteer
-#' pts <- data.frame(long = runif(n = 5, min = -5, max = 5),
-#'                   lat = runif(n = 5, min = -4, max = 4),
-#'                   dat = rep("A", 5))
-#' pts <- SpatialPointsDataFrame(coords = pts[,1:2], data = data.frame(pts[,3]))
-#'
-#' lin <- data.frame(long = seq(-5, 5, by = 1),
-#'                   lat = rep(2, times = 11))
-#' lin <- SpatialLinesDataFrame(sl = SpatialLines(list(Lines(Line(lin), ID="B1"))),
-#'                              data = data.frame("B", row.names = "B1"))
-#'
-#' gaz <- list(lines.strucutre = lin, point.structure = pts)
-#'
-#' out <- calculate_bias(x = occ, gaz = gaz, terrestrial = FALSE)
-#' summary(out)
+#'   out <- calculate_bias(x = occ, terrestrial = TRUE)
+#'   summary(out)
+#' }
 #' @method summary sampbias
 #' @export
 #'

### 2024-01-29 R/calculate_bias.R
fix cran checks

@@ -105,7 +105,7 @@
 #' @importFrom rlang .data
 #' @importFrom stats complete.cases
 #' @importFrom sf st_transform st_crop
-#' @importFrom rnaturalearth ne_download ne_file_name ne_countries ne_coastlines
+#' @importFrom rnaturalearth ne_download ne_file_name ne_countries ne_coastline
 #'
 calculate_bias <- function(x,
                           gaz = NULL,

### 2024-01-29 R/sampbias.R

@@ -46,7 +46,7 @@
 #' Borneo
 #'
 #' The outline of Borneo, as example data for the user-defined study area option
-#' of \code{\link{calculate_bias}}. From \url{www.naturalearth.org}.
+#' of \code{\link{calculate_bias}}. From \url{https://www.naturalearthdata.com}.
 #'
 #' @name borneo
 #' @docType data

### 2024-01-29 vignettes/Using_the_samp_bias_rpackage.R

@@ -1,40 +1,38 @@
 ## ----setup, include=FALSE-----------------------------------------------------
-knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE, fig.pos = 'H')
+knitr::opts_chunk$set(echo = TRUE, 
+                      warning = FALSE, 
+                      message = FALSE, 
+                      fig.pos = 'H',
+                      eval = TRUE)
 
 ## ---- eval = FALSE------------------------------------------------------------
 #  require(devtools)
 #  install_github(repo = "azizka/sampbias")
-#  library(sampbias)
-#  library(maptools)
-#  library(rgdal)
-#  library(raster)
 
-## ---- echo = FALSE------------------------------------------------------------
+## ----load_packages------------------------------------------------------------
 library(sampbias)
-library(maptools)
-library(rgdal)
-library(raster)
+library(terra)
 
 ## -----------------------------------------------------------------------------
 #loading a text file to R
-occ <-read.csv(system.file("extdata", 
+occ <- read.csv(system.file("extdata", 
                            "mammals_borneo.csv", 
-                           package="sampbias"),
+                           package = "sampbias"),
                sep = "\t")
 
 ## ---- warning = FALSE, message = FALSE----------------------------------------
-cit <- readOGR(dsn = system.file("extdata", package="sampbias"), 
-               layer = "Borneo_major_cities", verbose = FALSE)
-roa <- readOGR(dsn = system.file("extdata", package="sampbias"), 
-               layer = "Borneo_major_roads", verbose = FALSE)
+cit <- terra::vect(system.file("extdata/Borneo_major_cities.shp",
+                               package = "sampbias"))
+roa <- terra::vect(system.file("extdata/Borneo_major_roads.shp",
+                               package = "sampbias"))
 
 gazetteers <- list(cities = cit,
                    roads = roa)
 
-## -----------------------------------------------------------------------------
+## ---- message=FALSE-----------------------------------------------------------
 bias.out <- calculate_bias(x = occ, gaz = gazetteers)
 
[…]
